### 2020-03-03 README.Rmd
test

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below. 
Title <- "Title of your paper goes here"
```

# CTtps

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, CTtps, from GitHub with:

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2020-03-03 analysis/paper/paper.Rmd
test

---
title: "Title Goes Here"
author:
  - Author One
  - Author Two
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(CTtps) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

# Introduction

Here is a citation [@Marwick2017]

# Background

```{r demo, eval = FALSE}
1 + 1
```

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
git2r::repository(here::here())
```

### 2020-03-03 README.Rmd
Copied and adjusted the code and data to use in the project

@@
 
-You can download the compendium as a zip from from this URL: </archive/master.zip>
+You can download the compendium as a zip from from this URL: <https://github.com/parkgayoung/CTtps/archive/master.zip>
 
 Or you can install this compendium as an R package, CTtps, from GitHub with:
+```{r gh-installation, eval = FALSE}
+# install.packages("devtools")
+devtools::install_github("parkgayoung/CTtps")
+```
 

### 2020-03-03 analysis/paper/paper.Rmd
Copied and adjusted the code and data to use in the project

@@
 
+```{r}
+#reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
+
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG 
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T)) 
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T)) 
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP   
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+
+  landmarks = vector("list", l) # create an empty list
+
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file = data, header=F, skip = LM[i],
+                 nrows = nrows[i], col.names = c("X", "Y")),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i]))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}
+
+#need to figure out how to read the file inside the project
+landmarks <- read.tps("/Users/gayoungp/CTtps/analysis/data/raw_data/A_BG_2.tps")
+
+#Understanding each landmark
+#ML: maximum length, tip to bottom, perpendicular to the length axis 1- 7
+#BL: body length: tip to the closest wing, perpendicular to the length axis 1- 3 OR 1-10 depending on the artifact
+#TL: Tang length, the closest wing from the tip to bottom, perpendicular to the length axis 9 - 7 depending on the artifact
+#SL: max stem length, perpendicular to the length axis. Closer tang curve’s middle point from the tip to the most distant point of the basal end  9-7 depending on the artifact
+#MW: mid width: dimension from margin to margin at the mid-point of the length 2 - 11 axis, perpendicular to the width axis.
+#TW: tang width: dimension between each wing, perpendicular to the width  axis  3 - 10
+#SW: stem width: width of the basal end of the point, 5mm above the end  5 - 8
+
+#distance between landmarks
+
+landmark_dist = function(landmarks, pt1, pt2) {
+  sqrt(
+    (landmarks$X[pt1] - landmarks$X[pt2])^2 + 
+      (landmarks$Y[pt1] - landmarks$Y[pt2])^2
+  ) * as.numeric(levels(landmarks$SCALE))
+}
+
+#Maximum length of sp: ML, distance between landmark 1-7
+ML <- landmark_dist(landmarks, 1, 7)
+
+#Body length of sp is 1-3 or 1-10, depending on each artifact.
+bl_dist = function(landmarks) {
+  d1 = landmark_dist(landmarks, 1, 3)
+  d2 = landmark_dist(landmarks, 1, 10)
+  max(d1, d2)
+}
+
+BL <- bl_dist(landmarks)
+
+#Compare multiple artifacts
+
+#Open multiple tps files 
+#This will need to be changed to read files inside the project
+datadir <- "/Users/gayoungp/CTtps/analysis/data/raw_data/"
+files <- dir(datadir, pattern = "*.tps")
+
+#gather ML and BL of all artifacts
+ML = c()
+BL = c()
+for (i in 1:length(files)) {
+  filename <- paste(datadir, files[i], sep = "")
+  print(filename)
+  landmarks <- read.tps(filename)
+  ML <- c(ML, landmark_dist(landmarks, 1, 7))
+  BL <- c(BL, bl_dist(landmarks))
+}
+
+#list of ML and BL
+df <- data.frame(ML, BL, files)
+
+#Making a plot for ML and BL
+plot(df$ML, df$BL)
+
+```
+
 # Results

### 2020-03-03 analysis/paper/paper.Rmd
edited

@@
     landmarks[i] = list(data.frame(
-      read.table(file = data, header=F, skip = LM[i],
+      read.table(file = data, header = F, skip = LM[i],
                  nrows = nrows[i], col.names = c("X", "Y")),
@@
 #Compare multiple artifacts 
-
 #Open multiple tps files 

### 2020-03-03 analysis/paper/paper.Rmd
test

@@
 author:
-  - Author One
+  -gayoung Park
   - Author Two

### 2020-03-03 analysis/paper/paper.Rmd
added here function

@@
 #need to figure out how to read the file inside the project
-landmarks <- read.tps("/Users/gayoungp/CTtps/analysis/data/raw_data/A_BG_2.tps")
+landmarks <- read.tps(here("analysis/data/raw_data/A_BG_2.tps"))
 
@@
 #This will need to be changed to read files inside the project
-datadir <- "/Users/gayoungp/CTtps/analysis/data/raw_data/"
+datadir <- here("/analysis/data/raw_data/")
 files <- dir(datadir, pattern = "*.tps")

### 2020-03-19 analysis/paper/paper.Rmd
tried CV and pearson's r

@@
 
-library(CTtps) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
+#library(CTtps) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
+
+library(here)
 ```
@@
 
+# Correlations with significance levels
+library(Hmisc)
+rcorr(ML,BL, type = "pearson") # type can be pearson or spearman
+
+# Coefficient of variation (CV)
+library(goeveg)
+CV_ML <- cv(ML, na.rm = FALSE) #na.rm: logical. Should missing values be removed?
+CV_BL <- cv(BL, na.rm = FALSE)
+
+plot(CV_ML, CV_BL)
 ```

### 2020-03-26 analysis/paper/paper.Rmd
distance btw y axis

@@
 
+# distance between two y axis 
+vertical_dist = function(landmarks, pt1, pt2) {
+  d = abs(landmarks$Y[pt1] - landmarks$Y[pt2])
+  d * as.numeric(levels(landmarks$SCALE))
+}
+
+# Tang length of sp: TL, distance between landmark 10-7 or 3-7, (y axis)
+tl_dist <- function(landmarks) {
+  t1 = vertical_dist(landmarks, 10, 7)
+  t2 = vertical_dist(landmarks, 3, 7)
+  max(t1, t2)
+}
+
+TL <- tl_dist
+
+# Maximum length of stem:
+
 #Compare multiple artifacts 
@@
 plot(CV_ML, CV_BL)
+```
+
+```{r}
+# artifacts from SYG6 site
+
 ```

### 2020-03-26 analysis/paper/paper.Rmd
push before updating my r

@@
 
-#distance between landmarks
+#distance between landmarks(euclidean)
 
@@
   ) * as.numeric(levels(landmarks$SCALE))
+}
+
+# distance between two y axis 
+vertical_dist = function(landmarks, pt1, pt2) {
+  d = abs(landmarks$Y[pt1] - landmarks$Y[pt2])
+  d * as.numeric(levels(landmarks$SCALE))
 }
@@
 
-# distance between two y axis 
-vertical_dist = function(landmarks, pt1, pt2) {
-  d = abs(landmarks$Y[pt1] - landmarks$Y[pt2])
-  d * as.numeric(levels(landmarks$SCALE))
-}
-
 # Tang length of sp: TL, distance between landmark 10-7 or 3-7, (y axis)
@@
 
-# Maximum length of stem:
+# Maximum length of stem: SL, distance between landmark 9-7 or 4-7 (y axis)
+sl_dist <- function(landmarks) {
+  s1 = vertical_dist(landmarks, 9, 7)
+  s2 = vertical_dist(landmarks, 4, 7)
+  max(s1, s2)
+}
+
+SL <- sl_dist
+
+# Mid width: MW, distance between landmark 2-11
+MW <- landmark_dist(landmarks, 2, 11)
+
+# Maximum width of Tang: TW, distance between landmark 3-10
+TW <- landmark_dist(landmarks, 3, 10)
+
+# Stem width: SW, distance between landmark 5-8
+SW <- landmark_dist(landmarks, 5, 8)
 
@@
 BL = c()
+TL = c()
+SL = c()
+MW = c()
+TW = c()
+SW = c()
 for (i in 1:length(files)) {
@@
   BL <- c(BL, bl_dist(landmarks))
+  TL <- c(TL, tl_dist(landmarks))
+  SL <- c(SL, sl_dist(landmarks))
+  MW <- c(MW, landmark_dist(landmarks, 2, 11))
+  TW <- c(TW, landmark_dist(landmarks, 3, 10))
+  SW <- c(SW, landmark_dist(landmarks, 5, 8))
 }
 
-#list of ML and BL
-df <- data.frame(ML, BL, files)
-
-#Making a plot for ML and BL
-plot(df$ML, df$BL)
+#list of all attributes
+df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
 

### 2020-03-26 analysis/paper/paper.Rmd
gather data from SYG6

@@
 ```{r}
+
 # artifacts from SYG6 site
+SYG6_files <- dir(datadir, pattern = "SYG6_*")
+
+ML = c()
+BL = c()
+TL = c()
+SL = c()
+MW = c()
+TW = c()
+SW = c()
+for (i in 1:length(SYG6_files)) {
+  filename <- paste(datadir, SYG6_files[i], sep = "")
+  print(filename)
+  landmarks <- read.tps(filename)
+  ML <- c(ML, landmark_dist(landmarks, 1, 7))
+  BL <- c(BL, bl_dist(landmarks))
+  TL <- c(TL, tl_dist(landmarks))
+  SL <- c(SL, sl_dist(landmarks))
+  MW <- c(MW, landmark_dist(landmarks, 2, 11))
+  TW <- c(TW, landmark_dist(landmarks, 3, 10))
+  SW <- c(SW, landmark_dist(landmarks, 5, 8))
+}
 

### 2020-03-27 analysis/paper/paper.Rmd
tried PCA

@@
 df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
+#list of all attributes without ID
+DF <- df[,1:7]
 
@@
 rcorr(ML,BL, type = "pearson") # type can be pearson or spearman
+
+#simpler way to calculate Correlataion, you can use
+cor(DF)
+
+#a scatterplot matrix of Correlation
+pairs(DF)
 
@@
 
-plot(CV_ML, CV_BL)
+#simpler way to calculate CV
+sd(ML)/mean(ML)
+
+#PCA 
+library(ggbiplot)
+PCA <- prcomp(DF, center = TRUE,scale. = TRUE)
+
+ggbiplot(PCA, obs.scale = 1, var.scale = 1,          ellipse = TRUE, circle = TRUE)+
+scale_color_discrete (name = '') +
+  theme(legend.direction = 'horizontal', legend.position = 'top')
+
+###I need to figure out how to group them
+
 ```

### 2020-03-27 analysis/paper/paper.Rmd
tried ggfortify package

@@
 
-###I need to figure out how to group them
+#https://towardsdatascience.com/principal-component-analysis-pca-101-using-r-361f4c53a9ff <- another way of PCA, relationship with two variables
+
+#using ggfortify package
+library(ggfortify) 
+autoplot(prcomp(DF))
+
+###I need to figure out how to group them  -- by phase?
 

### 2020-03-30 analysis/paper/paper.Rmd
SYG6 site
This is what I did during our last meeting on Friday

@@
 df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
+library(tidyverse)
+df_sitename <- df %>% 
+  mutate(files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  separate(files, into = c("sitename", "files"), sep = "_") %>% 
+  select(-files)
+
 #list of all attributes without ID
@@
 
+autoplot(prcomp(DF), data = df_sitename, colour = 'sitename')
+
 ggbiplot(PCA, obs.scale = 1, var.scale = 1,          ellipse = TRUE, circle = TRUE)+
-scale_color_discrete (name = '') +
+scale_color_discrete (name = '') + groups (df$files)
   theme(legend.direction = 'horizontal', legend.position = 'top')
@@
 }
+
+#list of all attributes
+df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
+#list of all attributes without ID
+DF <- df[,1:7]
+
+# Correlations with significance levels
+library(Hmisc)
+rcorr(ML,BL, type = "pearson") # type can be pearson or spearman
+
+#simpler way to calculate Correlataion, you can use
+cor(DF)
+
+#a scatterplot matrix of Correlation
+pairs(DF)
+
+#simpler way to calculate Correlataion, you can use
+cor(DF)
+
+#a scatterplot matrix of Correlation
+pairs(DF)
+
+#simpler way to calculate CV
+sd(ML)/mean(ML)
 

### 2020-03-30 analysis/paper/paper.Rmd
Split the chunks

@@
 library(here)
+
+library(tidyverse)
 ```
@@
 landmarks <- read.tps(here("analysis/data/raw_data/A_BG_2.tps"))
+
+#Compare multiple artifacts 
+#Open multiple tps files. This will need to be changed to read files inside the project
+datadir <- here("/analysis/data/raw_data/")
+files <- dir(datadir, pattern = "*.tps")
+```
+
+```{r}
+#attributes of the points by calculating distance between landmakrs
 
@@
 
-#Compare multiple artifacts 
-#Open multiple tps files 
-#This will need to be changed to read files inside the project
-datadir <- here("/analysis/data/raw_data/")
-files <- dir(datadir, pattern = "*.tps")
-
 #gather ML and BL of all artifacts
@@
 df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
-library(tidyverse)
+
 df_sitename <- df %>% 
@@
 DF <- df[,1:7]
+```
 
+Correlation, Coefficient or variation, PCA
+```{r}
 # Correlations with significance levels
@@
 ```
+
+SYG6 site data and analysis
 
@@
 
+I'm working on how to read oulines and how to use them for anlaysis
+
+```{r}
+# reading landmarks and outline in tps, code from Morpho package, adjust readallTPS function. 
+readalltps = function (file) 
+{
+    out = list()
+    exOut <- FALSE
+    noutline <- 0
+    input <- readLines(file)
+    LM <- grep("LM=", input)
+    LMstring <- input[LM]
+    nLM <- sapply(LMstring, strsplit, split = "=")
+    nLM <- unlist(nLM)
+    nLM <- as.integer(nLM[-which(nLM == "LM")])
+    nobs <- length(nLM)
+    ID <- grep("ID=", input)
+    IDstring <- input[ID]
+    nID <- sapply(IDstring, function(x) {
+        x <- gsub("=", "_", x)
+    })
+    nID <- gsub(" ", "", unlist(nID))
+    out$ID <- nID
+    outline <- grep("CURVES=", input)
+    if (length(outline) > 0) {
+        exOut <- TRUE
+        outlinestring <- input[outline]
+        noutline <- sapply(outlinestring, strsplit, split = "=")
+        noutline <- unlist(noutline)
+        noutline <- as.integer(noutline[-which(noutline == "CURVES")])
+        LMoutline <- (sapply(outline, function(x) {
+            x <- max(which(LM < x))
+        }))
+    }
+    LMdata <- list()
+    for (i in 1:nobs) {
+        if (nLM[i] > 0) {
+            LMdata[[i]] <- as.numeric(unlist(strsplit(unlist(input[c((LM[i] + 
+                1):(LM[i] + nLM[i]))]), split = " ")))
+            LMdata[[i]] <- matrix(LMdata[[i]], nLM[i], 2, byrow = TRUE)
+        }
+        else LMdata[[i]] <- NA
+    }
+    names(LMdata) <- nID
+    out$LM <- LMdata
+    if (exOut) {
+        outlineData <- list()
+        for (i in 1:nobs) {
+            if (i %in% LMoutline) {
+                i1 <- grep(i, LMoutline)
+                outlinetmp <- list()
+                ptr <- outline[i1] + 1
+                j <- 1
+                while (j <= noutline[i1]) {
+                  tmpnr <- as.integer(unlist(strsplit(input[ptr], 
+                    split = "="))[2])
+                  outlinetmp[[j]] <- matrix(as.numeric(unlist(strsplit(input[(ptr + 
+                    1):(ptr + tmpnr)], split = " "))), tmpnr, 
+                    2, byrow = TRUE)
+                  ptr <- ptr + tmpnr + 1
+                  j <- j + 1
+                }
+                outlineData[[i]] <- outlinetmp
+            }
+            else outlineData[[i]] <- NA
+        }
+        names(outlineData) <- nID
+        out$outlines <- outlineData
+    }
+    cat(paste("Read", nobs, "datasets with", sum(noutline), "outlines\n"))
+    return(out)
+}
+
+lm_trial <- readalltps(here("analysis/data/raw_data/A_BG_2.tps"))
+
+```
+
 # Results

### 2020-03-30 analysis/paper/paper.Rmd
try to read all files

--- 2020-03-30 analysis/paper/paper.Rmd
@@
 
+I'm working on how to read oulines and how to use them for anlaysis
+
+```{r}
+# reading landmarks and outline in tps, code from Morpho package, adjust readallTPS function. 
+readalltps = function(file) 
+{
+    out = list()
+    exOut <- FALSE
+    noutline <- 0
+    input <- readLines(file)
+    LM <- grep("LM=", input)
+    LMstring <- input[LM]
+    nLM <- sapply(LMstring, strsplit, split = "=")
+    nLM <- unlist(nLM)
+    nLM <- as.integer(nLM[-which(nLM == "LM")])
+    nobs <- length(nLM)
+    ID <- grep("ID=", input)
+    IDstring <- input[ID]
+    nID <- sapply(IDstring, function(x) {
+        x <- gsub("=", "_", x)
+    })
+    nID <- gsub(" ", "", unlist(nID))
+    out$ID <- nID
+    outline <- grep("CURVES=", input)
+    if (length(outline) > 0) {
+        exOut <- TRUE
+        outlinestring <- input[outline]
+        noutline <- sapply(outlinestring, strsplit, split = "=")
+        noutline <- unlist(noutline)
+        noutline <- as.integer(noutline[-which(noutline == "CURVES")])
+        LMoutline <- (sapply(outline, function(x) {
+            x <- max(which(LM < x))
+        }))
+    }
+    LMdata <- list()
+    for (i in 1:nobs) {
+        if (nLM[i] > 0) {
+            LMdata[[i]] <- as.numeric(unlist(strsplit(unlist(input[c((LM[i] + 
+                1):(LM[i] + nLM[i]))]), split = " ")))
+            LMdata[[i]] <- matrix(LMdata[[i]], nLM[i], 2, byrow = TRUE)
+        }
+        else LMdata[[i]] <- NA
+    }
+    names(LMdata) <- nID
+    out$LM <- LMdata
+    if (exOut) {
+        outlineData <- list()
+        for (i in 1:nobs) {
+            if (i %in% LMoutline) {
+                i1 <- grep(i, LMoutline)
+                outlinetmp <- list()
+                ptr <- outline[i1] + 1
+                j <- 1
+                while (j <= noutline[i1]) {
+                  tmpnr <- as.integer(unlist(strsplit(input[ptr], 
+                    split = "="))[2])
+                  outlinetmp[[j]] <- matrix(as.numeric(unlist(strsplit(input[(ptr + 
+                    1):(ptr + tmpnr)], split = " "))), tmpnr, 
+                    2, byrow = TRUE)
+                  ptr <- ptr + tmpnr + 1
+                  j <- j + 1
+                }
+                outlineData[[i]] <- outlinetmp
+            }
+            else outlineData[[i]] <- NA
+        }
+        names(outlineData) <- nID
+        out$outlines <- outlineData
+    }
+    cat(paste("Read", nobs, "datasets with", sum(noutline), "outlines\n"))
+    return(out)
+}
+
+#try to read all data(landmarks and outlines) from one file
+lm_trial <- readalltps(here("analysis/data/raw_data/A_BG_2.tps"))
+
+#read all files from raw_data
+
+lm_landmarks <- readalltps(here("analysis/data/raw_data"))
+
+lm_landmarks = c()
+for (i in 1:length(files)) {
+  filename <- paste(datadir, files[i], sep = "")
+  print(filename)
+  landmarks <- readalltps(filename)
+ lm_landmarks <- c(lm_landmarks, landmarks)
+}
+
+```
+
 # Results

### 2020-04-02 analysis/paper/paper.Rmd
added phase for the SPs

@@
 landmarks <- read.tps(here("analysis/data/raw_data/A_BG_2.tps"))
+
+#getting info for phase of each site
+SPstage <- read.csv(here("analysis/data/SPstage.csv"))
 
@@
 #list of all attributes
-df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
+df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files, SPstage$Stage)
 
@@
 
-autoplot(prcomp(DF), data = df_sitename, colour = 'sitename')
-
-ggbiplot(PCA, obs.scale = 1, var.scale = 1,          ellipse = TRUE, circle = TRUE)+
-scale_color_discrete (name = '') + groups (df$files)
-  theme(legend.direction = 'horizontal', legend.position = 'top')
+library(ggfortify) 
+autoplot(prcomp(DF), data = df_sitename, colour = 'SPstage.Stage')
 
 #https://towardsdatascience.com/principal-component-analysis-pca-101-using-r-361f4c53a9ff <- another way of PCA, relationship with two variables
-
-#using ggfortify package
-library(ggfortify) 
-autoplot(prcomp(DF))
 

### 2020-04-03 analysis/paper/paper.Rmd
trimmed the chuncks a bit

@@
 library(here)
-
 library(tidyverse)
@@
 
+Reading tps data and phase information of each artifact 
 ```{r}
@@
 
+Computing size attributes of the points by calculating distance between landmakrs
 ```{r}
-#attributes of the points by calculating distance between landmakrs
 
@@
 
-#gather ML and BL of all artifacts
+```
+
+Gather attributes from  all artifacts
+```{r}
+
 ML = c()
@@
 
+#trim the name of each artifact.
 df_sitename <- df %>% 
-  mutate(files = str_remove_all(files, "A_|1|2|.tps")) %>% 
-  separate(files, into = c("sitename", "files"), sep = "_") %>% 
-  select(-files)
+  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  separate (files, into = c("sitename", "files"), sep = "_") %>% 
+  select (-files)
 
@@
 
-Correlation, Coefficient or variation, PCA
+Computing Pearson's correlation 
 ```{r}
 # Correlations with significance levels
+# I can move the library(pkg) to the top with other libraries but I'm leaving here for my reference for a while
 library(Hmisc) 
@@
 
+```
+
+Computing Coefficient or variation
+```{r}
 # Coefficient of variation (CV)
@@
 CV_BL <- cv(BL, na.rm = FALSE)
+CV_TL <- cv(TL, na.rm = FALSE)
+CV_SL <- cv(SL, na.rm = FALSE)
+CV_MW <- cv(MW, na.rm = FALSE)
+CV_TW <- cv(TW, na.rm = FALSE)
+CV_SW <- cv(SW, na.rm = FALSE)
 
@@
 sd(ML)/mean(ML)
+```
 
-#PCA 
+Principal component analysis
+```{r}
 library(ggbiplot)
+
 PCA <- prcomp(DF, center = TRUE,scale. = TRUE)
 
+#group them by three phase
 library(ggfortify) 
-autoplot(prcomp(DF), data = df_sitename, colour = 'SPstage.Stage')
-
-#https://towardsdatascience.com/principal-component-analysis-pca-101-using-r-361f4c53a9ff <- another way of PCA, relationship with two variables
-
-###I need to figure out how to group them  -- by phase?
+autoplot(prcomp(DF), data = df_sitename, colour = as.factor(df_sitename$SPstage.Stage))
 

### 2020-04-03 analysis/paper/paper.Rmd
computing CVs

@@
 
-#simpler way to calculate CV
-sd(ML)/mean(ML)
+#simpler way to calculate CV (%)
+sd(ML)/mean(ML)*100
+sd(BL)/mean(BL)*100
+sd(TL)/mean(TL)*100
+sd(SL)/mean(SL)*100
+sd(MW)/mean(MW)*100
+sd(TW)/mean(TW)*100
+sd(SW)/mean(SW)*100
+
 ```
+CV for all attributes are distributed from 23-31%
 
@@
 
-#simpler way to calculate CV
-sd(ML)/mean(ML)
+# CV%
+sd(ML)/mean(ML)*100
+sd(BL)/mean(BL)*100
+sd(TL)/mean(TL)*100
+sd(SL)/mean(SL)*100
+sd(MW)/mean(MW)*100
+sd(TW)/mean(TW)*100
+sd(SW)/mean(SW)*100
 
+#CV ranges are 20-33% more broad...
 ```

### 2020-04-04 analysis/paper/paper.Rmd
Work with Ben

@@
 
-Here is a citation [@Marwick2017]
+- What are our research questions?
+
+- What is our archaeological background?
+
+- What is our conceptual framework?
+
+- What are our methods?
+
+- What are our results?
+
+- What is the answer to our questions??
 
@@
 
+Make a table of the CV for all variables on all artefacts from all sites
+
+```{r}
+# all artefacts from all sites
+map_df(DF, cv)
+```
+
+Make a table of CVs for all variable grouped by site
+
+```{r}
+cv_by_site_df  <- 
+df_sitename %>% 
+  select(-SPstage.Stage) %>% 
+  group_by(sitename) %>% 
+  add_tally() %>% 
+  filter( n > 1) %>% 
+  select(-n) %>% 
+  nest(-sitename) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
+
+cv_by_site_df_unnest <-   
+cv_by_site_df %>% 
+  unnest(cv_by_site)
+```
+
+Make a table of CVs for all variable grouped by phase
+```{r}
+cv_by_stage_df  <- 
+df_sitename %>% 
+  select(-sitename) %>% 
+  group_by(SPstage.Stage) %>% 
+  add_tally() %>% 
+  filter( n > 1) %>% 
+  select(-n) %>% 
+  nest(-SPstage.Stage) %>% 
+  mutate(cv_by_stage = map(data, ~map_df(.x, cv)))
+
+cv_by_stage_df_unnest <-   
+cv_by_stage_df %>% 
+  unnest(cv_by_stage)
+```
+
 Computing Pearson's correlation 
@@
 # Coefficient of variation (CV)
-library(goeveg)
+
+cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
+
+# we've got DF a data frame with cols as variables
+# rows and specimens, and values as dimensions
+# need to add a column of site ID
+map(DF, cv)
+
 CV_ML <- cv(ML, na.rm = FALSE) #na.rm: logical. Should missing values be removed?
@@
 
-#simpler way to calculate CV (%)
-sd(ML)/mean(ML)*100
+sd(DF$ML)/mean(DF$ML)*100
 sd(BL)/mean(BL)*100
@@
 
-PCA <- prcomp(DF, center = TRUE,scale. = TRUE)
+PCA <- prcomp(DF, center = TRUE, scale = TRUE)
 
@@
 library(ggfortify) 
-autoplot(prcomp(DF), data = df_sitename, colour = as.factor(df_sitename$SPstage.Stage))
+ap <- 
+autoplot(prcomp(DF, center = TRUE, scale = TRUE), 
+         data = df_sitename, 
+         colour = as.factor(df_sitename$SPstage.Stage))
 
+pca_col <- data.frame(PCA$x, phase = as.factor(df_sitename$SPstage.Stage))
+
+gg <- 
+ggplot(pca_col ,
+       aes(PC1,
+           PC2,
+       colour = phase)) +
+  geom_point()
+
+gg
+
+# remove stage 1 artefact
+xx_pca <- 
+df_sitename %>% 
+  filter(SPstage.Stage != 1) %>% 
+  select(-sitename) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1 <- data.frame(xx_pca$x,stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
+
+gg_stage <- 
+ggplot(xx_pca1 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+gg_stage
 ```
+
+We see some differences in artefact shape and size. Now we need to examins the PCA output numbers to describe in more details.
 
@@
 #read all files from raw_data
-
-lm_landmarks <- readalltps(here("analysis/data/raw_data"))
+tps_files <- list.files(here("analysis/data/raw_data"), full.names = TRUE)
+lm_landmarks <- map(tps_files, readalltps)
+names(lm_landmarks) <- list.files(here("analysis/data/raw_data"))
 

### 2020-04-10 analysis/paper/paper.Rmd
create 3d pca
tried plotly package
what to write

@@
 
-- What are our research questions?
+1. What are our research questions?
+- Understanding technological transition during the Late Pleistocene in Korea base on Cultural transmission. 
+- What are the social contexts led to the technological transition
 
-- What is our archaeological background?
+2. What is our archaeological background?
+-SP was started in Korean Peninsula -> spread to Japan
+-Why SP important, Why we care
+-Current explantion of SP dispersal: migration vs in situ, in between
+ The evidence of the two main models: proportion of SP in assemblage, Y chromosome (two groups of people), composition of assemblage (blade to flake ratio)
+ -> no explanation about any mechanism/process/driver for the new technology
+ -> lack of details of the lithic technology, human behavior
 
-- What is our conceptual framework?
+3. What is our conceptual framework?
+- Evolutionary concepts that links stone artifacts to human behavior
+- Cultural Evolution : the idea that cultural change constitutes an evolutionary process
+- Same general concept of biological evolution: genetic change in population that is inherited over generations. Apply this concept to artifacts or cultural phenomena
+- In this paper, I will use Cultural Transmission 
+- Cultural Transmssion: way that people learn and pass on information
+- Transmission biases
+- similar study (CT in stone artifacts) : Mesoudi&O'Brien, Bettinger&Eerkens, Garvey
 
-- What are our methods?
+4. What are our methods?
+- Framework: Bettinger&Eerkens
+- Two models: Guided variation and Model-based bias =>  Multi-point Origin (in situ) and Single-point Origin (migration model)
+- Test by Correlation and Variation (Pearson's r, PCA, Coefficient of Variation) 
+- data collection, sites, artifact #
+- attributes: landmarks (tps software)
 
-- What are our results?
+5. What are our results? 
+- overall change through time, grouping (chronology, typology)
+- correlation
+- PCA
+- CV
 
-- What is the answer to our questions??
+6. What is the answer to our questions??
+- low correlation + high variation? High correlation + low variation?
+- How was SP spread in Korea, did they have single orgin or multiple origins?
+
+7. What is limitation of this study?
+- Maybe applying CV for SP is not ideal. It can be depending on individual knapper's skill and raw materials
+- CV: lithis vs pottery
 
@@
 
-Make a table of the CV for all variables on all artefacts from all sites
-
-```{r}
-# all artefacts from all sites
-map_df(DF, cv)
-```
-
-Make a table of CVs for all variable grouped by site
-
-```{r}
-cv_by_site_df  <- 
-df_sitename %>% 
-  select(-SPstage.Stage) %>% 
-  group_by(sitename) %>% 
-  add_tally() %>% 
-  filter( n > 1) %>% 
-  select(-n) %>% 
-  nest(-sitename) %>% 
-  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
-
-cv_by_site_df_unnest <-   
-cv_by_site_df %>% 
-  unnest(cv_by_site)
-```
-
-Make a table of CVs for all variable grouped by phase
-```{r}
-cv_by_stage_df  <- 
-df_sitename %>% 
-  select(-sitename) %>% 
-  group_by(SPstage.Stage) %>% 
-  add_tally() %>% 
-  filter( n > 1) %>% 
-  select(-n) %>% 
-  nest(-SPstage.Stage) %>% 
-  mutate(cv_by_stage = map(data, ~map_df(.x, cv)))
-
-cv_by_stage_df_unnest <-   
-cv_by_stage_df %>% 
-  unnest(cv_by_stage)
-```
-
 Computing Pearson's correlation 
@@
 
+Make a table of the CV for all variables on all artefacts from all sites
+
+```{r}
+# all artefacts from all sites
+map_df(DF, cv)
+```
+
+Make a table of CVs for all variable grouped by site
+
+```{r}
+cv_by_site_df  <- 
+df_sitename %>% 
+  select(-SPstage.Stage) %>% 
+  group_by(sitename) %>% 
+  add_tally() %>% 
+  filter( n > 1) %>% 
+  select(-n) %>% 
+  nest(-sitename) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
+
+cv_by_site_df_unnest <-   
+cv_by_site_df %>% 
+  unnest(cv_by_site)
+```
+
+Make a table of CVs for all variable grouped by phase
+```{r}
+cv_by_stage_df  <- 
+df_sitename %>% 
+  select(-sitename) %>% 
+  group_by(SPstage.Stage) %>% 
+  add_tally() %>% 
+  filter( n > 1) %>% 
+  select(-n) %>% 
+  nest(-SPstage.Stage) %>% 
+  mutate(cv_by_stage = map(data, ~map_df(.x, cv)))
+
+cv_by_stage_df_unnest <-   
+cv_by_stage_df %>% 
+  unnest(cv_by_stage)
+```
+
 Principal component analysis
@@
 
+PCA for 3 factors
+```{r}
+library(plot3D)
+
+scatter3D(pca_col$PC1, pca_col$PC2, pca_col$PC3, bty = "g", pch = 18, col = gg.col(100), 
+          main = "PCA", xlab = "PC1",
+          ylab = "PC2", zlab = "PC3")
+
+library(plotly)
+plot_ly(pca_col, x = ~PC1, y = ~PC2, z = ~PC3, colors = "Dark2", size = I(120)) %>%  add_markers(color = ~phase)
+
+```
+
 We see some differences in artefact shape and size. Now we need to examins the PCA output numbers to describe in more details.
+
+PCA with corrplot package to understand the individual variables
+```{r}
+library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
+
+fviz_pca_var(PCA, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+
+```
 

### 2020-04-10 analysis/paper/paper.Rmd
add more text

@@
 - Framework: Bettinger&Eerkens
-- Two models: Guided variation and Model-based bias =>  Multi-point Origin (in situ) and Single-point Origin (migration model)
+- Two models: Guided variation and Model-based bias =>  Multi-point Origin (in situ or in between one) and Single-point Origin (migration model)
 - Test by Correlation and Variation (Pearson's r, PCA, Coefficient of Variation) 
@@
 - overall change through time, grouping (chronology, typology)
-- correlation
-- PCA
-- CV
+- correlation: postively correlated(?)
+- PCA: 2,3 phases are highly overlapped, high redundancy. The individual phase are scattered, low redundancy. Variables of Dimension 1 is positively correlated
+- CV:values are around 20(ML)-34(SL). Compared to 2nd phase, 3rd phase has slightly lower variation.
 
@@
 - low correlation + high variation? High correlation + low variation?
+- What are the criteria for correlation and variation
 - How was SP spread in Korea, did they have single orgin or multiple origins?
@@
 - Maybe applying CV for SP is not ideal. It can be depending on individual knapper's skill and raw materials
+- CV of one site (SYG6) - no big difference with the value of all data
 - CV: lithis vs pottery
+
+- couldn't include thickness and weight
 
@@
 library(Hmisc) 
-rcorr(ML,BL, type = "pearson") # type can be pearson or spearman
+rcorr(ML, BL, type = "pearson") # type can be pearson or spearman
 

### 2020-04-10 analysis/paper/paper.Rmd
add one more plot

@@
 
-Computing Pearson's correlation 
-```{r}
-# Correlations with significance levels
-# I can move the library(pkg) to the top with other libraries but I'm leaving here for my reference for a while
-library(Hmisc) 
-rcorr(ML, BL, type = "pearson") # type can be pearson or spearman
-
-#simpler way to calculate Correlataion, you can use
-cor(DF)
-
-#a scatterplot matrix of Correlation
-pairs(DF)
-
-```
-
 Computing Coefficient or variation
@@
   unnest(cv_by_stage)
+```
+
+Computing Pearson's correlation 
+```{r}
+library(Hmisc)
+COR <- rcorr(as.matrix(DF))
+
+#a simple function for formatting a correlation matrix into a table with 4 columns containing from (http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software)
+flattenCorrMatrix <- function(cormat, pmat) {
+  ut <- upper.tri(cormat)
+  data.frame(
+    row = rownames(cormat)[row(cormat)[ut]],
+    column = rownames(cormat)[col(cormat)[ut]],
+    cor  =(cormat)[ut],
+    p = pmat[ut]
+    )
+}
+
+flattenCorrMatrix(COR$r, COR$P)
+
+#simpler way to calculate Correlataion, you can use
+SCOR <- cor(DF)
+
+#a scatterplot matrix of Correlation
+pairs(DF)
+
+# Plot for SCOR
+library(corrplot)
+corrplot(SCOR, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+
+corrplot(COR$r, type="upper", order="hclust", 
+         p.mat = COR$P, sig.level = 0.01, insig = "blank")
 ```
@@
 DF <- df[,1:7]
-
-# Correlations with significance levels
-library(Hmisc)
-rcorr(ML,BL, type = "pearson") # type can be pearson or spearman
 

### 2020-04-10 analysis/paper/paper.Rmd
editing PCA

@@
 library(ggfortify)
-ap <- 
-autoplot(prcomp(DF, center = TRUE, scale = TRUE), 
-         data = df_sitename, 
-         colour = as.factor(df_sitename$SPstage.Stage))
-
-pca_col <- data.frame(PCA$x, phase = as.factor(df_sitename$SPstage.Stage))
-
-gg <- 
-ggplot(pca_col ,
-       aes(PC1,
-           PC2,
-       colour = phase)) +
-  geom_point()
-
-gg
 
@@
 
-xx_pca1 <- data.frame(xx_pca$x,stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
+xx_pca1 <- 
+  data.frame(xx_pca$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
 
@@
 gg_stage
+```
+
+```{r}
+
+library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
+
+fviz_pca_var(xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+
 ```

### 2020-04-13 analysis/paper/paper.Rmd
trimmed a bit of PCA

@@
   filter(SPstage.Stage != 1) %>% 
-  select(-sitename) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
   prcomp(., center = TRUE, scale = TRUE)
@@
 
-fviz_pca_var(xx_pca, col.var = "contrib",
+xx_pca2 <- 
+df_sitename %>% 
+  filter(SPstage.Stage != 1) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+fviz_pca_var(xx_pca2, col.var = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

### 2020-04-13 analysis/paper/paper.Rmd
commit before update my R...

@@
 Principal component analysis
+
+Principal Component Analysis is a technique to extract relevant information from a multivariate dataset and to express this information as a set of few new variables called principal components or dimensions [@cascalheira2018use].
+
+In archaeology, PCA has been often used for~
+
+We applied PCA to our dataset to identify correlations between variables that would represent the existence of patterns in the points, which could possilby be translated into more ~ , in addition to correlation analysis.
+
 ```{r}
@@
 
+Contribution of variables for each of first two PCA demensions
+```{r}
+PC1_v <- fviz_contrib(xx_pca, choice = "var", axes = 1, top = 7)
+PC2_v <- fviz_contrib(xx_pca, choice = "var", axes = 2, top = 7)
+
+grid.arrange(PC1_v, PC2_v, nrow =1)
+
+```
+
+We see some differences in artefact shape and size. Now we need to examins the PCA output numbers to describe in more details.
+
 ```{r}
@@
 
-xx_pca2 <- 
-df_sitename %>% 
-  filter(SPstage.Stage != 1) %>% 
-  select(-sitename, -SPstage.Stage) %>% 
-  prcomp(., center = TRUE, scale = TRUE)
-
-fviz_pca_var(xx_pca2, col.var = "contrib",
+# Color variables by the continuous variable
+fviz_pca_var(xx_pca, col.var = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
@@
 plot_ly(pca_col, x = ~PC1, y = ~PC2, z = ~PC3, colors = "Dark2", size = I(120)) %>%  add_markers(color = ~phase)
-
-```
-
-We see some differences in artefact shape and size. Now we need to examins the PCA output numbers to describe in more details.
-
-PCA with corrplot package to understand the individual variables
-```{r}
-library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
-
-fviz_pca_var(PCA, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
 

### 2020-04-14 analysis/paper/paper.Rmd
plots for 4 dimentions

@@
 
-Contribution of variables for each of first two PCA demensions
+Contribution of variables for each of four PCA demensions
 ```{r}
+library(gridExtra)
+
 PC1_v <- fviz_contrib(xx_pca, choice = "var", axes = 1, top = 7)
 PC2_v <- fviz_contrib(xx_pca, choice = "var", axes = 2, top = 7)
+PC3_v <- fviz_contrib(xx_pca, choice = "var", axes = 3, top = 7)
+PC4_v <- fviz_contrib(xx_pca, choice = "var", axes = 4, top = 7)
 
-grid.arrange(PC1_v, PC2_v, nrow =1)
+grid.arrange(PC1_v, PC2_v, PC3_v, PC4_v, nrow =2)
 

### 2020-04-16 analysis/paper/paper.Rmd
add a boxplot
working on CV plots

@@
 
+Box plot for each attributes
+```{r}
+boxplot(DF, main = "Variaiton for each attributes", col = "gold")
+```
+
 Computing Coefficient or variation
@@
 map(DF, cv)
-
-CV_ML <- cv(ML, na.rm = FALSE) #na.rm: logical. Should missing values be removed?
-CV_BL <- cv(BL, na.rm = FALSE)
-CV_TL <- cv(TL, na.rm = FALSE)
-CV_SL <- cv(SL, na.rm = FALSE)
-CV_MW <- cv(MW, na.rm = FALSE)
-CV_TW <- cv(TW, na.rm = FALSE)
-CV_SW <- cv(SW, na.rm = FALSE)
-
-sd(DF$ML)/mean(DF$ML)*100
-sd(BL)/mean(BL)*100
-sd(TL)/mean(TL)*100
-sd(SL)/mean(SL)*100
-sd(MW)/mean(MW)*100
-sd(TW)/mean(TW)*100
-sd(SW)/mean(SW)*100
 
@@
   unnest(cv_by_site)
+
 ```
@@
   unnest(cv_by_stage)
+```
+
+making box plot by phase
+```{r]}
+cv_boxplot <- cv_by_stage_df_unnest %>% 
+  select(-data)
+
 ```

### 2020-04-16 analysis/paper/paper.Rmd
add pca for different phase

@@
 making box plot by phase
-```{r]}
-cv_boxplot <- cv_by_stage_df_unnest %>% 
+```{r}
+cv_plot <- cv_by_stage_df_unnest %>% 
   select(-data)
+
+library(tidyr)
+cv_plot %>% 
+pivot_longer(cols = -SPstage.Stage, names_to = "group") %>% 
+ggplot(aes(x = group, y = value, fill = as.factor(SPstage.Stage))) + 
+geom_bar(stat="identity", position=position_dodge())
 
@@
 
+#plot for both stage
 gg_stage <- 
@@
 
-gg_stage
+# phase 3
+xx_pca3 <- 
+df_sitename %>% 
+  filter(SPstage.Stage != 2) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1_3 <- 
+  data.frame(xx_pca3$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=2]))
+
+ggplot(xx_pca1_3 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+# phase 2
+
+xx_pca2 <- 
+df_sitename %>% 
+  filter(SPstage.Stage != 3) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1_2 <- 
+  data.frame(xx_pca2$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=3]))
+
+ggplot(xx_pca1_2 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
 ```
@@
 
+# phase 2
+fviz_pca_var(xx_pca2, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+# phase 3
+fviz_pca_var(xx_pca3, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
 ```

### 2020-04-17 analysis/paper/paper.Rmd
add more plots

@@
 library(ggbiplot)
+library(ggfortify)
 
 PCA <- prcomp(DF, center = TRUE, scale = TRUE)
-
-#group them by three phase
-library(ggfortify)
 
@@
 
+# phase 2
+
+xx_pca2 <- 
+df_sitename %>% 
+  filter(SPstage.Stage == 2) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1_2 <- 
+  data.frame(xx_pca2$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage ==2]))
+
+ggplot(xx_pca1_2 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
 # phase 3
+
 xx_pca3 <- 
 df_sitename %>% 
-  filter(SPstage.Stage != 2) %>% 
+  filter(SPstage.Stage == 3) %>% 
   select(-sitename, -SPstage.Stage) %>% 
@@
   data.frame(xx_pca3$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=2]))
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage ==3]))
 
@@
 
-# phase 2
-
-xx_pca2 <- 
-df_sitename %>% 
-  filter(SPstage.Stage != 3) %>% 
-  select(-sitename, -SPstage.Stage) %>% 
-  prcomp(., center = TRUE, scale = TRUE)
-
-xx_pca1_2 <- 
-  data.frame(xx_pca2$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=3]))
-
-ggplot(xx_pca1_2 ,
-       aes(PC1,
-           PC2,
-       colour = stage)) +
-  geom_point() +
-  stat_ellipse()
 ```
@@
 # artifacts from SYG6 site
-SYG6_files <- dir(datadir, pattern = "SYG6_*")
+library(stringr)
+SYG6_df <- 
+  df %>% 
+  filter(str_detect(files, "SYG6"))
 
-ML = c()
-BL = c()
-TL = c()
-SL = c()
-MW = c()
-TW = c()
-SW = c()
-for (i in 1:length(SYG6_files)) {
-  filename <- paste(datadir, SYG6_files[i], sep = "")
-  print(filename)
-  landmarks <- read.tps(filename)
-  ML <- c(ML, landmark_dist(landmarks, 1, 7))
-  BL <- c(BL, bl_dist(landmarks))
-  TL <- c(TL, tl_dist(landmarks))
-  SL <- c(SL, sl_dist(landmarks))
-  MW <- c(MW, landmark_dist(landmarks, 2, 11))
-  TW <- c(TW, landmark_dist(landmarks, 3, 10))
-  SW <- c(SW, landmark_dist(landmarks, 5, 8))
-}
+SYG6_df_sitename <- SYG6_df %>% 
+  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  separate (files, into = c("sitename", "files"), sep = "_") %>% 
+  select (-files)
 
-#list of all attributes
-df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files)
 #list of all attributes without ID
-DF <- df[,1:7]
+SYG6_DF <- SYG6_df[,1:7]
 
 #simpler way to calculate Correlataion, you can use
-cor(DF)
+cor(SYG6_DF)
 
 #a scatterplot matrix of Correlation
-pairs(DF)
+pairs(SYG6_DF)
 
-#simpler way to calculate Correlataion, you can use
-cor(DF)
+# CV
+map_df(SYG6_DF, cv)
 
-#a scatterplot matrix of Correlation
-pairs(DF)
+SYG6_PCA <- prcomp(SYG6_DF, center = TRUE, scale = TRUE)
 
-# CV%
-sd(ML)/mean(ML)*100
-sd(BL)/mean(BL)*100
-sd(TL)/mean(TL)*100
-sd(SL)/mean(SL)*100
-sd(MW)/mean(MW)*100
-sd(TW)/mean(TW)*100
-sd(SW)/mean(SW)*100
+# remove stage 1 artefact
+SYG6_xx_pca <- 
+SYG6_df_sitename %>% 
+  filter(SPstage.Stage != 1) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
 
-#CV ranges are 20-33% more broad...
+SYG6_xx_pca1 <- 
+  data.frame(SYG6_xx_pca$x,
+             stage = as.factor( SYG6_df_sitename$SPstage.Stage[SYG6_df_sitename$SPstage.Stage !=1]))
+
+#plot for both stage
+SYG6_gg_stage <- 
+ggplot(SYG6_xx_pca1 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
 ```

### 2020-04-17 analysis/paper/paper.Rmd
add plots

@@
 
+fviz_pca_var(SYG6_xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+
+```
+
+SYG_1 site data and analysis
+
+```{r}
+
+# artifacts from SYG_1 site
+library(stringr)
+SYG_1df <- 
+  df %>% 
+  filter(str_detect(files, "SYG_1"))
+
+SYG_1df_sitename <- SYG_1df %>% 
+  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  separate (files, into = c("sitename", "files"), sep = "_") %>% 
+  select (-files)
+
+#list of all attributes without ID
+SYG_1DF <- SYG_1df[,1:7]
+
+#simpler way to calculate Correlataion, you can use
+cor(SYG_1DF)
+
+#a scatterplot matrix of Correlation
+pairs(SYG_1DF)
+
+# CV
+map_df(SYG_1DF, cv)
+
+SYG_1PCA <- prcomp(SYG_1DF, center = TRUE, scale = TRUE)
+
+# remove stage 1 artefact
+SYG_1xx_pca <- 
+SYG_1df_sitename %>% 
+  filter(SPstage.Stage != 1) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+SYG_1xx_pca1 <- 
+  data.frame(SYG_1xx_pca$x,
+             stage = as.factor( SYG_1df_sitename$SPstage.Stage[SYG_1df_sitename$SPstage.Stage !=1]))
+
+#plot for both stage
+SYG_1gg_stage <- 
+ggplot(SYG_1xx_pca1 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+fviz_pca_var(SYG_1xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+```
+
+SYG_4 site data and analysis
+
+```{r}
+
+# artifacts from SYG_4 site
+library(stringr)
+SYG_4df <- 
+  df %>% 
+  filter(str_detect(files, "SYG_4"))
+
+SYG_4df_sitename <- SYG_4df %>% 
+  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  separate (files, into = c("sitename", "files"), sep = "_") %>% 
+  select (-files)
+
+#list of all attributes without ID
+SYG_4DF <- SYG_4df[,1:7]
+
+#simpler way to calculate Correlataion, you can use
+cor(SYG_4DF)
+
+#a scatterplot matrix of Correlation
+pairs(SYG_4DF)
+
+# CV
+map_df(SYG_4DF, cv)
+
+SYG_4PCA <- prcomp(SYG_4DF, center = TRUE, scale = TRUE)
+
+# remove stage 1 artefact
+SYG_4xx_pca <- 
+SYG_4df_sitename %>% 
+  filter(SPstage.Stage != 1) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+SYG_4xx_pca1 <- 
+  data.frame(SYG_4xx_pca$x,
+             stage = as.factor( SYG_4df_sitename$SPstage.Stage[SYG_4df_sitename$SPstage.Stage !=1]))
+
+#plot for both stage
+SYG_4gg_stage <- 
+ggplot(SYG_4xx_pca1 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+fviz_pca_var(SYG_4xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
 ```

### 2020-04-17 analysis/paper/paper.Rmd
done with today

@@
 
+Make a plot by site
+```{r}
+cv_plot_site <- cv_by_site_df_unnest %>% 
+  select(-data)
+
+library(tidyr)
+cv_plot_site %>% 
+pivot_longer(cols = -sitename, names_to = "group") %>% 
+ggplot(aes(x = group, y = value, fill = as.factor(sitename))) + 
+geom_bar(stat="identity", position=position_dodge())
+
+```
+
 Make a table of CVs for all variable grouped by phase
@@
 
-making box plot by phase
+Make a plot by phase
 ```{r}
-cv_plot <- cv_by_stage_df_unnest %>% 
+cv_plot_stage <- cv_by_stage_df_unnest %>% 
   select(-data)
@@
 library(tidyr)
-cv_plot %>% 
+cv_plot_stage %>% 
 pivot_longer(cols = -SPstage.Stage, names_to = "group") %>% 

### 2020-04-17 analysis/paper/paper.Rmd
add more plots

@@
 
+SYG6_PC1_v <- fviz_contrib(SYG6_xx_pca, choice = "var", axes = 1, top = 7)
+SYG6_PC2_v <- fviz_contrib(SYG6_xx_pca, choice = "var", axes = 2, top = 7)
+
+grid.arrange(SYG6_PC1_v, SYG6_PC2_v, nrow =1)
+
 ```
@@
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+
+SYG_1PC1_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 1, top = 7)
+SYG_1PC2_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 2, top = 7)
+
+grid.arrange(SYG_1PC1_v, SYG_1PC2_v, nrow =1)
+
 ```
@@
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+
+SYG_4PC1_v <- fviz_contrib(SYG_4xx_pca, choice = "var", axes = 1, top = 7)
+SYG_4PC2_v <- fviz_contrib(SYG_4xx_pca, choice = "var", axes = 2, top = 7)
+
+grid.arrange(SYG_4PC1_v, SYG_4PC2_v, nrow =1)
 ```

### 2020-04-21 analysis/paper/paper.Rmd
not working...

@@
 #a scatterplot matrix of Correlation
-pairs(DF)
+ALL<-pairs(DF)
+
+# correlation per phase
+cor2 <- 
+df_sitename %>% 
+  filter(SPstage.Stage == 2) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  cor(.)
+
+ACOR2 <- pairs(cor2)
+
+cor3 <- 
+df_sitename %>% 
+  filter(SPstage.Stage == 3) %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  cor(.)
+
+ACOR3 <- pairs(cor3)
 
@@
 library(corrplot)
-corrplot(SCOR, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
 
@@
          p.mat = COR$P, sig.level = 0.01, insig = "blank")
+
+# correlation for all
+cplotall <- corrplot(SCOR, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+# for phase 2
+cplot2 <- corrplot(cor2, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+# for phase 3
+cplot3 <- corrplot(cor3, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+
+library(patchwork)
+
+library(gridExtra)
+
 ```

### 2020-04-21 analysis/paper/paper.Rmd
correlation for site

@@
 
+par(mfrow=c(1,3))
 # correlation for all
@@
 library(gridExtra)
+
+par(mfrow=c(2,1)) #doesn't work
+
+ACOR2 <- pairs(cor2)
+ACOR3 <- pairs(cor3)
 
@@
 #simpler way to calculate Correlataion, you can use
-cor(SYG6_DF)
+SYG6_cor <- cor(SYG6_DF)
 
@@
 
+cplot_SYG6 <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+
 # CV
@@
 
-SYG6_PCA <- prcomp(SYG6_DF, center = TRUE, scale = TRUE)
+SYG6_PCA <- prcomp(SYG6_cor, center = TRUE, scale = TRUE)
 
@@
 #simpler way to calculate Correlataion, you can use
-cor(SYG_1DF)
+SYG_1cor <- cor(SYG_1DF)
 
@@
 pairs(SYG_1DF)
+
+cplot_SYG_1 <- corrplot(SYG_1cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+
+# CV
+map_df(SYG6_DF, cv)
+
+SYG6_PCA <- prcomp(SYG6_cor, center = TRUE, scale = TRUE)
 
@@
 #simpler way to calculate Correlataion, you can use
-cor(SYG_4DF)
+SYG_4cor <- cor(SYG_4DF)
 
@@
 pairs(SYG_4DF)
+
+cplot_SYG_4 <- corrplot(SYG_4cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
 
@@
 grid.arrange(SYG_4PC1_v, SYG_4PC2_v, nrow =1)
+```
+
+Correlations for each site
+```{r}
+par(mfrow=c(1,3))
+cplot_SYG6 <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+cplot_SYG_1 <- corrplot(SYG_1cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+cplot_SYG_4 <- corrplot(SYG_4cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
 ```

### 2020-04-24 analysis/paper/paper.Rmd
trimmed raw data

@@
 #need to figure out how to read the file inside the project
-landmarks <- read.tps(here("analysis/data/raw_data/A_BG_2.tps"))
+landmarks <- read.tps(here("analysis/data/raw_data/BG_0_2.tps"))
 
@@
 df_sitename <- df %>% 
-  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  mutate (files = str_remove_all(files, "_.tps")) %>% 
   separate (files, into = c("sitename", "files"), sep = "_") %>% 
@@
 
-# artifacts from SYG6 site
+# artifacts from SYG6 site: now it has two layers, two phases
 library(stringr)
@@
 SYG6_df_sitename <- SYG6_df %>% 
-  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  mutate (files = str_remove_all(files, "_.tps")) %>% 
   separate (files, into = c("sitename", "files"), sep = "_") %>% 
@@
 
-# artifacts from SYG_1 site
+# artifacts from SYG1 site
 library(stringr)
@@
   df %>% 
-  filter(str_detect(files, "SYG_1"))
+  filter(str_detect(files, "SYG1"))
 
 SYG_1df_sitename <- SYG_1df %>% 
-  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
+  mutate (files = str_remove_all(files, "_.tps")) %>% 
   separate (files, into = c("sitename", "files"), sep = "_") %>% 
@@
          tl.col = "black", tl.srt = 45)
-
-# CV
-map_df(SYG6_DF, cv)
-
-SYG6_PCA <- prcomp(SYG6_cor, center = TRUE, scale = TRUE)
 
@@
 
-SYG_4 site data and analysis
-
-```{r}
-
-# artifacts from SYG_4 site
-library(stringr)
-SYG_4df <- 
-  df %>% 
-  filter(str_detect(files, "SYG_4"))
-
-SYG_4df_sitename <- SYG_4df %>% 
-  mutate (files = str_remove_all(files, "A_|1|2|.tps")) %>% 
-  separate (files, into = c("sitename", "files"), sep = "_") %>% 
-  select (-files)
-
-#list of all attributes without ID
-SYG_4DF <- SYG_4df[,1:7]
-
-#simpler way to calculate Correlataion, you can use
-SYG_4cor <- cor(SYG_4DF)
-
-#a scatterplot matrix of Correlation
-pairs(SYG_4DF)
-
-cplot_SYG_4 <- corrplot(SYG_4cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
-
-# CV
-map_df(SYG_4DF, cv)
-
-SYG_4PCA <- prcomp(SYG_4DF, center = TRUE, scale = TRUE)
-
-# remove stage 1 artefact
-SYG_4xx_pca <- 
-SYG_4df_sitename %>% 
-  filter(SPstage.Stage != 1) %>% 
-  select(-sitename, -SPstage.Stage) %>% 
-  prcomp(., center = TRUE, scale = TRUE)
-
-SYG_4xx_pca1 <- 
-  data.frame(SYG_4xx_pca$x,
-             stage = as.factor( SYG_4df_sitename$SPstage.Stage[SYG_4df_sitename$SPstage.Stage !=1]))
-
-#plot for both stage
-SYG_4gg_stage <- 
-ggplot(SYG_4xx_pca1 ,
-       aes(PC1,
-           PC2,
-       colour = stage)) +
-  geom_point() +
-  stat_ellipse()
-
-fviz_pca_var(SYG_4xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
-
-SYG_4PC1_v <- fviz_contrib(SYG_4xx_pca, choice = "var", axes = 1, top = 7)
-SYG_4PC2_v <- fviz_contrib(SYG_4xx_pca, choice = "var", axes = 2, top = 7)
-
-grid.arrange(SYG_4PC1_v, SYG_4PC2_v, nrow =1)
-```
-
 Correlations for each site
 ```{r}
-par(mfrow=c(1,3))
+par(mfrow=c(1,2))
 cplot_SYG6 <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
@@
          tl.col = "black", tl.srt = 45)
-cplot_SYG_4 <- corrplot(SYG_4cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
+
 ```

### 2020-04-24 analysis/paper/paper.Rmd
delete some unnecessary codes

@@
 
-SYG6 site data and analysis
+SYG6 site data and analysis: artifacts from SYG6 site: now it has two layers, but the same phase (2nd)
 
@@
 
-# artifacts from SYG6 site: now it has two layers, two phases
 library(stringr)
@@
 SYG6_df_sitename %>% 
-  filter(SPstage.Stage != 1) %>% 
   select(-sitename, -SPstage.Stage) %>% 
@@
   data.frame(SYG6_xx_pca$x,
-             stage = as.factor( SYG6_df_sitename$SPstage.Stage[SYG6_df_sitename$SPstage.Stage !=1]))
+             stage = as.factor( SYG6_df_sitename$SPstage.Stage))
 
@@
 
-SYG_1 site data and analysis
+SYG1 site data and analysis: artifacts from SYG1 site, which include two differnt locations(excavation pits) and the artifacts are from different phase.
+SYG1_2 : 3rd phase
+SYG1_1: 2nd phase
 
@@
 
-# artifacts from SYG1 site
 library(stringr)
-SYG_1df <- 
+SYG1_2df <- 
   df %>% 
-  filter(str_detect(files, "SYG1"))
+  filter(str_detect(files, "SYG1_2"))
 
-SYG_1df_sitename <- SYG_1df %>% 
-  mutate (files = str_remove_all(files, "_.tps")) %>% 
-  separate (files, into = c("sitename", "files"), sep = "_") %>% 
-  select (-files)
+library(dplyr)
+SYG1_2df_sitename <- SYG1_2df %>% 
+  mutate (files = str_remove_all(files, "_\\d+.tps")) %>% 
+  rename(sitename = files)
 
 #list of all attributes without ID
-SYG_1DF <- SYG_1df[,1:7]
+SYG1_2DF <- SYG1_2df[,1:7]
 
 #simpler way to calculate Correlataion, you can use
-SYG_1cor <- cor(SYG_1DF)
+SYG1_2cor <- cor(SYG1_2DF)
 
 #a scatterplot matrix of Correlation
-pairs(SYG_1DF)
+pairs(SYG1_2DF)
 
-cplot_SYG_1 <- corrplot(SYG_1cor, type = "upper", order = "hclust", 
+cplot_SYG1_2 <- corrplot(SYG1_2cor, type = "upper", order = "hclust", 
          tl.col = "black", tl.srt = 45)
@@
 # CV
-map_df(SYG_1DF, cv)
+map_df(SYG1_2DF, cv)
 
-SYG_1PCA <- prcomp(SYG_1DF, center = TRUE, scale = TRUE)
+SYG1_2PCA <- prcomp(SYG1_2DF, center = TRUE, scale = TRUE)
 
 # remove stage 1 artefact
-SYG_1xx_pca <- 
-SYG_1df_sitename %>% 
-  filter(SPstage.Stage != 1) %>% 
+SYG1_2xx_pca <- 
+SYG1_2df_sitename %>% 
   select(-sitename, -SPstage.Stage) %>% 
@@
 
-SYG_1xx_pca1 <- 
-  data.frame(SYG_1xx_pca$x,
-             stage = as.factor( SYG_1df_sitename$SPstage.Stage[SYG_1df_sitename$SPstage.Stage !=1]))
+SYG1_2xx_pca1 <- 
+  data.frame(SYG1_2xx_pca$x,
+             stage = as.factor(SYG1_2df_sitename$SPstage.Stage))
 
 #plot for both stage
-SYG_1gg_stage <- 
-ggplot(SYG_1xx_pca1 ,
+SYG1_2gg_stage <- 
+ggplot(SYG1_2xx_pca1 ,
        aes(PC1,
@@
 
-fviz_pca_var(SYG_1xx_pca, col.var = "contrib",
+fviz_pca_var(SYG1_2xx_pca, col.var = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
 
-SYG_1PC1_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 1, top = 7)
-SYG_1PC2_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 2, top = 7)
+SYG1_2PC1_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 1, top = 7)
+SYG1_2PC2_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 2, top = 7)
 
-grid.arrange(SYG_1PC1_v, SYG_1PC2_v, nrow =1)
+grid.arrange(SYG1_2PC1_v, SYG1_2PC2_v, nrow =1)
 

### 2020-04-24 analysis/paper/paper.Rmd
done coding!

@@
 ```{r}
-library(Hmisc)
-COR <- rcorr(as.matrix(DF))
-
-#a simple function for formatting a correlation matrix into a table with 4 columns containing from (http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software)
-flattenCorrMatrix <- function(cormat, pmat) {
-  ut <- upper.tri(cormat)
-  data.frame(
-    row = rownames(cormat)[row(cormat)[ut]],
-    column = rownames(cormat)[col(cormat)[ut]],
-    cor  =(cormat)[ut],
-    p = pmat[ut]
-    )
-}
-
-flattenCorrMatrix(COR$r, COR$P)
 
@@
 ACOR3 <- pairs(cor3)
+```
 
+```{r}
+#using corrr package
+library(tidyverse)
+library(corrr)
+
+corrr_all <- correlate(DF) %>% 
+  shave()
+
+call<-rplot(corrr_all) +
+  ggtitle("Correlation for all phase")
+
+correlate(DF) %>% 
+  network_plot()
+
+#phase 2
+
+DF2<-df_sitename %>% 
+  filter(SPstage.Stage == 2) %>% 
+  select(-sitename, -SPstage.Stage)
+
+corrr_2 <-correlate(DF2) %>% 
+  shave()
+
+ctwo <- rplot(corrr_2) +
+  ggtitle("Correlation for 2nd phase")
+
+correlate(DF2) %>% 
+  network_plot()
+
+#phase 3
+
+DF3<-df_sitename %>% 
+  filter(SPstage.Stage == 3) %>% 
+  select(-sitename, -SPstage.Stage)
+
+corrr_3 <-correlate(DF3) %>% 
+  shave()
+
+cthree <- rplot(corrr_3) +
+  ggtitle("Correlation for 3rd phase")
+
+correlate(DF3) %>% 
+  network_plot()
+
+### three phases at once
+library(patchwork)
+call + ctwo + cthree + plot_layout(ncol=1)
 ```
@@
 # Color variables by the continuous variable
-fviz_pca_var(xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+cv_all <- fviz_pca_var(xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="All")
 
 # phase 2
-fviz_pca_var(xx_pca2, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+cv_two <- fviz_pca_var(xx_pca2, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="Phase 2")
 # phase 3
-fviz_pca_var(xx_pca3, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+cv_three <- fviz_pca_var(xx_pca3, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="Phase 3")
+
+library(patchwork)
+
+cv_all + cv_two + cv_three + plot_layout(ncol=1)
+
 ```
@@
 
+# using corrr
+
+corrr_SYG6_DF <-correlate(SYG6_DF) %>% 
+  shave()
+
+cSYG6_DF <- rplot(corrr_SYG6_DF) +
+  ggtitle("Correlation for corrr_SYG6_DF, 2nd phase")
+
+correlate(SYG6_DF) %>% 
+  network_plot()
+
 # CV
@@
 
-fviz_pca_var(SYG6_xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+cv_SYG6 <-fviz_pca_var(SYG6_xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="SYG6:2nd")
 
@@
 ```{r}
-
+#SYG1_2
 library(stringr)
@@
          tl.col = "black", tl.srt = 45)
+
+# using corrr
+
+corrr_SYG1_2DF <-correlate(SYG1_2DF) %>% 
+  shave()
+
+cSYG1_2DF <- rplot(corrr_SYG1_2DF) +
+  ggtitle("Correlation for corrr_SYG1_2DF, 3rd phase")
+
+correlate(SYG1_2DF) %>% 
+  network_plot()
 
@@
 
-fviz_pca_var(SYG1_2xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
+cv_SYG1_2 <- fviz_pca_var(SYG1_2xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="SYG1_2:3rd")
 
@@
 
+```{r}
+#SYG1_1
+library(stringr)
+SYG1_1df <- 
+  df %>% 
+  filter(str_detect(files, "SYG1_1"))
+
+library(dplyr)
+SYG1_1df_sitename <- SYG1_1df %>% 
+  mutate (files = str_remove_all(files, "_\\d+.tps")) %>% 
+  rename(sitename = files)
+
+#list of all attributes without ID
+SYG1_1DF <- SYG1_1df[,1:7]
+
+#simpler way to calculate Correlataion, you can use
+SYG1_1cor <- cor(SYG1_1DF)
+
+#a scatterplot matrix of Correlation
+pairs(SYG1_1DF)
+
+cplot_SYG1_1 <- corrplot(SYG1_1cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45)
+
+# using corrr
+
+corrr_SYG1_1DF <-correlate(SYG1_1DF) %>% 
+  shave()
+
+cSYG1_1DF <- rplot(corrr_SYG1_1DF) +
+  ggtitle("Correlation for corrr_SYG1_1DF, 2nd phase")
+
+correlate(SYG1_1DF) %>% 
+  network_plot()
+
+# CV
+map_df(SYG1_1DF, cv)
+
+SYG1_1PCA <- prcomp(SYG1_1DF, center = TRUE, scale = TRUE)
+
+# remove stage 1 artefact
+SYG1_1xx_pca <- 
+  SYG1_1df_sitename %>% 
+  select(-sitename, -SPstage.Stage) %>% 
+  prcomp(., center = TRUE, scale = TRUE)
+
+SYG1_1xx_pca1 <- 
+  data.frame(SYG1_1xx_pca$x,
+             stage = as.factor(SYG1_1df_sitename$SPstage.Stage))
+
+#plot for both stage
+SYG1_1gg_stage <- 
+ggplot(SYG1_1xx_pca1 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+cv_SYG1_1 <-fviz_pca_var(SYG1_1xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="SYG1_1:2nd")
+
+SYG1_1PC1_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 1, top = 7)
+SYG1_1PC2_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 2, top = 7)
+
+grid.arrange(SYG1_1PC1_v, SYG1_1PC2_v, nrow =1)
+
+```
+
 Correlations for each site
 ```{r}
-par(mfrow=c(1,2))
-cplot_SYG6 <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
-cplot_SYG_1 <- corrplot(SYG_1cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
+par(mfrow=c(1,3))
+cplot_SYG6_ <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45, title = "SYG6")
+cplot_SYG1_1 <- corrplot(SYG1_1cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45, title = "SYG1_1")
+cplot_SYG1_2 <- corrplot(SYG1_2cor, type = "upper", order = "hclust", 
+         tl.col = "black", tl.srt = 45, title = "SYG1_2")
 
+### three sites at once, using corrr
+library(patchwork)
+cSYG6_DF + cSYG1_2DF + cSYG1_1DF+ plot_layout(ncol=1)
+
+```
+
+CV for each site
+```{r}
+cv_SYG6 + cv_SYG1_1 + cv_SYG1_2 +  plot_layout(ncol=1)
 ```

### 2020-04-28 analysis/paper/paper.Rmd
working on putting n=?

@@
 5. What are our results? 
-- overall change through time, grouping (chronology, typology)
-- correlation: postively correlated(?)
+- overall change through time
+- correlation: postively correlated(?). We are using PCA too
 - PCA: 2,3 phases are highly overlapped, high redundancy. The individual phase are scattered, low redundancy. Variables of Dimension 1 is positively correlated
@@
 - How was SP spread in Korea, did they have single orgin or multiple origins?
+- What are the impact of combinaton with blade technology
 
@@
 - couldn't include thickness and weight
+
+** Expectation
+- The main measurements are correlation(+PCA) and CV.
+- The set of the two measurements would be like: 
+1) lower CV and higher correlation between sites, in this case, we will think that SPs were the result of model based bias and possibly had single origin. 
+2) higher CV and lower correlation between sites, for this case, we will thank that SPs were the result of guided variation and possibly had multiple origins. 
+- We have three phases, but there are only two artifacts for the first phase.
+There are more artifacts and sites for the first phase but not eligible for this research because we gathered complete or almost complete shape of SPs.
+We excluded broken pieces.
+Therefore, we focus on the change from the 2nd phase to 3rd phase for understanding the process of SP's dispersal in the Korean Peninsula.
+- Change from the 2nd to 3rd
+1) less correlated: the two groups were not very related/locally developed 
+2) highly correlated: the two groups were related, had single source and then spread to other places
+- Site by site
+- Regional relationship? Province/cities
 
@@
 ```{r}
-boxplot(DF, main = "Variaiton for each attributes", col = "gold")
+box1 <- boxplot(DF, main = "Variation for each attributes", col = "gold")
 ```
+
+This boxplot indicates the overview of each attribute.
+We can see that ML and BL have wider variation compared to other attribtues.
+Since TL has relativelly less variable, ML variation is more depending on BL.
+Throughout time and region, dimension of SW is consistant.
 
@@
 # need to add a column of site ID
-map(DF, cv)
+cvmap <- map(DF, cv)
+
+#Same as map but better view for summary
+map_df(DF, cv)
+box2 <- boxplot(cvmap) ##Is there any better way to show this data??
 
 ```
-CV for all attributes are distributed from 23-31%
-
-Make a table of the CV for all variables on all artefacts from all sites
-
-```{r}
-# all artefacts from all sites
-map_df(DF, cv)
-```
+The coefficient of variation (CV) is the ratio of the standard deviation devided by mean. It shows the extent of variability from the mean. 
+CV for all attributes are distributed from 23-31%. Interestingly, SL and SW have high CV values (36.3 and 32.0, respectively). These attributes show lower variation on boxplot. <= need to think about
 
@@
 
-Make a plot by site
+Make a plot for CV by site
 ```{r}
@@
 ```
+The plot shows CV values for different attributes The color of the bars represent different sites.
+Among tehs sites, we can see SC (Sachang) site has significantly high CV values.
+SC has two artifacts, one is from 1st phase and the other one is 2nd phase.
+We assume that this time difference and few number of artifacts may relate to the high CV value.
+
+```{r}
+cv_plot_site %>% 
+pivot_longer(cols = -sitename, names_to = "group") %>% 
+ggplot(aes(x = sitename, y = value, fill = as.factor(group))) + 
+geom_bar(stat="identity", position=position_dodge())
+
+```
+Unlike the previous plot, in this one, the bars represent different attributes. You can see that SC has highest CV values of most attributes. Overall, SW values are very high.
+
+```{r}
+#how to put n into the label
+mtcars2 <- mtcars %>% 
+  group_by(cyl, am) %>% mutate(n = n()) %>% 
+  mutate(label = paste0(cyl,'\nN = ',n))
+
+ggplot(mtcars2,
+   aes(x = factor(label), y = mpg, color = factor(label))) + 
+  geom_point() + 
+   xlab('cyl') + 
+facet_wrap(~am, scales = 'free_x') +
+  theme(legend.position = "none")
+
+## try with CV
+ cv_by_site_df_label <- 
+df_sitename %>% 
+  select(-SPstage.Stage) %>% 
+  group_by(sitename) %>% 
+  add_tally() %>% 
+  filter( n > 1) %>% 
+   mutate(label = as.factor(paste0(sitename,'\nN = ',n))) %>%   select(-n) %>% 
+ nest(-sitename, -label) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
+
+cv_by_site_df_unnest_label <-   
+cv_by_site_df_label %>% 
+  unnest(cv_by_site, label)
+
+cv_plot_site_label <- cv_by_site_df_unnest_label %>% 
+  select(-data)
+
+library(tidyr)
+cv_plot_site %>% 
+pivot_longer(cols = -sitename, names_to = "group") %>% 
+ggplot(aes(x = sitename, y = value, fill = as.factor(group))) + 
+geom_bar(stat="identity", position=position_dodge())
+
+```
 
@@
   add_tally() %>% 
-  filter( n > 1) %>% 
-  select(-n) %>% 
+  filter(SPstage.Stage > 1) %>% 
   nest(-SPstage.Stage) %>% 
@@
   unnest(cv_by_stage)
-```
 
-Make a plot by phase
-```{r}
 cv_plot_stage <- cv_by_stage_df_unnest %>% 
-  select(-data)
+  select(-data, -n)
 
@@
 ```
+This plot shows the difference of CV in between 2nd and 3rd phase.
+For TL, both two phases have almost same values.
+For MW and TW, 3rd phase has lower values than the 2nd.
+The other 4 attributes show higher valeues in 3rd phase, which means less standardized in their shape during the latter period.
+
+By the way, Previous research (Bettinger&Eerkins, Garvey) focused on Width of the projectile points along with weight to understand cultural transmssion.
+We haven't figured out which attibutes would be more infortant than others yet.
 
@@
 cplotall <- corrplot(SCOR, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
+         tl.col = "black", tl.srt = 45, main = "Correlatoin for ALL")
 # for phase 2
 cplot2 <- corrplot(cor2, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
+         tl.col = "black", tl.srt = 45, main = "Correlatoin for Phase 2")
 # for phase 3
 cplot3 <- corrplot(cor3, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
+         tl.col = "black", tl.srt = 45, main = "Correlatoin for Phase 3")
 
-library(patchwork)
-
-library(gridExtra)
-
-par(mfrow=c(2,1)) #doesn't work
-
-ACOR2 <- pairs(cor2)
-ACOR3 <- pairs(cor3)
 ```
+=> We may need to delete this chunk since we made
 

### 2020-04-28 analysis/paper/paper.Rmd
suceed to put n=x

@@
 
-## try with CV
+## CV with site (n=X)
  cv_by_site_df_label <- 
@@
 cv_by_site_df_label %>% 
-  unnest(cv_by_site, label)
+  unnest(cv_by_site)
 
 cv_plot_site_label <- cv_by_site_df_unnest_label %>% 
-  select(-data)
+  select(-data) %>%
 
 library(tidyr)
-cv_plot_site %>% 
-pivot_longer(cols = -sitename, names_to = "group") %>% 
-ggplot(aes(x = sitename, y = value, fill = as.factor(group))) + 
+p<- cv_plot_site_label %>% 
+pivot_longer(cols = -c(sitename,label), names_to = "group") %>% 
+ggplot(aes(label, value, fill = as.factor(group))) + 
 geom_bar(stat="identity", position=position_dodge())
 
+p + labs(x= "Site", fill = "Attributes")
 ```

### 2020-04-28 analysis/paper/paper.Rmd
added some descriptions

@@
 ggplot(aes(x = group, y = value, fill = as.factor(SPstage.Stage))) + 
-geom_bar(stat="identity", position=position_dodge())
+geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")
 
@@
 ```
-=> We may need to delete this chunk since we made
+=> We may need to delete this chunk since we made new plots below using corrr package and same dataset.
 
@@
 ```
+When it comes to the correlations between seven attributes, BL-ML, TW-MW, TL-TW, and TL-SL are relatively close to each other.
+There is no negative correlation among the attributes.
+Compared to the 2nd phase, in 3rd phase, correlations of SL with TW and MW, and TL with TWm SL, and MW get stronger.
+On the other hand, the correlation of TW-SW becomes weeker.
+Overall, the correlation between attributes became stronger from 2nd to 3rd phase.
+
+Combined the two results of CV and correlation from 2nd to 3rd,
+
+CV: lower in MW and TW, but higher in other 5 attribtues.
+
+COR: stronger
 

### 2020-04-28 analysis/paper/paper.Rmd
done with coding!

@@
 
-Combined the two results of CV and correlation from 2nd to 3rd,
+Combined the two results of CV and correlation from 2nd to 3rd...
 
@@
 
-COR: stronger
+COR: overall correlation became stronger.
+
+=> This combination doesn't fit into the models.
 
@@
              stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
+
+## CV with site (n=X)
+label_phase <-
+   df_sitename %>% 
+   group_by(SPstage.Stage) %>% 
+   add_tally() %>% 
+   mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
+
+count_phase <- length(which(label_phase[9]>0))
+count_phase2 <- length(which(label_phase[9] == 2))
+count_phase3 <- length(which(label_phase[9] == 3))
+
+###
 
@@
 
-Contribution of variables for each of four PCA demensions
+Contribution of variables for each of four PCA demensions of ALL phases except for the first. 
 ```{r}
 library(gridExtra)
-
+library(factoextra)
 PC1_v <- fviz_contrib(xx_pca, choice = "var", axes = 1, top = 7)
@@
 cv_all <- fviz_pca_var(xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="All")
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) + ggtitle(paste("All", "(n =", count_phase,")"))
 
@@
 cv_two <- fviz_pca_var(xx_pca2, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="Phase 2")
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) + ggtitle(paste("Phase 2", "(n =", count_phase2,")"))
 # phase 3
 cv_three <- fviz_pca_var(xx_pca3, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="Phase 3")
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) + ggtitle(paste("Phase 3", "(n =", count_phase3,")"))
 

### 2020-04-28 analysis/paper/paper.Rmd
done with the title

@@
 
-PCA for 3 factors
+PCA for 3 factors. We don't think we need this for now. 
 ```{r}
@@
 
+# make n=element number 
+count_SYG6 <- length(which(SYG6_df$SPstage.Stage == 2))
+
 #simpler way to calculate Correlataion, you can use
@@
 cSYG6_DF <- rplot(corrr_SYG6_DF) +
-  ggtitle("Correlation for corrr_SYG6_DF, 2nd phase")
+  ggtitle(paste("Correlation SYG6: 2nd Phase", "(n =", count_SYG6,")"))
 
@@
 cv_SYG6 <-fviz_pca_var(SYG6_xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="SYG6:2nd")
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) +
+             ggtitle(paste("SYG6: 2nd Phase", "(n =", count_SYG6,")"))
 
@@
 
+# make n=element number 
+count_SYG1_2 <- length(which(SYG1_2df$SPstage.Stage == 3))
+
 #simpler way to calculate Correlataion, you can use
@@
 cSYG1_2DF <- rplot(corrr_SYG1_2DF) +
-  ggtitle("Correlation for corrr_SYG1_2DF, 3rd phase")
+  ggtitle(paste("Correlation SYG1_2: 3rd Phase", "(n =", count_SYG1_2,")"))
 
@@
 cv_SYG1_2 <- fviz_pca_var(SYG1_2xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="SYG1_2:3rd")
-
-SYG1_2PC1_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 1, top = 7)
-SYG1_2PC2_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 2, top = 7)
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) +
+  ggtitle(paste("SYG1_2: 3rd Phase", "(n =", count_SYG1_2,")"))
+
+SYG1_2PC1_v <- fviz_contrib(SYG1_2xx_pca, choice = "var", axes = 1, top = 7)
+SYG1_2PC2_v <- fviz_contrib(SYG1_2xx_pca, choice = "var", axes = 2, top = 7)
 
@@
 
+# make n=element number 
+count_SYG1_1 <- length(which(SYG1_1df$SPstage.Stage == 2))
+
 #simpler way to calculate Correlataion, you can use
@@
 cSYG1_1DF <- rplot(corrr_SYG1_1DF) +
-  ggtitle("Correlation for corrr_SYG1_1DF, 2nd phase")
+  ggtitle(paste("Correlation SYG1_1: 2nd Phase", "(n =", count_SYG1_1,")"))
 
@@
 
-cv_SYG1_1 <-fviz_pca_var(SYG1_1xx_pca, col.var = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), title ="SYG1_1:2nd")
-
-SYG1_1PC1_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 1, top = 7)
-SYG1_1PC2_v <- fviz_contrib(SYG_1xx_pca, choice = "var", axes = 2, top = 7)
+cv_SYG1_1 <- fviz_pca_var(SYG1_1xx_pca, col.var = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) +
+            ggtitle(paste("SYG1_1: 2nd Phase", "(n =", count_SYG1_1,")"))
+
+SYG1_1PC1_v <- fviz_contrib(SYG1_1xx_pca, choice = "var", axes = 1, top = 7)
+SYG1_1PC2_v <- fviz_contrib(SYG1_1xx_pca, choice = "var", axes = 2, top = 7)
 
@@
 library(patchwork)
-cSYG6_DF + cSYG1_2DF + cSYG1_1DF+ plot_layout(ncol=1)
+cSYG6_DF + cSYG1_2DF + cSYG1_1DF + plot_layout(ncol=1)
 
@@
 
-I'm working on how to read oulines and how to use them for anlaysis
+I'm working on how to read oulines and how to use them for anlaysis. We may not use this for the research.
 

### 2020-04-28 analysis/paper/paper.Rmd
add more description

@@
 ```
+Overall, BL and ML have stronger influence than the other attributes and are related to each other but not with the other attributes throughout time.
+From the phase 2 to the phase 3, the proportion of TL and SL gets bigger and they are more related to SW.
 
@@
 ```
-
-CV for each site
+In this plot, we can see that the pattern of correlation for SYG6 and SYG1_1 are similar in terms of correlation based on ML/SL/TL with other attributes.
+Both are part of the phase 2.
+However, SYG1_2 is younger than the other two and the correlation based on SW become weaker.
+When it comes to the difference between SYG6 and SYG1_1, the correlation between artifacts of SYG1_1 is stronger than SYG6.
+
+PCA for each site
 ```{r}
@@
 ```
-
-I'm working on how to read oulines and how to use them for anlaysis. We may not use this for the research.
-
-```{r}
+Unlike the correlation of the three units, PCA of the three are more various. Especially, SW is related to SL/TL in SYG6, related to ML/BL in SYG1_1, and unrelated to the other attributes in SYG1_2.
+
+```{r}
+#I'm working on how to read oulines and how to use them for anlaysis. We may not use this for the research.
 # reading landmarks and outline in tps, code from Morpho package, adjust readallTPS function. 

### 2020-04-28 analysis/paper/paper.Rmd
trimmed

@@
 ```{r}
-#how to put n into the label
-mtcars2 <- mtcars %>% 
-  group_by(cyl, am) %>% mutate(n = n()) %>% 
-  mutate(label = paste0(cyl,'\nN = ',n))
-
-ggplot(mtcars2,
-   aes(x = factor(label), y = mpg, color = factor(label))) + 
-  geom_point() + 
-   xlab('cyl') + 
-facet_wrap(~am, scales = 'free_x') +
-  theme(legend.position = "none")
-
 ## CV with site (n=X)

### 2020-04-28 analysis/paper/paper.Rmd
changed csv
fixed one error
tps or TPS, we can handle it

@@
 #getting info for phase of each site
-SPstage <- read.csv(here("analysis/data/SPstage.csv"))
+SPstage <- read_csv(here("analysis/data/SPstage.csv"))
 
@@
 datadir <- here("/analysis/data/raw_data/")
-files <- dir(datadir, pattern = "*.tps")
+files <- dir(datadir, pattern = "*.tps$|*.TPS$")
 ```

### 2020-04-28 analysis/paper/paper.Rmd
put ###

--- 2020-04-28 analysis/paper/paper.Rmd
@@
 cv_plot_site_label <- cv_by_site_df_unnest_label %>%
-  select(-data) %>%
+                select(-data)
 
@@
 library(corrplot)
-
-corrplot(COR$r, type="upper", order="hclust", 
-         p.mat = COR$P, sig.level = 0.01, insig = "blank")
 
@@
 
-PCA for 3 factors. We don't think we need this for now. 
-```{r}
-library(plot3D)
-
-scatter3D(pca_col$PC1, pca_col$PC2, pca_col$PC3, bty = "g", pch = 18, col = gg.col(100), 
-          main = "PCA", xlab = "PC1",
-          ylab = "PC2", zlab = "PC3")
-
-library(plotly)
-plot_ly(pca_col, x = ~PC1, y = ~PC2, z = ~PC3, colors = "Dark2", size = I(120)) %>%  add_markers(color = ~phase)
-
-```
-
 SYG6 site data and analysis: artifacts from SYG6 site: now it has two layers, but the same phase (2nd)
@@
 # reading landmarks and outline in tps, code from Morpho package, adjust readallTPS function.
-readalltps = function(file) 
+
+#readalltps = function(file) 
 {
-    out = list()
-    exOut <- FALSE
-    noutline <- 0
-    input <- readLines(file)
-    LM <- grep("LM=", input)
-    LMstring <- input[LM]
-    nLM <- sapply(LMstring, strsplit, split = "=")
-    nLM <- unlist(nLM)
-    nLM <- as.integer(nLM[-which(nLM == "LM")])
-    nobs <- length(nLM)
-    ID <- grep("ID=", input)
-    IDstring <- input[ID]
-    nID <- sapply(IDstring, function(x) {
-        x <- gsub("=", "_", x)
-    })
-    nID <- gsub(" ", "", unlist(nID))
-    out$ID <- nID
-    outline <- grep("CURVES=", input)
-    if (length(outline) > 0) {
-        exOut <- TRUE
-        outlinestring <- input[outline]
-        noutline <- sapply(outlinestring, strsplit, split = "=")
-        noutline <- unlist(noutline)
-        noutline <- as.integer(noutline[-which(noutline == "CURVES")])
-        LMoutline <- (sapply(outline, function(x) {
-            x <- max(which(LM < x))
-        }))
+   # out = list()
+   # exOut <- FALSE
+  # noutline <- 0
+  # input <- readLines(file)
+   # LM <- grep("LM=", input)
+  # LMstring <- input[LM]
+   # nLM <- sapply(LMstring, strsplit, split = "=")
+  #  nLM <- unlist(nLM)
+   # nLM <- as.integer(nLM[-which(nLM == "LM")])
+  #  nobs <- length(nLM)
+   # ID <- grep("ID=", input)
+  #  IDstring <- input[ID]
+   # nID <- sapply(IDstring, function(x) {
+   #     x <- gsub("=", "_", x)
+  #  })
+   # nID <- gsub(" ", "", unlist(nID))
+  #  out$ID <- nID
+   # outline <- grep("CURVES=", input)
+  #  if (length(outline) > 0) {
+   #     exOut <- TRUE
+    #    outlinestring <- input[outline]
+     #   noutline <- sapply(outlinestring, strsplit, split = "=")
+      #  noutline <- unlist(noutline)
+      #  noutline <- as.integer(noutline[-which(noutline == "CURVES")])
+    #    LMoutline <- (sapply(outline, function(x) {
+   #         x <- max(which(LM < x))  })) }
+ #   LMdata <- list()
+  #  for (i in 1:nobs) {
+  #      if (nLM[i] > 0) {
+   #         LMdata[[i]] <- as.numeric(unlist(strsplit(unlist(input[c((LM[i] + 
+      #          1):(LM[i] + nLM[i]))]), split = " ")))
+       #     LMdata[[i]] <- matrix(LMdata[[i]], nLM[i], 2, byrow = TRUE)
         }
-    LMdata <- list()
-    for (i in 1:nobs) {
-        if (nLM[i] > 0) {
-            LMdata[[i]] <- as.numeric(unlist(strsplit(unlist(input[c((LM[i] + 
-                1):(LM[i] + nLM[i]))]), split = " ")))
-            LMdata[[i]] <- matrix(LMdata[[i]], nLM[i], 2, byrow = TRUE)
-        }
-        else LMdata[[i]] <- NA
-    }
-    names(LMdata) <- nID
-    out$LM <- LMdata
-    if (exOut) {
-        outlineData <- list()
-        for (i in 1:nobs) {
-            if (i %in% LMoutline) {
-                i1 <- grep(i, LMoutline)
-                outlinetmp <- list()
-                ptr <- outline[i1] + 1
-                j <- 1
-                while (j <= noutline[i1]) {
-                  tmpnr <- as.integer(unlist(strsplit(input[ptr], 
-                    split = "="))[2])
-                  outlinetmp[[j]] <- matrix(as.numeric(unlist(strsplit(input[(ptr + 
-                    1):(ptr + tmpnr)], split = " "))), tmpnr, 
-                    2, byrow = TRUE)
-                  ptr <- ptr + tmpnr + 1
-                  j <- j + 1
-                }
-                outlineData[[i]] <- outlinetmp
-            }
-            else outlineData[[i]] <- NA
-        }
-        names(outlineData) <- nID
-        out$outlines <- outlineData
-    }
-    cat(paste("Read", nobs, "datasets with", sum(noutline), "outlines\n"))
-    return(out)
-}
+   #     else LMdata[[i]] <- NA }
+   # names(LMdata) <- nID
+  #  out$LM <- LMdata
+   # if (exOut) {
+    #    outlineData <- list()
+     #   for (i in 1:nobs) {
+      #      if (i %in% LMoutline) {
+       #         i1 <- grep(i, LMoutline)
+        #        outlinetmp <- list()
+         #       ptr <- outline[i1] + 1
+          #      j <- 1
+           #     while (j <= noutline[i1]) {
+            #      tmpnr <- as.integer(unlist(strsplit(input[ptr], 
+        #            split = "="))[2])
+         #         outlinetmp[[j]] <- matrix(as.numeric(unlist(strsplit(input[(ptr + 
+         #           1):(ptr + tmpnr)], split = " "))), tmpnr, 
+          #          2, byrow = TRUE)
+           #       ptr <- ptr + tmpnr + 1
+            #      j <- j + 1  }
+             #   outlineData[[i]] <- outlinetmp  }
+          #  else outlineData[[i]] <- NA }
+       # names(outlineData) <- nID
+      #  out$outlines <- outlineData }
+ #   cat(paste("Read", nobs, "datasets with", sum(noutline), "outlines\n"))
+#    return(out)}
 
 #try to read all data(landmarks and outlines) from one file
-lm_trial <- readalltps(here("analysis/data/raw_data/A_BG_2.tps"))
+# lm_trial <- readalltps(here("analysis/data/raw_data/A_BG_2.tps"))
 
 #read all files from raw_data
-tps_files <- list.files(here("analysis/data/raw_data"), full.names = TRUE)
-lm_landmarks <- map(tps_files, readalltps)
-names(lm_landmarks) <- list.files(here("analysis/data/raw_data"))
-
-lm_landmarks = c()
-for (i in 1:length(files)) {
-  filename <- paste(datadir, files[i], sep = "")
-  print(filename)
-  landmarks <- readalltps(filename)
- lm_landmarks <- c(lm_landmarks, landmarks)
-}
+# tps_files <- list.files(here("analysis/data/raw_data"), full.names = TRUE)
+# lm_landmarks <- map(tps_files, readalltps)
+# names(lm_landmarks) <- list.files(here("analysis/data/raw_data"))
+
+#lm_landmarks = c()
+#for (i in 1:length(files)) {
+#  filename <- paste(datadir, files[i], sep = "")
+ # print(filename)
+#  landmarks <- readalltps(filename)
+# lm_landmarks <- c(lm_landmarks, landmarks)}
 

### 2020-04-28 analysis/paper/paper.Rmd
fix namespace clash

@@
   mutate (files = str_remove_all(files, "_\\d+.tps")) %>% 
-  rename(sitename = files)
+  dplyr::rename(sitename = files)
 

### 2020-05-05 analysis/paper/paper.Rmd
Chuck out the plots

@@
 Reading tps data and phase information of each artifact 
-```{r}
+```{r, echo = FALSE}
 #reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
@@
 
-```
-
-Make a plot for CV by site
-```{r}
 cv_plot_site <- cv_by_site_df_unnest %>% 
@@
 We assume that this time difference and few number of artifacts may relate to the high CV value.
-
-```{r}
-cv_plot_site %>% 
-pivot_longer(cols = -sitename, names_to = "group") %>% 
-ggplot(aes(x = sitename, y = value, fill = as.factor(group))) + 
-geom_bar(stat="identity", position=position_dodge())
-
-```
-Unlike the previous plot, in this one, the bars represent different attributes. You can see that SC has highest CV values of most attributes. Overall, SW values are very high.
 
@@
 ```
+Unlike the previous plot, in this one, the bars represent different attributes. You can see that SC has highest CV values of most attributes. Overall, SW values are very high.
 
@@
 
-Computing Pearson's correlation 
-```{r}
-
-#simpler way to calculate Correlataion, you can use
-SCOR <- cor(DF)
-
-#a scatterplot matrix of Correlation
-ALL<-pairs(DF)
-
-# correlation per phase
-cor2 <- 
-df_sitename %>% 
-  filter(SPstage.Stage == 2) %>% 
-  select(-sitename, -SPstage.Stage) %>% 
-  cor(.)
-
-ACOR2 <- pairs(cor2)
-
-cor3 <- 
-df_sitename %>% 
-  filter(SPstage.Stage == 3) %>% 
-  select(-sitename, -SPstage.Stage) %>% 
-  cor(.)
-
-ACOR3 <- pairs(cor3)
-
-# Plot for SCOR
-library(corrplot)
-
-par(mfrow=c(1,3))
-# correlation for all
-cplotall <- corrplot(SCOR, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45, main = "Correlatoin for ALL")
-# for phase 2
-cplot2 <- corrplot(cor2, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45, main = "Correlatoin for Phase 2")
-# for phase 3
-cplot3 <- corrplot(cor3, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45, main = "Correlatoin for Phase 3")
-
-```
-=> We may need to delete this chunk since we made new plots below using corrr package and same dataset.
-
 ```{r}
@@
 
-correlate(DF) %>% 
+direction_all <- correlate(DF) %>% 
   network_plot()
@@
 
-correlate(DF2) %>% 
+direction_2 <- correlate(DF2) %>% 
   network_plot()
@@
 
-correlate(DF3) %>% 
+direction_3 <-correlate(DF3) %>% 
   network_plot()
@@
 call + ctwo + cthree + plot_layout(ncol=1)
+
+direction_all+ direction_2 + direction_3 + plot_layout(ncol=1)
 ```
@@
 
-#simpler way to calculate Correlataion, you can use
-SYG6_cor <- cor(SYG6_DF)
-
-#a scatterplot matrix of Correlation
-pairs(SYG6_DF)
-
-cplot_SYG6 <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
-
 # using corrr
@@
   network_plot()
+
+SYG6_cor <- cor(SYG6_DF)
 
@@
 
-#simpler way to calculate Correlataion, you can use
-SYG1_2cor <- cor(SYG1_2DF)
-
-#a scatterplot matrix of Correlation
-pairs(SYG1_2DF)
-
-cplot_SYG1_2 <- corrplot(SYG1_2cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
-
 # using corrr
@@
   network_plot()
+
+SYG1_2cor <- cor(SYG1_2DF)
 
@@
 
-#simpler way to calculate Correlataion, you can use
-SYG1_1cor <- cor(SYG1_1DF)
-
-#a scatterplot matrix of Correlation
-pairs(SYG1_1DF)
-
-cplot_SYG1_1 <- corrplot(SYG1_1cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45)
-
 # using corrr
@@
   network_plot()
+
+SYG1_1cor <- cor(SYG1_1DF)
 
@@
 ```{r}
-par(mfrow=c(1,3))
-cplot_SYG6_ <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45, title = "SYG6")
-cplot_SYG1_1 <- corrplot(SYG1_1cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45, title = "SYG1_1")
-cplot_SYG1_2 <- corrplot(SYG1_2cor, type = "upper", order = "hclust", 
-         tl.col = "black", tl.srt = 45, title = "SYG1_2")
 
@@
 Unlike the correlation of the three units, PCA of the three are more various. Especially, SW is related to SL/TL in SYG6, related to ML/BL in SYG1_1, and unrelated to the other attributes in SYG1_2.
-
-```{r}
-#I'm working on how to read oulines and how to use them for anlaysis. We may not use this for the research.
-# reading landmarks and outline in tps, code from Morpho package, adjust readallTPS function.
-
-#readalltps = function(file) 
-{
-   # out = list()
-   # exOut <- FALSE
-  # noutline <- 0
-  # input <- readLines(file)
-   # LM <- grep("LM=", input)
-  # LMstring <- input[LM]
-   # nLM <- sapply(LMstring, strsplit, split = "=")
-  #  nLM <- unlist(nLM)
-   # nLM <- as.integer(nLM[-which(nLM == "LM")])
-  #  nobs <- length(nLM)
-   # ID <- grep("ID=", input)
-  #  IDstring <- input[ID]
-   # nID <- sapply(IDstring, function(x) {
-   #     x <- gsub("=", "_", x)
-  #  })
-   # nID <- gsub(" ", "", unlist(nID))
-  #  out$ID <- nID
-   # outline <- grep("CURVES=", input)
-  #  if (length(outline) > 0) {
-   #     exOut <- TRUE
-    #    outlinestring <- input[outline]
-     #   noutline <- sapply(outlinestring, strsplit, split = "=")
-      #  noutline <- unlist(noutline)
-      #  noutline <- as.integer(noutline[-which(noutline == "CURVES")])
-    #    LMoutline <- (sapply(outline, function(x) {
-   #         x <- max(which(LM < x))  })) }
- #   LMdata <- list()
-  #  for (i in 1:nobs) {
-  #      if (nLM[i] > 0) {
-   #         LMdata[[i]] <- as.numeric(unlist(strsplit(unlist(input[c((LM[i] + 
-      #          1):(LM[i] + nLM[i]))]), split = " ")))
-       #     LMdata[[i]] <- matrix(LMdata[[i]], nLM[i], 2, byrow = TRUE)
-        }
-   #     else LMdata[[i]] <- NA }
-   # names(LMdata) <- nID
-  #  out$LM <- LMdata
-   # if (exOut) {
-    #    outlineData <- list()
-     #   for (i in 1:nobs) {
-      #      if (i %in% LMoutline) {
-       #         i1 <- grep(i, LMoutline)
-        #        outlinetmp <- list()
-         #       ptr <- outline[i1] + 1
-          #      j <- 1
-           #     while (j <= noutline[i1]) {
-            #      tmpnr <- as.integer(unlist(strsplit(input[ptr], 
-        #            split = "="))[2])
-         #         outlinetmp[[j]] <- matrix(as.numeric(unlist(strsplit(input[(ptr + 
-         #           1):(ptr + tmpnr)], split = " "))), tmpnr, 
-          #          2, byrow = TRUE)
-           #       ptr <- ptr + tmpnr + 1
-            #      j <- j + 1  }
-             #   outlineData[[i]] <- outlinetmp  }
-          #  else outlineData[[i]] <- NA }
-       # names(outlineData) <- nID
-      #  out$outlines <- outlineData }
- #   cat(paste("Read", nobs, "datasets with", sum(noutline), "outlines\n"))
-#    return(out)}
-
-#try to read all data(landmarks and outlines) from one file
-# lm_trial <- readalltps(here("analysis/data/raw_data/A_BG_2.tps"))
-
-#read all files from raw_data
-# tps_files <- list.files(here("analysis/data/raw_data"), full.names = TRUE)
-# lm_landmarks <- map(tps_files, readalltps)
-# names(lm_landmarks) <- list.files(here("analysis/data/raw_data"))
-
-#lm_landmarks = c()
-#for (i in 1:length(files)) {
-#  filename <- paste(datadir, files[i], sep = "")
- # print(filename)
-#  landmarks <- readalltps(filename)
-# lm_landmarks <- c(lm_landmarks, landmarks)}
-
-```
 

### 2020-05-05 notes.Rmd
Chuck out the plots

---
title: "notes"
author: "Gayoung Park"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

I moved the leftover codes from paper, just in case I need them later.

```{r}

cv_plot_site %>% 
pivot_longer(cols = -sitename, names_to = "group") %>% 
ggplot(aes(x = sitename, y = value, fill = as.factor(group))) + 
geom_bar(stat="identity", position=position_dodge())

## I coded new one with N=artifact. 
```

Computing Pearson's correlation 
```{r}

#simpler way to calculate Correlataion, you can use
SCOR <- cor(DF)

#a scatterplot matrix of Correlation
ALL<-pairs(DF)

# correlation per phase
cor2 <- 
df_sitename %>% 
  filter(SPstage.Stage == 2) %>% 
  select(-sitename, -SPstage.Stage) %>% 
  cor(.)

ACOR2 <- pairs(cor2)

cor3 <- 
df_sitename %>% 
  filter(SPstage.Stage == 3) %>% 
  select(-sitename, -SPstage.Stage) %>% 
  cor(.)

ACOR3 <- pairs(cor3)

# Plot for SCOR
library(corrplot)

par(mfrow=c(1,3))
# correlation for all
cplotall <- corrplot(SCOR, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, main = "Correlatoin for ALL")
# for phase 2
cplot2 <- corrplot(cor2, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, main = "Correlatoin for Phase 2")
# for phase 3
cplot3 <- corrplot(cor3, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, main = "Correlatoin for Phase 3")

```

```{R}

#simpler way to calculate Correlataion, you can use
SYG6_cor <- cor(SYG6_DF)

#a scatterplot matrix of Correlation
pairs(SYG6_DF)

cplot_SYG6 <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

#simpler way to calculate Correlataion, you can use
SYG1_2cor <- cor(SYG1_2DF)

#a scatterplot matrix of Correlation
pairs(SYG1_2DF)

cplot_SYG1_2 <- corrplot(SYG1_2cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

#simpler way to calculate Correlataion, you can use
SYG1_1cor <- cor(SYG1_1DF)

#a scatterplot matrix of Correlation
pairs(SYG1_1DF)

cplot_SYG1_1 <- corrplot(SYG1_1cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

par(mfrow=c(1,3))
cplot_SYG6_ <- corrplot(SYG6_cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, title = "SYG6")
cplot_SYG1_1 <- corrplot(SYG1_1cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, title = "SYG1_1")
cplot_SYG1_2 <- corrplot(SYG1_2cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, title = "SYG1_2")

```

```{r}
#I'm working on how to read oulines and how to use them for anlaysis. We may not use this for the research.
# reading landmarks and outline in tps, code from Morpho package, adjust readallTPS function.

readalltps = function(file) 
{
   out = list()
   exOut <- FALSE
  noutline <- 0
  input <- readLines(file)
   LM <- grep("LM=", input)
  LMstring <- input[LM]
   nLM <- sapply(LMstring, strsplit, split = "=")
  nLM <- unlist(nLM)
   nLM <- as.integer(nLM[-which(nLM == "LM")])
   nobs <- length(nLM)
   ID <- grep("ID=", input)
   IDstring <- input[ID]
    nID <- sapply(IDstring, function(x) {
       x <- gsub("=", "_", x)
   })
   nID <- gsub(" ", "", unlist(nID))
   out$ID <- nID
   outline <- grep("CURVES=", input)
   if (length(outline) > 0) {
       exOut <- TRUE
       outlinestring <- input[outline]
      noutline <- sapply(outlinestring, strsplit, split = "=")
      noutline <- unlist(noutline)
       noutline <- as.integer(noutline[-which(noutline == "CURVES")])
       LMoutline <- (sapply(outline, function(x) {
           x <- max(which(LM < x))  })) }
  LMdata <- list()
   for (i in 1:nobs) {
       if (nLM[i] > 0) {
           LMdata[[i]] <- as.numeric(unlist(strsplit(unlist(input[c((LM[i] + 
               1):(LM[i] + nLM[i]))]), split = " ")))
           LMdata[[i]] <- matrix(LMdata[[i]], nLM[i], 2, byrow = TRUE)
        }
      else LMdata[[i]] <- NA }
    names(LMdata) <- nID
    out$LM <- LMdata
    if (exOut) {
        outlineData <- list()
       for (i in 1:nobs) {
            if (i %in% LMoutline) {
                i1 <- grep(i, LMoutline)
               outlinetmp <- list()
               ptr <- outline[i1] + 1
               j <- 1
                while (j <= noutline[i1]) {
                 tmpnr <- as.integer(unlist(strsplit(input[ptr], 
                   split = "="))[2])
                  outlinetmp[[j]] <- matrix(as.numeric(unlist(strsplit(input[(ptr + 
                  1):(ptr + tmpnr)], split = " "))), tmpnr, 
                   2, byrow = TRUE)
                 ptr <- ptr + tmpnr + 1
                 j <- j + 1  }
              outlineData[[i]] <- outlinetmp  }
          else outlineData[[i]] <- NA }
       names(outlineData) <- nID
       out$outlines <- outlineData }
   cat(paste("Read", nobs, "datasets with", sum(noutline), "outlines\n"))
    return(out)}

#try to read all data(landmarks and outlines) from one file
 lm_trial <- readalltps(here("analysis/data/raw_data/A_BG_2.tps"))

#read all files from raw_data
 tps_files <- list.files(here("analysis/data/raw_data"), full.names = TRUE)
lm_landmarks <- map(tps_files, readalltps)
 names(lm_landmarks) <- list.files(here("analysis/data/raw_data"))

lm_landmarks = c()
for (i in 1:length(files)) {
  filename <- paste(datadir, files[i], sep = "")
  print(filename)
  landmarks <- readalltps(filename)
 lm_landmarks <- c(lm_landmarks, landmarks)}

```

### 2020-05-05 analysis/paper/paper.Rmd
try to figure out the artifact number per site

@@
                 select(-data)
+
+## Reference: site and number of artifacts
+ cv_by_site_df_label_1 <- 
+df_sitename %>% 
+  select(-SPstage.Stage) %>% 
+  group_by(sitename) %>% 
+  add_tally() %>% 
+   mutate(label = as.factor(paste0(sitename,'\nN = ',n))) %>%   select(-n) %>% 
+ nest(-sitename, -label) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
+
+cv_by_site_df_unnest_label_1 <-   
+cv_by_site_df_label_1 %>% 
+  unnest(cv_by_site)
 

### 2020-05-05 analysis/paper/paper.Rmd
refine some plots of CVs, sites, phases

--- 2020-04-28 analysis/paper/paper.Rmd
@@
 
-```{r demo, eval = FALSE}
-1 + 1
-```
-
 # Methods
@@
 Reading tps data and phase information of each artifact 
-```{r}
+```{r results='hide', include=FALSE}
 #reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
@@
   filename <- paste(datadir, files[i], sep = "")
-  print(filename)
+ # print(filename)
   landmarks <- read.tps(filename)
@@
 cv_plot_site %>%
-pivot_longer(cols = -sitename, names_to = "group") %>% 
-ggplot(aes(x = sitename, y = value, fill = as.factor(group))) + 
-geom_bar(stat="identity", position=position_dodge())
+  pivot_longer(cols = -sitename, 
+               names_to = "group") %>%
+  ggplot(aes(x = sitename, 
+             y = value, 
+             fill = as.factor(group))) +
+  geom_bar(stat = "identity", 
+           position = position_dodge())
 
@@
 
-```{r}
-## CV with site (n=X)
- cv_by_site_df_label <- 
-df_sitename %>% 
-  select(-SPstage.Stage) %>% 
-  group_by(sitename) %>% 
-  add_tally() %>% 
-  filter( n > 1) %>% 
-   mutate(label = as.factor(paste0(sitename,'\nN = ',n))) %>%   select(-n) %>% 
- nest(-sitename, -label) %>% 
-  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
-
-cv_by_site_df_unnest_label <-   
-cv_by_site_df_label %>% 
-  unnest(cv_by_site)
-
-cv_plot_site_label <- cv_by_site_df_unnest_label %>%
-                select(-data)
-
-library(tidyr)
-p<- cv_plot_site_label %>% 
-pivot_longer(cols = -c(sitename,label), names_to = "group") %>% 
-ggplot(aes(label, value, fill = as.factor(group))) + 
-geom_bar(stat="identity", position=position_dodge())
-
-p + labs(x= "Site", fill = "Attributes")
-```
-
 Make a table of CVs for all variable grouped by phase
+
+BM: discuss this one first, this is the overview picture for our story:
+
 ```{r}
@@
 geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")
-
-```
+```
+
 This plot shows the difference of CV in between 2nd and 3rd phase.
@@
 We haven't figured out which attibutes would be more infortant than others yet.
+
+BM: this discuss this one second, this is where we get into the details
+
+```{r}
+## CV with site (n=X)
+
+cv_by_site_df_unnest_label %>% 
+  select(-SPstage.Stage) %>% 
+  group_by(sitename) %>% 
+  add_tally() %>% 
+  filter( n > 1) %>% 
+   mutate(label = as.factor(paste0(sitename,' (N = ', n, ")"))) %>%   select(-n) %>% 
+ nest(-sitename, -label) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
+
+cv_by_site_df_unnest_label <-   
+cv_by_site_df_label %>% 
+  unnest(cv_by_site)
+
+cv_plot_site_label <- cv_by_site_df_unnest_label %>%
+                select(-data)
+
+library(tidyr)
+p<- cv_plot_site_label %>% 
+pivot_longer(cols = -c(sitename,label), names_to = "group") %>% 
+ggplot(aes(label, value, fill = as.factor(group))) + 
+geom_bar(stat="identity", position=position_dodge())
+
+p + labs(x= "Site", fill = "Attributes")
+
+# plot CVs by site with number of artefacts showing >10
+# phase 2 and phase 3 sites only
+cv_by_site_df_label_more_than_10 <- 
+df_sitename %>% 
+  unite(col = "site_phase", 
+        c(sitename, 
+          SPstage.Stage), 
+        sep = "_") %>% 
+  group_by(site_phase) %>% 
+  add_tally() %>% 
+  filter( n > 10) %>% 
+   mutate(label = as.factor(paste0(site_phase,' (N = ', n, ")"))) %>%   select(-n) %>% 
+ nest(-site_phase, -label) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>% 
+  unnest(cv_by_site) %>% 
+  select(-data)
+
+cv_by_site_df_label_more_than_10 %>%
+  pivot_longer(cols = -c(site_phase, label), 
+               names_to = "lithic_attribute",
+               values_to = "CV") %>%
+  ggplot(aes(x = lithic_attribute, 
+             y = CV)) +
+  geom_col() +
+  facet_wrap(~  label) +
+  ggtitle("CVs of point attributes by site (with sites >10 artefacts")
+
+cv_by_site_df_label_more_than_10 %>%
+  pivot_longer(cols = -c(site_phase, label), 
+               names_to = "lithic_attribute",
+               values_to = "CV") %>%
+  filter(lithic_attribute %in% c("ML", "MW", "TW")) %>%
+
+  ggplot(aes(x = lithic_attribute, 
+             y = CV)) +
+  geom_col() +
+  facet_wrap(~  label) +
+  ggtitle("CVs of point attributes by site (with sites >10 artefacts)") +
+  theme_minimal()
+```
 
@@
 
-Contribution of variables for each of four PCA demensions of ALL phases except for the first. 
+Contribution of variables for each of four PCA dimensions of ALL phases except for the first. 
 ```{r}

### 2020-05-05 analysis/paper/paper.Rmd
resolve merge conflicts

--- 2020-05-05 analysis/paper/paper.Rmd
@@
 Reading tps data and phase information of each artifact
-```{r, echo = FALSE}
+
+```{r results='hide', include=FALSE}
 #reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
@@
 ```
+
 The coefficient of variation (CV) is the ratio of the standard deviation devided by mean. It shows the extent of variability from the mean.
+
 CV for all attributes are distributed from 23-31%. Interestingly, SL and SW have high CV values (36.3 and 32.0, respectively). These attributes show lower variation on boxplot. <= need to think about
@@
 ```
+
 The plot shows CV values for different attributes The color of the bars represent different sites.
@@
 ```{r}
+
+cv_plot_site %>%
+  pivot_longer(cols = -sitename, 
+               names_to = "group") %>%
+  ggplot(aes(x = sitename, 
+             y = value, 
+             fill = as.factor(group))) +
+  geom_bar(stat = "identity", 
+           position = position_dodge())
+
+```
+
+Unlike the previous plot, in this one, the bars represent different attributes. You can see that SC has highest CV values of most attributes. Overall, SW values are very high.
+
+Make a table of CVs for all variable grouped by phase
+
+BM: discuss this one first, this is the overview picture for our story:
+
+```{r}
+cv_by_stage_df  <- 
+df_sitename %>% 
+  select(-sitename) %>% 
+  group_by(SPstage.Stage) %>% 
+  add_tally() %>% 
+  filter(SPstage.Stage > 1) %>% 
+  nest(-SPstage.Stage) %>% 
+  mutate(cv_by_stage = map(data, ~map_df(.x, cv)))
+
+cv_by_stage_df_unnest <-   
+cv_by_stage_df %>% 
+  unnest(cv_by_stage)
+
+cv_plot_stage <- cv_by_stage_df_unnest %>% 
+  select(-data, -n)
+
+library(tidyr)
+cv_plot_stage %>% 
+pivot_longer(cols = -SPstage.Stage, names_to = "group") %>%
+ggplot(aes(x = group, y = value, fill = as.factor(SPstage.Stage))) + 
+geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")
+```
+
+This plot shows the difference of CV in between 2nd and 3rd phase.
+For TL, both two phases have almost same values.
+For MW and TW, 3rd phase has lower values than the 2nd.
+The other 4 attributes show higher valeues in 3rd phase, which means less standardized in their shape during the latter period.
+
+By the way, Previous research (Bettinger&Eerkins, Garvey) focused on Width of the projectile points along with weight to understand cultural transmssion.
+We haven't figured out which attibutes would be more infortant than others yet.
+
+BM: this discuss this one second, this is where we get into the details
+
+```{r}
 ## CV with site (n=X)
- cv_by_site_df_label <- 
-df_sitename %>% 
+
+cv_by_site_df_unnest_label %>% 
   select(-SPstage.Stage) %>% 
@@
   filter( n > 1) %>% 
-   mutate(label = as.factor(paste0(sitename,'\nN = ',n))) %>%   select(-n) %>% 
+   mutate(label = as.factor(paste0(sitename,' (N = ', n, ")"))) %>%   select(-n) %>% 
  nest(-sitename, -label) %>% 
@@
 ```
+
 Unlike the previous plot, in this one, the bars represent different attributes. You can see that SC has highest CV values of most attributes. Overall, SW values are very high.
 
-Make a table of CVs for all variable grouped by phase
 ```{r}
-cv_by_stage_df  <- 
+# plot CVs by site with number of artefacts showing >10
+# phase 2 and phase 3 sites only
+cv_by_site_df_label_more_than_10 <- 
 df_sitename %>% 
-  select(-sitename) %>% 
-  group_by(SPstage.Stage) %>% 
+  unite(col = "site_phase", 
+        c(sitename, 
+          SPstage.Stage), 
+        sep = "_") %>% 
+  group_by(site_phase) %>% 
   add_tally() %>% 
-  filter(SPstage.Stage > 1) %>% 
-  nest(-SPstage.Stage) %>% 
-  mutate(cv_by_stage = map(data, ~map_df(.x, cv)))
+  filter( n > 10) %>% 
+   mutate(label = as.factor(paste0(site_phase,' (N = ', n, ")"))) %>%   select(-n) %>% 
+ nest(-site_phase, -label) %>% 
+  mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>% 
+  unnest(cv_by_site) %>% 
+  select(-data)
 
-cv_by_stage_df_unnest <-   
-cv_by_stage_df %>% 
-  unnest(cv_by_stage)
+cv_by_site_df_label_more_than_10 %>%
+  pivot_longer(cols = -c(site_phase, label), 
+               names_to = "lithic_attribute",
+               values_to = "CV") %>%
+  ggplot(aes(x = lithic_attribute, 
+             y = CV)) +
+  geom_col() +
+  facet_wrap(~  label) +
+  ggtitle("CVs of point attributes by site (with sites >10 artefacts")
 
-cv_plot_stage <- cv_by_stage_df_unnest %>% 
-  select(-data, -n)
+cv_by_site_df_label_more_than_10 %>%
+  pivot_longer(cols = -c(site_phase, label), 
+               names_to = "lithic_attribute",
+               values_to = "CV") %>%
+  filter(lithic_attribute %in% c("ML", "MW", "TW")) %>%
 
-library(tidyr)
-cv_plot_stage %>% 
-pivot_longer(cols = -SPstage.Stage, names_to = "group") %>%
-ggplot(aes(x = group, y = value, fill = as.factor(SPstage.Stage))) + 
-geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")
-
+  ggplot(aes(x = lithic_attribute, 
+             y = CV)) +
+  geom_col() +
+  facet_wrap(~  label) +
+  ggtitle("CVs of point attributes by site (with sites >10 artefacts)") +
+  theme_minimal()
 ```
-This plot shows the difference of CV in between 2nd and 3rd phase.
-For TL, both two phases have almost same values.
-For MW and TW, 3rd phase has lower values than the 2nd.
-The other 4 attributes show higher valeues in 3rd phase, which means less standardized in their shape during the latter period.
-
-By the way, Previous research (Bettinger&Eerkins, Garvey) focused on Width of the projectile points along with weight to understand cultural transmssion.
-We haven't figured out which attibutes would be more infortant than others yet.
 
@@
 ```
+
 When it comes to the correlations between seven attributes, BL-ML, TW-MW, TL-TW, and TL-SL are relatively close to each other.
@@
 
-Contribution of variables for each of four PCA demensions of ALL phases except for the first. 
+Contribution of variables for each of four PCA dimensions of ALL phases except for the first.
+
 ```{r}
@@
 ```
+
 Overall, BL and ML have stronger influence than the other attributes and are related to each other but not with the other attributes throughout time.
@@
 Correlations for each site
+
 ```{r}
@@
 ```
+
 In this plot, we can see that the pattern of correlation for SYG6 and SYG1_1 are similar in terms of correlation based on ML/SL/TL with other attributes.
@@
 However, SYG1_2 is younger than the other two and the correlation based on SW become weaker.
+
 When it comes to the difference between SYG6 and SYG1_1, the correlation between artifacts of SYG1_1 is stronger than SYG6.
@@
 PCA for each site
+
 ```{r}
@@
 ```
+
 Unlike the correlation of the three units, PCA of the three are more various. Especially, SW is related to SL/TL in SYG6, related to ML/BL in SYG1_1, and unrelated to the other attributes in SYG1_2.

### 2020-05-05 analysis/paper/paper.Rmd
add mean cv lines

@@
   add_tally() %>% 
-  filter( n > 10) %>% 
+  filter( n > 5) %>% 
    mutate(label = as.factor(paste0(site_phase,' (N = ', n, ")"))) %>%   select(-n) %>% 
@@
 
+cv_by_site_df_label_more_than_10_ml_mw_tw <- 
 cv_by_site_df_label_more_than_10 %>%
@@
   filter(lithic_attribute %in% c("ML", "MW", "TW"))  %>% 
+  mutate(phase = parse_number(str_extract(site_phase, "_."))) %>% 
+  group_by(site_phase) %>% 
+  mutate(mean_cv = mean(CV))
 
-  ggplot(aes(x = lithic_attribute, 
+  ggplot(cv_by_site_df_label_more_than_10_ml_mw_tw,
+         aes(x = lithic_attribute, 
              y = CV)) +
   geom_col() +
+  geom_hline(aes(yintercept = mean_cv), 
+             colour = "red") +
   facet_wrap(~  label) +

### 2020-05-07 analysis/paper/paper.Rmd
fixed unmerged chunk

@@
 
-cv_by_site_df_unnest_label %>% 
+ cv_by_site_df_label <- 
+df_sitename %>% 
   select(-SPstage.Stage) %>% 

### 2020-05-15 analysis/paper/paper.Rmd
added new cvs file and trimed tps files
updated the sitelocation file

@@
 ---
-title: "Title Goes Here"
+title: "Understanding the technological transition based on cultural transmission biases during the Late Pleistocene in Korea"
 author:
@@
 Gather attributes from  all artifacts
+
 ```{r}

### 2020-05-15 analysis/paper/paper.Rmd
Move plots to note.rmd file

@@
 map_df(DF, cv)
-box2 <- boxplot(cvmap) ##Is there any better way to show this data??
 
@@
 We assume that this time difference and few number of artifacts may relate to the high CV value.
-
-```{r}
-
-cv_plot_site %>%
-  pivot_longer(cols = -sitename, 
-               names_to = "group") %>%
-  ggplot(aes(x = sitename, 
-             y = value, 
-             fill = as.factor(group))) +
-  geom_bar(stat = "identity", 
-           position = position_dodge())
-
-```
 
@@
 
-direction_all+ direction_2 + direction_3 + plot_layout(ncol=1)
 ```
@@
 
-xx_pca1_2 <- 
-  data.frame(xx_pca2$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage ==2]))
-
-ggplot(xx_pca1_2 ,
-       aes(PC1,
-           PC2,
-       colour = stage)) +
-  geom_point() +
-  stat_ellipse()
-
 # phase 3
@@
   prcomp(., center = TRUE, scale = TRUE)
-
-xx_pca1_3 <- 
-  data.frame(xx_pca3$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage ==3]))
-
-ggplot(xx_pca1_3 ,
-       aes(PC1,
-           PC2,
-       colour = stage)) +
-  geom_point() +
-  stat_ellipse()
-
-```
-
-Contribution of variables for each of four PCA dimensions of ALL phases except for the first.
-
-```{r}
-library(gridExtra)
-library(factoextra)
-PC1_v <- fviz_contrib(xx_pca, choice = "var", axes = 1, top = 7)
-PC2_v <- fviz_contrib(xx_pca, choice = "var", axes = 2, top = 7)
-PC3_v <- fviz_contrib(xx_pca, choice = "var", axes = 3, top = 7)
-PC4_v <- fviz_contrib(xx_pca, choice = "var", axes = 4, top = 7)
-
-grid.arrange(PC1_v, PC2_v, PC3_v, PC4_v, nrow =2)
 
@@
 
-correlate(SYG6_DF) %>% 
-  network_plot()
-
 SYG6_cor <- cor(SYG6_DF)
@@
 
-grid.arrange(SYG6_PC1_v, SYG6_PC2_v, nrow =1)
+#grid.arrange(SYG6_PC1_v, SYG6_PC2_v, nrow =1)
 
@@
   ggtitle(paste("Correlation SYG1_2: 3rd Phase", "(n =", count_SYG1_2,")"))
-
-correlate(SYG1_2DF) %>% 
-  network_plot()
 
@@
 
-grid.arrange(SYG1_2PC1_v, SYG1_2PC2_v, nrow =1)
+#grid.arrange(SYG1_2PC1_v, SYG1_2PC2_v, nrow =1)
 
@@
   ggtitle(paste("Correlation SYG1_1: 2nd Phase", "(n =", count_SYG1_1,")"))
-
-correlate(SYG1_1DF) %>% 
-  network_plot()
 
@@
 
-grid.arrange(SYG1_1PC1_v, SYG1_1PC2_v, nrow =1)
+#grid.arrange(SYG1_1PC1_v, SYG1_1PC2_v, nrow =1)
 

### 2020-05-15 analysis/paper/paper.Rmd
updated

@@
 ```{r}
+
+cv_by_site_df_label_more_than_10_ml_mw_bl <- 
+cv_by_site_df_label_more_than_10 %>%
+  pivot_longer(cols = -c(site_phase, label), 
+               names_to = "lithic_attribute",
+               values_to = "CV") %>%
+  filter(lithic_attribute %in% c("ML", "BL","MW"))  %>% 
+  mutate(phase = parse_number(str_extract(site_phase, "_."))) %>% 
+  group_by(site_phase) %>% 
+  mutate(mean_cv = mean(CV))
+
+  ggplot(cv_by_site_df_label_more_than_10_ml_mw_bl,
+         aes(x = lithic_attribute, 
+             y = CV)) +
+  geom_col() +
+  geom_hline(aes(yintercept = mean_cv), 
+             colour = "red") +
+  facet_wrap(~  label) +
+  ggtitle("CVs of point attributes by site (with sites >=7 artefacts)") +
+  theme_minimal()
+```
+
+```{r}
 #using corrr package

### 2020-05-15 analysis/paper/paper.Rmd
deal with stringsAsFactors change for R 4.0.0

@@
       read.table(file = data, header = F, skip = LM[i],
-                 nrows = nrows[i], col.names = c("X", "Y")),
+                 nrows = nrows[i], 
+                 col.names = c("X", "Y"),
+                 stringsAsFactors = TRUE),
       IMAGE = as.character(images[i]),
       ID = ID[i],
-      SCALE = SCALE[i]))
+      SCALE = SCALE[i],
+      stringsAsFactors = TRUE))
   }
@@
 
-Computing size attributes of the points by calculating distance between landmakrs
+Computing size attributes of the points by calculating distance between landmarks
 ```{r}

### 2020-05-16 analysis/paper/paper.Rmd
commit for pull Ben's version

--- 2020-05-15 analysis/paper/paper.Rmd
@@
 author:
-  -gayoung Park
+  -Gayoung Park
   - Author Two

### 2020-05-16 analysis/paper/paper.Rmd

@@
 
-Computing size attributes of the points by calculating distance between landmakrs
+Computing size attributes  of the points by calculating distance between landmarks
 ```{r}

### 2020-05-16 analysis/paper/paper.Rmd
tidy author names

@@
 author:
-  -Gayoung Park
-  - Author Two
+  - Gayoung Park
+  - Ben Marwick
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2020-05-26 analysis/paper/paper.Rmd
added several para for cultural evolution

@@
 - Same general concept of biological evolution: genetic change in population that is inherited over generations. Apply this concept to artifacts or cultural phenomena
+
 - In this paper, I will use Cultural Transmission 
@@
 - similar study (CT in stone artifacts) : Mesoudi&O'Brien, Bettinger&Eerkens, Garvey
+
+-------------
+Application the theoretical and methodological tenets of Darwinian evolutionary concpet to archaeological studies has helped understand technological transition related to human behavior (Dunnell 1980; Lipo et al.
+1997; Bettinger and Eerkens 1997, 1999, 2008, Mesoudi abd O'Brien 2008).
+
+biological concept to archaeological usage: (Richerson and Boyd 1992)
+
+Cultrual evolution that archaeologists adapted has three sub-categories: human behavioral ecology, phylogenetics and cultural transmission theory (Garvey 2018).
+Among them, we apply models and methods of cultural tranmssion to understand the technological change and socio cultural contexts.
+
+Cultural transmission is the idea that a transition can occur during the process of social learning and the transmission biases are important loci of changes in material cultural (Lycett 2015; Creanza et al.
+2017).
+In other words, this concept holds that instructions for making artifacts are transmitted between members of a group during social learning by observation of, or interaction with another member (Heyes 1994; Eerkens and Lipo 2007; Brantingham et al.
+2010; Lycett 2015).
+
+The transmission is subject to biases that result from different social contexts.
+Boyd and Richerson (1985) and Henrich and McElreath (2003) have defined many of these biases, including ‘guided variation’ (where individuals learn new behaviors and then modify them through trial and error), ‘content-based bias’ (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), ‘frequency-based bias’ (where an individual is biased to choose particular instructions based on their perceived frequency in the population, such as extremely popular or rare instructions), and ‘model-based bias’ (where a variant is transmitted because of its association with other attributes, such as the prestige or skill of other individuals).
+
+Bettinger and Eerkens (1997; 1999; 2008) have shown how these biases can be identified in stone artifact assemblages.
+They studied metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
+They equated guided transmission with high metric variation and low correlation between metric attributes, and equated model-based bias with less variation and more correlated variables.
+They found that in eastern California artifacts have poor correlations of basal width and mass, suggesting new technology was introduced and spread by guided transmission.On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by model-based bias.
+Since Bettiniger and Eerkens studied archaeological data so they are unable to directly observe the transmission process.
+To test the underlying assumptions that Bettinger and Eerkens used, Mesoudi and O’Brien (2008) have conducted experiments that simulated prestige bias and guided variation, and observed results that supported Bettinger and Eerkens’ hypothesis.
+
+Garvey
+
+-------------
 

### 2020-05-26 analysis/paper/paper.Rmd
add a few words for commit/push
stash

@@
 
-biological concept to archaeological usage: (Richerson and Boyd 1992)
+need to add 'biological concept to archaeological usage': (Richerson and Boyd 1992)
 
-Cultrual evolution that archaeologists adapted has three sub-categories: human behavioral ecology, phylogenetics and cultural transmission theory (Garvey 2018).
-Among them, we apply models and methods of cultural tranmssion to understand the technological change and socio cultural contexts.
+Cultural evolution that archaeologists adapted has three sub-categories: human behavioral ecology, phylogenetics and cultural transmission theory (Garvey 2018).
+Among them, we apply models and methods of cultural transmission to understand the technological change and socio cultural contexts.
 
@@
 Boyd and Richerson (1985) and Henrich and McElreath (2003) have defined many of these biases, including ‘guided variation’ (where individuals learn new behaviors and then modify them through trial and error), ‘content-based bias’ (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), ‘frequency-based bias’ (where an individual is biased to choose particular instructions based on their perceived frequency in the population, such as extremely popular or rare instructions), and ‘model-based bias’ (where a variant is transmitted because of its association with other attributes, such as the prestige or skill of other individuals).
+<!-- add some description of experimental and ethnographic evidence that these biases are real, some of this not archaeological, e.g.
+https://www.pnas.org/content/116/14/6726  -->
 

### 2020-05-26 analysis/paper/paper.Rmd
add more sentences

--- 2020-05-26 analysis/paper/paper.Rmd
@@
 
-biological concept to archaeological usage: (Richerson and Boyd 1992)
-
+need to add 'biological concept to archaeological usage': (Richerson and Boyd 1992)
 Cultrual evolution that archaeologists adapted has three sub-categories: human behavioral ecology, phylogenetics and cultural transmission theory (Garvey 2018).
@@
 Since Bettiniger and Eerkens studied archaeological data so they are unable to directly observe the transmission process.
+Eerkens and Bettinger (2001) claims a robust statistical technique to compare varations within different kinds of assemblages and propose actual values of the coefficient of variation for showing whether artifacts are randomly produced (CV = 57.7 percent) or the result of planned production (CV 1.7 percent).
+
 To test the underlying assumptions that Bettinger and Eerkens used, Mesoudi and O’Brien (2008) have conducted experiments that simulated prestige bias and guided variation, and observed results that supported Bettinger and Eerkens’ hypothesis.
 
-Garvey
+Garvey (2018)
 

### 2021-02-01 analysis/paper/paper.Rmd
moved writing related texts to my google notes

--- 2020-05-26 analysis/paper/paper.Rmd
@@
 ```
-
-# Introduction
-
-1. What are our research questions?
-- Understanding technological transition during the Late Pleistocene in Korea base on Cultural transmission. 
-- What are the social contexts led to the technological transition
-
-2. What is our archaeological background?
--SP was started in Korean Peninsula -> spread to Japan
--Why SP important, Why we care
--Current explantion of SP dispersal: migration vs in situ, in between
- The evidence of the two main models: proportion of SP in assemblage, Y chromosome (two groups of people), composition of assemblage (blade to flake ratio)
- -> no explanation about any mechanism/process/driver for the new technology
- -> lack of details of the lithic technology, human behavior
-
-3. What is our conceptual framework?
-- Evolutionary concepts that links stone artifacts to human behavior
-- Cultural Evolution : the idea that cultural change constitutes an evolutionary process
-- Same general concept of biological evolution: genetic change in population that is inherited over generations. Apply this concept to artifacts or cultural phenomena
-
-- In this paper, I will use Cultural Transmission 
-- Cultural Transmssion: way that people learn and pass on information
-- Transmission biases
-- similar study (CT in stone artifacts) : Mesoudi&O'Brien, Bettinger&Eerkens, Garvey
-
--------------
-Application the theoretical and methodological tenets of Darwinian evolutionary concpet to archaeological studies has helped understand technological transition related to human behavior (Dunnell 1980; Lipo et al.
-1997; Bettinger and Eerkens 1997, 1999, 2008, Mesoudi abd O'Brien 2008).
-
-biological concept to archaeological usage: (Richerson and Boyd 1992)
-
-Cultrual evolution that archaeologists adapted has three sub-categories: human behavioral ecology, phylogenetics and cultural transmission theory (Garvey 2018).
-Among them, we apply models and methods of cultural tranmssion to understand the technological change and socio cultural contexts.
-
-Cultural transmission is the idea that a transition can occur during the process of social learning and the transmission biases are important loci of changes in material cultural (Lycett 2015; Creanza et al.
-2017).
-In other words, this concept holds that instructions for making artifacts are transmitted between members of a group during social learning by observation of, or interaction with another member (Heyes 1994; Eerkens and Lipo 2007; Brantingham et al.
-2010; Lycett 2015).
-
-The transmission is subject to biases that result from different social contexts.
-Boyd and Richerson (1985) and Henrich and McElreath (2003) have defined many of these biases, including ‘guided variation’ (where individuals learn new behaviors and then modify them through trial and error), ‘content-based bias’ (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), ‘frequency-based bias’ (where an individual is biased to choose particular instructions based on their perceived frequency in the population, such as extremely popular or rare instructions), and ‘model-based bias’ (where a variant is transmitted because of its association with other attributes, such as the prestige or skill of other individuals).
-
-Bettinger and Eerkens (1997; 1999; 2008) have shown how these biases can be identified in stone artifact assemblages.
-They studied metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
-They equated guided transmission with high metric variation and low correlation between metric attributes, and equated model-based bias with less variation and more correlated variables.
-They found that in eastern California artifacts have poor correlations of basal width and mass, suggesting new technology was introduced and spread by guided transmission.On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by model-based bias.
-Since Bettiniger and Eerkens studied archaeological data so they are unable to directly observe the transmission process.
-To test the underlying assumptions that Bettinger and Eerkens used, Mesoudi and O’Brien (2008) have conducted experiments that simulated prestige bias and guided variation, and observed results that supported Bettinger and Eerkens’ hypothesis.
-
-Garvey
-
--------------
-
-4. What are our methods?
-- Framework: Bettinger&Eerkens
-- Two models: Guided variation and Model-based bias =>  Multi-point Origin (in situ or in between one) and Single-point Origin (migration model)
-- Test by Correlation and Variation (Pearson's r, PCA, Coefficient of Variation) 
-- data collection, sites, artifact #
-- attributes: landmarks (tps software)
-
-5. What are our results? 
-- overall change through time
-- correlation: postively correlated(?). We are using PCA too
-- PCA: 2,3 phases are highly overlapped, high redundancy. The individual phase are scattered, low redundancy. Variables of Dimension 1 is positively correlated
-- CV:values are around 20(ML)-34(SL). Compared to 2nd phase, 3rd phase has slightly lower variation.
-
-6. What is the answer to our questions??
-- low correlation + high variation? High correlation + low variation?
-- What are the criteria for correlation and variation
-- How was SP spread in Korea, did they have single orgin or multiple origins?
-- What are the impact of combinaton with blade technology
-
-7. What is limitation of this study?
-- Maybe applying CV for SP is not ideal. It can be depending on individual knapper's skill and raw materials
-- CV of one site (SYG6) - no big difference with the value of all data
-- CV: lithis vs pottery
-
-- couldn't include thickness and weight
-
-** Expectation
-- The main measurements are correlation(+PCA) and CV.
-- The set of the two measurements would be like: 
-1) lower CV and higher correlation between sites, in this case, we will think that SPs were the result of model based bias and possibly had single origin. 
-2) higher CV and lower correlation between sites, for this case, we will thank that SPs were the result of guided variation and possibly had multiple origins. 
-- We have three phases, but there are only two artifacts for the first phase.
-There are more artifacts and sites for the first phase but not eligible for this research because we gathered complete or almost complete shape of SPs.
-We excluded broken pieces.
-Therefore, we focus on the change from the 2nd phase to 3rd phase for understanding the process of SP's dispersal in the Korean Peninsula.
-- Change from the 2nd to 3rd
-1) less correlated: the two groups were not very related/locally developed 
-2) highly correlated: the two groups were related, had single source and then spread to other places
-- Site by site
-- Regional relationship? Province/cities
-
-# Background
-
-# Methods
-
-Reading tps data and phase information of each artifact
 
@@
 
+Attributes that we are using here:
+
+It must be noted that all of this previously undertaken work has focused exclusively on the size (i.e., scale) of artifactual attributes and the patterns of variation that may be produced via copying error rates due to limits in size perception.
+However, variation in the overall shape of artifacts is also evidently important, given that aspects of shape may have specific functional or aesthetic properties (Schillinger et al.
+2014).
+
+what other people use:
+
 Computing Coefficient or variation
@@
 pivot_longer(cols = -sitename, names_to = "group") %>% 
-ggplot(aes(x = group, y = value, fill = as.factor(sitename))) + 
+ggplot(aes(x = group, y = value, fill = sitename)) + 
 geom_bar(stat="identity", position=position_dodge())
@@
 geom_bar(stat="identity", position=position_dodge()) + labs(x="Attributes", fill= "Phase")
+
 ```
@@
   facet_wrap(~  label) +
-  ggtitle("CVs of point attributes by site (with sites >10 artefacts")
+  ggtitle("CVs of point attributes by site (with sites >10 artefacts)")
 
@@
                values_to = "CV") %>%
-  filter(lithic_attribute %in% c("ML", "MW", "TW"))  %>% 
+  filter(lithic_attribute %in% c("BL", "ML", "MW", "TW"))  %>% 
   mutate(phase = parse_number(str_extract(site_phase, "_."))) %>% 
@@
 => This combination doesn't fit into the models.
+
+Eerkens and Lipo (2005) demonstrated that the thickness of Rose Springs projectile points from Owens Valley (USA) varied according to the predictions of the random accumulated copying error model.
+However, the basal width of these projectile points exhibited less variation than expected, suggesting that some non-mutation process (i.e., some form of stabilizing selection such as conformity) was operating on this particular variable.
 

### 2021-02-01 analysis/paper/paper.Rmd
organize codes to each r script

@@
 
-Computing size attributes  of the points by calculating distance between landmarks
 ```{r}
+#Computing size attributes  of the points by calculating distance between landmarks
 
@@
 
-The coefficient of variation (CV) is the ratio of the standard deviation devided by mean. It shows the extent of variability from the mean.
+The coefficient of variation (CV) is the ratio of the standard deviation divided by mean. It shows the extent of variability from the mean.
 
@@
 
-library(tidyr)
 cv_plot_stage %>% 
@@
 
-library(tidyr)
 p<- cv_plot_site_label %>% 
@@
 
-We see some differences in artefact shape and size. Now we need to examins the PCA output numbers to describe in more details.
+We see some differences in artefact shape and size. Now we need to examines the PCA output numbers to describe in more details.
 

### 2021-02-01 analysis/paper/000-import-and-clean-data.R

@@ -1,0 +1,153 @@
+library(here)
+library(tidyverse)
+
+
+#reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
+
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T))
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T))
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+
+  landmarks = vector("list", l) # create an empty list
+
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file = data, header = F, skip = LM[i],
+                 nrows = nrows[i],
+                 col.names = c("X", "Y"),
+                 stringsAsFactors = TRUE),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i],
+      stringsAsFactors = TRUE))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}
+
+#need to figure out how to read the file inside the project
+landmarks <- read.tps(here("analysis/data/raw_data/BG_0_2.tps"))
+
+#getting info for phase of each site
+SPstage <- read_csv(here("analysis/data/SPstage.csv"))
+
+#Compare multiple artifacts
+#Open multiple tps files. This will need to be changed to read files inside the project
+datadir <- here("/analysis/data/raw_data/")
+files <- dir(datadir, pattern = "*.tps$|*.TPS$")
+
+
[…]

### 2021-02-01 analysis/paper/001-basic-analysis.R

@@ -1,0 +1,162 @@
+library(here)
+library(tidyverse)
+
+##################data treatment###################
+#reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
+
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T))
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T))
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+
+  landmarks = vector("list", l) # create an empty list
+
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file = data, header = F, skip = LM[i],
+                 nrows = nrows[i],
+                 col.names = c("X", "Y"),
+                 stringsAsFactors = TRUE),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i],
+      stringsAsFactors = TRUE))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}
+
+#need to figure out how to read the file inside the project
+landmarks <- read.tps(here("analysis/data/raw_data/BG_0_2.tps"))
+
+#getting info for phase of each site
+SPstage <- read_csv(here("analysis/data/SPstage.csv"))
+
+#Compare multiple artifacts
+#Open multiple tps files. This will need to be changed to read files inside the project
+datadir <- here("/analysis/data/raw_data/")
+files <- dir(datadir, pattern = "*.tps$|*.TPS$")
+
+
[…]

### 2021-02-01 analysis/paper/002-coefficient-variation.R

@@ -1,0 +1,328 @@
+library(here)
+library(tidyverse)
+library(tidyr)
+
+##################data treatment###################
+#reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
+
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T))
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T))
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+
+  landmarks = vector("list", l) # create an empty list
+
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file = data, header = F, skip = LM[i],
+                 nrows = nrows[i],
+                 col.names = c("X", "Y"),
+                 stringsAsFactors = TRUE),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i],
+      stringsAsFactors = TRUE))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}
+
+#need to figure out how to read the file inside the project
+landmarks <- read.tps(here("analysis/data/raw_data/BG_0_2.tps"))
+
+#getting info for phase of each site
+SPstage <- read_csv(here("analysis/data/SPstage.csv"))
+
+#Compare multiple artifacts
+#Open multiple tps files. This will need to be changed to read files inside the project
+datadir <- here("/analysis/data/raw_data/")
[…]

### 2021-02-01 analysis/paper/003-correlation.R

@@ -1,0 +1,203 @@
+library(here)
+library(tidyverse)
+library(tidyr)
+library(corrr)
+##################data treatment###################
+#reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
+
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T))
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T))
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+
+  landmarks = vector("list", l) # create an empty list
+
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file = data, header = F, skip = LM[i],
+                 nrows = nrows[i],
+                 col.names = c("X", "Y"),
+                 stringsAsFactors = TRUE),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i],
+      stringsAsFactors = TRUE))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}
+
+#need to figure out how to read the file inside the project
+landmarks <- read.tps(here("analysis/data/raw_data/BG_0_2.tps"))
+
+#getting info for phase of each site
+SPstage <- read_csv(here("analysis/data/SPstage.csv"))
+
+#Compare multiple artifacts
+#Open multiple tps files. This will need to be changed to read files inside the project
+datadir <- here("/analysis/data/raw_data/")
[…]

### 2021-02-01 analysis/paper/004-prinical-component-analysis.R

@@ -1,0 +1,442 @@
+library(here)
+library(tidyverse)
+library(tidyr)
+library(corrr)
+library(ggbiplot)
+library(ggfortify)
+
+##################data treatment###################
+#reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
+
+read.tps = function(data) {
+  # Reads the .tps file format produced by TPSDIG
+  # (http://life.bio.sunysb.edu/morph/ into a single data frame
+  # USAGE: R> read.tps("filename.tps")
+  a = readLines(data) # so we can do some searching and indexing
+  LM = grep("LM", a) # find the line numbers for LM
+  ID.ind = grep("ID", a) # find the line numbers for ID
+  # and the ID values, SCALE values, and image names
+  ID = gsub("(ID=)(.*)", "\\2", grep("ID", a, value=T))
+  SCALE = gsub("(SCALE=)(.*)", "\\2", grep("SCALE", a, value=T))
+  images = basename(gsub("(IMAGE=)(.*)", "\\2", a[ID.ind - 1]))
+  # FOR EACH LOOP
+  skip = LM # set how many lines to skip
+  # and how many rows to read
+  nrows = as.numeric(gsub("(LM=)(.*)", "\\2", grep("LM", a, value=T)))
+  l = length(LM) # number of loops we want
+
+  landmarks = vector("list", l) # create an empty list
+
+  for (i in 1:l) {
+    landmarks[i] = list(data.frame(
+      read.table(file = data, header = F, skip = LM[i],
+                 nrows = nrows[i],
+                 col.names = c("X", "Y"),
+                 stringsAsFactors = TRUE),
+      IMAGE = as.character(images[i]),
+      ID = ID[i],
+      SCALE = SCALE[i],
+      stringsAsFactors = TRUE))
+  }
+  do.call(rbind, landmarks) # rbind the list items into a data.frame
+}
+
+#need to figure out how to read the file inside the project
+landmarks <- read.tps(here("analysis/data/raw_data/BG_0_2.tps"))
+
+#getting info for phase of each site
[…]

### 2021-05-12 analysis/paper/000-import-and-clean-data.R
attempting to derive distributions

@@ -45,9 +45,8 @@ SPstage <- read_csv(here("analysis/data/SPstage.csv"))
 
 #Compare multiple artifacts
 #Open multiple tps files. This will need to be changed to read files inside the project
-datadir <- here("/analysis/data/raw_data/")
+datadir <- here("analysis/data/raw_data/")
 files <- dir(datadir, pattern = "*.tps$|*.TPS$")
-
 
 #Computing size attributes  of the points by calculating distance between landmarks
 
@@ -129,7 +128,7 @@ SW = c()
 for (i in 1:length(files)) {
   filename <- paste(datadir, files[i], sep = "")
   # print(filename)
-  landmarks <- read.tps(filename)
+  landmarks <- read.tps(here(filename))
   ML <- c(ML, landmark_dist(landmarks, 1, 7))
   BL <- c(BL, bl_dist(landmarks))
   TL <- c(TL, tl_dist(landmarks))

### 2021-05-12 analysis/paper/001-basic-analysis.R

@@ -4,6 +4,102 @@ load("data_main.RData")
 
 box1 <- boxplot(DF, main = "Variation for each attributes", col = "gold")
 
+library(tidyverse)
+# take a look at the distribution of each variable
+DF %>%
+  pivot_longer(everything()) %>%
+  ggplot() +
+  aes(value) +
+  geom_density() +
+  facet_wrap( ~ name) +
+  theme_minimal()
+
+# fit distributions for each of these, so we can simulate from those distributions
+library("fitdistrplus")
+library(distributions3)
+
+# what distributions will explore to see if they match with our data?
+the_distributions <- c("weibull",   "norm" )
+the_distributions3 <- c(Weibull,  Normal)
+names(the_distributions3) <- the_distributions
+
+fit_distributions <-
+DF %>%
+  pivot_longer(everything())  %>%
+  nest(-name) %>%
+  # compute fit of each variable with a selection of distributions
+  mutate(dist_out_w = map(data, ~fitdist(unlist(.x), the_distributions[1]))) %>%
+  mutate(dist_out_l = map(data, ~fitdist(unlist(.x), the_distributions[2]))) %>%
+ # mutate(dist_out_g = map(data, ~fitdist(unlist(.x), the_distributions[3]))) %>%
+ # mutate(dist_out_n = map(data, ~fitdist(unlist(.x), the_distributions[4]))) %>%
+  pivot_longer(cols = starts_with("dist"),
+               names_to = "dist") %>%
+  nest(-name) %>%
+  # extract goodness of fit statistics
+  mutate(gofstat_out = map(data, ~gofstat(.x$value,
+                                          fitnames = the_distributions))) %>%
+  # AIC, in particular, which distribution has the lowest value (so, best fit)
+  mutate(gofstat_out_aic = map_chr(gofstat_out,
+                                   ~names(which.min(.x$aic)))) %>%
+  unnest(data) %>%
+  distinct(name, .keep_all = TRUE) %>%
+  # estimate parameters of distribution to match our data
+  mutate(fitdist_out = map2(data, gofstat_out_aic,
+                            ~fitdist(deframe(.x), .y))) %>%
+  mutate(fit = map(fitdist_out, ~t(as.data.frame(.x$estimate)))) %>%
+  unnest_wider(fit) %>%
+  # create a function to generate a random distribution, using the distribution
+  # name we deduced earlier, and with the parameter values we got earlier
[…]

### 2021-05-12 analysis/paper/garvey_ct_simulation.Rmd

---
title: "CT simulation"
author: "R Garvey"
date: "6/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read data
```{r}
library(here)
data <- read.csv(here("analysis/data/garvey_data.csv"))
str(data)
```

### Multi panel Washita figure

```{r}

# Plot and simulation parameters

#######################

D <- data.frame(data$wash_width, data$wash_basewidth,
                data$wash_neck, data$wash_haftlength,
                data$wash_length, data$wash_midlength, 
                data$wash_thick, data$wash_bladelength, 
                data$wash_weight)

# set counter to # variables (i.e., # figure rows)
count <- c(1:9)

# set the attribute-specific n.measures
n.measures <- c(162, 157,
                249, 227,
                136, 136,
                259, 135,
                87)

# y-axis panel plot labels
labs <- c("width\n(mm)", "base width\n(mm)",
          "neck width\n(mm)", "haft length\n(mm)",
          "length\n(mm)", "mid length\n(mm)",  
          "thickness\n(mm)", "blade length\n(mm)",
          "weight\n(g)")

## simulation parameters
set.seed(1212)            # set seed
n.houses       <- 100     # set number of houses at site
y              <- 1000    # set number of simulation runs

## plotting parameters
col.density   <- "gray80"   # set density color
col.outline   <- "gray20"   # set density outline color
col.mean      <- "black"    # set col for mean of arch sample
cex.ylab      <- 1.2        # set y-axis label size
cex.ylabtick  <- 1.1        # set y-axis tick label size
cex.xlabtick  <- 1.1        # set x-axis tick label size

x.lim2   <- c(6, 7,         # set x limit for each plot
              3.5, 4,       # (this also affects tick mark placement)
              50, 40,
              1.4, 35,
              0.12)

y.lim2   <- c(1.2, 1,       # set y limit for each plot
              1.75, 1.2,    
              0.15, 0.15,
              5, 0.15,
              45)

###############################

# The plot:

pdf(file="/Users/##ENTER DESTINATION##/Garvey_Figure3.pdf", 
    height = 12, width = 12)

par(mfcol = c(length(count), 3))
par(oma = c (6, 10, 5, 2))

# column 1 (SD = 0.1)

for(x in count) {   # looping procedure for each plotting pannel

SD       <- 0.1    # set standard deviation multiplier for random draws 
d        <- na.omit(D[, x])     # select the data set to sample
N        <- n.measures[x]       # select the appropriate n.measures

metadig   <- NULL  #create empty vector to hold digs
var.Kdigs <- NULL  #create empty vector to hold variances
cv.Kdigs  <- NULL  #create empty vector to hold CVs

for(i in 1:y) {    # "master" loop that  repeats
                   # "archaeological sampling"" from 
                   # "generational learning sample"" one thousand times

    all.houses <-  NULL         # create an empty vector to hold results

    for(j in 1:n.houses) {       #loop through all houses

               house.seed <- sample(d, 1, replace = TRUE)   
               house.g1 <- rnorm(4, mean = house.seed, 
                                 sd = SD * mean(house.seed)) 
               house.g2 <- rnorm(4, mean = mean(house.g1), 
                                 sd = SD * mean(house.g1)) 
               house.g3 <- rnorm(4, mean = mean(house.g2), 
                                 sd = SD * mean(house.g2))
               house.g4 <- rnorm(4, mean = mean(house.g3), 
                                 sd = SD * mean(house.g3))
               total.house.sample <- as.vector(c(house.g1,
                                                  house.g2,
                                                  house.g3,
                                                  house.g4))
           all.houses <- c(total.house.sample, all.houses)

        }
           dig <- sample(all.houses, N, replace = TRUE)
           metadig <- c(metadig, dig)
           var.Kdigs <- c(var(dig), var.Kdigs)
           cv.Kdigs <- c(sd(dig)/mean(dig), cv.Kdigs)
}

#plotting
dp.md <- density(var.Kdigs)
par(mai = c(0.25, 0.5, 0.1, 0))

plot(dp.md, col = col.outline, bty = 'n',
     main = "", xlab = "", ylab = "",
     xlim = c (0, x.lim2[x]),
     ylim = c (0, y.lim2[x]),
     xaxt = "n", yaxt = "n")

axis(1, at = c(0, 0.25 * x.lim2[x],
               0.5 * x.lim2[x], 0.75 * x.lim2[x],
               x.lim2[x]),
     cex = cex.xlabtick)

# alternative, with only 3 x-axis ticks
# axis(1, at = c(0, 0.5 * x.lim2[x],
#                x.lim2[x]),
#      cex = cex.xlabtick)

axis(2, at = c(0, 0.5 * y.lim2[x],
               y.lim2[x]),
     line = 1.5, las = 2,
     cex = cex.ylabtick)

polygon(dp.md, col = col.density)
abline(v = var(d), col = col.mean, lwd = 2)
mtext (text = labs[x], side = 2, 
       las = 1, line = 5.5,
       cex = 1.2, outer = FALSE)

}

mtext (text = "A. CV = 10%", side = 3, 
       las = 1, line = 73,
       cex = 1.5, outer = FALSE)

# column 2 (CV = 5%)

for(x in count) {   # looping procedure for each plotting pannel

SD       <- 0.05    # set standard deviation multiplier for random draws 
d        <- na.omit(D[, x])     # select the data set to sample
N        <- n.measures[x]       # select the appropriate n.measures

metadig   <- NULL  #create empty vector to hold digs
var.Kdigs <- NULL  #create empty vector to hold variances
cv.Kdigs  <- NULL  #create empty vector to hold CVs

for(i in 1:y) {    # "master" loop that should repeat
                   # "archaeological sampling"" from 
                   # "generational learning sample"" one thousand times

    all.houses <-  NULL         # create an empty vector to hold results

    for(j in 1:n.houses) {       #loop through all houses

               house.seed <- sample(d, 1, replace = TRUE)   
               house.g1 <- rnorm(4, mean = house.seed, 
                                 sd = SD * mean(house.seed)) 
               house.g2 <- rnorm(4, mean = mean(house.g1), 
                                 sd = SD * mean(house.g1)) 
               house.g3 <- rnorm(4, mean = mean(house.g2), 
                                 sd = SD * mean(house.g2))
               house.g4 <- rnorm(4, mean = mean(house.g3), 
                                 sd = SD * mean(house.g3))
               total.house.sample <- as.vector(c(house.g1,
                                                  house.g2,
                                                  house.g3,
                                                  house.g4))
           all.houses <- c(total.house.sample, all.houses)

        }
           dig <- sample(all.houses, N, replace = TRUE)
           metadig <- c(metadig, dig)
           var.Kdigs <- c(var(dig), var.Kdigs)
           cv.Kdigs <- c(sd(dig)/mean(dig), cv.Kdigs)
}

#plotting
dp.md <- density(var.Kdigs)
par(mai = c(0.25, 0.5, 0.1, 0))

plot(dp.md, col = col.outline, bty = 'n',
     main = "", xlab = "", ylab = "",
     xlim = c (0, x.lim2[x]),
     ylim = c (0, y.lim2[x]),
     xaxt = "n", yaxt = "n")

axis(1, at = c(0, 0.25 * x.lim2[x],
               0.5 * x.lim2[x], 0.75 * x.lim2[x],
               x.lim2[x]),
     cex = cex.xlabtick)

polygon(dp.md, col = col.density)
abline(v = var(d), col = col.mean, lwd = 2)

}

mtext (text = "B. CV = 5%", side = 3, 
       las = 1, line = 73,
       cex = 1.5, outer = FALSE)

# column 3 (CV = 3%)

for(x in count) {   # looping procedure for each plotting pannel

SD       <- 0.03    # set standard deviation multiplier for random draws 
d        <- na.omit(D[, x])     # select the data set to sample
N        <- n.measures[x]       # select the appropriate n.measures

metadig   <- NULL  #create empty vector to hold digs
var.Kdigs <- NULL  #create empty vector to hold variances
cv.Kdigs  <- NULL  #create empty vector to hold CVs

for(i in 1:y) {    # "master" loop that should repeat
                   # "archaeological sampling"" from 
                   # "generational learning sample"" one thousand times

    all.houses <-  NULL         # create an empty vector to hold results

    for(j in 1:n.houses) {       #loop through all houses

               house.seed <- sample(d, 1, replace = TRUE)   
               house.g1 <- rnorm(4, mean = house.seed, 
                                 sd = SD * mean(house.seed)) 
               house.g2 <- rnorm(4, mean = mean(house.g1), 
                                 sd = SD * mean(house.g1)) 
               house.g3 <- rnorm(4, mean = mean(house.g2), 
                                 sd = SD * mean(house.g2))
               house.g4 <- rnorm(4, mean = mean(house.g3), 
                                 sd = SD * mean(house.g3))
               total.house.sample <- as.vector(c(house.g1,
                                                  house.g2,
                                                  house.g3,
                                                  house.g4))
           all.houses <- c(total.house.sample, all.houses)

        }
           dig <- sample(all.houses, N, replace = TRUE)
           metadig <- c(metadig, dig)
           var.Kdigs <- c(var(dig), var.Kdigs)
           cv.Kdigs <- c(sd(dig)/mean(dig), cv.Kdigs)
}

#plotting
dp.md <- density(var.Kdigs)
par(mai = c(0.25, 0.5, 0.1, 0))

plot(dp.md, col = col.outline, bty = 'n',
     main = "", xlab = "", ylab = "",
     xlim = c (0, x.lim2[x]),
     ylim = c (0, y.lim2[x]),
     xaxt = "n", yaxt = "n")

axis(1, at = c(0, 0.25 * x.lim2[x],
               0.5 * x.lim2[x], 0.75 * x.lim2[x],
               x.lim2[x]),
     cex = cex.xlabtick)

polygon(dp.md, col = col.density)
abline(v = var(d), col = col.mean, lwd = 2)

}

mtext (text = "C. CV = 3%", side = 3, 
       las = 1, line = 73,
       cex = 1.5, outer = FALSE)

dev.off()
```

### 2021-05-12 analysis/paper/garvey_ct_simulation.Rmd
start thinking about simulations

@@
 ```
+
+Ben's explorations of simulating artefacts over multiple generations. THis looks at one attribute only, so far
+
+```{r}
+library(tidyverse)
+
+data <- read.csv(here::here("analysis/data/garvey_data.csv"))
+
+# how many generations?
+generations <- 100
+
+# how many artefacts per generation?
+n_artefacts_per_generation <- 30
+
+# what CVs to simulate with?
+cvs <- seq(1, 50, by = 5) / 100
+length(cvs) # 50 different CV values
+
+# which dimension of the artefact?
+dimen <- names(data)[1]
+
+# create an empty list object to store the results
+generations_by_cv_lst <- vector("list", length = length(cvs))
+# start a new list to hold all the generations for this cv value
+generations_lst <- vector("list", length = generations)
+
+for(cv_i in 1:length(cvs)){
+
+  # assign the CV value for generating the simulated data
+  SD <- cvs[cv_i]
+
+# for each generation, simulate data based on the previous generation
+for(g in 1:generations){
+
+  # for generation 1, get the first generation from the archaeological data
+
+  if(g == 1){ 
+    d        <- na.omit(data[, dimen])    # select the data set to sample
+    generation_seed <-  sample(d, 1, replace = TRUE) 
+    generations_lst[[g]] <-  
+     abs( rnorm(n_artefacts_per_generation, 
+             mean(generation_seed), 
+             SD * mean(generation_seed)))
+
+  } else {
+
+  # for generations 2, 3, 4, etc. simulate the data from the previous generation
+  previous_generation_index <-  g - 1
+  previous_generation_data <-  generations_lst[[previous_generation_index]]
+
+  # and generate a new set of random normal variables from the previous generation
+  generations_lst[[g]] <- 
+    abs(rnorm(n_artefacts_per_generation, 
+            mean(previous_generation_data), 
+            SD * mean(previous_generation_data)))
+  }
+}
+
+# make into a single vector, and store in list element
+generations_by_cv_lst[[cv_i]] <- (generations_lst)
+
+}
+
+# take a look 
+#  generations_by_cv_lst
+
+# plot some
+library(viridis)
+
+plot_list <- 
+  map(1:length(cvs),
+      function(cv){
+library(ggridges)
+tibble(`001` =    generations_by_cv_lst[[cv]][[1]],
+       `025` =    generations_by_cv_lst[[cv]][[25]],
+       `050` =    generations_by_cv_lst[[cv]][[50]],
+       `100` =    generations_by_cv_lst[[cv]][[100]]) %>% 
+  pivot_longer(everything()) %>% 
+  mutate(name = fct_rev(name)) %>% 
+ggplot()  +
+  geom_density_ridges(aes(x = value, 
+                          y = name,
+                          fill = name)) +
+  theme_minimal()+
+  scale_fill_viridis_d() +
+  guides(fill = "none") +
+  labs(x = paste0("CV = ", cvs[cv]))
+})
+
+library(cowplot)
+plot_grid(plotlist = plot_list)
+
+# plot all generations
+
+map(1:length(cvs),
+    function(cv){
+  map(1:generations, 
+    function(g){
+      generations_by_cv_lst[[cv]][[g]]
+    }) %>% 
+  bind_cols() %>% 
+  pivot_longer(everything()) %>% 
+  mutate(gen = parse_number(str_remove_all (name, "\\.*"))) %>% 
+  arrange(gen) %>% 
+  group_by(gen) %>% 
+  summarise(mean_val = mean(value))
+    }) %>% 
+  bind_rows(.id = "CV") %>% 
+  mutate(CV = as.factor(cvs[as.numeric(CV)])) %>% 
+  ggplot() +
+  aes(gen, 
+      mean_val,
+      colour = CV,
+      group = CV) +
+  geom_line() +
+  scale_color_viridis_d()
+```
+
+Ben exploring ABC for parameter estimation, can we use this instead of distance metrics to find a good fit with our data?
+
+```{r}
+#-------------------------------------------------------------------
+# from https://theoreticalecology.wordpress.com/2012/12/02/the-easyabc-package-for-approximate-bayesian-computation-in-r/
+
+library(EasyABC)
+
+# assuming the data ...
+data =  d
+
+# we want to use ABC to infer the parameters that were used.
+# we sample from the same model and use mean and variance
+# as summary statstitics for the model and the data.
+
+# observed summary statistics
+summarydata = c(mean(data), sd(data))
+
+# stochastic model generates a sample for given par and returns summary statistics
+model <- function(par){ 
+  samples <- rnorm(100, mean =par[1], sd = par[2])
+  return(c(mean(samples), sd(samples)))
+}
+
+# prior list - this is where we assert our opinions about the data-generating
+# process
+prior_list <- list(c("unif", 0, 10),
+                   c("unif", 0, 10))
+
+# call to EasyABC with the ABC-MCMC algorithm Marjoram, P.; Molitor, J.; Plagnol, V. & 
+# Tavare, S. (2003) Markov chain Monte Carlo without likelihoods. Proc. Natl. Acad. Sci. USA, 100, 15324-15328.
+# with some automatic adjustment options 
+ABC_Marjoram_original <- ABC_mcmc(method = "Marjoram", 
+                                model = model, 
+                                prior = prior_list, 
+                                summary_stat_target = summarydata, 
+                                n_rec = 10000)
+
+str(ABC_Marjoram_original)
+par(mfrow=c(2,1))
+hist(ABC_Marjoram_original$param[5000:10000,1], 
+     main = "Posterior for mean")
+hist(ABC_Marjoram_original$param[5000:10000,2], 
+     main = "Posterior for standard deviation")
+
+##------------------------------------------------------
+library(EasyABC)
+
+# assuming the data are 10 samples of a normal distribution
+# with mean 5.3 and sd 2.7
+data =  rnorm(10, mean =5.3, sd = 2.7)
+
+# we want to use ABC to infer the parameters that were used.
+# we sample from the same model and use mean and variance
+# as summary statstitics for the model and the data.
+
+# observed summary statistics
+summarydata = c(mean(data), sd(data))
+
+# stochastic model generates a sample for given par and returns summary statistics
+model <- function(par){ 
+  samples <- rnorm(10, mean =par[1], sd = par[2])
+  return(c(mean(samples), sd(samples)))
+}
+
+# call to EasyABC with the ABC-MCMC algorithm Marjoram, P.; Molitor, J.; Plagnol, V. & 
+# Tavare, S. (2003) Markov chain Monte Carlo without likelihoods. Proc. Natl. Acad. Sci. USA, 100, 15324-15328.
+# with some automatic adjustment options 
+ABC_Marjoram_original <- 
+  ABC_mcmc(method="Marjoram", 
+           model=model, 
+           prior=list(c("unif",0,10),
+                      c("unif",1,5)), 
+           summary_stat_target=summarydata, 
+           n_rec = 10000)
+
+str(ABC_Marjoram_original)
+par(mfrow=c(2,1))
+hist(ABC_Marjoram_original$param[5000:10000,1], main = "Posterior for mean")
+hist(ABC_Marjoram_original$param[5000:10000,2], main = "Posterior for standard deviation")
+
+#-------------------------------------------------------------------
+#
+
+library(abc)
+data(human)
+
+par(mfcol = c(1,3), mar=c(5,3,4,.5))
+boxplot(stat.3pops.sim[,"pi"]~models, main="Mean nucleotide diversity")
+boxplot(stat.3pops.sim[,"TajD.m"]~models, main="Mean Tajima's D")
+boxplot(stat.3pops.sim[,"TajD.v"]~models, main="Var in Tajima's D")
+
+cv.modsel <- cv4postpr(models, 
+                       stat.3pops.sim, 
+                       nval=5, tol=.01, 
+                       method="mnlogistic")
+s <- summary(cv.modsel)
+
+modsel.ha <- postpr(target = stat.voight["hausa",], # observed data summary stats
+                    index = models,                 # model ID, to go with sumstat
+                    sumstat =  stat.3pops.sim,      # summary stats for simulated data 
+                    tol=.05, 
+                    method="mnlogistic")
+
+modsel.it <- postpr(stat.voight["italian",], models, stat.3pops.sim, tol=.05, method="mnlogistic")
+modsel.ch <- postpr(stat.voight["chinese",], models, stat.3pops.sim, tol=.05, method="mnlogistic")
+summary(modsel.ha)
+
+# ## an artifical example
+ss <- cbind(runif(1000),
+            rt(1000,df=20))
+# ss is a df of 1000 rows and two columns
+pp <- 
+postpr(target=c(3), 
+       index=c(rep("a",500),rep("b",500)),
+       sumstat=ss[,1], tol=.1, method="rejection")
+
+summary(pp)
+
+```

### 2022-08-05 analysis/paper/002-coefficient-variation.R
tested CV significance

@@ -165,6 +165,75 @@ cv_by_stage_df  <-
             cv_low_sharma =  sharma_int_low(value),
             cv_high_sharma = sharma_int_high(value))
 
+####### significant test between the two phases ######
+library(cvequality)
+
+raw_data_test_BL <- df_sitename %>%
+  select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  filter(SPstage.Stage > 1) %>%
+  select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  filter(variable == "BL")
+
+test_BL <- with(raw_data_test_BL,
+                mslr_test(nr = 1e4,
+                          value,
+                          SPstage.Stage)) #p_value = 0.547566
+
+
+raw_data_test_ML <- df_sitename %>%
+  select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  filter(SPstage.Stage > 1) %>%
+  select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  filter(variable == "ML")
+
+test_ML <- with(raw_data_test_ML,
+                mslr_test(nr = 1e4,
+                          value,
+                          SPstage.Stage)) #p_value = 0.5159077
+
+raw_data_test_MW <- df_sitename %>%
+  select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  filter(SPstage.Stage > 1) %>%
+  select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  filter(variable == "MW")
+
+test_MW <- with(raw_data_test_MW,
+                mslr_test(nr = 1e4,
+                          value,
+                          SPstage.Stage)) #p_value = 0.3779604
+
+raw_data_test_SL <- df_sitename %>%
+  select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  filter(SPstage.Stage > 1) %>%
+  select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
[…]

### 2022-08-05 analysis/paper/paper.Rmd
Added more references

@@
 ---
-title: "Understanding the technological transition based on cultural transmission biases during the Late Pleistocene in Korea"
+title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
 author:
-  - Gayoung Park
-  - Ben Marwick
+  - Gayoung Park: 
+      email: gayoungp@uw.edu
+      institute: [UofW]
+      correspondence: true
+  - Ben Marwick:
+      email: bmarwick@uw.edu
+      institute: [UofW]
+      correspondence: false
+
+institute:
+  - UofW: Department of Anthropology, University of Washington, Seattle. USA
+
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
+```
+
+```{r results='hide', include=FALSE}
 #library(CTtps) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
@@
 library(tidyverse)
-```
 
-```{r results='hide', include=FALSE}
 #reading landmarks in tps, code from https://gist.github.com/mrdwab/2062329
@@
 
-#grid.arrange(SYG6_PC1_v, SYG6_PC2_v, nrow =1)
+library(gridExtra)
+grid.arrange(SYG6_PC1_v, SYG6_PC2_v, nrow =1)
 

### 2022-08-05 analysis/paper/002-coefficient-variation.R
nested data frame and map to apply mslr test

@@ -184,6 +184,31 @@ test_BL <- with(raw_data_test_BL,
                           value,
                           SPstage.Stage)) #p_value = 0.547566
 
+# BM testing doing all the CV testing at once
+
+# created nested tibble, nest by variable
+df_sitename_nested <-
+df_sitename %>%
+  select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  filter(SPstage.Stage > 1) %>%
+  select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  ungroup() %>%
+  group_by(variable) %>%
+  nest()
+
+# apply mslr test to each variable (row with a tibble)
+df_sitename_nested %>%
+  mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
+                                            .$value,
+                                            .$SPstage.Stage))) %>%
+  unnest(mslr_result)
+
+
 
 raw_data_test_ML <- df_sitename %>%
   select(-sitename, -SPstage.Raw_material) %>%

### 2022-08-06 analysis/paper/002-coefficient-variation.R
updated R & rstudio

@@ -168,22 +168,6 @@ cv_by_stage_df  <-
 ####### significant test between the two phases ######
 library(cvequality)
 
-raw_data_test_BL <- df_sitename %>%
-  select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  filter(SPstage.Stage > 1) %>%
-  select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  filter(variable == "BL")
-
-test_BL <- with(raw_data_test_BL,
-                mslr_test(nr = 1e4,
-                          value,
-                          SPstage.Stage)) #p_value = 0.547566
-
 # BM testing doing all the CV testing at once
 
 # created nested tibble, nest by variable
@@ -202,61 +186,12 @@ df_sitename %>%
   nest()
 
 # apply mslr test to each variable (row with a tibble)
+#Krishnamoorthy and Lee’s (2014) modified signed-likelihood ratio test
 df_sitename_nested %>%
   mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
                                             .$value,
                                             .$SPstage.Stage))) %>%
   unnest(mslr_result)
-
-
-
-raw_data_test_ML <- df_sitename %>%
-  select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  filter(SPstage.Stage > 1) %>%
-  select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  filter(variable == "ML")
-
-test_ML <- with(raw_data_test_ML,
-                mslr_test(nr = 1e4,
-                          value,
-                          SPstage.Stage)) #p_value = 0.5159077
-
-raw_data_test_MW <- df_sitename %>%
-  select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  filter(SPstage.Stage > 1) %>%
-  select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  filter(variable == "MW")
-
-test_MW <- with(raw_data_test_MW,
-                mslr_test(nr = 1e4,
-                          value,
[…]

### 2022-08-19 analysis/paper/paper-doc.qmd
add name & affiliation

---
title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
format: docx
editor: visual

author:
  - name: Gayoung Park
    affiliations:
      - University of Washington

  - name: Ben Marwick
    affiliations:
      - University of Washington

bibliography: references.bib

---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-08-19 analysis/paper/paper-doc.qmd
add abstract

@@
   - name: Gayoung Park
-    affiliations:
-      - University of Washington
+    email: gayoungp@uw.edu
+    affiliations: University of Washington
+  - name: Ben Marwick
+    email: bmarwick@uw.edu
+    affiliations: University of Washington
 
-  - name: Ben Marwick
-    affiliations:
-      - University of Washington
+abstract: |
+  Archaeologists have applied evolutionary concepts linking material evidence and cultural phenomena to understand human behavior.
+Evolutionary approaches suggest that technological transitions can occur through changes in social learning processes, and transmission biases are important loci of changes.
+The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, is a key event in understanding modern human dispersal in Northeast Asia.
+Previous studies mainly focus on possible origin locations of new technologies, but they rarely address the process of the technological transition.
+In this research we use a cultural transmission framework to investigate the social contexts that can give us insights into the emergence of these new technologies.
+Our main question is: what was the dominant mode of cultural transmission during the time of technological innovation in the Korean Late Paleolithic?
+Inspired by Bettinger and Eerkens [-@bettinger1999point], we build two models using guided variation and indirect bias.
+To test the models and understand the transmission processes, we use coefficients of variation (CV), and Principal Component Analysis (PCA).
+Here we show that the information about the new technology was transmitted via selective combinations of guided variation and indirect bias.
+We found that some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more dependent on raw materials or other factors.
 

### 2022-08-19 analysis/paper/paper-doc.qmd
add backgrounds

@@
 
-## Quarto
+# Keywords {-}
+ Stemmed points; Korean Paleolithic; Cultural Transmission; Learning process, Transmission Biases
 
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
+# Introduction {-}
 
-## Running Code
+The application of evolutionary theory to archaeological research has guided the interpretation of technological transitions and related human behaviors [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
+For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual @garvey2018current].
+We use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
+This technological innovation is not just the appearance of new tools such as stemmed points and blades, but may also represent a key event related to modern human dispersal in East Asia, if the appearance of these new technologies represents the first arrival of modern humans in this region [@seong2009emergence].
+Previous studies mainly focussed on the possible origin locations of stemmed points, to connect Korea with global patterns of modern human dispersal.
+Questions about the processes of technological change remain largely unanswered [@seong2008tanged;@bae2010origin; @bae2017origin].
+In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic.
+Our main question is: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+Additionally we ask: Do the modes of cultural transmission vary over space?
+Do the modes of cultural transmission vary over time?
+We consider three possibilities, guided variation (trial and error), or indirect bias (copying others), or a combination of the two.
+Our results can help to determine if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.
 
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
+# The Late Paleolithic of the Korean Peninsula {-}
 
-```{r}
-1 + 1
-```
+The onset of the Late Paleolithic period is marked by the appearance of stemmed points around 40-35 ka.
+The stemmed point is a projectile point made out of an elongated flake or blade with slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
+The stemmed point is the first composite tool type and represents new hunting strategies in the Korean Peninsula, as well as other adjacent regions [@seong2008tanged; @lee2019were].
+Stemmed points, combined with blade technology and multi-stage production sequences, and evidence of being resharpened and reused, are important aspects of Late Paleolithic technological innovation in Korea because these features are rarely seen in earlier periods [@chang2013human; @bamforth2009projectile; @seong2015diversity].
+Currently the earliest stemmed points in Northeast Asia are from Yongho-doing site in Korea dated to 38.5ka and made on elongate flakes  [@seong2009emergence; @seong2015diversity; @bae2017origin].
+After their first appearance in Korea, the stemmed points spread to the Japanese archipelago [@chang2013human].
 
-You can add options to executable code like this
+Previous studies of technological transitions during the Late Paleolithic in Korea mainly focussed on possible origin locations of stemmed points, along with the issue of the timing and routes of the dispersal of modern human dispersals in eastern Asia.
+The debate about the origin of stemmed points can be summarized into two competing models: in situ evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
+The in situ model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
+To support his claim of an in situ development, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages.
+In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one.
+He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention.
+In addition, Seong argues that increased numbers of stemmed points over time, and standardization of their shape, supports the idea of gradual, local, evolutionary process.
 
-```{r}
-#| echo: false
-2 * 2
-```
+The migration model argues that the new blade industry including stemmed points and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
+While the in situ model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
+Specifically, the blade technology was introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP while the simple flake tool tradition came from southern China [@bae2013early].
+The migration from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia including Korea in 30-25 ka BP [@shi2005chromosome].
+Bae et al.
+[-@bae2012nature] assume this southern migration was related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
 
-The `echo: false` option disables the printing of code (only output is displayed).
+Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than those models of migration or in situ development.
+He partly agrees with the in situ model that simple flake tools had continuously been used in Korea as the result of ancestor-descendent relationships, under conditions of low effective population size.
+With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration.
+We explore these three options, in situ, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.
+
+# Cultural transmission and transmission biases {-}

### 2022-08-22 analysis/paper/paper-doc.qmd
add a table and image

@@
 bibliography: references.bib
-
 ---
 
-# Keywords {-}
+# Keywords {.unnumbered}
+
 Stemmed points; Korean Paleolithic; Cultural Transmission; Learning process, Transmission Biases
 
-# Introduction {-}
+# Introduction {.unnumbered}
 
@@
 Previous studies mainly focussed on the possible origin locations of stemmed points, to connect Korea with global patterns of modern human dispersal.
-Questions about the processes of technological change remain largely unanswered [@seong2008tanged;@bae2010origin; @bae2017origin].
+Questions about the processes of technological change remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
 In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic.
@@
 
-# The Late Paleolithic of the Korean Peninsula {-}
+# The Late Paleolithic of the Korean Peninsula {.unnumbered}
 
@@
 
-# Cultural transmission and transmission biases {-}
+# Cultural transmission and transmission biases {.unnumbered}
+
+Darwinian evolutionary theory has helped archaeologists to understand technological innovation and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+Cultural evolutionary approaches can explain cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
+The mechanisms are similar to biological evolution, but unique to cultural systems.
+We focus here on how different social contexts result in different modes of social transmission of tool-making skills.
+Social transmission can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
+
+Cultural transmission theory holds that information about how to behave is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
+Individuals learn by themselves (e.g.
+trial and error), or from each other by sharing information.
+Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire through recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual].
+Transmission biases can be important loci of changes in material culture, and can be influenced by social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social].
+These biases include guided variation (where individuals learn new behaviors and then modify them through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular instructions based on their perceived frequency in the population, such as extremely popular or rare instructions), and indirect bias (where a variant is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].
+
+Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how these biases can be identified in stone artifact assemblages.
+Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
+They equated guided transmission with high metric variation and low correlation between metric attributes.
+Conversely, they inferred indirect bias from less variation and more correlated variables.
+They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools.
+From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly, with new technologies developed largely by trial and error.
+On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.
+
+Garvey [-@garvey2018current] also explores Bettinger and Eerkens' framework and uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points.
+With the two types of projectile points, Washita and Fresno points, from Henderson site, she measured CV values of weight, thickness, width and lengths.
+She then computed simulated CV values according to various learning scenarios of knapper skill, raw material quality and human perception (i.e.
+CV = 10%, 5%, and 3%).
+The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying".
+Garvey's modeling and simulation shows how the concept of cultural transmission can be useful to interpret the archaeological record.
+
+Although Garvey and Bettinger and Eerkens were unable to directly observe transmission processes, Mesoudi and O'Brien [-@mesoudi2008cultural] conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
+They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants.
+They also simulated guided variation by allowing their participants to explore their own designs.
+They observed that the majority of participants who were able to choose the previous design copied the most-successful model.
+Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
+Mesoudi and O'Brien's results demonstrate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture variation.
+
+# Modeling the social context of the appearance of stemmed points {.unnumbered}
+
+Inspired by the previous studies that used Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect, to understand the spread of stemmed point technology during the Korean Late Paleolithic period.
+We propose a spectrum for foraging groups and how they started to make stemmed points: on one end we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.
+
+To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups.
+Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement.
+This trial and error behavior leaves a distinctive signature on metric variables of the stemmed points.
+If trial and error was the dominant bias in the appearance of stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation between sites.
+
+On the other end of our spectrum, socially connected groups occupied places close to other groups or had regular contact with other groups over long or short distances.
+Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
+The model stemmed point might be a highly successful and frequently chosen one, and the learners copy all information about the point design as a package.
+In this social context, indirect bias is the dominant influence on transmission of lithic technology.
+If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will be less varied.
+
++-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
+| Phase | SP                 | Blades  | Micro-blade | SP Blank           | Newly added raw materials             | New changes                                              | Example sites | Radiocarbon ages |
++=======+====================+=========+=============+====================+=======================================+==========================================================+===============+==================+
+| 1     | Present            | Absent  | Absent      | Flake              | Higher-quality quartzite, vein quartz | Increasing number of small flake tools, appearance of SP | Hwadaeri\     | \~35 ka\         |
+|       |                    |         |             |                    |                                       |                                                          | Yonghodong\   | \                |
+|       |                    |         |             |                    |                                       |                                                          | Sachang       |                  |
+|       |                    |         |             |                    |                                       |                                                          |               |                  |
+|       |                    |         |             |                    |                                       |                                                          | \             |                  |
++-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
+| 2     | Present            | Present | Present     | Blade              | Porphyry, siliceous shale, hitherto   | Appearance of blades, increasing number of SP            |               | 35-25 ka         |
++-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
+| 3     | Present until 16ka | Present | Present     | Blade, Micro-blade | Obsidian                              | Appearance of Micro-blade, microliths                    |               | 25-12 ka         |
+|       |                    |         |             |                    |                                       |                                                          |               |                  |
+|       |                    |         |             |                    |                                       |                                                          |               | \                |
++-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
+
+: Korean Paleolithic Chronology edited based on \@seong2015diversity; \@Park_2013
+
+# Materials and methods {.unnumbered}
+
+1.  Materials and stemmed points chronology
+
+After the first discovery in Seokjangri site in the 1960s, around 300 stemmed points have been found in Korea.
+Most sites contain only a few points but a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
+Among these stemmed points, we selected those that are unbroken from the tip to the stem.
+We excluded artifacts that were recorded as stemmed points but lack a stem.
+This resulted in a sample of 152 stemmed points from 28 assemblages unearthed from 20 sites spanning the period 40-17 ka (Figure 1).
+The images of the stemmed points were obtained from published excavation reports and by direct photography during our research on local museum collections.
+We defined multiple assemblages in a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
+
+We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
+These chronologies were built on the presence of blades, stemmed point blanks, toolkit composition and radiocarbon ages.
+We arranged our 28 assemblages containing stemmed points in chronological order.
+For example, Sachang site has two assemblages that are fitted into phase 1 and 2.
+Applying the chronology is useful because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research situations and limitations, chronological order help us estimate the time periods of those assemblages as well as understand the morphological change of stemmed points associated with the introduction of blade technology.
+Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages.
+
+![Figure 1. Korean Paleolithic sites mentioned in this study](images/001-site-map.png)

### 2022-08-22 analysis/paper/paper-doc.qmd
figure out cross-referencing

@@
 
-+-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
++-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
 | Phase | SP                 | Blades  | Micro-blade | SP Blank           | Newly added raw materials           | New changes                                              | Example sites | Radiocarbon ages |
-+=======+====================+=========+=============+====================+=======================================+==========================================================+===============+==================+
-| 1     | Present            | Absent  | Absent      | Flake              | Higher-quality quartzite, vein quartz | Increasing number of small flake tools, appearance of SP | Hwadaeri\     | \~35 ka\         |
-|       |                    |         |             |                    |                                       |                                                          | Yonghodong\   | \                |
++=======+====================+=========+=============+====================+=====================================+==========================================================+===============+==================+
+| 1     | Present            | Absent  | Absent      | Flake              | Higher-quality quartzite,\          | Increasing number of small flake tools, appearance of SP | Hwadaeri\     | \~35 ka\         |
+|       |                    |         |             |                    | Vein quartz                         |                                                          | Yonghodong\   | \                |
 |       |                    |         |             |                    |                                     |                                                          | Sachang       |                  |
@@
 |       |                    |         |             |                    |                                     |                                                          | \             |                  |
-+-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
-| 2     | Present            | Present | Present     | Blade              | Porphyry, siliceous shale, hitherto   | Appearance of blades, increasing number of SP            |               | 35-25 ka         |
-+-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
++-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
+| 2     | Present            | Present | Present     | Blade              | Porphyry, Siliceous shale, Hitherto | Appearance of blades, increasing number of SP            |               | 35-25 ka         |
++-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
 | 3     | Present until 16ka | Present | Present     | Blade, Micro-blade | Obsidian                            | Appearance of Micro-blade, microliths                    |               | 25-12 ka         |
@@
 |       |                    |         |             |                    |                                     |                                                          |               | \                |
-+-------+--------------------+---------+-------------+--------------------+---------------------------------------+----------------------------------------------------------+---------------+------------------+
++-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
 
@@
 
-![Figure 1. Korean Paleolithic sites mentioned in this study](images/001-site-map.png)
+![Korean Paleolithic sites mentioned in this study](images/001-site-map.png){#fig-one width="477"}
+
+2.  Methods
+
+![(a) Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks:\
+ML: Maximum length, tip to bottom, perpendicular to the length axis 1- 7\
+BL: Body length, tip to the closest wing, perpendicular to the length axis 1- 3 or 1-10 depending on the artifact\
+TL: Tang length, the closest wing from the tip to bottom,perpendicular to the length axis 3 - 7 or 10-7 depending on the artifact\
+SL: (Maximum) stem length, perpendicular to the length axis. Closer tang curve's middle point from the tip to the most distant point of the basal end 9-7 or 4-7 depending on the artifact\
+MW: mid width, dimension from margin to margin at the mid-point of the length 2 - 11 axis, perpendicular to the width axis.\
+TW: tang width, dimension between each wing, perpendicular to the width axis 3 - 10\
+SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="221"}
+
+In addition to attributes from previous studies of cultural transmission and projectile points, i.e.
+maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-two).
+We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts.
+Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+We did not include weight and thickness as these measurements were not available to us.
+For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses.
+The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.

### 2022-08-22 analysis/paper/paper-doc.qmd
finished methods

@@
 We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.
+
+To measure the variations between stemmed points, we used Coefficients of Variation (CV) with confidence intervals.
+CV is the ratio of the standard deviation to the mean.
+It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc.
+in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence].
+In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes including distinguishing different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
+Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2% (i.e.
+CV for planned production = 1.7 %) and about 60% (i.e.
+CV for random production = 57.7%) and the values over the upper limit likely reflect intended attempts to differentiate the products.
+Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
+Considering the lower quality of raw materials in Korea, we sat roughly 25 as the threshold to decide the type of transmission biases.
+
+Sample size can impact on accuracy of estimated CV that reflects the corresponding population size.
+[@toebe2018sample; @kelley2007sample; @vanpool2011quantitative].
+In archeological research, sample size for CV values varies from n=5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
+A variety of methods are available for improving the reliability of CV measurements on small samples.
+@vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e.
+n\<25).
+Statistical research has results in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
+Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates @kelley2007sample; @albatineh2014con fidence].
+
+We chose Principal Component Analysis (PCA) to compute Pearson's r correlations between the attributes as previous studies used [@bettinger1999point; @eerkens2008cultural; @garvey2018current] and to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
+PCA is often used to detect relevant information from dataset, indicate relationships between variables as well as cases, extract general patterns in data, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life].
+PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].
+
+All data treatment, analyses and visualization were performed in the R environment[@core2021team].
+Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well at DOI to enable full reproducibility [@marwick2017computational].
+
+# Results {.unnumbered}
+
+@fig-three shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points.
+This result implies a similar thickness of a wooden shaft that is attached to the stem.
+
+![Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value.](images/003-cv-sharma.png){#fig-three width="397"}

### 2022-08-22 analysis/paper/paper-doc.qmd
edit pca plots

@@
 
+In addition to attributes from previous studies of cultural transmission and projectile points, i.e.
+maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-two).
+We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts.
+Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+We did not include weight and thickness as these measurements were not available to us.
+For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses.
+The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.
+
 ![(a) Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks:\
@@
 SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="221"}
-
-In addition to attributes from previous studies of cultural transmission and projectile points, i.e.
-maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-two).
-We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts.
-Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
-We did not include weight and thickness as these measurements were not available to us.
-For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses.
-The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.
 
@@
 
+1.  Distribution of attributes
+
 @fig-three shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
@@
 
-![Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value.](images/003-cv-sharma.png){#fig-three width="397"}
+![Boxplot for each attribute measured on the stemmed points.
+The black bold horizontal lines in the middle of boxes represent the median value.](images/002-attributes-variation-01.png){#fig-three width="324"}
+
+2.  Coefficient of variation
+
+We used CV to measure the variations between stemmed points.
+If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias.
+Our results show that CV values for all attributes are distributed from 23.8 to 36.3 (@fig-four).
+Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.
+Unlike the distribution of actual attributes (@fig-three), the CV of body length and maximum length have the lowest values.
+The lower CV of maximum length indicates highly standardized with narrowest confidence intervals.
+Tang and stem related attributes (e.g.
+SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes on the artifacts.
+Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.
+
+![The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
+95% confidence intervals for the CVs were computed using Sharma & Krishna's method.
+The vertical lines indicate the range of confidence intervals.](images/003-cv-sharma-01.png){#fig-four width="339"}
+
+3.  Principal Component Analysis
+
+PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.Figure 5 shows positive relationship between attributes.
+Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length.
+Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other.
+Other width and tang related attributes are close to each other but a bit far from body length and maximum length.
+Stem width contributes the least to the overall shape.
+We can assume that

### 2022-08-22 analysis/paper/002-coefficient-variation.R
results done

@@ -124,7 +124,8 @@ cv_plot_site_label %>%
   theme(legend.position = "none") +
   coord_cartesian(ylim = c(0,130)) +
   theme_bw() +
-  labs(color = "Attributes")
+  labs(color = "Attributes") +
+  theme(strip.text.x = element_text(size = 6.5))
 
 
 # checking SC site values

### 2022-08-22 analysis/paper/002-coefficient-variation.R
stash

@@ -184,31 +184,6 @@ test_BL <- with(raw_data_test_BL,
                           value,
                           SPstage.Stage)) #p_value = 0.547566
 
-# BM testing doing all the CV testing at once
-
-# created nested tibble, nest by variable
-df_sitename_nested <-
-df_sitename %>%
-  select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  filter(SPstage.Stage > 1) %>%
-  select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  ungroup() %>%
-  group_by(variable) %>%
-  nest()
-
-# apply mslr test to each variable (row with a tibble)
-df_sitename_nested %>%
-  mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
-                                            .$value,
-                                            .$SPstage.Stage))) %>%
-  unnest(mslr_result)
-
-
 
 raw_data_test_ML <- df_sitename %>%
   select(-sitename, -SPstage.Raw_material) %>%

### 2022-08-22 analysis/paper/paper-doc.qmd

@@
 Statistical research has results in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
-Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates @kelley2007sample; @albatineh2014con fidence].
+Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates [@kelley2007sample; @albatineh2014confidence].
 
@@
 
-PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.Figure 5 shows positive relationship between attributes.
+PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.
+@fig-five shows positive relationship between attributes.
 Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length.
@@
 Stem width contributes the least to the overall shape.
-We can assume that
+We can assume that the body length and maximum length were prioritized when the stemmed points were transmitted.
+
+![PCA analysis for stemmed points.
+The direction of the arrow shows the relationship between attributes.
+The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA-1.png){#fig-five width="284"}
+
+4.  Variation Over Time
+
+To understand the temporal patterns in the modes of cultural transmission, we applied Korea Late Paleolithic chronological phases into our samples and computed CV and PCA.
+Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute.
+However, this does not mean that there were only two stemmed points for the Phase 1 because we excluded broken pieces from our sample to conduct the morphological analysis.
+
+After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-six).
+Body length of Phase 2, maximum length, mid width, and tang width are under 25, which indicates as a result of indirect bias, while only maximum length are under 25 in Figure 4.
+On the other hand, Body length of Phase 3, stem length, stem width, and tang length are relatively higher then 25 and mostly over 30 indicating guided variation.
+
+Compared to Phase 2, except for maximum width and tang width, the other CV values of Phase 3 tend to be higher.
+In addition, Phase 3 tends to have wider confidence intervals.
+Stem width and tang length values are relatively steady over time.
+Our results show that stemmed points from Phase 2 are less varied than Phase 3 as well as Phase 1.
+
+![CV values for the second and third chronological phases in the Korean Late Paleolithic period.
+The points represent CV values for each attribute.
+The vertical lines indicate the range of confidence intervals.](images/004-cv-phases-01.png){#fig-six width="346"}
+
+Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-seven).
+However, there are no notable changes in correlations over time.
+
+![PCA analysis for the second and third chronological phases in the Korean Late Paleolithic period.
+The direction of the arrow shows the relationship between attributes.
+The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA.png){#fig-seven width="534"}
+
+We explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people.
+There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain seven or more stemmed points.
+SYG1_2 and SYG6_2 are from Phase 2 and SYG1_3 and YS_3 are from Phase 3.
+Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width.
+@fig-eight shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias.
+Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals.
+Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-six results of body length and maximum length from all assemblages.
+It is assumed that those attributes became more standardized within the same assemblages.
+Compared to Suyanggae assemblages, Yongsandong has smaller CV values.
+
+![CV values of four attributes from assemblages that have 7 or more stemmed points.
+Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.](images/006-cv-four-assemblage.png){#fig-eight width="386"}
+
+5.  Variation Between sites
+
+We examined the CV values per site to address our research question about the cultural transmission over space.
+We excluded assemblages that contain a single point.
+@fig-nine shows that Sachang site has the highest CV values with the widest confidence level.
+Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2.
+We excluded sites that have only one stemmed point for the analysis.
+Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points.
+CV values of body length is the lowest while stem width the highest among all assemblages.
+Overall, there are no clear patterns between the sites.
+
+![CV values per site that has more than one stemmed point.
+The points represent CV values for each attribute.
+The vertical lines indicate the range of confidence intervals.](images/004-cv-sites-01.png){#fig-nine width="519"}
+
+6.  Variation Between Raw Materials
+
+We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials.
+We exclude raw materials that were used for single point.
+@fig-ten shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals.
+Quartz has high variations of CV among the attributes.
+On the other hand, shale has the lower CV, which is very stable throughout the attributes.
+
+![CV values for raw materials that were used for making more than one stemmed point.
+The points represent CV values for each attribute.
+The vertical lines indicate the range of confidence intervals.](images/00000-cv-raw-materials.png){#fig-ten width="526"}
+
+# Discussion {.unnumbered}
+
+# References {.unnumbered}

### 2022-08-22 analysis/paper/002-coefficient-variation.R
fix merge conflict, remove redundant code

@@ -26,9 +26,9 @@ cv_all <- map_df(DF, cv)
 #arrange the data properly
 cv_data_all <- gather(cv_all)
 
-hist(DF$ML)
+# hist(DF$ML)
 
-hist(DFCV)
+# hist(DFCV)
 
 # to install MKmisc
 ## Install package BiocManager
@@ -169,32 +169,9 @@ cv_by_stage_df  <-
 ####### significant test between the two phases ######
 library(cvequality)
 
-<<<<<<< HEAD
-raw_data_test_BL <- df_sitename %>%
-  select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  filter(SPstage.Stage > 1) %>%
-  select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  filter(variable == "BL")
-
-test_BL <- with(raw_data_test_BL,
-                mslr_test(nr = 1e4,
-                          value,
-                          SPstage.Stage)) #p_value = 0.547566
-
-
-raw_data_test_ML <- df_sitename %>%
-=======
-# BM testing doing all the CV testing at once
-
 # created nested tibble, nest by variable
 df_sitename_nested <-
 df_sitename %>%
->>>>>>> 602ceeb7953d98ce8aefe3d214027b9d8367a727
   select(-sitename, -SPstage.Raw_material) %>%
   group_by(SPstage.Stage) %>%
   add_tally() %>%
@@ -203,46 +180,6 @@ df_sitename %>%
   pivot_longer(-SPstage.Stage,
                names_to = "variable",
                values_to = "value") %>%
-<<<<<<< HEAD
-  filter(variable == "ML")
-
-test_ML <- with(raw_data_test_ML,
-                mslr_test(nr = 1e4,
-                          value,
-                          SPstage.Stage)) #p_value = 0.5159077
-
-raw_data_test_MW <- df_sitename %>%
-  select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  filter(SPstage.Stage > 1) %>%
-  select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  filter(variable == "MW")
-
-test_MW <- with(raw_data_test_MW,
-                mslr_test(nr = 1e4,
[…]

### 2022-08-22 analysis/paper/000-draw-the-map.R
reorganise some file locations, review code to map the map

@@ -1,26 +1,18 @@
+# This script draws the map, this is figure 1 in the paper.
+
+
 korean_archaeological_site_locations <-
-  readr::read_csv(here::here("analysis/data/sitelocation.csv"))
+  readr::read_csv(here::here("analysis/data/raw_data/sitelocation.csv"))
 
 korean_archaeological_site_locations <-
   tibble::rowid_to_column(korean_archaeological_site_locations, "ID")
 
-
-library(ggmap)
-
-# bounding box
-# 39.534214, 124.159154 ... 39.657415, 129.138604
-# 33.694662, 123.491142 ... 34.338454, 130.569658
-
-map <-
-  get_stamenmap(bbox = c(left = 124,
-                         bottom = 33,
-                         right = 	130,
-                         top = 39),
-                zoom = 10)
-
 library(ggrepel)
 # devtools::install_github('3wen/legendMap')
 library(legendMap)
+
+# load map tiles we previously saved
+map <- readRDS(here::here("analysis/data/raw_data/map_tiles.rds"))
 
 # map with site names next to the points
 map_of_sites <-
@@ -136,7 +128,6 @@ ggdraw() +
   ) +
   theme(panel.background = element_rect(fill='white', colour="white"),
         plot.background = element_rect(fill='white', colour="white"))
-
 
 ggsave(here::here("analysis/figures/001-site-map.png"),
        width = 4.45,

### 2022-08-22 analysis/paper/paper-doc.qmd
review and minor edits to import-and-clean-data

@@
 
-The onset of the Late Paleolithic period is marked by the appearance of stemmed points around 40-35 ka.
+The onset of the Late Paleolithic period in Korea is marked by the appearance of stemmed points around 40-35 ka.
 The stemmed point is a projectile point made out of an elongated flake or blade with slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
@@
 The in situ model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
-To support his claim of an in situ development, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages.
+To support his claim of an *in situ* development, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages.
 In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one.
@@
 
-Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than those models of migration or in situ development.
-He partly agrees with the in situ model that simple flake tools had continuously been used in Korea as the result of ancestor-descendent relationships, under conditions of low effective population size.
+Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than those models of migration or *in situ* development.
+He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size.
 With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration.
-We explore these three options, in situ, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.
+We explore these three options, *in situ*, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.
 
@@
 
-After the first discovery in Seokjangri site in the 1960s, around 300 stemmed points have been found in Korea.
+After the first discovery in Seokjangri site in the 1960s, around 300 stemmed points have been found in nearly 30 sites across Korea (@fig-map).
 Most sites contain only a few points but a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
@@
 
-![Korean Paleolithic sites mentioned in this study](images/001-site-map.png){#fig-one width="477"}
+![Korean Paleolithic sites mentioned in this study](images/001-site-map.png){#fig-map width="477"}
 

### 2022-08-22 analysis/paper/000-import-and-clean-data.R

@@ -42,7 +42,7 @@ landmarks <- read.tps(here::here("analysis/data/raw_data/original-landmark-data/
 
 #getting info for phase of each site
 # SPstage <- read_csv(here("analysis/data/SPstage.csv"))
-SPstage <- read_excel(here("analysis/data/SPstage_raw_materials.xls"))
+SPstage <- readxl::read_excel(here("analysis/data/raw_data/SPstage_raw_materials.xls"))
 
 #Compare multiple artifacts
 #Open multiple tps files. This will need to be changed to read files inside the project
@@ -50,7 +50,7 @@ datadir <- here("analysis/data/raw_data/original-landmark-data/")
 files <- dir(datadir, pattern = "*.tps$|*.TPS$")
 
 #Getting site info
-korean_archaeological_site_locations <- read_csv(here("analysis/data/sitelocation.csv"))
+korean_archaeological_site_locations <- read_csv(here("analysis/data/raw_data/sitelocation.csv"))
 
 #Computing size attributes  of the points by calculating distance between landmarks
 
@@ -147,19 +147,8 @@ landmarks_list <- vector("list", length = length(files))
 no_curves_files <- list.files(here::here("analysis/data/raw_data/no-curves-landmark-data"),
                               full.names = TRUE)
 
-# no_curves_files_no_bad_ones <- # remove bad files or fix them
-#
-# for (i in 1:length(no_curves_files)) {
-#   filename <- no_curves_files[i]
-#   print(filename)
-#
-#   # read in landmark files
-#   landmarks_list[[i]] <- geomorph::readland.tps(filename)
-# }
-
 # this is the one we want, it produces the output in a format we can easily work with
 x2 <- geomorph::readmulti.tps(no_curves_files)
-
 
 # attach artefact ID's to landmarks
 names(landmarks_list) <- files
@@ -169,12 +158,12 @@ landmarks_list_df <- bind_rows(landmarks_list, .id = "artefact")
 
 library(Momocs)
 
+look_at_the_outlines <-
 Out(x2, fac = no_curves_files) %>%
   coo_align() %>%
   coo_center() %>%
   coo_scale() %>%
   pile(transp = 0.9)
-
 
 #Gather attributes from  all artifacts
 
@@ -185,6 +174,7 @@ SL = c()
 MW = c()
 TW = c()
 SW = c()
+
 for (i in 1:length(files)) {
   filename <- paste(datadir, files[i], sep = "")
[…]

### 2022-08-22 analysis/paper/001-basic-analysis.R
review 'basic-analysis' and delete code we don't use

@@ -1,14 +1,12 @@
-load("data_main.RData")
+
 library(tidyverse)
 library(ggplot2)
 library(ggbeeswarm)
-#Box plot for each attributes
 
-box1 <- boxplot(DF,
-                main = "Variation for each attributes",
-                col = "gold",
-                xlab = "Attributes",
-                ylab = "Length (cm)")
+# import the cleaned data ready to work on
+load(here::here("analysis/data/derived_data/data_main.RData"))
+
+# Box plot for each attributes
 
 # mutate data to make plot with ggplot
 data_box <- gather(DF)
@@ -27,102 +25,4 @@ ggsave(here::here("analysis/figures/002-attributes-variation.png"),
        width = 4.45,
        height = 4.5,
        units = "in")
-
-# take a look at the distribution of each variable
-DF %>%
-  pivot_longer(everything()) %>%
-  ggplot() +
-  aes(value) +
-  geom_density() +
-  facet_wrap( ~ name) +
-  theme_minimal()
-
-# fit distributions for each of these, so we can simulate from those distributions
-library("fitdistrplus")
-library(distributions3)
-
-# what distributions will explore to see if they match with our data?
-the_distributions <- c("weibull",   "norm" )
-the_distributions3 <- c(Weibull,  Normal)
-names(the_distributions3) <- the_distributions
-
-fit_distributions <-
-DF %>%
-  pivot_longer(everything())  %>%
-  nest(-name) %>%
-  # compute fit of each variable with a selection of distributions
-  mutate(dist_out_w = map(data, ~fitdist(unlist(.x), the_distributions[1]))) %>%
-  mutate(dist_out_l = map(data, ~fitdist(unlist(.x), the_distributions[2]))) %>%
- # mutate(dist_out_g = map(data, ~fitdist(unlist(.x), the_distributions[3]))) %>%
- # mutate(dist_out_n = map(data, ~fitdist(unlist(.x), the_distributions[4]))) %>%
-  pivot_longer(cols = starts_with("dist"),
-               names_to = "dist") %>%
-  nest(-name) %>%
-  # extract goodness of fit statistics
-  mutate(gofstat_out = map(data, ~gofstat(.x$value,
-                                          fitnames = the_distributions))) %>%
-  # AIC, in particular, which distribution has the lowest value (so, best fit)
-  mutate(gofstat_out_aic = map_chr(gofstat_out,
[…]

### 2022-08-22 analysis/paper/paper-doc.qmd
review 'basic-analysis' and delete code we don't use
minor edits around fig 3

@@
 
+```{r}
+
+```
+
 1.  Materials and stemmed points chronology
@@
 ![Boxplot for each attribute measured on the stemmed points.
-The black bold horizontal lines in the middle of boxes represent the median value.](images/002-attributes-variation-01.png){#fig-three width="324"}
+The black bold horizontal lines in the middle of boxes represent the median value.
+The dots represent individual artefacts.](images/002-attributes-variation-01.png){#fig-three width="324"}
 

### 2022-08-22 analysis/paper/001-basic-analysis.R

@@ -1,3 +1,4 @@
+# this code makes Fig 3: Boxplot for each attribute measured on the stemmed points.
 
 library(tidyverse)
 library(ggplot2)
@@ -11,7 +12,7 @@ load(here::here("analysis/data/derived_data/data_main.RData"))
 # mutate data to make plot with ggplot
 data_box <- gather(DF)
 
-# make box plot for each attributes with better visulazation options with ggplot
+# make box plot for each attributes
 ggplot(data_box, aes(key, value)) +
   geom_boxplot() +
   geom_beeswarm(alpha = 0.3,

### 2022-08-22 analysis/paper/000-import-and-clean-data.R
add space to site names so plot labels are neater

@@ -199,7 +199,9 @@ df_sitename <- df %>%
 
 library(dplyr)
 df_full_sitename <-
-  left_join(df_sitename, korean_archaeological_site_locations, by = c("sitename"="sitename"))
+  left_join(df_sitename,
+            korean_archaeological_site_locations,
+            by = c("sitename"="sitename"))
 
 #list of all attributes without ID
 DF <- df[,1:7]

### 2022-08-22 analysis/paper/002-coefficient-variation.R

@@ -1,10 +1,7 @@
-library(tidyr)
-library(purrr)
-library(dplyr)
-library(readr)
-library(stringr)
-library(ggplot2)
-load("data_main.RData")
+library(tidyverse)
+
+# import the cleaned data ready to work on
+load(here::here("analysis/data/derived_data/data_main.RData"))
 
 # Coefficient of variation (CV): the ratio of the standard deviation divided by mean
 
@@ -19,7 +16,7 @@ mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100 * (1 + 1/(4*length(x)))
 # rows and specimens, and values as dimensions
 # need to add a column of site ID
 
-DFCV<- map_dbl(DF, cv)
+DFCV <- map_dbl(DF, cv)
 
 #Same as map but better view for summary
 cv_all <- map_df(DF, cv)
@@ -56,7 +53,6 @@ cv_data_sharma <- data.frame(all_cv_sharma, all_low_sharma, all_high_sharma)
 library(tibble)
 cv_data_sharma_named <- rownames_to_column(cv_data_sharma, var = "attribute")
 
-
 library(ggplot2)
 ggplot(cv_data_sharma_named, aes(attribute, all_cv_sharma)) +        # ggplot2 plot with confidence intervals
   geom_point() +
@@ -75,11 +71,11 @@ ggsave(here::here("analysis/figures/003-cv-sharma.png"),
 #Make a table of CVs for all variable grouped by site
 cv_by_site_df  <-
   df_full_sitename %>%
-  select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -SPstage.Raw_material) %>%
+  dplyr::select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -SPstage.Raw_material) %>%
   group_by(full_sitename) %>%
   add_tally() %>%
-  filter( n > 1) %>%
-  select(-n) %>%
+  dplyr::filter( n > 1) %>%
+  dplyr::select(-n) %>%
   pivot_longer(-full_sitename,
                names_to = "variable",
                values_to = "value") %>%
@@ -93,28 +89,28 @@ cv_by_site_df  <-
 
 cv_by_full_site_df_label <-
   df_full_sitename %>%
-  select(-SPstage.Stage) %>%
+  dplyr::select(-SPstage.Stage) %>%
   group_by(full_sitename) %>%
   add_tally() %>%
   filter( n > 1) %>%
[…]

### 2022-08-22 analysis/paper/000-import-and-clean-data.R
review of 002-coefficient-variation to neaten the plots, remove the colour

@@ -195,7 +195,7 @@ df <- data.frame(ML, BL, TL, SL, MW, TW, SW, files, SPstage$Stage, SPstage$Raw_m
 df_sitename <- df %>%
   mutate (files = str_remove_all(files, "_.tps")) %>%
   separate (files, into = c("sitename", "files"), sep = "_") %>%
-  select (-files)
+  dplyr::select (-files)
 
 library(dplyr)
 df_full_sitename <-

### 2022-08-22 analysis/paper/002-coefficient-variation.R

@@ -1,3 +1,15 @@
+# this script makes figure 4: CV values for each attribute measured on the
+# stemmed points. 95% confidence intervals for the CVs were computed using
+# Sharma & Krishna's method.
+
+# This script also makes figure 9: CV values per site that has more than one stemmed point.
+# The points represent CV values for each attribute. The vertical lines
+# indicate the range of confidence intervals.
+
+# This script also makes figure 10: CV values for raw materials that were
+# used for making more than one stemmed point. The points represent CV
+# values for each attribute. The vertical lines indicate the range of confidence interval
+
 library(tidyverse)
 
 # import the cleaned data ready to work on
@@ -56,7 +68,7 @@ cv_data_sharma_named <- rownames_to_column(cv_data_sharma, var = "attribute")
 library(ggplot2)
 ggplot(cv_data_sharma_named, aes(attribute, all_cv_sharma)) +        # ggplot2 plot with confidence intervals
   geom_point() +
-  geom_errorbar(aes(ymin = all_low_sharma, ymax = all_high_sharma), width = .2) +
+  geom_errorbar(aes(ymin = all_low_sharma, ymax = all_high_sharma), width = 0.2) +
   ylab("CV") +
   xlab("Variables") +
   geom_text(aes(label = round(all_cv_sharma,1)), col="blue", hjust = -0.3, size = 3) +
@@ -67,8 +79,8 @@ ggsave(here::here("analysis/figures/003-cv-sharma.png"),
        height = 4.5,
        units = "in")
 
-#####################
-#Make a table of CVs for all variable grouped by site
+#-Site by site -------------------------------
+
 cv_by_site_df  <-
   df_full_sitename %>%
   dplyr::select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -SPstage.Raw_material) %>%
@@ -99,30 +111,28 @@ cv_by_full_site_df_label <-
   mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>%
   dplyr::select (-cv_by_site)
 
-
 ## add label to main dataframe
 
 cv_plot_site_label <- cv_by_site_df %>%
   left_join (cv_by_full_site_df_label) %>%
   dplyr::select (- data,-full_sitename)
 
-
 ## facet plot
 
 cv_plot_site_label %>%
-  ggplot(aes(x = variable, y = cv_by_site)) +
+  ggplot(aes(x = variable,
[…]

### 2022-08-22 analysis/paper/paper-doc.qmd
remove correlation script as we are not using that in the paper

@@
 This technological innovation is not just the appearance of new tools such as stemmed points and blades, but may also represent a key event related to modern human dispersal in East Asia, if the appearance of these new technologies represents the first arrival of modern humans in this region [@seong2009emergence].
-Previous studies mainly focussed on the possible origin locations of stemmed points, to connect Korea with global patterns of modern human dispersal.
+Previous studies mainly focused on the possible origin locations of stemmed points, to connect Korea with global patterns of modern human dispersal.
 Questions about the processes of technological change remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
@@
 
-1.  Materials and stemmed points chronology
+## Materials and stemmed points chronology
 
@@
 
-2.  Methods
+## Methods
 
@@
 [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative].
-In archeological research, sample size for CV values varies from n=5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
+In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
 A variety of methods are available for improving the reliability of CV measurements on small samples.
@@
 n\<25).
-Statistical research has results in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
+Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
 Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates [@kelley2007sample; @albatineh2014confidence].
@@
 If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias.
-Our results show that CV values for all attributes are distributed from 23.8 to 36.3 (@fig-four).
+Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-four).
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.

### 2022-08-22 analysis/paper/001-basic-analysis.R

@@ -7,13 +7,20 @@ library(ggbeeswarm)
 # import the cleaned data ready to work on
 load(here::here("analysis/data/derived_data/data_main.RData"))
 
+# how many stemmed artefacts per site?
+stemmed_points_per_site <-
+df_full_sitename %>%
+  count(full_sitename) %>%
+  arrange(desc(n))
+
 # Box plot for each attributes
 
 # mutate data to make plot with ggplot
 data_box <- gather(DF)
 
 # make box plot for each attributes
-ggplot(data_box, aes(key, value)) +
+ggplot(data_box,
+       aes(key, value)) +
   geom_boxplot() +
   geom_beeswarm(alpha = 0.3,
                 size = 0.8,

### 2022-08-22 analysis/paper/002-coefficient-variation.R

@@ -47,7 +47,8 @@ cv_data_all <- gather(cv_all)
 # install.packages("MKmisc")
 
 library(MKmisc)
-#CV value
+
+# CV value
 sharma_cv <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100
 
 #CV sharma interval values
@@ -65,8 +66,14 @@ cv_data_sharma <- data.frame(all_cv_sharma, all_low_sharma, all_high_sharma)
 library(tibble)
 cv_data_sharma_named <- rownames_to_column(cv_data_sharma, var = "attribute")
 
+# for in-text numbers
+max_cv <- round(max(cv_data_sharma_named$all_cv_sharma), 1)
+min_cv <- round(min(cv_data_sharma_named$all_cv_sharma), 1)
+
 library(ggplot2)
-ggplot(cv_data_sharma_named, aes(attribute, all_cv_sharma)) +        # ggplot2 plot with confidence intervals
+ggplot(cv_data_sharma_named,
+       aes(attribute,
+           all_cv_sharma)) +        # ggplot2 plot with confidence intervals
   geom_point() +
   geom_errorbar(aes(ymin = all_low_sharma, ymax = all_high_sharma), width = 0.2) +
   ylab("CV") +

### 2022-08-22 analysis/paper/003-correlation.R

@@ -1,56 +1,0 @@
-
-library(tidyr)
-library(corrr)
-load("data_main.RData")
-
-#using corrr package
-corrr_all <- correlate(DF) %>%
-  shave()
-
-call<-rplot(corrr_all) +
-  ggtitle("Correlation for all phase") +
-  theme(legend.position = "none")
-
-direction_all <- correlate(DF) %>%
-  network_plot()
-
-#phase 2
-
-DF2<- df_sitename %>%
-  filter(SPstage.Stage == 2) %>%
-  select(-sitename, -SPstage.Stage)
-
-corrr_2 <-correlate(DF2) %>%
-  shave()
-
-ctwo <- rplot(corrr_2) +
-  ggtitle("Correlation for 2nd phase") +
-  theme(legend.position = "none")
-
-direction_2 <- correlate(DF2) %>%
-  network_plot()
-
-#phase 3
-
-DF3<-df_sitename %>%
-  filter(SPstage.Stage == 3) %>%
-  select(-sitename, -SPstage.Stage)
-
-corrr_3 <-correlate(DF3) %>%
-  shave()
-
-cthree <- rplot(corrr_3) +
-  ggtitle("Correlation for 3rd phase")
-
-direction_3 <-correlate(DF3) %>%
-  network_plot()
-
-### three phases at once
-library(patchwork)
-call + ctwo + cthree + plot_layout(nrow=1)
-
-ggsave(here::here("analysis/figures/006-correlation.png"),
-       width = 10,
-       height = 3,
-       units = "in")
-

### 2022-08-23 analysis/paper/004-principal-component-analysis.R
delete duplicate images, very confusing

@@ -4,10 +4,10 @@ library(corrr)
 library(ggbiplot)
 library(ggfortify)
 
-load("data_main.RData")
+# import the cleaned data ready to work on
+load(here::here("analysis/data/derived_data/data_main.RData"))
 
 ### PCA analysis using ggbiplot, ggfortify pkgs
-
 
 PCA <- prcomp(DF, center = TRUE, scale = TRUE)
 
@@ -30,15 +30,14 @@ label_phase <-
   add_tally() %>%
   mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
 
-
 count_phase <- length(which(label_phase[9]>0))
 count_phase2 <- length(which(label_phase[9] == 2))
 count_phase3 <- length(which(label_phase[9] == 3))
 
-
 ###
 
-#plot for both stage
+# plot for both stage
+
 gg_stage <-
   ggplot(xx_pca1 ,
          aes(PC1,
@@ -55,7 +54,6 @@ xx_pca2 <-
   select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
   prcomp(., center = TRUE, scale = TRUE)
 
-
 # phase 3
 
 xx_pca3 <-
@@ -68,31 +66,25 @@ xx_pca3 <-
 library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
 
 # Color variables by the continuous variable
-fviz_pca_var(xx_pca, col.var = "contrib",
-                       gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) +
-  ggtitle(paste0("All", " (n =", count_phase, ")"))
-           # + theme(legend.position = "none")
-
-ggsave(here::here("analysis/figures/007-PCA-1.png"),
-       width = 3,
-       height = 3,
-       units = "in")
-
-# phase 2
-cv_two <- fviz_pca_var(xx_pca2, col.var = "contrib",
-                       gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) +
-  ggtitle(paste0("Phase 2", " (n =", count_phase2,")"))+
+cv_all <-
+fviz_pca_var(xx_pca) +
+  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
   theme(legend.position = "none")
 
+# phase 2
+cv_two <-
[…]

### 2022-08-23 analysis/paper/paper-doc.qmd
move all code from 001 into the qmd and delete that script file

@@
 title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
-format: docx
-editor: visual
-
 author:
@@
     affiliations: University of Washington
+format: 
+  docx:
+    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
+execute:
+  echo: false
+  warning: false
+  message: false
+  comment: "#>"
+  fig-path: "../figures/"
+  fig-dpi: 600
+editor: visual
+date: now
+date-format: long
+title-block-published: "Last updated" 
+filters:
+  - ../templates/scholarly-metadata.lua
+  - ../templates/author-info-blocks.lua
+  - ../templates/pagebreak.lua
+bibliography: references.bib
+csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
 
@@
 We found that some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more dependent on raw materials or other factors.
-
-bibliography: references.bib
+keywords: |
+  Stemmed points; Korean Paleolithic; Cultural Transmission; Learning process, Transmission Biases
 ---
-
-# Keywords {.unnumbered}
-
-Stemmed points; Korean Paleolithic; Cultural Transmission; Learning process, Transmission Biases
 
@@
 ```{r}
+#| label: summary-of-data
 
+library(tidyverse)
+library(ggplot2)
+library(ggbeeswarm)
+
+# import the cleaned data ready to work on
+load(here::here("analysis/data/derived_data/data_main.RData"))
+
+# how many stemmed artefacts per site?
+stemmed_points_per_site <-
+df_full_sitename %>%
+  dplyr::count(full_sitename) %>%
+  arrange(desc(n))
+
+# prepare some values to use in the text below
+total_number_of_sp <- sum(stemmed_points_per_site$n)
+total_number_of_sites <-nrow(stemmed_points_per_site)
 ```
@@
 We excluded artifacts that were recorded as stemmed points but lack a stem.
-This resulted in a sample of 152 stemmed points from 28 assemblages unearthed from 20 sites spanning the period 40-17 ka (Figure 1).
+This resulted in a sample of `r total_number_of_sp` stemmed points from 28 assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@fig-map).
 The images of the stemmed points were obtained from published excavation reports and by direct photography during our research on local museum collections.
@@
 
-![Korean Paleolithic sites mentioned in this study](images/001-site-map.png){#fig-map width="477"}
+![Korean Paleolithic sites mentioned in this study](images/site-map.png){#fig-map width="477"}
 
@@
 
-@fig-three shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+@fig-plot-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
 Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points.
@@
 
-![Boxplot for each attribute measured on the stemmed points.
-The black bold horizontal lines in the middle of boxes represent the median value.
-The dots represent individual artefacts.](images/002-attributes-variation-01.png){#fig-three width="324"}
+```{r}
+#| label: fig-plot-attributes
+#| fig-cap: Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value. The dots represent individual artefacts.
+#| fig-width:   4.45
+#| fig-height:  4.5
+
+# Box plot for each attribute
+
+# mutate data to make plot with ggplot
+data_box <- gather(DF)
+
+# make box plot for each attributes
+ggplot(data_box,
+       aes(key, value)) +
+  geom_boxplot() +
+  geom_beeswarm(alpha = 0.3,
+                size = 0.8,
+                cex = 0.7) +
+  ylab("Length (cm)") +
+  xlab("Variables") +
+  theme_bw()
+```
 
@@
 If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias.
-Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-four).
+Our results show that CV values for all attributes are distributed from ` min_cv` to ` max_cv` (@fig-four).
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.
@@
 
+# Acknowledgements
+
+<!-- The following line inserts a page break  -->
+
+\newpage
+
 # References {.unnumbered}
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2022-08-23 analysis/paper/000-draw-the-map.R

@@ -129,7 +129,7 @@ ggdraw() +
   theme(panel.background = element_rect(fill='white', colour="white"),
         plot.background = element_rect(fill='white', colour="white"))
 
-ggsave(here::here("analysis/figures/001-site-map.png"),
+ggsave(here::here("analysis/figures/site-map.png"),
        width = 4.45,
        height = 4.5,
        units = "in")

### 2022-08-23 analysis/paper/001-basic-analysis.R

@@ -1,36 +1,0 @@
-# this code makes Fig 3: Boxplot for each attribute measured on the stemmed points.
-
-library(tidyverse)
-library(ggplot2)
-library(ggbeeswarm)
-
-# import the cleaned data ready to work on
-load(here::here("analysis/data/derived_data/data_main.RData"))
-
-# how many stemmed artefacts per site?
-stemmed_points_per_site <-
-df_full_sitename %>%
-  count(full_sitename) %>%
-  arrange(desc(n))
-
-# Box plot for each attributes
-
-# mutate data to make plot with ggplot
-data_box <- gather(DF)
-
-# make box plot for each attributes
-ggplot(data_box,
-       aes(key, value)) +
-  geom_boxplot() +
-  geom_beeswarm(alpha = 0.3,
-                size = 0.8,
-                cex = 0.7) +
-  ylab("Length (cm)") +
-  xlab("Variables") +
-  theme_bw()
-
-ggsave(here::here("analysis/figures/002-attributes-variation.png"),
-       width = 4.45,
-       height = 4.5,
-       units = "in")
-

### 2022-08-23 analysis/paper/002-coefficient-variation.R

@@ -81,7 +81,7 @@ ggplot(cv_data_sharma_named,
   geom_text(aes(label = round(all_cv_sharma,1)), col="blue", hjust = -0.3, size = 3) +
   theme_bw()
 
-ggsave(here::here("analysis/figures/003-cv-sharma.png"),
+ggsave(here::here("analysis/figures/all-artefacts-cv-sharma.png"),
        width = 4.8,
        height = 4.5,
        units = "in")
@@ -153,7 +153,7 @@ cv_plot_site_label %>%
 # sharma_int_low(Sac$SL)
 # sharma_int_high(Sac$SL)
 
-ggsave(here::here("analysis/figures/004-cv-sites.png"),
+ggsave(here::here("analysis/figures/all-sites-cv-sites.png"),
        width = 8.5,
        height = 4.5,
        units = "in")
@@ -221,7 +221,7 @@ cv_by_stage_df %>%
   facet_wrap( ~ variable, scales = "free_y") +
   theme_bw()
 
-ggsave(here::here("analysis/figures/005-cv-phases.png"),
+ggsave(here::here("analysis/figures/all-phases-cv-phases.png"),
        width = 4.5,
        height = 4.5,
        units = "in")
@@ -263,7 +263,7 @@ cv_by_site_df_label_more_than_5 %>%
   theme_bw()
   # caption should be "CVs of point attributes by site (with sites >=7 artifacts)"
 
-ggsave(here::here("analysis/figures/006-cv-four-assemblage.png"),
+ggsave(here::here("analysis/figures/cv-four-assemblage.png"),
        width = 5,
        height = 4,
        units = "in")

### 2022-08-23 analysis/paper/paper-doc.qmd
move all 002-coefficient-variation code into the qmd and delete that file

@@
 author:
-  - name: Gayoung Park
+  - Gayoung Park:
+      institute: uw
+      correspondence: "yes"
       email: gayoungp@uw.edu
-    affiliations: University of Washington
-  - name: Ben Marwick
-    email: bmarwick@uw.edu
-    affiliations: University of Washington
+  - Ben Marwick:
+      institute: uw
+institute:
+  - uw:
+      name: Department of Anthropology, University of Washington
+      address: Seattle, WA, USA
 format: 
@@
 
-1.  Distribution of attributes
+## Distribution of attributes
 
@@
 
-2.  Coefficient of variation
+## Coefficient of variation
+
+```{r}
+#| label: prepare-for-cv-calcs
+
+# Coefficient of variation (CV): the ratio of the standard deviation divided by mean
+
+cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
+
+#try corrected cv with different methods
+
+library(cvcqv)
+ccv <- function(x, ... ) cv_versatile(x, method = "kelly", ...)
+mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100 * (1 + 1/(4*length(x)))
+
+# we've got DF a data frame with cols as variables
+# rows and specimens, and values as dimensions
+# need to add a column of site ID
+
+DFCV <- map_dbl(DF, cv)
+
+#Same as map but better view for summary
+cv_all <- map_df(DF, cv)
+#arrange the data properly
+cv_data_all <- gather(cv_all)
+
+# to install MKmisc
+## Install package BiocManager
+# install.packages("BiocManager")
+## Use BiocManager to install limma
+# BiocManager::install("limma")
+# install.packages("MKmisc")
+
+library(MKmisc)
+
+# CV value
+sharma_cv <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100
+
+#CV sharma interval values
+sharma_int_low <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1]*100
+sharma_int_high <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2]*100
+
+all_cv_sharma <- map_dbl(DF, sharma_cv)
+all_low_sharma <- map_dbl(DF, sharma_int_low)
+all_high_sharma <- map_dbl(DF, sharma_int_high)
+
+#create data frame
+cv_data_sharma <- data.frame(all_cv_sharma, all_low_sharma, all_high_sharma)
+
+#add column name
+library(tibble)
+cv_data_sharma_named <- rownames_to_column(cv_data_sharma, var = "attribute")
+
+# for in-text numbers
+max_cv <- round(max(cv_data_sharma_named$all_cv_sharma), 1)
+min_cv <- round(min(cv_data_sharma_named$all_cv_sharma), 1)
+
+```
 
@@
 If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias.
-Our results show that CV values for all attributes are distributed from ` min_cv` to ` max_cv` (@fig-four).
+Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes).
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.
-Unlike the distribution of actual attributes (@fig-three), the CV of body length and maximum length have the lowest values.
-The lower CV of maximum length indicates highly standardized with narrowest confidence intervals.
+Unlike the distribution of actual attributes (@fig-plot-attributes), the CVs of body length and maximum length have the lowest values.
+The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals.
 Tang and stem related attributes (e.g.
-SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes on the artifacts.
-Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.
+SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.
 
-![The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
+```{r}
+#| label: fig-plot-cv-all-attributes
+#| fig-cap: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
 95% confidence intervals for the CVs were computed using Sharma & Krishna's method.
-The vertical lines indicate the range of confidence intervals.](images/003-cv-sharma-01.png){#fig-four width="339"}
+The vertical lines indicate the range of confidence intervals.
+#| fig-width:   4.8
+#| fig-height:  4.5
 
-3.  Principal Component Analysis
+ggplot(cv_data_sharma_named,
+       aes(attribute,
+           all_cv_sharma)) +        # ggplot2 plot with confidence intervals
+  geom_point() +
+  geom_errorbar(aes(ymin = all_low_sharma, 
+                    ymax = all_high_sharma), 
+                width = 0.2) +
+  ylab("CV") +
+  xlab("Variables") +
+  geom_text(aes(label = round(all_cv_sharma,1)), 
+            col="blue", 
+            hjust = -0.3, 
+            size = 3) +
+  theme_bw()
+```
+
+## Principal Component Analysis
 
@@
 
-After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-six).
+After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-plot-cv-over-time).
 Body length of Phase 2, maximum length, mid width, and tang width are under 25, which indicates as a result of indirect bias, while only maximum length are under 25 in Figure 4.
@@
 
-![CV values for the second and third chronological phases in the Korean Late Paleolithic period.
+```{r}
+#| label: tab-cv-over-time
+
+#-- temporal patterns ----------------------------------------------------------------
+# Make a table of CVs for all variable grouped by stage, according to
+# stemmed point chronology
+# Exclude the first phase that has only two artefacts
+
+cv_by_stage_df  <-
+  df_sitename %>%
+  dplyr::select(-sitename,
+         -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  dplyr::filter(SPstage.Stage > 1) %>%
+  dplyr::select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  group_by(SPstage.Stage, variable) %>%
+  dplyr::summarise(cv_by_site =   sharma_cv(value),
+            cv_low_sharma =  sharma_int_low(value),
+            cv_high_sharma = sharma_int_high(value))
+
+####### significant test between the two phases ######
+library(cvequality)
+
+# created nested tibble, nest by variable
+df_sitename_nested <-
+  df_sitename %>%
+  dplyr::select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  dplyr::filter(SPstage.Stage > 1) %>%
+  dplyr::select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  ungroup() %>%
+  group_by(variable) %>%
+  nest()
+
+# apply mslr test to each variable (row with a tibble)
+# Krishnamoorthy and Lee’s (2014) modified signed-likelihood ratio test
+# this is a table we can show in the text
+table_mslr_test_results_for_phase_differences <-
+  df_sitename_nested %>%
+  mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
+                                               .$value,
+                                               .$SPstage.Stage))) %>%
+  unnest(mslr_result) %>%
+  dplyr::select(-data)
+
+names(table_mslr_test_results_for_phase_differences) <- 
+  c("Variable", "MSLR statistic", "p value")
+
+knitr::kable(table_mslr_test_results_for_phase_differences,
+             caption = "CV values for the second and third chronological phases in the Korean Late Paleolithic period.
 The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.](images/004-cv-phases-01.png){#fig-six width="346"}
+The vertical lines indicate the range of confidence intervals.")
+```
+
+```{r}
+#| label: fig-plot-cv-over-time
+#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period.
+The points represent CV values for each attribute.
+The vertical lines indicate the range of confidence intervals.
+#| fig-width:   4.5
+#| fig-height:  4.5
+
+# facet plot for two phases
+cv_by_stage_df %>%
+  ggplot(aes(x = factor(SPstage.Stage),
+             y = cv_by_site)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = cv_low_sharma,
+                    ymax = cv_high_sharma),
+                width = 0.2) +
+  xlab("Phase") +
+  ylab("Coefficient of Variation on Attributes") +
+  facet_wrap( ~ variable, scales = "free_y") +
+  theme_bw()
+```
 
@@
 We excluded assemblages that contain a single point.
-@fig-nine shows that Sachang site has the highest CV values with the widest confidence level.
+@fig-plot-cv-by-site shows that Sachang site has the highest CV values with the widest confidence level.
 Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2.
@@
 
-![CV values per site that has more than one stemmed point.
-The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.](images/004-cv-sites-01.png){#fig-nine width="519"}
+```{r}
+#| label: fig-plot-cv-by-site
+#| fig-cap: CV values per site that has more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
+#| fig-width:   8.5
+#| fig-height:  4.5
+
+cv_by_site_df  <-
+  df_full_sitename %>%
+  dplyr::select(-SPstage.Stage, 
+                -lat_dd, 
+                -long_dd, 
+                -sitename, 
+                -SPstage.Raw_material) %>%
+  group_by(full_sitename) %>%
+  add_tally() %>%
+  dplyr::filter( n > 1) %>%
+  dplyr::select(-n) %>%
+  pivot_longer(-full_sitename,
+               names_to = "variable",
+               values_to = "value") %>%
+  group_by(full_sitename, variable) %>%
+  dplyr::summarise(cv_by_site =   sharma_cv(value),
+            cv_low_sharma =  sharma_int_low(value),
+            cv_high_sharma = sharma_int_high(value))
+
+## CV for per each site (n=X) with the full site name : label
+
+cv_by_full_site_df_label <-
+  df_full_sitename %>%
+  dplyr::select(-SPstage.Stage) %>%
+  group_by(full_sitename) %>%
+  add_tally() %>%
+  filter( n > 1) %>%
+  mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
+  dplyr::select(-n) %>%
+  nest(-full_sitename, -label) %>%
+  mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>%
+  dplyr::select (-cv_by_site)
+
+## add label to main dataframe
+
+cv_plot_site_label <- cv_by_site_df %>%
+  left_join (cv_by_full_site_df_label) %>%
+  dplyr::select (- data,-full_sitename)
+
+## facet plot
+
+cv_plot_site_label %>%
+  ggplot(aes(x = variable,
+             y = cv_by_site)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = cv_low_sharma,
+                    ymax = cv_high_sharma),
+                width = 0.2) +
+  xlab("Attribute") +
+  ylab("Coefficient of Variation on Attributes") +
+  facet_wrap( ~ label, scales = "free_y") +
+  theme(legend.position = "none") +
+  coord_cartesian(ylim = c(0,250)) +
+  theme_bw() +
+  theme(strip.text.x = element_text(size = 6.5))
+```
 

### 2022-08-23 analysis/paper/002-coefficient-variation.R

@@ -1,325 +1,0 @@
-# this script makes figure 4: CV values for each attribute measured on the
-# stemmed points. 95% confidence intervals for the CVs were computed using
-# Sharma & Krishna's method.
-
-# This script also makes figure 9: CV values per site that has more than one stemmed point.
-# The points represent CV values for each attribute. The vertical lines
-# indicate the range of confidence intervals.
-
-# This script also makes figure 10: CV values for raw materials that were
-# used for making more than one stemmed point. The points represent CV
-# values for each attribute. The vertical lines indicate the range of confidence interval
-
-library(tidyverse)
-
-# import the cleaned data ready to work on
-load(here::here("analysis/data/derived_data/data_main.RData"))
-
-# Coefficient of variation (CV): the ratio of the standard deviation divided by mean
-
-cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
-
-#try corrected cv with different methods
-library(cvcqv)
-ccv <- function(x, ... ) cv_versatile(x, method = "kelly", ...)
-mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100 * (1 + 1/(4*length(x)))
-
-# we've got DF a data frame with cols as variables
-# rows and specimens, and values as dimensions
-# need to add a column of site ID
-
-DFCV <- map_dbl(DF, cv)
-
-#Same as map but better view for summary
-cv_all <- map_df(DF, cv)
-#arrange the data properly
-cv_data_all <- gather(cv_all)
-
-# hist(DF$ML)
-
-# hist(DFCV)
-
-# to install MKmisc
-## Install package BiocManager
-# install.packages("BiocManager")
-## Use BiocManager to install limma
-# BiocManager::install("limma")
-# install.packages("MKmisc")
-
-library(MKmisc)
-
-# CV value
-sharma_cv <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100
-
-#CV sharma interval values
-sharma_int_low <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1]*100
[…]

### 2022-08-23 analysis/paper/003-principal-component-analysis.R
move all PCA code into the qmd and delete that file

@@ -1,92 +1,0 @@
-library(tidyverse)
-library(tidyr)
-library(corrr)
-library(ggbiplot)
-library(ggfortify)
-
-# import the cleaned data ready to work on
-load(here::here("analysis/data/derived_data/data_main.RData"))
-
-### PCA analysis using ggbiplot, ggfortify pkgs
-
-PCA <- prcomp(DF, center = TRUE, scale = TRUE)
-
-# remove stage 1 artefact
-xx_pca <-
-  df_sitename %>%
-  filter(SPstage.Stage != 1) %>%
-  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
-  prcomp(., center = TRUE, scale = TRUE)
-
-xx_pca1 <-
-  data.frame(xx_pca$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
-
-
-## CV with site (n=X)
-label_phase <-
-  df_sitename %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
-
-count_phase <- length(which(label_phase[9]>0))
-count_phase2 <- length(which(label_phase[9] == 2))
-count_phase3 <- length(which(label_phase[9] == 3))
-
-###
-
-# plot for both stage
-
-gg_stage <-
-  ggplot(xx_pca1 ,
-         aes(PC1,
-             PC2,
-             colour = stage)) +
-  geom_point() +
-  stat_ellipse()
-
-# phase 2
-
-xx_pca2 <-
-  df_sitename %>%
-  filter(SPstage.Stage == 2) %>%
-  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
-  prcomp(., center = TRUE, scale = TRUE)
-
-# phase 3
-
-xx_pca3 <-
-  df_sitename %>%
-  filter(SPstage.Stage == 3) %>%
-  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
-  prcomp(., center = TRUE, scale = TRUE)
-
-##### using factoextra pkg
-library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
-
-# Color variables by the continuous variable
-cv_all <-
[…]

### 2022-08-23 analysis/paper/paper-doc.qmd
split captions to landmarks out from schematic figure

@@
 
-![Korean Paleolithic sites mentioned in this study](images/site-map.png){#fig-map width="477"}
+```{r}
+#| label: fig-map
+#| fig-cap: Korean Paleolithic sites mentioned in this study.
+#| fig-width:   4.45
+#| fig-height:  4.5
+
+# Thisdraws the map
+
+korean_archaeological_site_locations <-
+  readr::read_csv(here::here("analysis/data/raw_data/sitelocation.csv"))
+
+korean_archaeological_site_locations <-
+  tibble::rowid_to_column(korean_archaeological_site_locations, "ID")
+
+library(ggmap)
+library(ggrepel)
+# devtools::install_github('3wen/legendMap')
+library(legendMap)
+
+# load map tiles we previously saved, we got these using the code in 
+# 000-download-map-tiles.R
+map <- readRDS(here::here("analysis/data/raw_data/map_tiles.rds"))
+
+# map with site names next to the points
+map_of_sites <-
+  ggmap(map)  +
+  geom_point(data = korean_archaeological_site_locations,
+             aes(long_dd ,
+                 lat_dd),
+             colour = "red",
+             size = 2) +
+  # label points with site name
+  geom_text_repel(data = korean_archaeological_site_locations,
+                  aes(long_dd ,
+                      lat_dd,
+                      label = full_sitename),
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.15) +
+  theme_minimal(base_size = 6) +
+  labs(x = "Longitute",
+       y = "Latitude") +
+  legendMap::scale_bar(
+    # edit these numbers to select a suitable location
+    # for the scale bar where it does not cover
+    # important details on the map
+    lon = 124.5,
+    lat = 33.5,
+    legend_size = 2,
+    # distance of one section of scale bar, in km
+    distance_lon = 50,
+    # height of the scale bar, in km
+    distance_lat = 5,
+    # distance between scale bar and units, in km
+    distance_legend = 20,
+    # units of scale bar
+    dist_unit = "km",
+    # add the north arrow
+    orientation = TRUE,
+    # length of N arrow, in km
+    arrow_length = 60,
+    # distance between scale bar & base of N arrow, in km
+    arrow_distance = 50,
+    # size of letter 'N' on N arrow, in km
+    arrow_north_size = 10) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+# inset world map, from
+# https://stackoverflow.com/a/62104396/1036500
+library(maps)
+library(magrittr)
+library(maptools)
+library(raster)
+
+#Defining a general CRS
+mycrs <- "+proj=longlat +datum=WGS84 +no_defs"
+
+#Using the original maps package, then converting map into SpatialPolygons object
+world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
+  maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))
+
+#The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
+while(rgeos::gIsValid(world)==FALSE){
+  world <- rgeos::gBuffer(world, byid = TRUE, width = 0, quadsegs = 5, capStyle = "ROUND")
+}
+
+#Dissolving polygon's limits
+world <- raster::aggregate(world)
+
+#Plotting. I add theme_void to your code to erase any axis, etc
+worldplot <-
+  ggplot() +
+  geom_polygon(data = world,
+               aes(x=long, y=lat,
+                   group=group),
+               fill='NA',
+               color='black',
+               size=0.1) +
+  annotate("rect",
+           xmin = 118,
+           xmax = 137,
+           ymin = 28,
+           ymax = 47,
+           fill = NA,
+           colour = "red",
+           size = 0.5
+  ) +
+  theme_void() +
+  coord_fixed(1) +
+  labs(x = "", y = "") +
+  labs(x = NULL, y = NULL, title = NULL) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(panel.background = element_rect(fill = "white"),
+        panel.border = element_blank(),
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
+
+# combine site map and inset map
+library(cowplot)
+ggdraw() +
+  draw_plot(map_of_sites) +
+  draw_plot(plot = worldplot,
+            x = 0.6, # x location of inset placement
+            y = -0.09, # y location of inset placement
+            width = .4, # Inset width
+            height = .35, # Inset height
+            scale = 0.5 # Inset scale
+  ) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
+```
 
@@
 In addition to attributes from previous studies of cultural transmission and projectile points, i.e.
-maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-two).
+maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
 We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts.
@@
 
-![(a) Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks:\
-ML: Maximum length, tip to bottom, perpendicular to the length axis 1- 7\
-BL: Body length, tip to the closest wing, perpendicular to the length axis 1- 3 or 1-10 depending on the artifact\
-TL: Tang length, the closest wing from the tip to bottom,perpendicular to the length axis 3 - 7 or 10-7 depending on the artifact\
-SL: (Maximum) stem length, perpendicular to the length axis. Closer tang curve's middle point from the tip to the most distant point of the basal end 9-7 or 4-7 depending on the artifact\
-MW: mid width, dimension from margin to margin at the mid-point of the length 2 - 11 axis, perpendicular to the width axis.\
-TW: tang width, dimension between each wing, perpendicular to the width axis 3 - 10\
-SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="221"}
+```{r}
+#| label: fig-labelled-schematic
+#| fig-height: 2
+#| fig-cap: Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks
+
+knitr::include_graphics(here::here("analysis/figures/hand-made-figures/ISKM-talk-2021-cut.png"))
+```
+
+```{r}
+#| label: tab-landmark-abbreviations
+
+here::here("analysis/data/raw_data/table-of-landmark-abbreviations.csv") %>% 
+  read_csv() %>% 
+  knitr::kable(., caption = "Description of artefact landmarks used in this study")
+
+```
 
@@
 
-All data treatment, analyses and visualization were performed in the R environment[@core2021team].
+All data treatment, analyses and visualization were performed in the R environment [@core2021team].
 Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well at DOI to enable full reproducibility [@marwick2017computational].
@@
 
-@fig-plot-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+@fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
 Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points.
@@
 #| label: fig-plot-attributes
-#| fig-cap: Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value. The dots represent individual artefacts.
-#| fig-width:   4.45
-#| fig-height:  4.5
 
@@
 # make box plot for each attributes
+plot_raw_measurements_for_attributes <- 
 ggplot(data_box,
@@
   ylab("Length (cm)") +
-  xlab("Variables") +
+  xlab("") +
   theme_bw()
@@
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.
-Unlike the distribution of actual attributes (@fig-plot-attributes), the CVs of body length and maximum length have the lowest values.
+Unlike the distribution of actual attributes (@fig-plot-cv-all-attributes panel A), the CVs of body length and maximum length have the lowest values.
 The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals.
@@
 
+(ref:cap-fig-plot-cv-all-attributes) A: Boxplot for each attribute measured on the stemmed points.
+The black bold horizontal lines in the middle of boxes represent the median value.
+The dots represent individual artefacts.
+B: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
+95 p confidence intervals for the CVs were computed using Sharma and Krishna's method.
+The vertical lines indicate the range of confidence intervals.
+
 ```{r}
 #| label: fig-plot-cv-all-attributes
-#| fig-cap: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
-95% confidence intervals for the CVs were computed using Sharma & Krishna's method.
-The vertical lines indicate the range of confidence intervals.
-#| fig-width:   4.8
-#| fig-height:  4.5
+#| fig-cap: '(ref:cap-fig-plot-cv-all-attributes)'
+#| fig-width:   7
+#| fig-height:  4
 
+plot_cv_all_attributes <- 
 ggplot(cv_data_sharma_named,
@@
   ylab("CV") +
-  xlab("Variables") +
+  xlab("Variable") +
   geom_text(aes(label = round(all_cv_sharma,1)), 
@@
   theme_bw()
+
+library(cowplot)
+plot_grid(plot_raw_measurements_for_attributes,
+          plot_cv_all_attributes,
+          ncol = 1, 
+          labels = "AUTO")
 ```
@@
 PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.
-@fig-five shows positive relationship between attributes.
+@fig-pca-attributes shows positive relationship between attributes.
 Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length.
@@
 
-![PCA analysis for stemmed points.
-The direction of the arrow shows the relationship between attributes.
-The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA-1.png){#fig-five width="284"}
+```{r}
+#| label: fig-pca-attributes
+#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three.
+The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated.
+The length of the arrow indicates the strength of contribution to the stemmed point shape.
+#| fig-width:   7
+#| fig-height:  3
 
-4.  Variation Over Time
+library(ggbiplot)
+library(ggfortify)
+### PCA analysis using ggbiplot, ggfortify pkgs
+
+PCA <- prcomp(DF, center = TRUE, scale = TRUE)
+
+# remove stage 1 artefact
+xx_pca <-
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage != 1) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1 <-
+  data.frame(xx_pca$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
+
+## CV with site (n=X)
+label_phase <-
+  df_sitename %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
+
+count_phase <- length(which(label_phase[9]>0))
+count_phase2 <- length(which(label_phase[9] == 2))
+count_phase3 <- length(which(label_phase[9] == 3))
+
+# phase 2
+
+xx_pca2 <-
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage == 2) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+# phase 3
+
+xx_pca3 <-
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage == 3) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+##### using factoextra pkg
+library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
+
+# Color variables by the continuous variable
+cv_all <-
+fviz_pca_var(xx_pca) +
+  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
+  theme(legend.position = "none")
+
+# phase 2
+cv_two <-
+  fviz_pca_var(xx_pca2) +
+  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))
+
+# phase 3
+cv_three <-
+  fviz_pca_var(xx_pca3) +
+  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))
+
+library(patchwork)
+cv_all + cv_two + cv_three
+```
+
+## Variation Over Time
 
@@
 Our results show that stemmed points from Phase 2 are less varied than Phase 3 as well as Phase 1.
+
+\newpage
 
@@
 
-Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-seven).
+Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-pca-attributes).
 However, there are no notable changes in correlations over time.
-
-![PCA analysis for the second and third chronological phases in the Korean Late Paleolithic period.
-The direction of the arrow shows the relationship between attributes.
-The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA.png){#fig-seven width="534"}
 
@@
 
-5.  Variation Between sites
+## Variation Between sites
 
@@
   add_tally() %>%
-  filter( n > 1) %>%
+  dplyr::filter( n > 1) %>%
   mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
@@
   nest(-full_sitename, -label) %>%
-  mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>%
-  dplyr::select (-cv_by_site)
+  dplyr::mutate(cv_by_site = purrr::map(data, ~map_df(.x, cv))) %>%
+  dplyr::select(-cv_by_site)
 
@@
 cv_plot_site_label <- cv_by_site_df %>%
-  left_join (cv_by_full_site_df_label) %>%
-  dplyr::select (- data,-full_sitename)
+  left_join(cv_by_full_site_df_label) %>%
+  dplyr::select(- data,-full_sitename)
 
@@
 
-6.  Variation Between Raw Materials
+## Variation Between Raw Materials
 
@@
 We exclude raw materials that were used for single point.
-@fig-ten shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals.
+@fig-plot-cv-by-raw-material shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals.
 Quartz has high variations of CV among the attributes.
@@
 
-![CV values for raw materials that were used for making more than one stemmed point.
+```{r}
+#| label: fig-plot-cv-by-raw-material
+#| fig-cap: CV values for raw materials that were used for making more than one stemmed point.
 The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.](images/00000-cv-raw-materials.png){#fig-ten width="526"}
+The vertical lines indicate the range of confidence intervals.
+#| fig-width:   8.5
+#| fig-height:  4.5
+
+# CV by raw materials
+### Make a table of CVs for all variable grouped by raw materials
+cv_by_raw_materials <-
+  df_full_sitename %>%
+  dplyr::select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -full_sitename) %>%
+  group_by(SPstage.Raw_material) %>%
+  add_tally() %>%
+  dplyr::filter( n > 1) %>%
+  dplyr::select(-n) %>%
+  pivot_longer(-SPstage.Raw_material,
+               names_to = "variable",
+               values_to = "value") %>%
+  group_by(SPstage.Raw_material, variable) %>%
+  dplyr::summarise(cv_by_materials =   sharma_cv(value),
+            cv_low_sharma =  sharma_int_low(value),
+            cv_high_sharma = sharma_int_high(value))
+
+######CV for per raw materials (n=X) with the full site name : label
+cv_by_raw_materials_df_label <-
+  df_sitename %>%
+  group_by(SPstage.Raw_material) %>%
+  add_tally() %>%
+  dplyr::filter( n > 1) %>%
+  mutate(label = as.factor(paste0(SPstage.Raw_material,' (N = ', n, ")"))) %>%
+  dplyr::select(-n) %>%
+  nest(-SPstage.Raw_material, -label) %>%
+  mutate(cv_by_site = purrr::map(data, ~map_df(.x, cv))) %>%
+  dplyr::select (-cv_by_site)
+
+######add label to main dataframe
+cv_plot_raw_materials_label <- cv_by_raw_materials %>%
+  left_join (cv_by_raw_materials_df_label) %>%
+  dplyr::select (- data,-SPstage.Raw_material)
+
+###### facet plot
+cv_plot_raw_materials_label %>%
+  ggplot(aes(x = variable, y = cv_by_materials)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = cv_low_sharma,
+                    ymax = cv_high_sharma), width = .2) +
+  xlab("Law Materials") +
+  ylab("Coefficient of Variation on Attributes") +
+  facet_wrap( ~ label, scales = "free_y") +
+  coord_cartesian(ylim = c(0,130)) +
+  theme_bw() 
+```
 

### 2022-08-25 analysis/paper/paper-doc.qmd

--- 2022-08-23 analysis/paper/paper-doc.qmd
@@
 
-![PCA analysis for stemmed points.
-The direction of the arrow shows the relationship between attributes.
-The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA-1.png){#fig-five width="284"}
+```{r}
+#| label: fig-pca-attributes
+#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three.
+The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated.
+The length of the arrow indicates the strength of contribution to the stemmed point shape.
+#| fig-width:   7
+#| fig-height:  3
 
-4.  Variation Over Time
+library(ggbiplot)
+library(ggfortify)
+### PCA analysis using ggbiplot, ggfortify pkgs
+
+PCA <- prcomp(DF, center = TRUE, scale = TRUE)
+
+# remove stage 1 artefact
+xx_pca <-
+  df_sitename %>%
+  filter(SPstage.Stage != 1) %>%
+  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1 <-
+  data.frame(xx_pca$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
+
+## CV with site (n=X)
+label_phase <-
+  df_sitename %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
+
+count_phase <- length(which(label_phase[9]>0))
+count_phase2 <- length(which(label_phase[9] == 2))
+count_phase3 <- length(which(label_phase[9] == 3))
+
+###
+
+# phase 2
+
+xx_pca2 <-
+  df_sitename %>%
+  filter(SPstage.Stage == 2) %>%
+  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+# phase 3
+
+xx_pca3 <-
+  df_sitename %>%
+  filter(SPstage.Stage == 3) %>%
+  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+##### using factoextra pkg
+library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
+
+# Color variables by the continuous variable
+cv_all <-
+fviz_pca_var(xx_pca) +
+  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
+  theme(legend.position = "none")
+
+# phase 2
+cv_two <-
+  fviz_pca_var(xx_pca2) +
+  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))
+
+# phase 3
+cv_three <-
+  fviz_pca_var(xx_pca3) +
+  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))
+
+library(patchwork)
+cv_all + cv_two + cv_three
+```
+
+## Variation Over Time
 
@@
 
-6.  Variation Between Raw Materials
+##  Variation Between Raw Materials
 

### 2022-08-28 analysis/paper/paper-doc.qmd
add credit authorship

@@
 
+# CRediT authorship contribution statement
+Gayoung Park: Conceptualization, Software, Validation, Formal analysis,  Methodology, Resources, Data curation,  Writing - original draft, Writing - Review & Editing, Visualization, Project administration.
+
+Ben Marwick: Software, Validation, Formal analysis, Investigation, Writing - Review & Editing, Visualization, Supervision
+
 \newpage

### 2022-08-28 analysis/paper/paper-doc.qmd
add map function chunck, then cv_by_site_df cause some issues

@@
 
-: Korean Paleolithic Chronology edited based on \@seong2015diversity; \@Park_2013
+: Korean Paleolithic Chronology edited based on Seong[-@seong2015diversity] and Park [-@Park_2013]
 
@@
 
-![Korean Paleolithic sites mentioned in this study](images/site-map.png){#fig-map width="477"}
+```{r}
+#| label: fig-map
+#| fig-cap: Korean Paleolithic sites mentioned in this study
+#| fig-width:   4.45
+#| fig-height:  4.5
+
+# This script draws the map, this is figure 1 in the paper.
+
+korean_archaeological_site_locations <-
+  readr::read_csv(here::here("analysis/data/raw_data/sitelocation.csv"))
+
+korean_archaeological_site_locations <-
+  tibble::rowid_to_column(korean_archaeological_site_locations, "ID")
+
+library(ggrepel)
+# devtools::install_github('3wen/legendMap')
+library(legendMap)
+library(sp)
+library(ggmap)
+# load map tiles we previously saved
+map <- readRDS(here::here("analysis/data/raw_data/map_tiles.rds"))
+
+# map with site names next to the points
+map_of_sites <-
+  ggmap(map)  +
+  geom_point(data = korean_archaeological_site_locations,
+             aes(long_dd ,
+                 lat_dd),
+             colour = "red",
+             size = 2) +
+# label points with site name
+  geom_text_repel(data = korean_archaeological_site_locations,
+                  aes(long_dd ,
+                      lat_dd,
+                      label = full_sitename),
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.15) +
+  theme_minimal(base_size = 6) +
+  labs(x = "Longitute",
+       y = "Latitude") +
+  legendMap::scale_bar(
+    # edit these numbers to select a suitable location
+    # for the scale bar where it does not cover
+    # important details on the map
+    lon = 124.5,
+    lat = 33.5,
+    legend_size = 2,
+    # distance of one section of scale bar, in km
+    distance_lon = 50,
+    # height of the scale bar, in km
+    distance_lat = 5,
+    # distance between scale bar and units, in km
+    distance_legend = 20,
+    # units of scale bar
+    dist_unit = "km",
+    # add the north arrow
+    orientation = TRUE,
+    # length of N arrow, in km
+    arrow_length = 60,
+    # distance between scale bar & base of N arrow, in km
+    arrow_distance = 50,
+    # size of letter 'N' on N arrow, in km
+    arrow_north_size = 10) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+# inset world map, from
+# https://stackoverflow.com/a/62104396/1036500
+library(maps)
+library(magrittr)
+library(maptools)
+library(raster)
+library(ggplot2)
+
+#Defining a general CRS
+mycrs <- "+proj=longlat +datum=WGS84 +no_defs"
+
+#Using the original maps package, then converting map into SpatialPolygons object
+world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
+  maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))
+
+#The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
+while(rgeos::gIsValid(world)==FALSE){
+  world <- rgeos::gBuffer(world, byid = TRUE, width = 0, quadsegs = 5, capStyle = "ROUND")
+}
+
+#Dissolving polygon's limits
+world <- raster::aggregate(world)
+
+#Plotting. I add theme_void to your code to erase any axis, etc
+worldplot <-
+  ggplot() +
+  geom_polygon(data = world,
+               aes(x=long, y=lat,
+                   group=group),
+               fill='NA',
+               color='black',
+               size=0.1) +
+  annotate("rect",
+           xmin = 118,
+           xmax = 137,
+           ymin = 28,
+           ymax = 47,
+           fill = NA,
+           colour = "red",
+           size = 0.5
+  ) +
+  theme_void() +
+  coord_fixed(1) +
+  labs(x = "", y = "") +
+  labs(x = NULL, y = NULL, title = NULL) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(panel.background = element_rect(fill = "white"),
+        panel.border = element_blank(),
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
+
+# combine site map and inset map
+library(cowplot)
+ggdraw() +
+  draw_plot(map_of_sites) +
+  draw_plot(plot = worldplot,
+            x = 0.6, # x location of inset placement
+            y = -0.09, # y location of inset placement
+            width = .4, # Inset width
+            height = .35, # Inset height
+            scale = 0.5 # Inset scale
+  ) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
+
+```
 
@@
 library(MKmisc)
-
+library(purrr)
 # CV value
@@
   filter(SPstage.Stage != 1) %>%
-  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
   prcomp(., center = TRUE, scale = TRUE)
@@
   filter(SPstage.Stage == 2) %>%
-  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
   prcomp(., center = TRUE, scale = TRUE)
@@
   filter(SPstage.Stage == 3) %>%
-  select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
   prcomp(., center = TRUE, scale = TRUE)

### 2022-08-28 analysis/paper/paper-doc.qmd
restore attribute image (redirection)

@@
 TW: tang width, dimension between each wing, perpendicular to the width axis 3 - 10\
-SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="221"}
+SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="212"}
 
@@
 # CRediT authorship contribution statement
+
 Gayoung Park: Conceptualization, Software, Validation, Formal analysis, Methodology, Resources, Data curation, Writing - original draft, Writing - Review & Editing, Visualization, Project administration.

### 2022-08-28 analysis/paper/paper-doc.qmd
moved discussion and conclusion text to qmd

@@
 Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals.
-Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-six results of body length and maximum length from all assemblages.
+Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-plot-cv-over-time results of body length and maximum length from all assemblages.
 It is assumed that those attributes became more standardized within the same assemblages.
@@
 
+To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission theory.
+We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+Do the modes of cultural transmission vary over space?
+and Do the modes of cultural transmission vary over time?
+We focussed on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
+We set 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques] and the relatively lower quality of lithic raw materials in Korea.
+If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.
+
+The CV values for all stemmed points are mostly over 25, except for maximum length which is 23.8.
+PCA analysis shows positive relationships between attributes.
+We also examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time).
+After excluding stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower.
+In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25.
+We observe the different degree of contribution by each attribute to the shape of stemmed points.
+For example, tang and stem related attributes have higher CV values compared to the body related attributes.
+We expect that people chose to emulate the body related attributes and had more freedom to produce the tang and stem related attributes.
+
+Compared to the  Phase 2 and 3, we found that there are only minor differences between the two phases.
+The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3.
+However the increase is not significant to generalize (p > 0.05), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
+The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (Figure 8<== not included yet).
+This could indicate the stronger connections within the groups during Phase 3.
+
+There is no clear pattern variation in CV between the sites.
+Some sites such as Sachang have very high CV values while other sites including Suyanggae and Jingeunuel have lower CV (Figure 9).
+We assumed that the CV could be highly impacted by the choice of raw materials (Figure 10).
+However, similar to the CV variation for the Phase 2, the results of site variation show that CV of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, indicating the combination of the two biases.
+The attributes with lower CV values could be taught through indirect bias while the higher CV attributes were depending more on various manufacturing situations.
+The lower CV values of the four attributes could imply that those four attributes are closely associated with the function of the tool so that the knappers wanted to standardize the shape.
+For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
+This combination model describes that there were social connections between groups that share the tool production technique.
+Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias).
+Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability.
+
+# Conclusion {.unnumbered}
+
+In this research, we examine the dominant mode of cultural transmission that might impact technological innovation in the Korean Late Paleolithic.
+Following the previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we use two transmission biases, guided variation and indirect bias to understand the transmission process and social contexts.
+We claim the two scenarios that explains the introduction of new technology: socially isolated groups made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others (indirect bias).
+We ask three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+do the modes of cultural transmission vary over space?
+and do the modes of cultural transmission vary over time?
+To determine transmission biases and answer our research questions, we computed CV and PCA.
+The higher CV values (equals to or over 25) and low correlation indicate guided variation while the lower CV while the lower CV values (under 25) and high correlations represent indirect bias.
+Due to the small sample size, we applied confidence intervals to CV for better estimation.
+
+Our results show that some attributes such as body length, maximum length, mid width, and tang width were transmitted through indirect bias and the other attributes were developed through guided variation.
+All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points.
+We assume that the guided variation played a more important role at the beginning of the transition.
+We observe a slight decrease in CV values over time.
+No clear regional patterns are confirmed.
+We found that CV values vary depending on raw materials.
+For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals.
+We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission.
+We used landmarks to earn the dimensional attributes, which has potential for future research with different dimensions by calculating the distance between other sets of the existing landmarks.
+In our future research, we will focus on understanding actual connection between the assemblages and examining the social networks for sharing technology between the groups such as trades or physical migration to different locations.
+
 # Acknowledgements

### 2022-08-28 analysis/paper/paper-doc.qmd
GP tried t-test

@@
 The vertical lines indicate the range of confidence intervals.")
+
+# apply t test to each variable
+# prepare df for t test : ML
+t_test_df_ML_2<- 
+df_sitename %>%
+  dplyr::select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  dplyr::filter(SPstage.Stage > 1)%>%
+  dplyr::select(-n) %>% 
+   pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>% 
+  filter(variable == "ML", SPstage.Stage == 2)
+
+t_test_df_ML_3<- 
+df_sitename %>%
+  dplyr::select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  dplyr::filter(SPstage.Stage > 1)%>%
+  dplyr::select(-n) %>% 
+   pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>% 
+  filter(variable == "ML", SPstage.Stage == 3)
+
+t.test(t_test_df_ML_2$value, t_test_df_ML_3$value)
+#t = 3.7637, df = 135.36, p-value = 0.0002485
+
 ```

### 2022-08-29 analysis/paper/paper-doc.qmd
resolve merge conflict and re-render

@@
 
+# load map tiles we previously saved, we got these using the code in 
+# 000-download-map-tiles.R
+
 library(ggrepel)
@@
 library(ggmap)
+
 # load map tiles we previously saved
@@
              size = 2) +
+
 # label points with site name
@@
 library(raster)
-library(ggplot2)
 
@@
         plot.background = element_rect(fill='white', colour="white"))
-
 ```
@@
 
-![(a) Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks:\
-ML: Maximum length, tip to bottom, perpendicular to the length axis 1- 7\
-BL: Body length, tip to the closest wing, perpendicular to the length axis 1- 3 or 1-10 depending on the artifact\
-TL: Tang length, the closest wing from the tip to bottom,perpendicular to the length axis 3 - 7 or 10-7 depending on the artifact\
-SL: (Maximum) stem length, perpendicular to the length axis. Closer tang curve's middle point from the tip to the most distant point of the basal end 9-7 or 4-7 depending on the artifact\
-MW: mid width, dimension from margin to margin at the mid-point of the length 2 - 11 axis, perpendicular to the width axis.\
-TW: tang width, dimension between each wing, perpendicular to the width axis 3 - 10\
-SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="212"}
+```{r}
+#| label: fig-labelled-schematic
+#| fig-height: 2
+#| fig-cap: Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks
+
+knitr::include_graphics(here::here("analysis/figures/hand-made-figures/ISKM-talk-2021-cut.png"))
+```
+
+```{r}
+#| label: tab-landmark-abbreviations
+
+here::here("analysis/data/raw_data/table-of-landmark-abbreviations.csv") %>% 
+  read_csv() %>% 
+  knitr::kable(., caption = "Description of artefact landmarks used in this study")
+
+```
 
@@
   df_sitename %>%
-  filter(SPstage.Stage != 1) %>%
+  dplyr::filter(SPstage.Stage != 1) %>%
   dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
@@
 
-###
-
 # phase 2
@@
   df_sitename %>%
-  filter(SPstage.Stage == 2) %>%
+  dplyr::filter(SPstage.Stage == 2) %>%
   dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
@@
   df_sitename %>%
-  filter(SPstage.Stage == 3) %>%
+  dplyr::filter(SPstage.Stage == 3) %>%
   dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%

### 2022-08-29 analysis/paper/gp-to-review-and-delete/notes.Rmd
add badges and binder

--- 2020-05-05 notes.Rmd
@@
 I moved the leftover codes from paper, just in case I need them later. 
+```{r}
+#box plot for CV for each attributes
+box2 <- boxplot(cvmap) ##Is there any better way to show this data??
+```
+
+```{r}
+
+cv_plot_site %>%
+  pivot_longer(cols = -sitename, 
+               names_to = "group") %>%
+  ggplot(aes(x = sitename, 
+             y = value, 
+             fill = as.factor(group))) +
+  geom_bar(stat = "identity", 
+           position = position_dodge())
+
+```
 
@@
 
-```{R}
+```{r}
+### three phases at once
+library(patchwork)
+call + ctwo + cthree + plot_layout(ncol=1)
+
+direction_all+ direction_2 + direction_3 + plot_layout(ncol=1)
+
+```
+
+```{r}
+xx_pca1_2 <- 
+  data.frame(xx_pca2$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage ==2]))
+
+ggplot(xx_pca1_2 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+xx_pca1_3 <- 
+  data.frame(xx_pca3$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage ==3]))
+
+ggplot(xx_pca1_3 ,
+       aes(PC1,
+           PC2,
+       colour = stage)) +
+  geom_point() +
+  stat_ellipse()
+
+```
+
+Contribution of variables for each of four PCA dimensions of ALL phases except for the first.
+
+```{r}
+library(gridExtra)
+library(factoextra)
+PC1_v <- fviz_contrib(xx_pca, choice = "var", axes = 1, top = 7)
+PC2_v <- fviz_contrib(xx_pca, choice = "var", axes = 2, top = 7)
+PC3_v <- fviz_contrib(xx_pca, choice = "var", axes = 3, top = 7)
+PC4_v <- fviz_contrib(xx_pca, choice = "var", axes = 4, top = 7)
+
+grid.arrange(PC1_v, PC2_v, PC3_v, PC4_v, nrow =2)
+
+```
+
+```{r}
 
@@
 
+correlate(SYG6_DF) %>% 
+  network_plot()
+
 #simpler way to calculate Correlataion, you can use
@@
          tl.col = "black", tl.srt = 45)
+
+correlate(SYG1_2DF) %>% 
+  network_plot()
+
+correlate(SYG1_1DF) %>% 
+  network_plot()
 

### 2022-08-29 README.Rmd

@@
 # CTtps
+
+  <!-- badges: start -->
+[![R-CMD-check](https://github.com/parkgayoung/CTtps/actions/workflows/render-in-docker.yaml/badge.svg)](https://github.com/parkgayoung/CTtps/actions/workflows/render-in-docker.yaml)
+[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/parkgayoung/CTtps/master?urlpath=rstudio)
+  <!-- badges: end -->
 

### 2022-08-29 analysis/paper/paper.qmd
update dockerfile

--- 2022-08-29 analysis/paper/paper-doc.qmd
@@
 #| label: fig-labelled-schematic
-#| fig-height: 2
+#| fig-height: 1
 #| fig-cap: Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks
@@
 The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3.
-However the increase is not significant to generalize (p > 0.05), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
-The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (Figure 8<== not included yet).
+However the increase is not significant to generalize (p \> 0.05), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
+The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (Figure 8\<== not included yet).
 This could indicate the stronger connections within the groups during Phase 3.

### 2022-10-10 analysis/paper/gp-to-review-and-delete/999-scraps.R
deleted unused file

@@ -1,235 +1,0 @@
-
-## CV per each site (n=X)
-
-cv_by_site_df_label <-
-  df_sitename %>%
-  select(-SPstage.Stage) %>%
-  group_by(sitename) %>%
-  add_tally() %>%
-  filter( n > 1) %>%
-  mutate(label = as.factor(paste0(sitename,' (N = ', n, ")"))) %>%   select(-n) %>%
-  nest(-sitename, -label) %>%
-  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
-
-cv_by_site_df_unnest_label <-
-  cv_by_site_df_label %>%
-  unnest(cv_by_site)
-
-cv_plot_site_label <- cv_by_site_df_unnest_label %>%
-  select(-data)
-
-## Reference: site and number of artifacts
-cv_by_site_df_label_1 <-
-  df_sitename %>%
-  select(-SPstage.Stage) %>%
-  group_by(sitename) %>%
-  add_tally() %>%
-  mutate(label = as.factor(paste0(sitename,'\nN = ',n))) %>%   select(-n) %>%
-  nest(-sitename, -label) %>%
-  mutate(cv_by_site = map(data, ~map_df(.x, cv)))
-
-cv_by_site_df_unnest_label_1 <-
-  cv_by_site_df_label_1 %>%
-  unnest(cv_by_site)
-
-
-p<- cv_plot_site_label %>%
-  pivot_longer(cols = -c(sitename,label), names_to = "group") %>%
-  ggplot(aes(label, value, fill = as.factor(group))) +
-  geom_bar(stat="identity", position=position_dodge())
-
-p + labs(x= "Site", fill = "Attributes")
-
-
-######### explore SYG6 site which contains more than 1 stage
-library(stringr)
-SYG6_df <-
-  df %>%
-  filter(str_detect(files, "SYG6"))
-
-
-SYG6_df_sitename <- SYG6_df %>%
-  mutate (files = str_remove_all(files, "_.tps")) %>%
-  separate (files, into = c("sitename", "files"), sep = "_") %>%
-  select (-files)
-
-#list of all attributes without ID
-SYG6_DF <- SYG6_df[,1:7]
-
[…]

### 2022-10-10 analysis/paper/paper.qmd
test table caption

@@
 ```{r}
-#| label: tab-landmark-abbreviations
+#| label: tbl-landmark-abbreviations
+#| tbl-cap: Description of artefact landmarks used in this study
 
 here::here("analysis/data/raw_data/table-of-landmark-abbreviations.csv") %>% 
-  read_csv() %>% 
-  knitr::kable(., caption = "Description of artefact landmarks used in this study")
+  read.csv() %>% 
+  knitr::kable(.)
 
 ```
+@tbl-landmark-abbreviations sth special~
 
@@
 ```{r}
-#| label: tab-cv-over-time
+#| label: tbl-cv-over-time
 
@@
 
-# apply t test to each variable
-# prepare df for t test : ML
-t_test_df_ML_2<- 
-df_sitename %>%
-  dplyr::select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  dplyr::filter(SPstage.Stage > 1)%>%
-  dplyr::select(-n) %>% 
-   pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>% 
-  filter(variable == "ML", SPstage.Stage == 2)
-
-t_test_df_ML_3<- 
-df_sitename %>%
-  dplyr::select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  dplyr::filter(SPstage.Stage > 1)%>%
-  dplyr::select(-n) %>% 
-   pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>% 
-  filter(variable == "ML", SPstage.Stage == 3)
-
-t.test(t_test_df_ML_2$value, t_test_df_ML_3$value)
-#t = 3.7637, df = 135.36, p-value = 0.0002485
-
 ```
+@tbl-cv-over-time says sth interesting
 

### 2022-10-10 analysis/paper/paper.qmd
add csv table

@@
 If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will be less varied.
-
-+-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
-| Phase | SP                 | Blades  | Micro-blade | SP Blank           | Newly added raw materials           | New changes                                              | Example sites | Radiocarbon ages |
-+=======+====================+=========+=============+====================+=====================================+==========================================================+===============+==================+
-| 1     | Present            | Absent  | Absent      | Flake              | Higher-quality quartzite,\          | Increasing number of small flake tools, appearance of SP | Hwadaeri\     | \~35 ka\         |
-|       |                    |         |             |                    | Vein quartz                         |                                                          | Yonghodong\   | \                |
-|       |                    |         |             |                    |                                     |                                                          | Sachang       |                  |
-|       |                    |         |             |                    |                                     |                                                          |               |                  |
-|       |                    |         |             |                    |                                     |                                                          | \             |                  |
-+-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
-| 2     | Present            | Present | Present     | Blade              | Porphyry, Siliceous shale, Hitherto | Appearance of blades, increasing number of SP            |               | 35-25 ka         |
-+-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
-| 3     | Present until 16ka | Present | Present     | Blade, Micro-blade | Obsidian                            | Appearance of Micro-blade, microliths                    |               | 25-12 ka         |
-|       |                    |         |             |                    |                                     |                                                          |               |                  |
-|       |                    |         |             |                    |                                     |                                                          |               | \                |
-+-------+--------------------+---------+-------------+--------------------+-------------------------------------+----------------------------------------------------------+---------------+------------------+
-
-: Korean Paleolithic Chronology edited based on Seong[-@seong2015diversity] and Park [-@Park_2013]
 
@@
 
+```{r}
+#| label: tbl-korean-chronology
+#| tbl-cap: Korean Paleolithic Chronology edited based on Seong[-@seong2015diversity] and Park [-@Park_2013]
+
+here::here("analysis/data/raw_data/kr-paleolithic-chronology.csv") %>% 
+  read.csv() %>% 
+  knitr::kable(.)
+
+```
+
 ## Methods
@@
 ```
-@tbl-landmark-abbreviations sth special~
+
+@tbl-landmark-abbreviations sth special\~
 
@@
 ```
+
 @tbl-cv-over-time says sth interesting

### 2022-10-14 analysis/paper/paper.qmd
Added a table for sample size, ajusted inner (world) map size for the site map

@@
 The application of evolutionary theory to archaeological research has guided the interpretation of technological transitions and related human behaviors [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
-For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual @garvey2018current].
+For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual; @garvey2018current].
 We use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
@@
 Our main question is: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-Additionally we ask: Do the modes of cultural transmission vary over space?
-Do the modes of cultural transmission vary over time?
+Additionally we ask: Do the modes of cultural transmission vary over time?
+Do the modes of cultural transmission vary over space?
 We consider three possibilities, guided variation (trial and error), or indirect bias (copying others), or a combination of the two.
@@
 
-Previous studies of technological transitions during the Late Paleolithic in Korea mainly focussed on possible origin locations of stemmed points, along with the issue of the timing and routes of the dispersal of modern human dispersals in eastern Asia.
-The debate about the origin of stemmed points can be summarized into two competing models: in situ evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
-The in situ model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
-To support his claim of an *in situ* development, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages.
+Previous studies of technological transitions during the Late Paleolithic in Korea mainly focused on possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
+The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
+The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
+To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages.
 In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one.
 He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention.
-In addition, Seong argues that increased numbers of stemmed points over time, and standardization of their shape, supports the idea of gradual, local, evolutionary process.
+In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the idea of gradual, local, evolutionary process.
 
 The migration model argues that the new blade industry including stemmed points and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
-While the in situ model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
-Specifically, the blade technology was introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP while the simple flake tool tradition came from southern China [@bae2013early].
+While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
+Specifically, the blade technology was introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
 The migration from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia including Korea in 30-25 ka BP [@shi2005chromosome].
 Bae et al.
-[-@bae2012nature] assume this southern migration was related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
+[-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
 
@@
 
-Darwinian evolutionary theory has helped archaeologists to understand technological innovation and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+Darwinian evolutionary theory has helped archaeologists understand technological innovation and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
 Cultural evolutionary approaches can explain cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
@@
 The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying".
-Garvey's modeling and simulation shows how the concept of cultural transmission can be useful to interpret the archaeological record.
+Garvey's modeling and simulation show how the concept of cultural transmission can be useful to interpret the archaeological record.
 
@@
 Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
-Mesoudi and O'Brien's results demonstrate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture variation.
+Mesoudi and O'Brien's results demonstrate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture.
 
@@
 
-Inspired by the previous studies that used Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect, to understand the spread of stemmed point technology during the Korean Late Paleolithic period.
+Inspired by the previous studies that used Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to understand the spread of stemmed point technology during the Korean Late Paleolithic period.
 We propose a spectrum for foraging groups and how they started to make stemmed points: on one end we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.
@@
 We excluded artifacts that were recorded as stemmed points but lack a stem.
-This resulted in a sample of `r total_number_of_sp` stemmed points from 28 assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@fig-map).
+This resulted in a sample of `r total_number_of_sp` stemmed points from 28 assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number).
 The images of the stemmed points were obtained from published excavation reports and by direct photography during our research on local museum collections.
@@
 Applying the chronology is useful because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research situations and limitations, chronological order help us estimate the time periods of those assemblages as well as understand the morphological change of stemmed points associated with the introduction of blade technology.
-Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages.
+Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
 
@@
   draw_plot(plot = worldplot,
-            x = 0.6, # x location of inset placement
-            y = -0.09, # y location of inset placement
-            width = .4, # Inset width
-            height = .35, # Inset height
+            x = 0.45, # x location of inset placement
+            y = -0.27, # y location of inset placement
+            width = .6, # Inset width
+            height = .75, # Inset height
             scale = 0.5 # Inset scale
@@
 ```{r}
+#| label: tbl-sp-number
+#| tbl-cap: Total number of stemmed points found in sites and the number of stemmed points we used in the research. We excluded broken and unclear artifacts.
+
+here::here("analysis/data/raw_data/site-total-artifacts.csv") %>% 
+  read.csv() %>% 
+  knitr::kable(.)
+
+```
+
+```{r}
 #| label: tbl-korean-chronology
-#| tbl-cap: Korean Paleolithic Chronology edited based on Seong[-@seong2015diversity] and Park [-@Park_2013]
+#| tbl-cap: Korean Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]
 
-here::here("analysis/data/raw_data/kr-paleolithic-chronology.csv") %>% 
+here::here("analysis/data/raw_data/Korean-paleolithic-chronology.csv") %>% 
   read.csv() %>% 

### 2022-10-15 analysis/paper/paper.qmd
Changed the order of the results and edit the text and plot

@@
 
-In addition to attributes from previous studies of cultural transmission and projectile points, i.e.
+In addition to the metric attributes from previous studies of cultural transmission and projectile points, i.e.
 maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
-We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts.
+We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
 Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
@@
 The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data (@tbl-landmark-abbreviations).
 
@@
 
-@tbl-landmark-abbreviations sth special\~
-
 To measure the variations between stemmed points, we used Coefficients of Variation (CV) with confidence intervals.
@@
 Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
-Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to show uncertainty around our CV estimates [@kelley2007sample; @albatineh2014confidence].
+Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results.
+[@kelley2007sample; @albatineh2014confidence].
 
-We chose Principal Component Analysis (PCA) to compute Pearson's r correlations between the attributes as previous studies used [@bettinger1999point; @eerkens2008cultural; @garvey2018current] and to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
+We chose Principal Component Analysis (PCA) to compute Pearson's r correlations between the attributes as previous studies did [@bettinger1999point; @eerkens2008cultural; @garvey2018current] and to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
 PCA is often used to detect relevant information from dataset, indicate relationships between variables as well as cases, extract general patterns in data, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life].
@@
 All data treatment, analyses and visualization were performed in the R environment [@core2021team].
-Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well at DOI to enable full reproducibility [@marwick2017computational].
+Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well as DOI to enable full reproducibility [@marwick2017computational].
 
@@
 
-(ref:cap-fig-plot-cv-all-attributes) A: Boxplot for each attribute measured on the stemmed points.
+```{r}
+#| label: fig-plot-cv-all-attributes
+#| fig-cap: A.
+Boxplot for each attribute measured on the stemmed points.
 The black bold horizontal lines in the middle of boxes represent the median value.
 The dots represent individual artefacts.
-B: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
-95 p confidence intervals for the CVs were computed using Sharma and Krishna's method.
+B.
+The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
+95% confidence intervals for the CVs were computed using Sharma and Krishna's method.
 The vertical lines indicate the range of confidence intervals.
 
-```{r}
-#| label: fig-plot-cv-all-attributes
-#| fig-cap: '(ref:cap-fig-plot-cv-all-attributes)'
-#| fig-width:   7
-#| fig-height:  4
+#| fig-width:7
+#| fig-height:4
 
@@
 
-## Principal Component Analysis
-
-PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.
-@fig-pca-attributes shows positive relationship between attributes.
-Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length.
-Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other.
-Other width and tang related attributes are close to each other but a bit far from body length and maximum length.
-Stem width contributes the least to the overall shape.
-We can assume that the body length and maximum length were prioritized when the stemmed points were transmitted.
-
-```{r}
-#| label: fig-pca-attributes
-#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three.
-The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated.
-The length of the arrow indicates the strength of contribution to the stemmed point shape.
-#| fig-width:   7
-#| fig-height:  3
-
-library(ggbiplot)
-library(ggfortify)
-### PCA analysis using ggbiplot, ggfortify pkgs
-
-PCA <- prcomp(DF, center = TRUE, scale = TRUE)
-
-# remove stage 1 artefact
-xx_pca <-
-  df_sitename %>%
-  dplyr::filter(SPstage.Stage != 1) %>%
-  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
-  prcomp(., center = TRUE, scale = TRUE)
-
-xx_pca1 <-
-  data.frame(xx_pca$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
-
-## CV with site (n=X)
-label_phase <-
-  df_sitename %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
-
-count_phase <- length(which(label_phase[9]>0))
-count_phase2 <- length(which(label_phase[9] == 2))
-count_phase3 <- length(which(label_phase[9] == 3))
-
-# phase 2
-
-xx_pca2 <-
-  df_sitename %>%
-  dplyr::filter(SPstage.Stage == 2) %>%
-  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
-  prcomp(., center = TRUE, scale = TRUE)
-
-# phase 3
-
-xx_pca3 <-
-  df_sitename %>%
-  dplyr::filter(SPstage.Stage == 3) %>%
-  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
-  prcomp(., center = TRUE, scale = TRUE)
-
-##### using factoextra pkg
-library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
-
-# Color variables by the continuous variable
-cv_all <-
-fviz_pca_var(xx_pca) +
-  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
-  theme(legend.position = "none")
-
-# phase 2
-cv_two <-
-  fviz_pca_var(xx_pca2) +
-  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))
-
-# phase 3
-cv_three <-
-  fviz_pca_var(xx_pca3) +
-  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))
-
-library(patchwork)
-cv_all + cv_two + cv_three
-```
-
 ## Variation Over Time
 
-To understand the temporal patterns in the modes of cultural transmission, we applied Korea Late Paleolithic chronological phases into our samples and computed CV and PCA.
+To understand the temporal patterns in the modes of cultural transmission, we applied Korea Late Paleolithic chronological phases into our samples to  compute CV (@tbl-korean-chronology).
 Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute.
@@
 ```{r}
-#| label: tbl-cv-over-time
+#| label: fig-plot-cv-over-time
+#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period.
+The points represent CV values for each attribute.
+The vertical lines indicate the range of confidence intervals.
+#| fig-width:   4.5
+#| fig-height:  4.5
 
@@
             cv_high_sharma = sharma_int_high(value))
-
-####### significant test between the two phases ######
-library(cvequality)
-
-# created nested tibble, nest by variable
-df_sitename_nested <-
-  df_sitename %>%
-  dplyr::select(-sitename, -SPstage.Raw_material) %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  dplyr::filter(SPstage.Stage > 1) %>%
-  dplyr::select(-n) %>%
-  pivot_longer(-SPstage.Stage,
-               names_to = "variable",
-               values_to = "value") %>%
-  ungroup() %>%
-  group_by(variable) %>%
-  nest()
-
-# apply mslr test to each variable (row with a tibble)
-# Krishnamoorthy and Lee’s (2014) modified signed-likelihood ratio test
-# this is a table we can show in the text
-table_mslr_test_results_for_phase_differences <-
-  df_sitename_nested %>%
-  mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
-                                               .$value,
-                                               .$SPstage.Stage))) %>%
-  unnest(mslr_result) %>%
-  dplyr::select(-data)
-
-names(table_mslr_test_results_for_phase_differences) <- 
-  c("Variable", "MSLR statistic", "p value")
-
-knitr::kable(table_mslr_test_results_for_phase_differences,
-             caption = "CV values for the second and third chronological phases in the Korean Late Paleolithic period.
-The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.")
-
-```
-
-@tbl-cv-over-time says sth interesting
-
-```{r}
-#| label: fig-plot-cv-over-time
-#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period.
-The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.
-#| fig-width:   4.5
-#| fig-height:  4.5
 
@@
 
+## Principal Component Analysis
+
+PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.
+@fig-pca-attributes shows positive relationship between attributes.
+Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length.
+Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other.
+Other width and tang related attributes are close to each other but a bit far from body length and maximum length.
+Stem width contributes the least to the overall shape.
+We can assume that the body length and maximum length were prioritized when the manufacturing technique of stemmed points were transmitted.
+
+```{r}
+#| label: fig-pca-attributes
+#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three.
+The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated.
+The length of the arrow indicates the strength of contribution to the stemmed point shape.
+#| fig-width:   7
+#| fig-height:  3
+
+library(ggbiplot)
+library(ggfortify)
+### PCA analysis using ggbiplot, ggfortify pkgs
+
+PCA <- prcomp(DF, center = TRUE, scale = TRUE)
+
+# remove stage 1 artefact
+xx_pca <-
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage != 1) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+xx_pca1 <-
+  data.frame(xx_pca$x,
+             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
+
+## CV with site (n=X)
+label_phase <-
+  df_sitename %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
+
+count_phase <- length(which(label_phase[9]>0))
+count_phase2 <- length(which(label_phase[9] == 2))
+count_phase3 <- length(which(label_phase[9] == 3))
+
+# phase 2
+
+xx_pca2 <-
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage == 2) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+# phase 3
+
+xx_pca3 <-
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage == 3) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
+  prcomp(., center = TRUE, scale = TRUE)
+
+##### using factoextra pkg
+library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
+
+# Color variables by the continuous variable
+cv_all <-
+fviz_pca_var(xx_pca) +
+  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
+  theme(legend.position = "none")
+
+# phase 2
+cv_two <-
+  fviz_pca_var(xx_pca2) +
+  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))
+
+# phase 3
+cv_three <-
+  fviz_pca_var(xx_pca3) +
+  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))
+
+library(patchwork)
+cv_all + cv_two + cv_three
+```
+
 # Discussion {.unnumbered}
@@
 The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3.
-However the increase is not significant to generalize (p \> 0.05), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
+However the increase is not significant to generalize (p \> 0.05, @tbl-cv-over-time), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
 The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (Figure 8\<== not included yet).
 This could indicate the stronger connections within the groups during Phase 3.
+
+```{r}
+#| label: tbl-cv-over-time
+
+####### significant test between the two phases ######
+library(cvequality)
+
+# created nested tibble, nest by variable
+df_sitename_nested <-
+  df_sitename %>%
+  dplyr::select(-sitename, -SPstage.Raw_material) %>%
+  group_by(SPstage.Stage) %>%
+  add_tally() %>%
+  dplyr::filter(SPstage.Stage > 1) %>%
+  dplyr::select(-n) %>%
+  pivot_longer(-SPstage.Stage,
+               names_to = "variable",
+               values_to = "value") %>%
+  ungroup() %>%
+  group_by(variable) %>%
+  nest()
+
+# apply mslr test to each variable (row with a tibble)
+# Krishnamoorthy and Lee’s (2014) modified signed-likelihood ratio test
+# this is a table we can show in the text
+table_mslr_test_results_for_phase_differences <-
+  df_sitename_nested %>%
+  mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
+                                               .$value,
+                                               .$SPstage.Stage))) %>%
+  unnest(mslr_result) %>%
+  dplyr::select(-data)
+
+names(table_mslr_test_results_for_phase_differences) <- 
+  c("Variable", "MSLR statistic", "p value")
+
+knitr::kable(table_mslr_test_results_for_phase_differences)
+
+```
 

### 2022-10-15 analysis/paper/paper.qmd
restore missing plot

@@
 
-Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-pca-attributes).
-However, there are no notable changes in correlations over time.
-
-We explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people.
-There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain seven or more stemmed points.
+We then explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people.
+There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain more than 5 stemmed points.
 SYG1_2 and SYG6_2 are from Phase 2 and SYG1_3 and YS_3 are from Phase 3.
 Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width.
-@fig-eight shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias.
+@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias.
 Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals.
@@
 
-![CV values of four attributes from assemblages that have 7 or more stemmed points.
-Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.](images/006-cv-four-assemblage.png){#fig-eight width="386"}
+```{r}
+#| label: fig-plot-cv-four-assemblage
+#| fig-cap: CV values of four attributes from assemblages that more than 5 stemmed points.
+Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
+#| fig-width:   4.5
+#| fig-height:  4.5
+
+#### plot CVs by site with number of artefacts showing >5
+
+cv_by_site_df_label_more_than_5 <-
+  df_sitename %>%
+  unite(col = "site_phase",
+        c(sitename,
+          SPstage.Stage),
+        sep = "_") %>%
+  dplyr::select(-SPstage.Raw_material) %>%
+  group_by(site_phase) %>%
+  add_tally() %>%
+  filter( n > 5) %>%
+  mutate(label = as.factor(paste0(site_phase,' (N = ', n, ")"))) %>%
+  ungroup() %>%
+  dplyr::select(-n, -site_phase) %>%
+  pivot_longer(-label,
+               names_to = "variable",
+               values_to = "value") %>%
+  group_by(label, variable) %>%
+  summarise(cv_by_site =   sharma_cv(value),
+            cv_low_sharma =  sharma_int_low(value),
+            cv_high_sharma = sharma_int_high(value)) %>%
+  filter(variable %in% c("ML", "BL","MW", "TW"))
+
+cv_by_site_df_label_more_than_5 %>%
+  ggplot(aes(x = factor(variable),
+             y = cv_by_site)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = cv_low_sharma,
+                    ymax = cv_high_sharma), width = 0.2) +
+  xlab("Attribute") +
+  ylab("Coefficient of Variation on Attributes") +
+  facet_wrap( ~ label, scales = "free_y") +
+  coord_cartesian(ylim = c(15,60)) +
+  theme_bw()
+
+```
 
@@
 We can assume that the body length and maximum length were prioritized when the manufacturing technique of stemmed points were transmitted.
+
+Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective.
+However, there are no notable changes in correlations over time.
 

### 2022-10-15 analysis/paper/paper.qmd
Finish editing

@@
 All data treatment, analyses and visualization were performed in the R environment [@core2021team].
-Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well as DOI to enable full reproducibility [@marwick2017computational].
+Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well as https://osf.io/eb8mx/ to enable full reproducibility [@marwick2017computational].
 
@@
 Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2.
-We excluded sites that have only one stemmed point for the analysis.
 Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points.
@@
 We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials.
-We exclude raw materials that were used for single point.
+We excluded raw materials that were used for single point.
 @fig-plot-cv-by-raw-material shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals.
@@
 
+For the temporal variation, we computed PCA for difference periods.
 Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective.
@@
 
-To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission theory.
+To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission.
 We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-Do the modes of cultural transmission vary over space?
-and Do the modes of cultural transmission vary over time?
-We focussed on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
+Do the modes of cultural transmission vary over time?
+and Do the modes of cultural transmission vary over space?
+We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
 We set 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques] and the relatively lower quality of lithic raw materials in Korea.
@@
 PCA analysis shows positive relationships between attributes.
-We also examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time).
-After excluding stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower.
+We then examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time).
+After excluding two stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower.
 In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25.
@@
 The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3.
-However the increase is not significant to generalize (p \> 0.05, @tbl-cv-over-time), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
-The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (Figure 8\<== not included yet).
+The increase is not significant to generalize (p \> 0.05, @tbl-cv-over-time), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
+The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (@fig-plot-cv-four-assemblage).
 This could indicate the stronger connections within the groups during Phase 3.
@@
 There is no clear pattern variation in CV between the sites.
-Some sites such as Sachang have very high CV values while other sites including Suyanggae and Jingeunuel have lower CV (Figure 9).
-We assumed that the CV could be highly impacted by the choice of raw materials (Figure 10).
+Some sites such as Sachang have very high CV values while other sites including Suyanggae and Jingeunuel have lower CV (@fig-plot-cv-by-site).
+We assumed that the CV could be impacted by other factors such as the choice of raw materials (@fig-plot-cv-by-raw-material).
 However, similar to the CV variation for the Phase 2, the results of site variation show that CV of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, indicating the combination of the two biases.
-The attributes with lower CV values could be taught through indirect bias while the higher CV attributes were depending more on various manufacturing situations.
-The lower CV values of the four attributes could imply that those four attributes are closely associated with the function of the tool so that the knappers wanted to standardize the shape.
+The attributes with lower CV values could be taught through indirect bias while the higher CV attributes could be depending more on various manufacturing situations.
+We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the function of the tool so that the knappers wanted to standardize the shape.
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
@@
 Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability.
+Our results recalls -@lee2013current's model but his model explains combination of the *in situ* tradition for flake tools and introduction of blade-associate toolkit including stemmed points.
+On the other hand, we saw the combination of the *in situ* tradition and introduced stemmed point technology within the single artifact.
 
@@
 In this research, we examine the dominant mode of cultural transmission that might impact technological innovation in the Korean Late Paleolithic.
-Following the previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we use two transmission biases, guided variation and indirect bias to understand the transmission process and social contexts.
+Following the previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we use two transmission biases, guided variation and indirect bias to scrutinize the transmission process and social contexts.
 We claim the two scenarios that explains the introduction of new technology: socially isolated groups made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others (indirect bias).
 We ask three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-do the modes of cultural transmission vary over space?
-and do the modes of cultural transmission vary over time?
+do the modes of cultural transmission vary over time?
+and do the modes of cultural transmission vary over space?
 To determine transmission biases and answer our research questions, we computed CV and PCA.
-The higher CV values (equals to or over 25) and low correlation indicate guided variation while the lower CV while the lower CV values (under 25) and high correlations represent indirect bias.
-Due to the small sample size, we applied confidence intervals to CV for better estimation.
+The higher CV values (i.e.
+>=25) and low correlation indicate guided variation while the lower CV while the lower CV values (i.e.
+<25) and high correlations represent indirect bias.
+To offset the small sample issue and increase the credibility of the comparisons, we applied confidence intervals to CV for better estimation.
 
@@
 All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points.
-We assume that the guided variation played a more important role at the beginning of the transition.
+We assume that the guided variation played a more important role at the beginning of the transition based on high CV values of Phase 1.
 We observe a slight decrease in CV values over time.
 No clear regional patterns are confirmed.
-We found that CV values vary depending on raw materials.
+CV values could be vary depending on raw materials.
 For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals.
+Overall, we conclude that combination of the two transmission biases well explains the technological transition during the Late Paleolithic.
 We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission.

### 2022-10-16 analysis/paper/paper.qmd
add doi

@@
 All data treatment, analyses and visualization were performed in the R environment [@core2021team].
-Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well as https://osf.io/eb8mx/ to enable full reproducibility [@marwick2017computational].
+Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well as DOI: 10.17605/OSF.IO/EB8MX to enable full reproducibility [@marwick2017computational].
 

### 2022-10-16 analysis/paper/paper.qmd
update doi address

--- 2022-10-15 analysis/paper/paper.qmd
@@
 All data treatment, analyses and visualization were performed in the R environment [@core2021team].
-Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well as https://osf.io/eb8mx/ to enable full reproducibility [@marwick2017computational].
+Our R code and data including the raw images are openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].
 

### 2022-10-21 analysis/paper/paper.qmd
editing the abstract and intro

@@
 abstract: |
-  Archaeologists have applied evolutionary concepts linking material evidence and cultural phenomena to understand human behavior.
-Evolutionary approaches suggest that technological transitions can occur through changes in social learning processes, and transmission biases are important loci of changes.
-The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, is a key event in understanding modern human dispersal in Northeast Asia.
-Previous studies mainly focus on possible origin locations of new technologies, but they rarely address the process of the technological transition.
-In this research we use a cultural transmission framework to investigate the social contexts that can give us insights into the emergence of these new technologies.
+  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia.
+Previous studies have mainly focussed on possible origin locations of these new technologies.
+The specific cultural processes of the appearance stemmed points and blades has rarely been considered.
+We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
 Our main question is: what was the dominant mode of cultural transmission during the time of technological innovation in the Korean Late Paleolithic?
-Inspired by Bettinger and Eerkens [-@bettinger1999point], we build two models using guided variation and indirect bias.
-To test the models and understand the transmission processes, we use coefficients of variation (CV), and Principal Component Analysis (PCA).
-Here we show that the information about the new technology was transmitted via selective combinations of guided variation and indirect bias.
+Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and  indirect bias.
+To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA).
+We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias.
 We found that some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more dependent on raw materials or other factors.
+I'M A BIT CONFUSED ABOUT THIS LAST SENTENCE, SEEMS LIKE IT SHOULD REFER TO SOCIAL CONTEXTS OF TRANSMISSION, BUT IT TALKS ABOUT RAW MATERIALS?
+CAN YOU LINK THE TWO?
+
 keywords: |
@@
 
-The application of evolutionary theory to archaeological research has guided the interpretation of technological transitions and related human behaviors [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
+The application of evolutionary theory to archaeological research has been productive for the investigation of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
 For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual; @garvey2018current].
-We use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
-This technological innovation is not just the appearance of new tools such as stemmed points and blades, but may also represent a key event related to modern human dispersal in East Asia, if the appearance of these new technologies represents the first arrival of modern humans in this region [@seong2009emergence].
-Previous studies mainly focused on the possible origin locations of stemmed points, to connect Korea with global patterns of modern human dispersal.
-Questions about the processes of technological change remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
+In this paper we used cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
+This technological innovation is not just the appearance of new tools such as stemmed points and blades, but if the appearance of these new technologies represents the first arrival of modern humans in this region, this transition may also represent a key event related to modern human dispersal throughout East Asia [@seong2009emergence].
+Previous studies on the origin of stemmed points have mainly focused on their possible origin locations, connecting Korea with global patterns of modern human dispersal.
+However, questions about the specific processes of this technological change in Korea remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
 In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic.
-Our main question is: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-Additionally we ask: Do the modes of cultural transmission vary over time?
-Do the modes of cultural transmission vary over space?
-We consider three possibilities, guided variation (trial and error), or indirect bias (copying others), or a combination of the two.
+Our main question is: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+Additionally we ask: Do the modes of cultural transmission vary over time and space?
+We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two.
 Our results can help to determine if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.
@@
 The higher CV values (i.e.
->=25) and low correlation indicate guided variation while the lower CV while the lower CV values (i.e.
-<25) and high correlations represent indirect bias.
+\>=25) and low correlation indicate guided variation while the lower CV while the lower CV values (i.e.
+\<25) and high correlations represent indirect bias.
 To offset the small sample issue and increase the credibility of the comparisons, we applied confidence intervals to CV for better estimation.

### 2022-10-21 analysis/paper/paper.qmd
add caching, and ignore cache in git (some files are a bit large)

@@
   fig-dpi: 600
+  cache: true
 editor: visual
@@
 The stemmed point is a projectile point made out of an elongated flake or blade with slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
-The stemmed point is the first composite tool type and represents new hunting strategies in the Korean Peninsula, as well as other adjacent regions [@seong2008tanged; @lee2019were].
-Stemmed points, combined with blade technology and multi-stage production sequences, and evidence of being resharpened and reused, are important aspects of Late Paleolithic technological innovation in Korea because these features are rarely seen in earlier periods [@chang2013human; @bamforth2009projectile; @seong2015diversity].
+The stemmed point was the first composite tool type appearing on the Korean Peninsula, and represents new hunting strategies there, as well as other adjacent regions where it appeared later [@seong2008tanged; @lee2019were].
+Stemmed points, combined with blade technology, multi-stage production sequences, and evidence of being resharpened and reused, are important aspects of the Late Paleolithic technological innovations in Korea because these features are rarely seen in earlier periods [@chang2013human; @bamforth2009projectile; @seong2015diversity].
 Currently the earliest stemmed points in Northeast Asia are from Yongho-doing site in Korea dated to 38.5ka and made on elongate flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
@@
 
-Previous studies of technological transitions during the Late Paleolithic in Korea mainly focused on possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
+Previous studies of technological transitions during the Late Paleolithic in Korea mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
 The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].

### 2022-10-21 analysis/paper/paper.qmd
editing evo theory section

@@
 
-Previous studies of technological transitions during the Late Paleolithic in Korea mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
+Previous studies of technological transitions during the Late Paleolithic in Korea have mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
 The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
 The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
-To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages.
+To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea.
 In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one.
 He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention.
-In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the idea of gradual, local, evolutionary process.
-
-The migration model argues that the new blade industry including stemmed points and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
+In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies .
+
+The migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
 While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
-Specifically, the blade technology was introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
-The migration from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia including Korea in 30-25 ka BP [@shi2005chromosome].
+Specifically, the blade technology in claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
+The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome].
 Bae et al.
@@
 
-Darwinian evolutionary theory has helped archaeologists understand technological innovation and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
-Cultural evolutionary approaches can explain cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
-The mechanisms are similar to biological evolution, but unique to cultural systems.
-We focus here on how different social contexts result in different modes of social transmission of tool-making skills.
-Social transmission can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
+To measure transmission biases in tool making, we draw on Darwinian evolutionary theory.
+This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
+The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties.
+Social transmission strongly affects these key evolutionary processes [@Whiten_2017], and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
+We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gather populations.
 
@@
 trial and error), or from each other by sharing information.
-Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire through recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual].
+Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire.
+Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual].
 Transmission biases can be important loci of changes in material culture, and can be influenced by social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social].
-These biases include guided variation (where individuals learn new behaviors and then modify them through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular instructions based on their perceived frequency in the population, such as extremely popular or rare instructions), and indirect bias (where a variant is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].
-
-Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how these biases can be identified in stone artifact assemblages.
+These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].
+
+Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two these biases can be identified in stone artifact assemblages.
 Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
-They equated guided transmission with high metric variation and low correlation between metric attributes.
-Conversely, they inferred indirect bias from less variation and more correlated variables.
+They equated guided transmission (where individuals learn new behaviors through trial and error) with high metric variation and low correlation between metric attributes.
+Conversely, they inferred indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) from less variation and more correlated variables.
 They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools.
-From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly, with new technologies developed largely by trial and error.
+From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly.
+In this context of limited contact, they argue that the new technologies developed largely by trial and error.
 On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.
 
-Garvey [-@garvey2018current] also explores Bettinger and Eerkens' framework and uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points.
-With the two types of projectile points, Washita and Fresno points, from Henderson site, she measured CV values of weight, thickness, width and lengths.
-She then computed simulated CV values according to various learning scenarios of knapper skill, raw material quality and human perception (i.e.
+Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
+Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
+She then computed simulated CV values according to various three scenarios of three levels of transmission fidelity (i.e.
 CV = 10%, 5%, and 3%).
 The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying".
-Garvey's modeling and simulation show how the concept of cultural transmission can be useful to interpret the archaeological record.
-
-Although Garvey and Bettinger and Eerkens were unable to directly observe transmission processes, Mesoudi and O'Brien [-@mesoudi2008cultural] conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
+A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being to subtle for people to notice [@eerkens2000practice].
+Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.
+
+Although Garvey and Bettinger and Eerkens were unable to directly observe transmission processes because of their focus on archaeological data, Mesoudi and O'Brien [-@mesoudi2008cultural] conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
 They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants.
@@
 Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
-Mesoudi and O'Brien's results demonstrate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture.
+Mesoudi and O'Brien's experimental results validate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture.
 

### 2022-10-22 analysis/paper/paper.qmd
editing 'Modeling the social context' section

@@
 
-To measure transmission biases in tool making, we draw on Darwinian evolutionary theory.
+To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory.
 This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
@@
 
-Although Garvey and Bettinger and Eerkens were unable to directly observe transmission processes because of their focus on archaeological data, Mesoudi and O'Brien [-@mesoudi2008cultural] conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
+Due to their focus on archaeological data, Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
+Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
 They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants.
@@
 Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
-Mesoudi and O'Brien's experimental results validate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture.
-
-# Modeling the social context of the appearance of stemmed points {.unnumbered}
-
-Inspired by the previous studies that used Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to understand the spread of stemmed point technology during the Korean Late Paleolithic period.
-We propose a spectrum for foraging groups and how they started to make stemmed points: on one end we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.
+Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.
+
+# Modeling the social context of the appearance of stemmed points in the Late Palaeolithic of Korea {.unnumbered}
+
+Inspired by Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period.
+We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness.
+On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.
 
@@
 Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement.
-This trial and error behavior leaves a distinctive signature on metric variables of the stemmed points.
-If trial and error was the dominant bias in the appearance of stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation between sites.
-
-On the other end of our spectrum, socially connected groups occupied places close to other groups or had regular contact with other groups over long or short distances.
+This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points.
+If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.
+
+On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups.
 Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
-The model stemmed point might be a highly successful and frequently chosen one, and the learners copy all information about the point design as a package.
+Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information.
+In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package.
 In this social context, indirect bias is the dominant influence on transmission of lithic technology.
-If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will be less varied.
+If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.
 

### 2022-10-22 analysis/paper/paper.qmd
editing methods: materials

@@
 On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups.
+Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information.
 Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
-Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information.
 In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package.
@@
 total_number_of_sites <-nrow(stemmed_points_per_site)
+total_number_of_assemblages <- NULL # ADD CODE HERE
 ```
@@
 
-After the first discovery in Seokjangri site in the 1960s, around 300 stemmed points have been found in nearly 30 sites across Korea (@fig-map).
+After the first discovery of stemmed points at the Seokjangri site in the 1960s, around 300 have been found in nearly 30 sites across Korea (@fig-map).
 Most sites contain only a few points but a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
@@
 We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were built on the presence of blades, stemmed point blanks, toolkit composition and radiocarbon ages.
-We arranged our 28 assemblages containing stemmed points in chronological order.
-For example, Sachang site has two assemblages that are fitted into phase 1 and 2.
-Applying the chronology is useful because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research situations and limitations, chronological order help us estimate the time periods of those assemblages as well as understand the morphological change of stemmed points associated with the introduction of blade technology.
+These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition .
 Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
+Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations.
+Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
+We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to facilitate observations of change over time.
 

### 2022-10-22 analysis/paper/paper.qmd
editing methods

@@
 In addition to the metric attributes from previous studies of cultural transmission and projectile points, i.e.
-maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
+maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we also used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
@@
 For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses.
-The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data (@tbl-landmark-abbreviations).
+The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artefacts, and export them as XY coordinate data for further analysis.
 
@@
 
-To measure the variations between stemmed points, we used Coefficients of Variation (CV) with confidence intervals.
-CV is the ratio of the standard deviation to the mean.
+To measure the variation among the measurements of the stemmed points, we used Coefficients of Variation (CV) with confidence intervals.
+The CV is the ratio of the standard deviation to the mean.
 It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc.
 in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence].
-In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes including distinguishing different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2% (i.e.
-CV for planned production = 1.7 %) and about 60% (i.e.
-CV for random production = 57.7%) and the values over the upper limit likely reflect intended attempts to differentiate the products.
+In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
+Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e.
+CV for planned production = 1.7) and about 60 (i.e.
+CV for random production = 57.7) and the values over the upper limit likely reflect intended attempts to differentiate the products.
 Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
-Considering the lower quality of raw materials in Korea, we sat roughly 25 as the threshold to decide the type of transmission biases.
+Considering the lower quality of raw materials in Korea, we selected roughly 25 as the threshold to decide between guided variation (\<25) and indirect bias (\>25).
+I THINK WE NEED TO EXPLAIN THIS DECISION OF THE THRESHOLD VALUE IN MORE DETAIL.
+I EXPECT REVIEWERS WILL HAVE QUESTIONS ABOUT WHY 25 IS THE NUMBER YOU CHOSE.
 
@@
 Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
-Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results.
+Since our sample size is `r total_number_of_sp` artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results.
 [@kelley2007sample; @albatineh2014confidence].
-
-We chose Principal Component Analysis (PCA) to compute Pearson's r correlations between the attributes as previous studies did [@bettinger1999point; @eerkens2008cultural; @garvey2018current] and to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
-PCA is often used to detect relevant information from dataset, indicate relationships between variables as well as cases, extract general patterns in data, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life].
+I THINK THIS 'standard CV formula' IS A BAD CHOICE AND CONFUSING TO REVIEWERS.
+THIS IS BECAUSE IN SOME OF OUR PLOTS BELOW THERE ARE AS FEW AS 2 ARTEFACTS, THAT IS A VERY SMALL SAMPLE SIZE.
+GIVEN WHAT YOU HAVE WRITTEN ABOVE, WE SHOULD USE THE 'CORRECTED CV' BECAUSE MANY OF OUR COMPARISONS ARE WITH N\<25, FOR EXAMPLE, MOST SITES HAVE \<25 SP, SO WE *DO* HAVE SMALL SAMPLES AND WE NEED TO DEAL WITH IT USING THE BEST METHODS AVAILABLE.
+
+To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we chose Principal Component Analysis (PCA).
+PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
+PCA is often used to reduce dimensionality, detect patterns and clusters in data, indicate relationships between variables as well as cases, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life].
 PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].
@@
 All data treatment, analyses and visualization were performed in the R environment [@core2021team].
-Our R code and data including the raw images are openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].
+Our R code and data, including the original artefact images, are fully and openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].
 

### 2022-10-31 analysis/paper/paper.qmd
replace cv() with mccv() and cvCI() with our custome sharma_cvCI() that uses mccv() instead of CV()

@@
 
-cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
-
-#try corrected cv with different methods
-
-library(cvcqv)
-ccv <- function(x, ... ) cv_versatile(x, method = "kelly", ...)
-mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100 * (1 + 1/(4*length(x)))
+# cv <- function(x, ... ) sd(x, ...) / mean(x, ...) * 100
+
+# use this one that is optimized for small samples
+
+mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) * 100 * (1 + 1/(4*length(x)))
+
+# BM: this is the cvCI function from https://github.com/stamats/MKmisc, I 
+# removed all the non-sharma methods and replaced CV() with mccv() above
+
+sharma_cvCI <- function(x, 
+                 conf.level = 0.95, 
+                 method = "sharma", 
+                 na.rm = FALSE){
+    stopifnot(is.numeric(x))
+
+    if(length(conf.level) != 1)
+        stop("'conf.level' has to be of length 1 (confidence level)")
+    if(conf.level < 0.5 | conf.level > 1)
+        stop("'conf.level' has to be in [0.5, 1]")
+
+    alpha <- 1 - conf.level
+    zquant <- qnorm(1-alpha/2)
+    cv <- mccv(x = x, na.rm = na.rm)
+    Infos <- NULL
+    if(na.rm) x <- x[!is.na(x)]
+    n <- length(x)
+
+     # Sharma and Krishna 1994
+      CI.lower <- 1/(1/cv + zquant/sqrt(n))
+      CI.upper <- 1/(1/cv - zquant/sqrt(n))
+      METH <- "Sharma and Krishna (1994) confidence interval"
+
+    CI <- matrix(c(CI.lower, CI.upper), nrow = 1)
+    rownames(CI) <- "CV"
+    colnames(CI) <- c(paste(alpha/2*100, "%"), paste((1-alpha/2)*100, "%"))
+    attr(CI, "conf.level") <- conf.level
+    names(cv) <- "CV"
+
+    return(structure(list("estimate" = cv, "conf.int" = CI, "Infos" = Infos,
+                          "method" = METH),
+                     class = "confint"))
+}
+```
+
+```{r}
 
@@
 
-DFCV <- map_dbl(DF, cv)
+DFCV <- map_dbl(DF, mccv)
 
 #Same as map but better view for summary
-cv_all <- map_df(DF, cv)
+cv_all <- map_df(DF, mccv)
 #arrange the data properly
@@
 # CV value
-sharma_cv <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100
+sharma_cv <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100
 
 #CV sharma interval values
-sharma_int_low <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1]*100
-sharma_int_high <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2]*100
+sharma_int_low <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1] * 100
+
+sharma_int_high <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2] * 100
 

### 2022-10-31 analysis/paper/paper.qmd
update a few more places where cv is calculated, change output of fn from % to 0-1 value

@@
 
-# use this one that is optimized for small samples
-
-mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) * 100 * (1 + 1/(4*length(x)))
+# use this one that is optimized for small samples, returns value 0-1, not %
+
+mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) * (1 + 1/(4*length(x)))
 
@@
   nest(-full_sitename, -label) %>%
-  dplyr::mutate(cv_by_site = purrr::map(data, ~map_df(.x, cv))) %>%
+  dplyr::mutate(cv_by_site = purrr::map(data, ~map_df(.x, mccv))) %>%
   dplyr::select(-cv_by_site)
@@
   nest(-SPstage.Raw_material, -label) %>%
-  mutate(cv_by_site = purrr::map(data, ~map_df(.x, cv))) %>%
+  mutate(cv_by_site = purrr::map(data, ~map_df(.x, mccv))) %>%
   dplyr::select (-cv_by_site)

### 2022-10-31 analysis/paper/paper.qmd
make the artefact schematic smaller, and the sig tests table show only 3 digits

@@
 here::here("analysis/data/raw_data/site-total-artifacts.csv") %>% 
-  read.csv() %>% 
+  read_csv() %>% 
   knitr::kable(.)
@@
 here::here("analysis/data/raw_data/Korean-paleolithic-chronology.csv") %>% 
-  read.csv() %>% 
+  read_csv() %>% 
   knitr::kable(.)
@@
 #| label: fig-labelled-schematic
-#| fig-height: 1
 #| fig-cap: Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks
+#| out-width: "300px"
 
 knitr::include_graphics(here::here("analysis/figures/hand-made-figures/ISKM-talk-2021-cut.png"))
+
 ```
@@
 here::here("analysis/data/raw_data/table-of-landmark-abbreviations.csv") %>% 
-  read.csv() %>% 
+  read_csv() %>% 
   knitr::kable(.)
@@
 #| label: tbl-cv-over-time
+#| tbl-cap: "Summary of significance tests for CV values between the two chronological phases"
 
@@
   unnest(mslr_result) %>%
-  dplyr::select(-data)
+  dplyr::select(-data) %>% 
+  # round numeric columns to 3 decimal places
+  mutate(across(where(is.numeric), round, 3)) %>% 
+  arrange(desc(p_value))
 
@@
 
-knitr::kable(table_mslr_test_results_for_phase_differences)
+knitr::kable(table_mslr_test_results_for_phase_differences,
+             digits = 3)
 

### 2022-11-09 analysis/paper/paper.qmd
Dealt with Ben's comments

@@
 We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias.
-We found that some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more dependent on raw materials or other factors.
-I'M A BIT CONFUSED ABOUT THIS LAST SENTENCE, SEEMS LIKE IT SHOULD REFER TO SOCIAL CONTEXTS OF TRANSMISSION, BUT IT TALKS ABOUT RAW MATERIALS?
-CAN YOU LINK THE TWO?
+Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more associated to guided variation.
+Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group and then modified the shape according to their specific circumstances.
 
@@
 
-# Modeling the social context of the appearance of stemmed points in the Late Palaeolithic of Korea {.unnumbered}
+# Modeling the social context of the appearance of stemmed points in the Late Paleolithic of Korea {.unnumbered}
 
@@
 We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition .
+These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
 Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
@@
 Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
-Considering the lower quality of raw materials in Korea, we selected roughly 25 as the threshold to decide between guided variation (\<25) and indirect bias (\>25).
-I THINK WE NEED TO EXPLAIN THIS DECISION OF THE THRESHOLD VALUE IN MORE DETAIL.
-I EXPECT REVIEWERS WILL HAVE QUESTIONS ABOUT WHY 25 IS THE NUMBER YOU CHOSE.
+Considering the lower quality of raw materials in Korea and the median of the CV range (30.5), we selected roughly 25 as the threshold to decide between guided variation and indirect bias.
+We premised that CV values that are lower or similar to 25 represent guided variation while CV values over 25 reflect indirect bias.
 
@@
 Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
-Since our sample size is `r total_number_of_sp` artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results.
+Although we have a total of `r total_number_of_sp` artifacts, we used the corrected CV formula since we computed CV per assemblages to examine the temporal and regional pattern.
+We also computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results.
 [@kelley2007sample; @albatineh2014confidence].
-I THINK THIS 'standard CV formula' IS A BAD CHOICE AND CONFUSING TO REVIEWERS.
-THIS IS BECAUSE IN SOME OF OUR PLOTS BELOW THERE ARE AS FEW AS 2 ARTEFACTS, THAT IS A VERY SMALL SAMPLE SIZE.
-GIVEN WHAT YOU HAVE WRITTEN ABOVE, WE SHOULD USE THE 'CORRECTED CV' BECAUSE MANY OF OUR COMPARISONS ARE WITH N\<25, FOR EXAMPLE, MOST SITES HAVE \<25 SP, SO WE *DO* HAVE SMALL SAMPLES AND WE NEED TO DEAL WITH IT USING THE BEST METHODS AVAILABLE.
 
@@
 
-## Distribution of attributes
-
-@fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+## Coefficient of variation
+
+@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
 Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points.
 This result implies a similar thickness of a wooden shaft that is attached to the stem.
-
-```{r}
-#| label: fig-plot-attributes
-
-# Box plot for each attribute
-
-# mutate data to make plot with ggplot
-data_box <- gather(DF)
-
-# make box plot for each attributes
-plot_raw_measurements_for_attributes <- 
-ggplot(data_box,
-       aes(key, value)) +
-  geom_boxplot() +
-  geom_beeswarm(alpha = 0.3,
-                size = 0.8,
-                cex = 0.7) +
-  ylab("Length (cm)") +
-  xlab("") +
-  theme_bw()
-```
-
-## Coefficient of variation
 
@@
 
-We used CV to measure the variations between stemmed points.
+We used the corrected CV to measure the variations between stemmed points.
 If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias.
-Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes).
+Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes panel B).
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.
-Unlike the distribution of actual attributes (@fig-plot-cv-all-attributes panel A), the CVs of body length and maximum length have the lowest values.
+Unlike the distribution of actual attributes, the CVs of body length and maximum length have the lowest values.
 The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals.
@@
 We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
-We set 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques] and the relatively lower quality of lithic raw materials in Korea.
+We set roughly 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
 If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.
 
-The CV values for all stemmed points are mostly over 25, except for maximum length which is 23.8.
-PCA analysis shows positive relationships between attributes.
+The CV values for all stemmed points are mostly over 25, except for maximum length.
+PCA analysis shows positive relationships between attributes (@fig-plot-cv-all-attributes panel B).
 We then examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time).
@@
 No clear regional patterns are confirmed.
-CV values could be vary depending on raw materials.
+CV values could be depending on raw materials.
 For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals.

### 2022-11-28 analysis/paper/paper.qmd
editing the results section

@@
 We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
-Our main question is: what was the dominant mode of cultural transmission during the time of technological innovation in the Korean Late Paleolithic?
-Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and  indirect bias.
+Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic?
+Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points.
 To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA).
 We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias.
-Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more associated to guided variation.
-Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group and then modified the shape according to their specific circumstances.
+Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more due to guided variation.
+Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group, and then modified the shape according to their specific circumstances.
 
@@
 In this paper we used cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
-This technological innovation is not just the appearance of new tools such as stemmed points and blades, but if the appearance of these new technologies represents the first arrival of modern humans in this region, this transition may also represent a key event related to modern human dispersal throughout East Asia [@seong2009emergence].
+The primary technological innovation is during this time was the appearance of new tools such as stemmed points and blades.
+The appearance of these new lithic technologies may represent the first arrival of modern humans in this region, and so they may also represent a key event related to modern human dispersal throughout East Asia [@seong2009emergence].
 Previous studies on the origin of stemmed points have mainly focused on their possible origin locations, connecting Korea with global patterns of modern human dispersal.
-However, questions about the specific processes of this technological change in Korea remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
+However, questions about the specific cultural processes and social contexts of this technological change in Korea remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
 In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic.
@@
 We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two.
-Our results can help to determine if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.
+Our results have implications for determining if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.
 
@@
 
-The migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
+In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
 While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
@@
 
-Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than those models of migration or *in situ* development.
+Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development.
 He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size.
@@
 Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual].
-Transmission biases can be important loci of changes in material culture, and can be influenced by social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social].
+Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social].
 These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].
 
-Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two these biases can be identified in stone artifact assemblages.
+Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two of these biases can be identified in stone artifact assemblages.
 Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
@@
 In this context of limited contact, they argue that the new technologies developed largely by trial and error.
-On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.
+On the other hand, in central Nevada they found that metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.
 
@@
 Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
-She then computed simulated CV values according to various three scenarios of three levels of transmission fidelity (i.e.
+She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e.
 CV = 10%, 5%, and 3%).
 The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying".
-A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being to subtle for people to notice [@eerkens2000practice].
+A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
 Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.
 
-Due to their focus on archaeological data, Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
+Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
 Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
@@
 
-Inspired by Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period.
+Inspired by Bettinger and Eerkens' approach, we use two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period.
 We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness.
@@
 To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups.
-Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement.
+Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf.
+Seong's *in situ* model).
 This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points.
@@
 Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
-In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package.
+In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package (cf.
+Bae's migration model).
 In this social context, indirect bias is the dominant influence on transmission of lithic technology.
@@
 
-We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
+To analyse change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
 These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
@@
 Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
-We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to facilitate observations of change over time.
+We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to analyse them using these three phases to facilitate observations of change over time.
 
@@
 
+\newpage
+
 ```{r}
@@
 ```
+
+\newpage
 
@@
 
-To measure the variation among the measurements of the stemmed points, we used Coefficients of Variation (CV) with confidence intervals.
+\newpage
+
+To measure the variation among the measurements of the stemmed points, we computed Coefficients of Variation (CV) with confidence intervals.
 The CV is the ratio of the standard deviation to the mean.
@@
 in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence].
-In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e.
+In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
+Eerkens and Bettinger [-@eerkens2001techniques] summarized CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e.
 CV for planned production = 1.7) and about 60 (i.e.
-CV for random production = 57.7) and the values over the upper limit likely reflect intended attempts to differentiate the products.
-Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
-Considering the lower quality of raw materials in Korea and the median of the CV range (30.5), we selected roughly 25 as the threshold to decide between guided variation and indirect bias.
-We premised that CV values that are lower or similar to 25 represent guided variation while CV values over 25 reflect indirect bias.
-
-Sample size can impact on accuracy of estimated CV that reflects the corresponding population size.
+CV for random production = 57.7) and the values over the upper limit likely reflect  attempts to intentionally differentiate the products.
+Among the Great Basin projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
+Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected 25 as the threshold to decide between guided variation and indirect bias.
+We premised that CV values that are lower than 25 represent guided variation while CV values over 25 reflect indirect bias.
+
+Sample size can impact the accuracy of CV estimates.
 [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative].
-In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
-A variety of methods are available for improving the reliability of CV measurements on small samples.
-@vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e.
+In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
+A variety of methods are available for improving the reliability of CV measurements when using small samples.
+For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e.
 n\<25).
 Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
-Although we have a total of `r total_number_of_sp` artifacts, we used the corrected CV formula since we computed CV per assemblages to examine the temporal and regional pattern.
-We also computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results.
-[@kelley2007sample; @albatineh2014confidence].
-
-To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we chose Principal Component Analysis (PCA).
+Although we have a total of `r total_number_of_sp` artifacts, most of our analyses here involves comparisons of smaller subsets of this sample.
+For example to explore temporal and regional patterns we compared samples of well under 25 pieces.
+Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use.
+We also computed confidence intervals using method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].
+
+To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we used Principal Component Analysis (PCA).
 PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
@@
 
-All data treatment, analyses and visualization were performed in the R environment [@core2021team].
-Our R code and data, including the original artefact images, are fully and openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].
+All data preparation, analyses and visualization were performed in the R environment [@core2021team].
+Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].
 
 # Results {.unnumbered}
-
-## Coefficient of variation
-
-@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
-Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points.
-This result implies a similar thickness of a wooden shaft that is attached to the stem.
 
@@
 
-We used the corrected CV to measure the variations between stemmed points.
-If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias.
-Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes panel B).
-Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher.
-Unlike the distribution of actual attributes, the CVs of body length and maximum length have the lowest values.
+@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
+This result may imply that the stem width was the most constrained dimension of the artefact.
+We might speculate that this is due to the requirements of halfting with wooden shafts, where the shafts were selected from a small range of diameters that optimised strength, weight and flexibility.
+
+## Coefficient of variation
+
+Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes).
+Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artefacts is higher.
+Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values.
 The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals.
 Tang and stem related attributes (e.g.
-SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.
+SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was influenced by indirect bias.
 
@@
 
-To understand the temporal patterns in the modes of cultural transmission, we applied Korea Late Paleolithic chronological phases into our samples to compute CV (@tbl-korean-chronology).
+To understand the temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases and computed CV values for each phase (@tbl-korean-chronology).
 Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute.

### 2022-11-29 analysis/paper/paper.qmd
working on results section 'Variation Over Time'

@@
 
-@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+```{r}
+#| label: fig-plot-attributes
+# Box plot for each attribute
+# mutate data to make plot with ggplot
+data_box <- gather(DF)
+# make box plot for each attributes
+plot_raw_measurements_for_attributes <- 
+ggplot(data_box,
+       aes(key, value)) +
+  geom_boxplot() +
+  geom_beeswarm(alpha = 0.3,
+                size = 0.8,
+                cex = 0.7) +
+  ylab("Length (cm)") +
+  xlab("") +
+  theme_bw()
+```
+
+Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
 Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
@@
   ylab("CV") +
-  xlab("Variable") +
+  xlab("Artefact measurement") +
   geom_text(aes(label = round(all_cv_sharma,1)), 
@@
             size = 3) +
+  # draw a line to show our threshold value
+  geom_hline(yintercept = 25, colour = "red") +
   theme_bw()
@@
 
-To understand the temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases and computed CV values for each phase (@tbl-korean-chronology).
-Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute.
-However, this does not mean that there were only two stemmed points for the Phase 1 because we excluded broken pieces from our sample to conduct the morphological analysis.
-
-After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-plot-cv-over-time).
-Body length of Phase 2, maximum length, mid width, and tang width are under 25, which indicates as a result of indirect bias, while only maximum length are under 25 in Figure 4.
-On the other hand, Body length of Phase 3, stem length, stem width, and tang length are relatively higher then 25 and mostly over 30 indicating guided variation.
-
-Compared to Phase 2, except for maximum width and tang width, the other CV values of Phase 3 tend to be higher.
-In addition, Phase 3 tends to have wider confidence intervals.
-Stem width and tang length values are relatively steady over time.
-Our results show that stemmed points from Phase 2 are less varied than Phase 3 as well as Phase 1.
+To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarised in @tbl-korean-chronology, and computed CV values for all complete stemmed points each phase.
+We excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.
+
+@fig-plot-cv-over-time shows that the directional trend in CV values of artefact attributes is complicated.
+Half of the attributes are below our threshold value of 25, and half are above.
+Only body length changes from <25 to >25 over time, indicating more guided variation.
+For those attributes where the CV is <25, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
+For stem length, stem width and tang length, CV values remain above 25, and show a slight trend to increase over times, suggesting increased influence of guided variation for these attributes.
 
@@
   xlab("Phase") +
-  ylab("Coefficient of Variation on Attributes") +
+  ylab("Coefficient of Variation for all Attributes") +
+    # draw a line to show our threshold value
+  geom_hline(yintercept = 25, colour = "red") +
   facet_wrap( ~ variable, scales = "free_y") +
@@
   dplyr::select(-SPstage.Raw_material) %>%
-  group_by(site_phase) %>%
+  dplyr::group_by(site_phase) %>%
   add_tally() %>%
-  filter( n > 5) %>%
+  dplyr::filter( n > 5) %>%
   mutate(label = as.factor(paste0(site_phase,' (N = ', n, ")"))) %>%
@@
                values_to = "value") %>%
-  group_by(label, variable) %>%
-  summarise(cv_by_site =   sharma_cv(value),
+  dplyr::group_by(label, variable) %>%
+  dplyr::summarise(cv_by_site =   sharma_cv(value),
             cv_low_sharma =  sharma_int_low(value),
             cv_high_sharma = sharma_int_high(value)) %>%
-  filter(variable %in% c("ML", "BL","MW", "TW"))
+  dplyr::filter(variable %in% c("ML", "BL","MW", "TW"))
 
@@
                     ymax = cv_high_sharma), width = 0.2) +
+    # draw a line to show our threshold value
+  geom_hline(yintercept = 25, colour = "red") +
   xlab("Attribute") +
@@
   ylab("Coefficient of Variation on Attributes") +
+    # draw a line to show our threshold value
+  geom_hline(yintercept = 25, colour = "red") +
   facet_wrap( ~ label, scales = "free_y") +
@@
                     ymax = cv_high_sharma), width = .2) +
+    # draw a line to show our threshold value
+  geom_hline(yintercept = 25, colour = "red") +
   xlab("Law Materials") +

### 2022-11-30 analysis/paper/paper.qmd
still editing

@@
   fig-dpi: 600
-  cache: true
+  cache: false
 editor: visual
@@
 
+threshold_value <- 25
+
 ```
@@
 Among the Great Basin projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
-Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected 25 as the threshold to decide between guided variation and indirect bias.
-We premised that CV values that are lower than 25 represent guided variation while CV values over 25 reflect indirect bias.
-
-Sample size can impact the accuracy of CV estimates.
-[@toebe2018sample; @kelley2007sample; @vanpool2011quantitative].
+Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected `r threshold_value` as the threshold to decide between guided variation and indirect bias.
+We premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
+<!-- I STILL THIS WE NEED A MORE COMPELLING JUSTIFICATION FOR THE THRESHOLD VALUE, BASED ON SOME THEORY IN THE LITERATURE.
+I EXPECT REVIEWERS WILL GET STUCK ON THIS AND DEMAND MORE DETAILS.
+THE CHOICE OF VALUE HAS A HUGE EFFECT ON OUR INTERPRETATIONS BELOW -->
+
+Sample size can impact the accuracy of CV estimates [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative].
 In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
@@
 
-All data preparation, analyses and visualization were performed in the R environment [@core2021team].
+All data preparation, analyses and visualization were performed in the R environment [@RCoreTeam].
 Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].
@@
   # draw a line to show our threshold value
-  geom_hline(yintercept = 25, colour = "red") +
+  geom_hline(yintercept = threshold_value, colour = "red") +
   theme_bw()
@@
 @fig-plot-cv-over-time shows that the directional trend in CV values of artefact attributes is complicated.
-Half of the attributes are below our threshold value of 25, and half are above.
-Only body length changes from <25 to >25 over time, indicating more guided variation.
-For those attributes where the CV is <25, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
-For stem length, stem width and tang length, CV values remain above 25, and show a slight trend to increase over times, suggesting increased influence of guided variation for these attributes.
+Half of the attributes are below our threshold value of `r threshold_value`, and half are above.
+Only body length crosses the threshold value, changing from <`r threshold_value` to >`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
+For those attributes where the CV is <`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
+For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over times, suggesting increased influence of guided variation for these attributes.
 
@@
     # draw a line to show our threshold value
-  geom_hline(yintercept = 25, colour = "red") +
+  geom_hline(yintercept = threshold_value, colour = "red") +
   facet_wrap( ~ variable, scales = "free_y") +
@@
 
-We then explored assemblages that contain multiple stemmed points to examine the temporal patterns with the premise that stemmed points from one assemblage were made by the same group of people.
-There are four assemblages, three assemblages from the Suyanggae site (SYG) and one from the Yongsandong site (YS) that contain more than 5 stemmed points.
-SYG1_2 and SYG6_2 are from Phase 2 and SYG1_3 and YS_3 are from Phase 3.
+One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
+These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-over-time).
+These four assemblages are more likely to represent a consistent way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making.
+The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
+SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure).
 Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width.
-@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below 25, which represent transmissions through indirect bias.
-Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest range of confidence intervals.
-Compared to the Phase 2, CV values of Phase 3 are lower, which is different from @fig-plot-cv-over-time results of body length and maximum length from all assemblages.
-It is assumed that those attributes became more standardized within the same assemblages.
-Compared to Suyanggae assemblages, Yongsandong has smaller CV values.
+@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below `r threshold_value`, which represent transmissions dominated by indirect bias.
+Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here.
+Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-over-time.
+This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of indirect bias, rather than substantial changes in the type of bias dominating cultural transmission.
 
@@
     # draw a line to show our threshold value
-  geom_hline(yintercept = 25, colour = "red") +
+  geom_hline(yintercept = threshold_value, colour = "red") +
   xlab("Attribute") +

### 2022-11-30 analysis/paper/paper.qmd
edited up to Principal Component Analysis

@@
 
-All data preparation, analyses and visualization were performed in the R environment [@RCoreTeam].
+All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam].
 Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].
@@
 For those attributes where the CV is <`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
-For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over times, suggesting increased influence of guided variation for these attributes.
+For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these attributes.
 
@@
 These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-over-time).
-These four assemblages are more likely to represent a consistent way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making.
+These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making.
 The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
@@
 We examined the CV values per site to address our research question about the cultural transmission over space.
-We excluded assemblages that contain a single point.
-@fig-plot-cv-by-site shows that Sachang site has the highest CV values with the widest confidence level.
-Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2.
-Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points.
-CV values of body length is the lowest while stem width the highest among all assemblages.
-Overall, there are no clear patterns between the sites.
+We excluded assemblages that contain less than five points.
+@fig-plot-cv-by-site shows higher CV values for stem length and stem width, indicating high variability at the stem of the points from these sites.
+Overall size variability is similar between these three assemblages, suggesting a consistent approach to constraining shape variation over space, rather than site-specific variances.
 
@@
 #| label: fig-plot-cv-by-site
-#| fig-cap: CV values per site that has more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
+#| fig-cap: CV values per site that has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
 #| fig-width:   8.5
@@
   add_tally() %>%
-  dplyr::filter( n > 1) %>%
+  dplyr::filter( n > 5) %>%
   dplyr::select(-n) %>%
@@
   theme(legend.position = "none") +
-  coord_cartesian(ylim = c(0,250)) +
+  coord_cartesian(ylim = c(0,100)) +
   theme_bw() +
@@
 
-We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials.
-We excluded raw materials that were used for single point.
-@fig-plot-cv-by-raw-material shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals.
-Quartz has high variations of CV among the attributes.
-On the other hand, shale has the lower CV, which is very stable throughout the attributes.
+We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was highly dependent on raw material.
+We excluded raw materials that were used for less than three points.
+@fig-plot-cv-by-raw-material shows that quartzite, the raw materials for stemmed points at Sachang, has the highest CV value with a wide range of conference intervals.
+Other raw materials are generally stable in variation throughout the attributes.
 
@@
   add_tally() %>%
-  dplyr::filter( n > 1) %>%
+  dplyr::filter( n >= 3) %>%
   dplyr::select(-n) %>%

### 2022-11-30 analysis/data/raw_data/b-and-e-table-5.R

@@ -1,0 +1,61 @@
+library(magrittr)
+library(magick)
+library(tesseract)
+library(tidyverse)
+
+# preprocessing
+img <-
+  image_read(here::here("analysis/data/raw_data/b-and-e-table-5.png")) %>%
+  image_transparent("white", fuzz=82) %>%
+  image_background("white") %>%
+  image_negate() %>%
+  image_crop(geometry_area(0, 0, 25, 25))
+
+img
+
+# read img and ocr
+data <- img %>%
+  image_ocr()
+
+# some wrangling
+tbl <-
+data %>%
+  stringi::stri_split(fixed = "\n") %>%
+  .[[1]]
+
+
+Monitor_Valley <- tbl[9:11] # Nevada - indirect bias
+Monitor_Valley_tbl <-
+tibble(mean = stringi::stri_split(Monitor_Valley[1], fixed = " ")[[1]],
+       std =  stringi::stri_split(Monitor_Valley[2], fixed = " ")[[1]],
+       n =    stringi::stri_split(Monitor_Valley[3], fixed = " ")[[1]]) %>%
+  slice(-1) %>%
+  mutate(cv = parse_number(std) / parse_number(mean))
+
+Eastern_California <- tbl[13:15] # Eastern California - guided variation
+Eastern_California_tbl <-
+  tibble(mean = stringi::stri_split(Eastern_California[1], fixed = " ")[[1]],
+         std =  stringi::stri_split(Eastern_California[2], fixed = " ")[[1]],
+         n =    stringi::stri_split(Eastern_California[3], fixed = " ")[[1]]) %>%
+  slice(-1) %>%
+  mutate(cv = parse_number(std) / parse_number(mean))
+
+Rose_Spring_Site <- tbl[17:19] # Eastern California - guided variation
+Rose_Spring_Site_tbl <-
+  tibble(mean = stringi::stri_split(str_squish(str_remove(Rose_Spring_Site[1], "no data")), fixed = " ")[[1]],
+         std =  stringi::stri_split(str_squish(str_remove(Rose_Spring_Site[2], "no data")), fixed = " ")[[1]][1:7],
+         n =    stringi::stri_split(str_squish(str_remove(Rose_Spring_Site[3], "no data")), fixed = " ")[[1]]) %>%
+  slice(-1) %>%
+  mutate(cv = parse_number(std) / parse_number(mean))
+
+bind_rows(list(Monitor_Valley = Monitor_Valley_tbl,
[…]

### 2022-11-30 analysis/paper/paper.qmd
remove PCA

@@
 
-To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we used Principal Component Analysis (PCA).
-PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points.
-PCA is often used to reduce dimensionality, detect patterns and clusters in data, indicate relationships between variables as well as cases, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life].
-PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].
-
 All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam].
@@
 #| fig-width:   4.5
-#| fig-height:  4.5
+#| fig-height:  4
 
@@
 #| label: fig-plot-cv-four-assemblage
-#| fig-cap: CV values of four attributes from assemblages that more than 5 stemmed points.
+#| fig-cap: CV values of four attributes from assemblages with more than 5 stemmed points.
 Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
 #| fig-width:   4.5
-#| fig-height:  4.5
+#| fig-height:  4
 
@@
 #| label: fig-plot-cv-by-site
-#| fig-cap: CV values per site that has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
+#| fig-cap: CV values per site for each site has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
 #| fig-width:   8.5
-#| fig-height:  4.5
+#| fig-height:  4
 
@@
   theme(legend.position = "none") +
-  coord_cartesian(ylim = c(0,100)) +
+  coord_cartesian(ylim = c(0,70)) +
   theme_bw() +
@@
 We excluded raw materials that were used for less than three points.
-@fig-plot-cv-by-raw-material shows that quartzite, the raw materials for stemmed points at Sachang, has the highest CV value with a wide range of conference intervals.
+@fig-plot-cv-by-raw-material shows that quartzite, the raw material for the stemmed points at Sachang, has the highest CV values, and also has wide conference intervals.
 Other raw materials are generally stable in variation throughout the attributes.
@@
 #| fig-width:   8.5
-#| fig-height:  4.5
+#| fig-height:  4
 
@@
   facet_wrap( ~ label, scales = "free_y") +
-  coord_cartesian(ylim = c(0,130)) +
+  coord_cartesian(ylim = c(0,250)) +
   theme_bw() 
@@
 
-## Principal Component Analysis
-
-PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission.
-@fig-pca-attributes shows positive relationship between attributes.
-Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length.
-Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other.
-Other width and tang related attributes are close to each other but a bit far from body length and maximum length.
-Stem width contributes the least to the overall shape.
-We can assume that the body length and maximum length were prioritized when the manufacturing technique of stemmed points were transmitted.
-
-For the temporal variation, we computed PCA for difference periods.
-Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective.
-However, there are no notable changes in correlations over time.
-
-```{r}
-#| label: fig-pca-attributes
-#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three.
-The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated.
-The length of the arrow indicates the strength of contribution to the stemmed point shape.
+## Correlation analysis of attributes
+
+```{r}
+#| label: fig-corr-analysis
+#| fig-cap: Correlation analysis of point attributes, calculated using Pear
 #| fig-width:   7
-#| fig-height:  3
-
-library(ggbiplot)
-library(ggfortify)
-### PCA analysis using ggbiplot, ggfortify pkgs
-
-PCA <- prcomp(DF, center = TRUE, scale = TRUE)
-
-# remove stage 1 artefact
-xx_pca <-
+#| fig-height:  4
+
+library(corrr)
+
+# correlation of all artefacts
+df_sitename_corr <- 
   df_sitename %>%
-  dplyr::filter(SPstage.Stage != 1) %>%
   dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
-  prcomp(., center = TRUE, scale = TRUE)
-
-xx_pca1 <-
-  data.frame(xx_pca$x,
-             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))
-
-## CV with site (n=X)
-label_phase <-
-  df_sitename %>%
-  group_by(SPstage.Stage) %>%
-  add_tally() %>%
-  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))
-
-count_phase <- length(which(label_phase[9]>0))
-count_phase2 <- length(which(label_phase[9] == 2))
-count_phase3 <- length(which(label_phase[9] == 3))
-
-# phase 2
-
-xx_pca2 <-
+  correlate() %>% 
+  shave()
+
+all_corr_plot <- rplot(df_sitename_corr, legend = FALSE)
+
+df_sitename_corr_long <- 
+  df_sitename_corr %>% 
+    pivot_longer(-term)
+
+# correlation of phase 2 artefacts
+df_sitename_corr_p2 <- 
   df_sitename %>%
@@
   dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
-  prcomp(., center = TRUE, scale = TRUE)
-
-# phase 3
-
-xx_pca3 <-
+  correlate() %>% 
+  shave()
+
+p2_corr_plot <- rplot(df_sitename_corr_p2, legend = FALSE)
+
+df_sitename_corr_p2_long <- 
+  df_sitename_corr_p2 %>% 
+    pivot_longer(-term)
+
+# correlation of phase 3 artefacts
+df_sitename_corr_p3 <- 
   df_sitename %>%
@@
   dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
-  prcomp(., center = TRUE, scale = TRUE)
-
-##### using factoextra pkg
-library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
-
-# Color variables by the continuous variable
-cv_all <-
-fviz_pca_var(xx_pca) +
-  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
-  theme(legend.position = "none")
-
-# phase 2
-cv_two <-
-  fviz_pca_var(xx_pca2) +
-  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))
-
-# phase 3
-cv_three <-
-  fviz_pca_var(xx_pca3) +
-  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))
-
-library(patchwork)
-cv_all + cv_two + cv_three
+  correlate() %>% 
+  shave()
+
+p3_corr_plot <-  
+  rplot(df_sitename_corr_p3) +
+   theme(legend.position = c(0.8, 0.8))
+
+df_sitename_corr_p3_long <- 
+  df_sitename_corr_p3 %>% 
+    pivot_longer(-term)
+
+# compare p2 and p3 correlations
+compare_p2_and_p3_correlations_plot <- 
+list(`phase 2` = df_sitename_corr_p2_long,
+     `phase 3` = df_sitename_corr_p3_long) %>% 
+  bind_rows(.id = "phase") %>% 
+  ggplot() +
+  aes(phase,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 4,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient")
+
+library(cowplot)
+plot_grid(
+          plot_grid(p2_corr_plot, 
+                    p3_corr_plot,
+                    ncol = 1,
+                    labels = c("A", "B")),
+          compare_p2_and_p3_correlations_plot,
+          ncol = 2,
+          labels = c("", "C"),
+          rel_heights = c(1, 1.5))
+
 ```

### 2022-11-30 analysis/data/raw_data/b-and-e-table-9.R

@@ -1,0 +1,33 @@
+library(magrittr)
+library(magick)
+library(tesseract)
+library(tidyverse)
+library(readxl)
+
+# data from table 9 of
+# https://www.researchgate.net/publication/277543080_Point_Typologies_Cultural_Transmission_and_the_Spread_of_Bow-and-Arrow_Technology_in_the_Prehistoric_Great_Basin
+
+# Central Nevada: highly correlated - indirect bias
+c_nv <- read_excel("analysis/data/raw_data/b-and-e-table-9.xlsx",
+                   sheet = 1) %>%
+  pivot_longer(-`C. NV` ) %>%
+  dplyr::rename(dim = `C. NV`)
+
+# Eastern California: poorly correlated - guided variation
+e_ca <- read_excel("analysis/data/raw_data/b-and-e-table-9.xlsx",
+                   sheet = 2) %>%
+  pivot_longer(-`E. CA`) %>%
+  dplyr::rename(dim = `E. CA`)
+
+list(
+c_nv = c_nv ,
+e_ca = e_ca
+) %>%
+  bind_rows( .id = "location") %>%
+  ggplot() +
+  aes(location,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 4,
+                               alpha = 0.3) +
+  theme_minimal()

### 2022-11-30 analysis/paper/paper.qmd
editing discussion

@@
 
+Following @bettinger1999point we computed Pearson's correlation coefficients for all measurements of the seven attributes that we measured on the stemmed points.
+The correlation of artefact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission.
+This is based on the assumption that assumed that artefact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
+In their work points from Eastern California and Central Nevada, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.
+
 All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam].
@@
 Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
-This result may imply that the stem width was the most constrained dimension of the artefact.
-We might speculate that this is due to the requirements of halfting with wooden shafts, where the shafts were selected from a small range of diameters that optimised strength, weight and flexibility.
 
@@
 
+<!-- GP to write up the results in detail, that you see in the plot made by the code below.
+Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology" -->
+
 ```{r}
@@
 We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
-We set roughly 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
-If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.
-
-The CV values for all stemmed points are mostly over 25, except for maximum length.
-PCA analysis shows positive relationships between attributes (@fig-plot-cv-all-attributes panel B).
-We then examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time).
-After excluding two stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower.
-In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25.
-We observe the different degree of contribution by each attribute to the shape of stemmed points.
-For example, tang and stem related attributes have higher CV values compared to the body related attributes.
-We expect that people chose to emulate the body related attributes and had more freedom to produce the tang and stem related attributes.
-
-Compared to the Phase 2 and 3, we found that there are only minor differences between the two phases.
-The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3.
-The increase is not significant to generalize (p \> 0.05, @tbl-cv-over-time), we explored individual assemblages including multiple stemmed points that could show a more clear pattern.
-The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (@fig-plot-cv-four-assemblage).
-This could indicate the stronger connections within the groups during Phase 3.
+We set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
+If the CV value is lower than 25, we interpreted this as indicating indirect bias while values higher than 25 were interpreted as guided variation.
+
+The CV values for all stemmed point attributes are mostly over 25, except for maximum length, which is just below.
+Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time).
+Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases.
+Applying the Modified signed-likelihood ratio test (SLRT) to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
+In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).
 
@@
 
-There is no clear pattern variation in CV between the sites.
-Some sites such as Sachang have very high CV values while other sites including Suyanggae and Jingeunuel have lower CV (@fig-plot-cv-by-site).
-We assumed that the CV could be impacted by other factors such as the choice of raw materials (@fig-plot-cv-by-raw-material).
-However, similar to the CV variation for the Phase 2, the results of site variation show that CV of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, indicating the combination of the two biases.
-The attributes with lower CV values could be taught through indirect bias while the higher CV attributes could be depending more on various manufacturing situations.
-We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the function of the tool so that the knappers wanted to standardize the shape.
+In exploring geographical variation we found no clear pattern variation in CV between sites with more than five stemmed points (@fig-plot-cv-by-site).
+The most striking pattern in this site comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
+The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations.
+We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape.
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
-This combination model describes that there were social connections between groups that share the tool production technique.
+This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artefact.
+These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
+
+The most striking differences in CV values is evident in comparisons of points grouped by raw materials.
+Artefacts made from quartzite have much higher CV values, and wider confidence intervals, than all other raw materials (@fig-plot-cv-by-raw-material).
+This likely reflects the relatively unpredictable knapping quality of quartzite due to the presence of visible grains.
+These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals.
+This makes is harder for the knapper to repeatedly produce the same size and shape end product.
+This hints at the possibility that physical properties of the stone artefact raw materials were a key mechanism in determining shape and size variability.
+However with only three artefacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.
+
+However, similar to the CV variation for the Phase 2,  This combination model describes that there were social connections between groups that share the tool production technique.
 Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias).

### 2022-12-01 analysis/paper/paper.qmd
added comments for GP's tasks to revise the manuscript

@@
 Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points.
-To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA).
+To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and correlation coefficients.
 We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias.
@@
 Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group, and then modified the shape according to their specific circumstances.
+<!-- GP to revise and edit the abstract after completing the other tasks below -->
 
@@
 Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
-We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to analyse them using these three phases to facilitate observations of change over time.
+We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.
+
+Panel A of @fig-raw-materials-by-phase summarises the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases .
+We exclude phase 1 from our analyses below because of the small sample size.
+Shale is the dominant raw material in phase 2 and 3.
+In phase 3 there is an increase in the proportions of rhyolite and porphyry, driven largely by finds from Wolpyeng and Yongsangdong.
+Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at sites with more than five stemmed points.
+Much of the raw material diversity comes from isolated finds, with most sites with more than five points showing high homogeneity in raw materials.
+
+<!-- BM: I have left many captions with only xxx. GP please fill them out with detailed and specific captions -->
+
+```{r}
+#| label: fig-raw-materials-by-phase
+#| fig-cap: "A: xxx. B: Frequencies of stemmed points by site and raw material, for sites with five or more stemmed points "
+#| fig-height: 4
+
+df_full_sitename_fct <- 
+df_full_sitename %>% 
+  mutate(SPstage.Raw_material = factor(SPstage.Raw_material))
+
+#Create a custom color scale
+library(RColorBrewer)
+myColors <- 
+  brewer.pal(length(unique(df_full_sitename_fct$SPstage.Raw_material)),
+             "RdBu")
+names(myColors) <- levels(df_full_sitename_fct$SPstage.Raw_material)
+colScale <- scale_fill_manual(name = "Raw material",
+                                values = myColors)
+
+p1 <- 
+ggplot(df_full_sitename_fct) +
+  aes(SPstage.Stage,
+      fill = SPstage.Raw_material) +
+  geom_bar() +
+  theme_bw() +
+  labs(fill = "Raw material",
+       x = "Chronological Phase") +
+  theme(legend.position = "top",
+        legend.key.size = unit(0.5, 'cm'), #change legend key size
+        legend.key.height = unit(0.5, 'cm'), #change legend key height
+        legend.key.width = unit(0.5, 'cm'), #change legend key width
+        legend.title = element_text(size=8), #change legend title font size
+        legend.text = element_text(size=6)
+        ) +
+colScale
+
+p2 <- 
+  df_full_sitename_fct %>%
+  mutate(SPstage.Stage = paste0("Phase ", SPstage.Stage)) %>% 
+  group_by(full_sitename) %>% 
+  filter(n() >= 5) %>% 
+    ggplot() +
+    aes(full_sitename,
+        fill = SPstage.Raw_material) +
+    geom_bar() +
+    facet_wrap(~ SPstage.Stage) +
+  theme_bw () +
+  labs(fill = "Raw material",
+       x = "Site") +
+ theme(legend.position = "none",
+       axis.text.x = element_text(size = 4)  ) +
+  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
+   colScale
+
+library(cowplot)
+plot_grid(p1, p2,
+          ncol = 1,
+          rel_heights = c(1.5, 0.75),
+          labels = c("A", "B"))
+```
 
@@
 
+<!-- BM: I think these tables should appear before the figures above, what do you think? What do you think a typical reader is expecting? -->
+
 ```{r}
@@
 
+# to ensure we don't have any copy-paste errors
 threshold_value <- 25
-
 ```
@@
 We premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
-<!-- I STILL THIS WE NEED A MORE COMPELLING JUSTIFICATION FOR THE THRESHOLD VALUE, BASED ON SOME THEORY IN THE LITERATURE.
+<!-- BM: I STILL THIS WE NEED A MORE COMPELLING AND STRONGER JUSTIFICATION FOR THE THRESHOLD VALUE, BASED ON SOME THEORY IN THE LITERATURE.
+PLEASE ADD SOMETHING.
 I EXPECT REVIEWERS WILL GET STUCK ON THIS AND DEMAND MORE DETAILS.
@@
 
+<!-- BM: overall I think the CV values are really not that interesting or informative.
+They are difficult to interpret because they all cluster around 15-30, which seems to me to be the midpoint between 'highly standardised' (3-5) and 'highly different' (40-50).
+So I think we will struggle to convince any readers that we have any strong signals of cultural transmission biases in the CV data.
+Probably we can satisfy most readers that guided variation is the main bias in these data.
+That's my perspective after reading this paper several times, and re-reading several related papers by Bettiner and Eerkens and Lipo, and others  -->
+
 Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes).
@@
 Tang and stem related attributes (e.g.
-SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was influenced by indirect bias.
+SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was more influenced by indirect bias than the others.
 
@@
 95% confidence intervals for the CVs were computed using Sharma and Krishna's method.
-The vertical lines indicate the range of confidence intervals.
+The vertical lines indicate the confidence intervals.
 
@@
 The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.
+The vertical lines indicate the confidence intervals.
 #| fig-width:   4.5
@@
   add_tally() %>%
-  dplyr::filter(SPstage.Stage > 1) %>%
+  dplyr::filter(n >= 5) %>%
   dplyr::select(-n) %>%
@@
 One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
-These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-over-time).
+These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-four-assemblage).
 These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making.
@@
 Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width.
-@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below `r threshold_value`, which represent transmissions dominated by indirect bias.
+@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`.
 Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here.
-Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-over-time.
-This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of indirect bias, rather than substantial changes in the type of bias dominating cultural transmission.
+Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
+This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
 
@@
 Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
+The points represent CV values for each attribute.
+The vertical lines indicate the confidence interval.
 #| fig-width:   4.5
@@
 #| label: fig-plot-cv-by-raw-material
-#| fig-cap: CV values for raw materials that were used for making more than one stemmed point.
-The points represent CV values for each attribute.
-The vertical lines indicate the range of confidence intervals.
+#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the confidence intervals.
 #| fig-width:   8.5
@@
   geom_hline(yintercept = 25, colour = "red") +
-  xlab("Law Materials") +
+  xlab("Raw Materials") +
   ylab("Coefficient of Variation on Attributes") +
@@
 
-<!-- GP to write up the results in detail, that you see in the plot made by the code below.
-Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology" -->
+<!-- BM: You'll see I've completely removed the PCA content.
+I did some more reading into what PCA calculates, and it is not correlation coefficients of the variables with each other, but with the Principal Components that it calculates.
+So that's a very different kind of correlation, compared to what we see in the previous literature on transmission biases.
+I suppose we could use it and interpret it, if you want to investigate that further.
+I could not think of anything straightforward over the several days I have been editing this.
+I saw that you did not give them the same treatment as the CV values with exploring variation by raw material and by site, so I guess you were not very interested in the PCA results.
+My perspective is that we should look for the quickest and simplest way to complete this paper.
+So I have replaced the PCA with the same kind of correlation analysis that we see elsewhere so we can use the previously published papers to guide our interpretation of our results -->
+
+<!-- BM: GP to write up the correlation coefficients results in detail, according to what you see in the plot made by the code below.
+I've already done all the analysis here.
+Your task is to describe it here, and then contextualise and interpret it in the next section.
+Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology".
+Aim for a style of writing that is detailed, direct, specific, short sentences.
+-->
 
@@
 #| label: fig-corr-analysis
-#| fig-cap: Correlation analysis of point attributes, calculated using Pear
+#| fig-cap: "A: xxx. B. xxx. C. xxx"
 #| fig-width:   7
-#| fig-height:  4
+#| fig-height:  9
+
+# BM: I added this section of analysis to replace the PCA sections.
 
@@
   rplot(df_sitename_corr_p3) +
-   theme(legend.position = c(0.8, 0.8))
+   theme(legend.position = c(0.9, 0.9),
+         legend.key.size = unit(0.3, 'cm'), #change legend key size
+        legend.key.height = unit(0.3, 'cm'), #change legend key height
+        legend.key.width = unit(0.3, 'cm'), #change legend key width
+        legend.title = element_text(size=6), #change legend title font size
+        legend.text = element_text(size=8))
 
@@
   geom_boxplot() +
-  ggbeeswarm::geom_quasirandom(size = 4,
+  ggbeeswarm::geom_quasirandom(size = 2,
                                alpha = 0.3) +
@@
 
+# correlation by raw material
+df_sitename_corr_raw_material <- 
+  df_sitename %>%
+  group_by(SPstage.Raw_material) %>% 
+  filter(n() >= 3) %>% 
+  dplyr::select(-sitename, 
+                -SPstage.Stage) %>% 
+  nest(-SPstage.Raw_material) %>% 
+  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+                                   shave() %>% 
+                                  pivot_longer(-term))) %>% 
+  unnest(corrr) %>% 
+  filter()
+
+correlations_by_raw_material <- 
+  ggplot(df_sitename_corr_raw_material) +
+  aes(SPstage.Raw_material,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient",
+       x = "Raw material")
+
+# correlation by site
+df_sitename_corr_site <- 
+  df_sitename %>%
+  group_by(sitename) %>% 
+  filter(n() >= 5) %>% 
+  dplyr::select(-sitename, 
+                -SPstage.Stage) %>% 
+  nest(-sitename) %>% 
+  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+                                   shave() %>% 
+                                  pivot_longer(-term))) %>% 
+  unnest(corrr) %>% 
+  filter()
+
+correlations_by_site <- 
+  ggplot(df_sitename_corr_site) +
+  aes(sitename,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient",
+       x = "Site")
+
 library(cowplot)
-plot_grid(
+plot_grid(plot_grid(
           plot_grid(p2_corr_plot, 
@@
           labels = c("", "C"),
-          rel_heights = c(1, 1.5))
+          rel_heights = c(1, 1.5)),
+          correlations_by_raw_material,
+          correlations_by_site,
+          ncol = 1,
+          labels = c("", "D", "E"))
 
@@
 Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases.
-Applying the Modified signed-likelihood ratio test (SLRT) to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
+Applying the Modified Signed-Likelihood Ratio Test (SLRT) <!-- GP to add a citation for this test here --> to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
 In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).
@@
 
-However, similar to the CV variation for the Phase 2,  This combination model describes that there were social connections between groups that share the tool production technique.
-Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias).
-Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability.
-Our results recalls -@lee2013current's model but his model explains combination of the *in situ* tradition for flake tools and introduction of blade-associate toolkit including stemmed points.
-On the other hand, we saw the combination of the *in situ* tradition and introduced stemmed point technology within the single artifact.
+<!-- BM: GP to write up the discussion of the correlation analysis in detail.
+Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology".
+Aim to improve your writing, make it better than your usual style.
+-->
 
@@
 
-In this research, we examine the dominant mode of cultural transmission that might impact technological innovation in the Korean Late Paleolithic.
-Following the previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we use two transmission biases, guided variation and indirect bias to scrutinize the transmission process and social contexts.
-We claim the two scenarios that explains the introduction of new technology: socially isolated groups made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others (indirect bias).
-We ask three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-do the modes of cultural transmission vary over time?
-and do the modes of cultural transmission vary over space?
-To determine transmission biases and answer our research questions, we computed CV and PCA.
-The higher CV values (i.e.
-\>=25) and low correlation indicate guided variation while the lower CV while the lower CV values (i.e.
-\<25) and high correlations represent indirect bias.
-To offset the small sample issue and increase the credibility of the comparisons, we applied confidence intervals to CV for better estimation.
-
-Our results show that some attributes such as body length, maximum length, mid width, and tang width were transmitted through indirect bias and the other attributes were developed through guided variation.
-All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points.
-We assume that the guided variation played a more important role at the beginning of the transition based on high CV values of Phase 1.
-We observe a slight decrease in CV values over time.
-No clear regional patterns are confirmed.
-CV values could be depending on raw materials.
-For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals.
-Overall, we conclude that combination of the two transmission biases well explains the technological transition during the Late Paleolithic.
-We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission.
-We used landmarks to earn the dimensional attributes, which has potential for future research with different dimensions by calculating the distance between other sets of the existing landmarks.
-In our future research, we will focus on understanding actual connection between the assemblages and examining the social networks for sharing technology between the groups such as trades or physical migration to different locations.
+In this research, we investigated the social context of technological innovation in the Korean Late Paleolithic.
+Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we investigated two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
+We proposed  two scenarios for explaining the introduction of new technology: socially isolated groups that made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came in contact with (indirect bias).
+We asked three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+Do the modes of cultural transmission vary over time?
+And do the modes of cultural transmission vary over space?
+
+We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error).
+This is indicated by relatively high CV values of around and above 25 and mostly low correlation coefficients of less than 0.5.
+A social context that favoured trial and error for artefact making is consistent with Seong's _in situ_ model for the appearance of stemmed points in Korea.
+This model proposes that people developed stemmed points by experimenting with existing blade technologies.
+Our results support experimentation and trial and error as important processes in the cultural transmission of this artefact technology.
+
+The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values at 0.5 and higher in chronological phase 3, after 25 ka.
+This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
+This may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model.
+We might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure <!-- GP to find something relevant by Fitzhugh to cite here -->.
+With more frequent contacts and connections with members of the social network, there may have been more opportunities to learn point-making directly from socially successful individuals.
+
+In comparing sites with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
+We did not find strong support for distinctive forms of cultural transmission of tool-making at different sites.
+Correlation coefficient values across the sites affirm the overall chronological trend indicating indirect bias in phase three.
+Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Wolpyeng and Yonsangdong that date to this period.
+One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favoured an expanded range for searching for raw materials, leading to a higher diversity.
+Similarly, higher mobility may have resulted in more frequent contacts and connections with members of social networks.
+This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.
+
+<!-- BM: GP to check where we can cite our QSR paper in the conclusion.
+Does what we say here agree with what we said in the QSR paper?
+GP to edit the conclusion here to show good harmony with our previous paper -->.
 

### 2023-01-20 analysis/paper/paper.qmd
tackled the threshold issue

@@
 Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points.
-To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and correlation coefficients.
-We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias.
-Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more due to guided variation.
-Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group, and then modified the shape according to their specific circumstances.
-<!-- GP to revise and edit the abstract after completing the other tasks below -->
+To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV).
+We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias over time.
+Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation.
+Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation and then indirect bias was adapted to carry out the technological change.
+We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial part of a successful model to ensure the quality to optimize tool usage.
+As a result, the shape of stemmed points became standardized among their social groups.
 
 keywords: |
-  Stemmed points; Korean Paleolithic; Cultural Transmission; Learning process, Transmission Biases
+  Stemmed points; Korean Paleolithic; Cultural Transmission; Learning process, Copying errors, Transmission Biases, Guided variation, Indirect bias
 ---
@@
 
-Previous studies of technological transitions during the Late Paleolithic in Korea have mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
+Previous studies of the technological transitions during the Late Paleolithic in Korea have mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
 The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
@@
 He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention.
-In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies .
+In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies.
 
@@
 
-# how many stemmed artefacts per site?
+# how many stemmed artifacts per site?
 stemmed_points_per_site <-
@@
 
-To analyse change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
-Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
-Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations.
-Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
-We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.
-
-Panel A of @fig-raw-materials-by-phase summarises the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases .
-We exclude phase 1 from our analyses below because of the small sample size.
-Shale is the dominant raw material in phase 2 and 3.
-In phase 3 there is an increase in the proportions of rhyolite and porphyry, driven largely by finds from Wolpyeng and Yongsangdong.
-Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at sites with more than five stemmed points.
-Much of the raw material diversity comes from isolated finds, with most sites with more than five points showing high homogeneity in raw materials.
-
-<!-- BM: I have left many captions with only xxx. GP please fill them out with detailed and specific captions -->
-
-```{r}
-#| label: fig-raw-materials-by-phase
-#| fig-cap: "A: xxx. B: Frequencies of stemmed points by site and raw material, for sites with five or more stemmed points "
-#| fig-height: 4
-
-df_full_sitename_fct <- 
-df_full_sitename %>% 
-  mutate(SPstage.Raw_material = factor(SPstage.Raw_material))
-
-#Create a custom color scale
-library(RColorBrewer)
-myColors <- 
-  brewer.pal(length(unique(df_full_sitename_fct$SPstage.Raw_material)),
-             "RdBu")
-names(myColors) <- levels(df_full_sitename_fct$SPstage.Raw_material)
-colScale <- scale_fill_manual(name = "Raw material",
-                                values = myColors)
-
-p1 <- 
-ggplot(df_full_sitename_fct) +
-  aes(SPstage.Stage,
-      fill = SPstage.Raw_material) +
-  geom_bar() +
-  theme_bw() +
-  labs(fill = "Raw material",
-       x = "Chronological Phase") +
-  theme(legend.position = "top",
-        legend.key.size = unit(0.5, 'cm'), #change legend key size
-        legend.key.height = unit(0.5, 'cm'), #change legend key height
-        legend.key.width = unit(0.5, 'cm'), #change legend key width
-        legend.title = element_text(size=8), #change legend title font size
-        legend.text = element_text(size=6)
-        ) +
-colScale
-
-p2 <- 
-  df_full_sitename_fct %>%
-  mutate(SPstage.Stage = paste0("Phase ", SPstage.Stage)) %>% 
-  group_by(full_sitename) %>% 
-  filter(n() >= 5) %>% 
-    ggplot() +
-    aes(full_sitename,
-        fill = SPstage.Raw_material) +
-    geom_bar() +
-    facet_wrap(~ SPstage.Stage) +
-  theme_bw () +
-  labs(fill = "Raw material",
-       x = "Site") +
- theme(legend.position = "none",
-       axis.text.x = element_text(size = 4)  ) +
-  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
-   colScale
-
-library(cowplot)
-plot_grid(p1, p2,
-          ncol = 1,
-          rel_heights = c(1.5, 0.75),
-          labels = c("A", "B"))
-```
+\newpage
 
@@
 #| label: fig-map
-#| fig-cap: Korean Paleolithic sites mentioned in this study
+#| fig-cap: "Korean Paleolithic sites mentioned in this study."
 #| fig-width:   4.45
@@
 
-<!-- BM: I think these tables should appear before the figures above, what do you think? What do you think a typical reader is expecting? -->
-
 ```{r}
@@
 
+To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
+These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
+Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
+Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations.
+Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
+We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.
+
 ```{r}
@@
 
+Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases.
+We exclude phase 1 from our analyses below because of the small sample size.
+Shale is the dominant raw material in both phase 2 and 3.
+In phase 3 there is an increase in the proportions of rhyolite and porphyry, driven largely by finds from Wolpyeng and Yongsangdong.
+Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at assemblages with more than five stemmed points.
+Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.
+
+```{r}
+#| label: fig-raw-materials-by-phase
+#| fig-cap: "A: Three chronological phases of the Korean Late Paleolithic, and frequecies of stemmed points and distribution of raw materials in each phase.
+B: Frequencies of stemmed points by assemblage and raw material, for assemblages with five or more stemmed points."
+#| fig-height: 4
+
+df_full_sitename_fct <- 
+df_full_sitename %>% 
+  mutate(SPstage.Raw_material = factor(SPstage.Raw_material))
+
+#Create a custom color scale
+library(RColorBrewer)
+myColors <- 
+  brewer.pal(length(unique(df_full_sitename_fct$SPstage.Raw_material)),
+             "RdBu")
+names(myColors) <- levels(df_full_sitename_fct$SPstage.Raw_material)
+colScale <- scale_fill_manual(name = "Raw material",
+                                values = myColors)
+
+p1 <- 
+ggplot(df_full_sitename_fct) +
+  aes(SPstage.Stage,
+      fill = SPstage.Raw_material) +
+  geom_bar() +
+  theme_bw() +
+  labs(fill = "Raw material",
+       x = "Chronological Phase") +
+  theme(legend.position = "top",
+        legend.key.size = unit(0.5, 'cm'), #change legend key size
+        legend.key.height = unit(0.5, 'cm'), #change legend key height
+        legend.key.width = unit(0.5, 'cm'), #change legend key width
+        legend.title = element_text(size=8), #change legend title font size
+        legend.text = element_text(size=6)
+        ) +
+colScale
+
+p2 <- 
+  df_full_sitename_fct %>%
+  mutate(SPstage.Stage = paste0("Phase ", SPstage.Stage)) %>% 
+  group_by(full_sitename) %>% 
+  filter(n() >= 5) %>% 
+    ggplot() +
+    aes(full_sitename,
+        fill = SPstage.Raw_material) +
+    geom_bar() +
+    facet_wrap(~ SPstage.Stage) +
+  theme_bw () +
+  labs(fill = "Raw material",
+       x = "Site") +
+ theme(legend.position = "none",
+       axis.text.x = element_text(size = 5)  ) +
+  scale_x_discrete(guide = guide_axis(n.dodge= 2.3)) +
+   colScale
+
+library(cowplot)
+plot_grid(p1, p2,
+          ncol = 1,
+          rel_heights = c(1.5, 0.75),
+          labels = c("A", "B"))
+```
+
+\newpage
+
 ## Methods
 
-In addition to the metric attributes from previous studies of cultural transmission and projectile points, i.e.
-maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we also used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
+Similar to the metric attributes from previous studies of cultural transmission and projectile points, i.e.
+maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
@@
 The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artefacts, and export them as XY coordinate data for further analysis.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for current and further analyses.
 
@@
 #| label: fig-labelled-schematic
-#| fig-cap: Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks
+#| fig-cap: Stemmed point from Yongsandong site with landmarks showing the attributes considered in this study and further research
 #| out-width: "300px"
@@
 #| label: tbl-landmark-abbreviations
-#| tbl-cap: Description of artefact landmarks used in this study
+#| tbl-cap: Description of artifact landmarks used in this study
 
@@
 \newpage
+
+Following @bettinger1999point we computed Pearson's correlation coefficients for all measurements of the seven attributes that we measured on the stemmed points.
+The correlation of artifact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission.
+This is based on the assumption that assumed that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
+In their work points from Eastern California and Central Nevada, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.
 
@@
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-Eerkens and Bettinger [-@eerkens2001techniques] summarized CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e.
-CV for planned production = 1.7) and about 60 (i.e.
-CV for random production = 57.7) and the values over the upper limit likely reflect  attempts to intentionally differentiate the products.
-Among the Great Basin projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55.
-Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected `r threshold_value` as the threshold to decide between guided variation and indirect bias.
+Based on the Weber Fraction for visual measurement of line lenght, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
+They also presented average CV values (%) for artifacts made from various materials (See Table 1 from @eerkens2001techniques).
+Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22.
+Additionally, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current].
+We sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points.
+Specifically, we calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point).
+With the exception of weight, the CV range for metric attributes is 17.01 to 24.59.
+Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
+The errors are also highly associated with raw materials.
+Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques].
+We chose higher side on the CV range of Rosegate points as the threshold of our research considering the lower quality raw materials of stemmed points and longer duration of their transmission process, which is roughly whole period of the Korean Late Paleolithic (approx.30,000yrs).
 We premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
-<!-- BM: I STILL THIS WE NEED A MORE COMPELLING AND STRONGER JUSTIFICATION FOR THE THRESHOLD VALUE, BASED ON SOME THEORY IN THE LITERATURE.
-PLEASE ADD SOMETHING.
-I EXPECT REVIEWERS WILL GET STUCK ON THIS AND DEMAND MORE DETAILS.
-THE CHOICE OF VALUE HAS A HUGE EFFECT ON OUR INTERPRETATIONS BELOW -->
 
@@
 
-Following @bettinger1999point we computed Pearson's correlation coefficients for all measurements of the seven attributes that we measured on the stemmed points.
-The correlation of artefact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission.
-This is based on the assumption that assumed that artefact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
-In their work points from Eastern California and Central Nevada, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.
-
 All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam].
-Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].
+Our R code and data, including the original artifact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].
 
@@
 Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes).
-Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artefacts is higher.
+Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher.
 Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values.
@@
 The black bold horizontal lines in the middle of boxes represent the median value.
-The dots represent individual artefacts.
+The dots represent individual artifacts.
 B.
@@
   ylab("CV") +
-  xlab("Artefact measurement") +
+  xlab("Artifact measurement") +
   geom_text(aes(label = round(all_cv_sharma,1)), 
@@
 
-@fig-plot-cv-over-time shows that the directional trend in CV values of artefact attributes is complicated.
+@fig-plot-cv-over-time shows that the directional trend in CV values of artifact attributes is complicated.
 Half of the attributes are below our threshold value of `r threshold_value`, and half are above.
-Only body length crosses the threshold value, changing from <`r threshold_value` to >`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
-For those attributes where the CV is <`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
+Only body length crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
+For those attributes where the CV is \<`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
 For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these attributes.
@@
 # stemmed point chronology
-# Exclude the first phase that has only two artefacts
+# Exclude the first phase that has only two artifacts
 
@@
 One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
-These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-four-assemblage).
-These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making.
+These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-four-assemblage).
+These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
 The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
@@
 
-#### plot CVs by site with number of artefacts showing >5
+#### plot CVs by site with number of artifacts showing >5
 
@@
 
-# correlation of all artefacts
+# correlation of all artifacts
 df_sitename_corr <- 
@@
 
-# correlation of phase 2 artefacts
+# correlation of phase 2 artifacts
 df_sitename_corr_p2 <- 
@@
 
-# correlation of phase 3 artefacts
+# correlation of phase 3 artifacts
 df_sitename_corr_p3 <- 
@@
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
-This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artefact.
+This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact.
 These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
@@
 The most striking differences in CV values is evident in comparisons of points grouped by raw materials.
-Artefacts made from quartzite have much higher CV values, and wider confidence intervals, than all other raw materials (@fig-plot-cv-by-raw-material).
+Artifacts made from quartzite have much higher CV values, and wider confidence intervals, than all other raw materials (@fig-plot-cv-by-raw-material).
 This likely reflects the relatively unpredictable knapping quality of quartzite due to the presence of visible grains.
@@
 This makes is harder for the knapper to repeatedly produce the same size and shape end product.
-This hints at the possibility that physical properties of the stone artefact raw materials were a key mechanism in determining shape and size variability.
-However with only three artefacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.
+This hints at the possibility that physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
+However with only three artifacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.
 
@@
 This is indicated by relatively high CV values of around and above 25 and mostly low correlation coefficients of less than 0.5.
-A social context that favoured trial and error for artefact making is consistent with Seong's _in situ_ model for the appearance of stemmed points in Korea.
+A social context that favoured trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
 This model proposes that people developed stemmed points by experimenting with existing blade technologies.
-Our results support experimentation and trial and error as important processes in the cultural transmission of this artefact technology.
+Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.
 
@@
 Does what we say here agree with what we said in the QSR paper?
-GP to edit the conclusion here to show good harmony with our previous paper -->.
+GP to edit the conclusion here to show good harmony with our previous paper -->
+
+.
 

### 2023-01-20 analysis/paper/paper.qmd
re-arranged the results

@@
 
+## Correlation analysis of attributes
+
+```{r}
+#| label: fig-corr-analysis
+#| fig-cap: "A: xxx. B. xxx. C. xxx"
+#| fig-width:   7
+#| fig-height:  9
+
+# BM: I added this section of analysis to replace the PCA sections.
+
+library(corrr)
+
+# correlation of all artifacts
+df_sitename_corr <- 
+  df_sitename %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
+  correlate() %>% 
+  shave()
+
+all_corr_plot <- rplot(df_sitename_corr, legend = FALSE)
+
+df_sitename_corr_long <- 
+  df_sitename_corr %>% 
+    pivot_longer(-term)
+
+# correlation of phase 2 artifacts
+df_sitename_corr_p2 <- 
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage == 2) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
+  correlate() %>% 
+  shave()
+
+p2_corr_plot <- rplot(df_sitename_corr_p2, legend = FALSE)
+
+df_sitename_corr_p2_long <- 
+  df_sitename_corr_p2 %>% 
+    pivot_longer(-term)
+
+# correlation of phase 3 artifacts
+df_sitename_corr_p3 <- 
+  df_sitename %>%
+  dplyr::filter(SPstage.Stage == 3) %>%
+  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
+  correlate() %>% 
+  shave()
+
+p3_corr_plot <-  
+  rplot(df_sitename_corr_p3) +
+   theme(legend.position = c(0.9, 0.9),
+         legend.key.size = unit(0.3, 'cm'), #change legend key size
+        legend.key.height = unit(0.3, 'cm'), #change legend key height
+        legend.key.width = unit(0.3, 'cm'), #change legend key width
+        legend.title = element_text(size=6), #change legend title font size
+        legend.text = element_text(size=8))
+
+df_sitename_corr_p3_long <- 
+  df_sitename_corr_p3 %>% 
+    pivot_longer(-term)
+
+# compare p2 and p3 correlations
+compare_p2_and_p3_correlations_plot <- 
+list(`phase 2` = df_sitename_corr_p2_long,
+     `phase 3` = df_sitename_corr_p3_long) %>% 
+  bind_rows(.id = "phase") %>% 
+  ggplot() +
+  aes(phase,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient")
+
+# correlation by raw material
+df_sitename_corr_raw_material <- 
+  df_sitename %>%
+  group_by(SPstage.Raw_material) %>% 
+  filter(n() >= 3) %>% 
+  dplyr::select(-sitename, 
+                -SPstage.Stage) %>% 
+  nest(-SPstage.Raw_material) %>% 
+  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+                                   shave() %>% 
+                                  pivot_longer(-term))) %>% 
+  unnest(corrr) %>% 
+  filter()
+
+correlations_by_raw_material <- 
+  ggplot(df_sitename_corr_raw_material) +
+  aes(SPstage.Raw_material,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient",
+       x = "Raw material")
+
+# correlation by site
+df_sitename_corr_site <- 
+  df_sitename %>%
+  group_by(sitename) %>% 
+  filter(n() >= 5) %>% 
+  dplyr::select(-sitename, 
+                -SPstage.Stage) %>% 
+  nest(-sitename) %>% 
+  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+                                   shave() %>% 
+                                  pivot_longer(-term))) %>% 
+  unnest(corrr) %>% 
+  filter()
+
+correlations_by_site <- 
+  ggplot(df_sitename_corr_site) +
+  aes(sitename,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient",
+       x = "Site")
+
+library(cowplot)
+plot_grid(plot_grid(
+          plot_grid(p2_corr_plot, 
+                    p3_corr_plot,
+                    ncol = 1,
+                    labels = c("A", "B")),
+          compare_p2_and_p3_correlations_plot,
+          ncol = 2,
+          labels = c("", "C"),
+          rel_heights = c(1, 1.5)),
+          correlations_by_raw_material,
+          correlations_by_site,
+          ncol = 1,
+          labels = c("", "D", "E"))
+
+```
+
 ```{r}
@@
 
-Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
-Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
-
 ## Coefficient of variation
@@
 SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was more influenced by indirect bias than the others.
+
+Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
 
@@
 
-## Correlation analysis of attributes
-
-<!-- BM: You'll see I've completely removed the PCA content.
-I did some more reading into what PCA calculates, and it is not correlation coefficients of the variables with each other, but with the Principal Components that it calculates.
-So that's a very different kind of correlation, compared to what we see in the previous literature on transmission biases.
-I suppose we could use it and interpret it, if you want to investigate that further.
-I could not think of anything straightforward over the several days I have been editing this.
-I saw that you did not give them the same treatment as the CV values with exploring variation by raw material and by site, so I guess you were not very interested in the PCA results.
-My perspective is that we should look for the quickest and simplest way to complete this paper.
-So I have replaced the PCA with the same kind of correlation analysis that we see elsewhere so we can use the previously published papers to guide our interpretation of our results -->
-
-<!-- BM: GP to write up the correlation coefficients results in detail, according to what you see in the plot made by the code below.
-I've already done all the analysis here.
-Your task is to describe it here, and then contextualise and interpret it in the next section.
-Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology".
-Aim for a style of writing that is detailed, direct, specific, short sentences.
--->
-
-```{r}
-#| label: fig-corr-analysis
-#| fig-cap: "A: xxx. B. xxx. C. xxx"
-#| fig-width:   7
-#| fig-height:  9
-
-# BM: I added this section of analysis to replace the PCA sections.
-
-library(corrr)
-
-# correlation of all artifacts
-df_sitename_corr <- 
-  df_sitename %>%
-  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
-  correlate() %>% 
-  shave()
-
-all_corr_plot <- rplot(df_sitename_corr, legend = FALSE)
-
-df_sitename_corr_long <- 
-  df_sitename_corr %>% 
-    pivot_longer(-term)
-
-# correlation of phase 2 artifacts
-df_sitename_corr_p2 <- 
-  df_sitename %>%
-  dplyr::filter(SPstage.Stage == 2) %>%
-  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
-  correlate() %>% 
-  shave()
-
-p2_corr_plot <- rplot(df_sitename_corr_p2, legend = FALSE)
-
-df_sitename_corr_p2_long <- 
-  df_sitename_corr_p2 %>% 
-    pivot_longer(-term)
-
-# correlation of phase 3 artifacts
-df_sitename_corr_p3 <- 
-  df_sitename %>%
-  dplyr::filter(SPstage.Stage == 3) %>%
-  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>% 
-  correlate() %>% 
-  shave()
-
-p3_corr_plot <-  
-  rplot(df_sitename_corr_p3) +
-   theme(legend.position = c(0.9, 0.9),
-         legend.key.size = unit(0.3, 'cm'), #change legend key size
-        legend.key.height = unit(0.3, 'cm'), #change legend key height
-        legend.key.width = unit(0.3, 'cm'), #change legend key width
-        legend.title = element_text(size=6), #change legend title font size
-        legend.text = element_text(size=8))
-
-df_sitename_corr_p3_long <- 
-  df_sitename_corr_p3 %>% 
-    pivot_longer(-term)
-
-# compare p2 and p3 correlations
-compare_p2_and_p3_correlations_plot <- 
-list(`phase 2` = df_sitename_corr_p2_long,
-     `phase 3` = df_sitename_corr_p3_long) %>% 
-  bind_rows(.id = "phase") %>% 
-  ggplot() +
-  aes(phase,
-      value) +
-  geom_boxplot() +
-  ggbeeswarm::geom_quasirandom(size = 2,
-                               alpha = 0.3) +
-  theme_minimal() +
-  labs(y = "Correlation coefficient")
-
-# correlation by raw material
-df_sitename_corr_raw_material <- 
-  df_sitename %>%
-  group_by(SPstage.Raw_material) %>% 
-  filter(n() >= 3) %>% 
-  dplyr::select(-sitename, 
-                -SPstage.Stage) %>% 
-  nest(-SPstage.Raw_material) %>% 
-  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
-                                   shave() %>% 
-                                  pivot_longer(-term))) %>% 
-  unnest(corrr) %>% 
-  filter()
-
-correlations_by_raw_material <- 
-  ggplot(df_sitename_corr_raw_material) +
-  aes(SPstage.Raw_material,
-      value) +
-  geom_boxplot() +
-  ggbeeswarm::geom_quasirandom(size = 2,
-                               alpha = 0.3) +
-  theme_minimal() +
-  labs(y = "Correlation coefficient",
-       x = "Raw material")
-
-# correlation by site
-df_sitename_corr_site <- 
-  df_sitename %>%
-  group_by(sitename) %>% 
-  filter(n() >= 5) %>% 
-  dplyr::select(-sitename, 
-                -SPstage.Stage) %>% 
-  nest(-sitename) %>% 
-  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
-                                   shave() %>% 
-                                  pivot_longer(-term))) %>% 
-  unnest(corrr) %>% 
-  filter()
-
-correlations_by_site <- 
-  ggplot(df_sitename_corr_site) +
-  aes(sitename,
-      value) +
-  geom_boxplot() +
-  ggbeeswarm::geom_quasirandom(size = 2,
-                               alpha = 0.3) +
-  theme_minimal() +
-  labs(y = "Correlation coefficient",
-       x = "Site")
-
-library(cowplot)
-plot_grid(plot_grid(
-          plot_grid(p2_corr_plot, 
-                    p3_corr_plot,
-                    ncol = 1,
-                    labels = c("A", "B")),
-          compare_p2_and_p3_correlations_plot,
-          ncol = 2,
-          labels = c("", "C"),
-          rel_heights = c(1, 1.5)),
-          correlations_by_raw_material,
-          correlations_by_site,
-          ncol = 1,
-          labels = c("", "D", "E"))
-
-```
-
 # Discussion {.unnumbered}

### 2023-01-20 analysis/paper/paper.qmd
update

@@
 
+@fig-corr-analysis
+
 ```{r}
 #| label: fig-corr-analysis
-#| fig-cap: "A: xxx. B. xxx. C. xxx"
+#| fig-cap: "Correlation analysis by chronological phases. A: xxx. B. xxx. C. xxx"
 #| fig-width: 7
-#| fig-height:  9
+#| fig-height: 3
 
@@
 
+plot_grid(plot_grid(
+          plot_grid(p2_corr_plot, 
+                    p3_corr_plot,
+                    ncol = 1,
+                    labels = c("A", "B")),
+          compare_p2_and_p3_correlations_plot,
+          ncol = 2,
+          labels = c("", "C"),
+          rel_heights = c(1, 1.5)))
+
+```
+
+@fig-corr-raw-ass
+
+```{r}
+#| label: fig-corr-raw-ass
+#| fig-cap: A. sth sth B. sth sth.
+
+#| fig-width:7
+#| fig-height:5
+
 # correlation by raw material
@@
 library(cowplot)
-plot_grid(plot_grid(
-          plot_grid(p2_corr_plot, 
-                    p3_corr_plot,
-                    ncol = 1,
-                    labels = c("A", "B")),
-          compare_p2_and_p3_correlations_plot,
-          ncol = 2,
-          labels = c("", "C"),
-          rel_heights = c(1, 1.5)),
-          correlations_by_raw_material,
+plot_grid(correlations_by_raw_material,
           correlations_by_site,
                     ncol = 1,
-          labels = c("", "D", "E"))
-
-```
+                    labels = c("A", "B"))
+
+```
+
+## Coefficient of variation
 
@@
 
-## Coefficient of variation
-
 <!-- BM: overall I think the CV values are really not that interesting or informative.
-They are difficult to interpret because they all cluster around 15-30, which seems to me to be the midpoint between 'highly standardised' (3-5) and 'highly different' (40-50).
+They are difficult to interpret because they all cluster around 15-30, which seems to me to be the midpoint between 'highly standardized' (3-5) and 'highly different' (40-50).
 So I think we will struggle to convince any readers that we have any strong signals of cultural transmission biases in the CV data.

### 2023-01-25 analysis/paper/paper.qmd
commit to merge

@@
   theme_minimal() +
-  labs(y = "Correlation coefficient")
+  labs(y = "Correlation coefficient") +
+theme(axis.title.y = element_text(size = 13))
 
@@
   labs(y = "Correlation coefficient",
-       x = "Raw material")
+       x = "Raw material") +
+theme(axis.title.y = element_text(size = 7.5))
 
@@
   labs(y = "Correlation coefficient",
-       x = "Site")
+       x = "Assemblage") +
+theme(axis.title.y = element_text(size = 7.5))
 

### 2023-01-25 analysis/paper/paper.qmd
simulation for comparing cv and mccv results

@@
 
-# use this one that is optimized for small samples, returns value 0-1, not %
+# use this one that is optimized for small samples, returns value 0-1, not %. 
+# We multiply 100 (%) later, altogether with confident interval.
 
@@
 
+# GP's test for comparing CV vs corrected CV
+# mock_test <-
+#   readr::read_csv(here::here("analysis/data/raw_data/mock_test.csv"))
+# mock_test_cv <-map_dbl(mock_test, cv)
+# mock_test_mccv <- map_dbl(mock_test, mccv)
+
 #Same as map but better view for summary
@@
 # CV value
-sharma_cv <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100
+sharma_cv <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate *100
 
 #CV sharma interval values
-sharma_int_low <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1] * 100
-
-sharma_int_high <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2] * 100
+sharma_int_low <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1] *100
+
+sharma_int_high <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2] *100
 

### 2023-01-26 analysis/paper/paper.qmd
Added another plot for correlation

@@
 We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial part of a successful model to ensure the quality to optimize tool usage.
-As a result, the shape of stemmed points became standardized among their social groups.
+As a result, the shape of stemmed points became more standardized among their social groups.
 
@@
 
+# how many assemblages do we use?
+stemmed_points_ass <- df_sitename %>%
+  unite(col = "site_phase",
+        c(sitename,
+          SPstage.Stage),
+        sep = "_") %>%
+  dplyr::select(-SPstage.Raw_material) %>%
+  dplyr::count(site_phase)
+
 # prepare some values to use in the text below
@@
 total_number_of_sites <-nrow(stemmed_points_per_site)
-total_number_of_assemblages <- NULL # ADD CODE HERE
+total_number_of_assemblages <- nrow(stemmed_points_ass)
 ```
@@
 We excluded artifacts that were recorded as stemmed points but lack a stem.
-This resulted in a sample of `r total_number_of_sp` stemmed points from 28 assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number).
+This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number).
 The images of the stemmed points were obtained from published excavation reports and by direct photography during our research on local museum collections.
@@
 These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
-Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
-Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations.
+Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
+Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to various research limitations.
 Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
@@
 #| label: tbl-korean-chronology
-#| tbl-cap: Korean Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]
+#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]
 
@@
 Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases.
-We exclude phase 1 from our analyses below because of the small sample size.
+We exclude phase 1 from our analyses below because of the small sample size to compute correlation and CV.
 Shale is the dominant raw material in both phase 2 and 3.
-In phase 3 there is an increase in the proportions of rhyolite and porphyry, driven largely by finds from Wolpyeng and Yongsangdong.
+In phase 3 there is an increase in the proportions of porphyry, driven largely by finds from Yongsangdong, and other raw materials.
 Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at assemblages with more than five stemmed points.
@@
   group_by(full_sitename) %>% 
-  filter(n() >= 5) %>% 
+  filter(n() > 5) %>% 
     ggplot() +
@@
 # to ensure we don't have any copy-paste errors 
+# Correlation threshold
+threshold_value_cor <- 0.5
+
+# CV threshold
 threshold_value <- 25
@@
 
-Following @bettinger1999point we computed Pearson's correlation coefficients for all measurements of the seven attributes that we measured on the stemmed points.
+Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points.
 The correlation of artifact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission.
-This is based on the assumption that assumed that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
-In their work points from Eastern California and Central Nevada, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.
-
-To measure the variation among the measurements of the stemmed points, we computed Coefficients of Variation (CV) with confidence intervals.
-The CV is the ratio of the standard deviation to the mean.
+This is based on the assumption that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
+In their work, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.
+In this research, we examine correlation between attributes and explore other variables such as different phases, raw materials and assemblages.
+
+We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points.
+The CV is the ratio (%) of the standard deviation to the mean.
 It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc.
@@
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-Based on the Weber Fraction for visual measurement of line lenght, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
-They also presented average CV values (%) for artifacts made from various materials (See Table 1 from @eerkens2001techniques).
-Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22.
-Additionally, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current].
-We sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points.
-Specifically, we calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point).
-With the exception of weight, the CV range for metric attributes is 17.01 to 24.59.
-Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
-The errors are also highly associated with raw materials.
-Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques].
-We chose higher side on the CV range of Rosegate points as the threshold of our research considering the lower quality raw materials of stemmed points and longer duration of their transmission process, which is roughly whole period of the Korean Late Paleolithic (approx.30,000yrs).
-We premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
+Based on the Weber Fraction for visual measurement of line length, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
 
@@
 
+-@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
+Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22.
+Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current].
+Unlike correlation, @bettinger1999point had vaguely mention about the threshold for CV values.
+Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points.
+We calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point).
+With the exception of weight, the CV range for metric attributes on the points is 17.01 to 24.59.
+We chose higher side on the CV range of Rosegate points as the threshold of our research based on following reasons.
+1) Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
+The duration of transmission process for stemmed points is much longer (approx.30,000yrs) than the case of Rosegate (approx.500yrs in Fremont region) [@bischoff2021rosegate].
+2) The errors are also highly associated with raw materials.
+Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques].
+We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to raw materials for stemmed points such as shale, rhyolite, porphyry, etc.
+(@fig-raw-materials-by-phase).
+3) Compared to normal CV, corrected CV has slightly higher values [@vanpool2011quantitative].
+Based on the threshold, we premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
+
 All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam].
@@
 
-## Correlation analysis of attributes
-
-@fig-corr-analysis
+## Correlation coefficients over time
+
+Our results show positive correlation between attributes.
+Panel A and B of @fig-corr-analysis indicate that some attributes such as BL and ML have stronger correlation than others.
+The correlations between attributes became stronger over time as we can clearly see in the Panel C of @fig-corr-analysis.
+The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases.
+Therefore, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
+In Phase 3, the median is close to `r threshold_value_cor`, which means the Phase 3 could be more associated with indirect bias.
 
@@
 #| label: fig-corr-analysis
-#| fig-cap: "Correlation analysis by chronological phases. A: xxx. B. xxx. C. xxx"
+#| fig-cap: "Correlation analysis by chronological phases in the Korean Late Paleolithic period.
+A: Correlation coefficient between attributes for Phase 1.
+B.
+Correlation coefficient between attributes for Phase 2.
+C.
+Correlation coefficient for the second and third chronological phases.
+The grey points represent correlation between two attributes.
+"
 #| fig-width: 7
 #| fig-height: 3
-
-# BM: I added this section of analysis to replace the PCA sections.
 
@@
   labs(y = "Correlation coefficient") +
-theme(axis.title.y = element_text(size = 13))
+theme(axis.title.y = element_text(size = 13)) +
+  # draw a line to show our threshold value
+  geom_hline(yintercept = threshold_value_cor, colour = "red") +
+  theme_bw()
 
@@
 
+```{r}
+#| label: fig-plot-cor-four-assemblage
+#| fig-cap: Correlation for assemblages with more than 5 stemmed points.
+Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
+#| fig-width:   4.5
+#| fig-height:  4
+
+#### plot cors by site with number of artifacts showing >5
+
+cor_by_site_df_label_more_than_5 <-
+  df_sitename %>%
+  unite(col = "site_phase",
+        c(sitename,
+          SPstage.Stage),
+        sep = "_") %>%
+  dplyr::select(-SPstage.Raw_material) %>%
+  dplyr::group_by(site_phase) %>%
+  add_tally() %>%
+  dplyr::filter( n > 5) %>%
+  mutate(label = as.factor(paste0(site_phase,' (N = ', n, ")"))) %>%
+  nest(-site_phase) %>%
+  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+                                   shave() %>% 
+                                  pivot_longer(-term))) %>% 
+  unnest(corrr) %>% 
+  filter()
+
+#better arrangement for x axis for the plot
+level_order <- c('SYG1_2', 'SYG6_2', 'SYG1_3', 'YS_3')
+
+  ggplot(cor_by_site_df_label_more_than_5) +
+  aes(x=factor(site_phase, level=level_order), y=value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient",
+       x = "Assemblage") +
+theme(axis.title.y = element_text(size = 7.5)) +
+  # draw a line to show our threshold value
+  geom_hline(yintercept = threshold_value_cor, colour = "red") +
+  theme_bw()
+
+```
+
+## Correlation between sites and raw materials
+
 @fig-corr-raw-ass
@@
 #| fig-height:5
+
+# correlation by site
+df_sitename_corr_site <- 
+  df_sitename %>%
+  group_by(sitename) %>% 
+  filter(n() > 5) %>% 
+  dplyr::select(-sitename, 
+                -SPstage.Stage) %>% 
+  nest(-sitename) %>% 
+  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+                                   shave() %>% 
+                                  pivot_longer(-term))) %>% 
+  unnest(corrr) %>% 
+  filter()
+
+correlations_by_site <- 
+  ggplot(df_sitename_corr_site) +
+  aes(sitename,
+      value) +
+  geom_boxplot() +
+  ggbeeswarm::geom_quasirandom(size = 2,
+                               alpha = 0.3) +
+  theme_minimal() +
+  labs(y = "Correlation coefficient",
+       x = "Site") +
+theme(axis.title.y = element_text(size = 7.5)) +
+  # draw a line to show our threshold value
+  geom_hline(yintercept = threshold_value_cor, colour = "red") +
+  theme_bw()
 
@@
   group_by(SPstage.Raw_material) %>% 
-  filter(n() >= 3) %>% 
+  filter(n() > 2) %>% 
   dplyr::select(-sitename, 
@@
        x = "Raw material") +
-theme(axis.title.y = element_text(size = 7.5))
-
-# correlation by site
-df_sitename_corr_site <- 
-  df_sitename %>%
-  group_by(sitename) %>% 
-  filter(n() >= 5) %>% 
-  dplyr::select(-sitename, 
-                -SPstage.Stage) %>% 
-  nest(-sitename) %>% 
-  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
-                                   shave() %>% 
-                                  pivot_longer(-term))) %>% 
-  unnest(corrr) %>% 
-  filter()
-
-correlations_by_site <- 
-  ggplot(df_sitename_corr_site) +
-  aes(sitename,
-      value) +
-  geom_boxplot() +
-  ggbeeswarm::geom_quasirandom(size = 2,
-                               alpha = 0.3) +
-  theme_minimal() +
-  labs(y = "Correlation coefficient",
-       x = "Assemblage") +
-theme(axis.title.y = element_text(size = 7.5))
+theme(axis.title.y = element_text(size = 7.5)) +
+  # draw a line to show our threshold value
+  geom_hline(yintercept = threshold_value_cor, colour = "red") +
+  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  theme_bw()
 
 library(cowplot)
-plot_grid(correlations_by_raw_material,
-          correlations_by_site,
+plot_grid(correlations_by_site,
+          correlations_by_raw_material,
                     ncol = 1,

### 2023-01-26 analysis/paper/paper.qmd
edit correlation plots
render and then minor change

@@
 # CV value
-sharma_cv <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate *100
+sharma_cv <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate * 100
 
 #CV sharma interval values
-sharma_int_low <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1] *100
-
-sharma_int_high <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2] *100
+sharma_int_low <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1] * 100
+
+sharma_int_high <- function(x,...) sharma_cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2] * 100
 

### 2023-02-07 analysis/paper/paper.qmd
Edit correlation part in results

@@
 
-Our results show positive correlation between attributes.
-Panel A and B of @fig-corr-analysis indicate that some attributes such as BL and ML have stronger correlation than others.
-The correlations between attributes became stronger over time as we can clearly see in the Panel C of @fig-corr-analysis.
-The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases.
-Therefore, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
-In Phase 3, the median is close to `r threshold_value_cor`, which means the Phase 3 could be more associated with indirect bias.
+To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarized in @tbl-korean-chronology, and computed correlation for all complete stemmed points each phase.
+Among the three choronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.
+
+Our results show positive correlation between attributes throughout the Late Paleolithic.
+Panel A and B of @fig-corr-analysis indicate that some attributes such as BL and ML have stronger correlation (i.e.
+darker-blue points) than others (i.e.
+lighter-blue or almost invisible points).
+The correlations between attributes became stronger over time as we can clearly see in the the Panel C of @fig-corr-analysis.
+The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold.
+Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
+On the other hand, Phase 3 could be more associated with indirect bias.
 
@@
 Correlation coefficient between attributes for Phase 2.
+The individual point represesnts correlation between two attributes.
+Some points are invisible due to their weak relationship.
 C.
@@
 
+One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
+These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase
+
 ```{r}
@@
 ## Variation Over Time
-
-To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarised in @tbl-korean-chronology, and computed CV values for all complete stemmed points each phase.
-We excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.
 

### 2023-02-08 analysis/paper/paper.qmd
Before deleting "variation between sites"

@@
 
--@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
+@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
 Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22.
 Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current].
-Unlike correlation, @bettinger1999point had vaguely mention about the threshold for CV values.
+Unlike correlation, @bettinger1999point had vaguely mentioned about the threshold for CV values.
 Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points.
@@
 
-## Correlation coefficients over time
-
-To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarized in @tbl-korean-chronology, and computed correlation for all complete stemmed points each phase.
-Among the three choronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.
-
-Our results show positive correlation between attributes throughout the Late Paleolithic.
-Panel A and B of @fig-corr-analysis indicate that some attributes such as BL and ML have stronger correlation (i.e.
+## Correlation coefficients and Variation Over Time
+
+Our results show positive correlation between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis).
+Some relationships such as BL and ML have stronger correlation (i.e.
 darker-blue points) than others (i.e.
 lighter-blue or almost invisible points).
-The correlations between attributes became stronger over time as we can clearly see in the the Panel C of @fig-corr-analysis.
+To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarized in @tbl-korean-chronology.
+Among the three chronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.
+
+The correlations between attributes became stronger over time as we can clearly see in the the Panel D of @fig-corr-analysis.
 The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold.
@@
 #| fig-cap: "Correlation analysis by chronological phases in the Korean Late Paleolithic period.
-A: Correlation coefficient between attributes for Phase 1.
+A.
+Correlation coefficient between attributes for all periods.
 B.
+Correlation coefficient between attributes for Phase 1.
+C.
 Correlation coefficient between attributes for Phase 2.
@@
 Some points are invisible due to their weak relationship.
-C.
+D.
 Correlation coefficient for the second and third chronological phases.
-The grey points represent correlation between two attributes.
-"
+The grey points represent correlation between two attributes."
 #| fig-width: 7
-#| fig-height: 3
+#| fig-height: 5
 
@@
 plot_grid(plot_grid(
-          plot_grid(p2_corr_plot, 
+          plot_grid(
+                    all_corr_plot,        
+                    p2_corr_plot, 
                     p3_corr_plot,
                     ncol = 1,
-                    labels = c("A", "B")),
+                    labels = c("A", "B", "C")),
           compare_p2_and_p3_correlations_plot,
           ncol = 2,
-          labels = c("", "C"),
-          rel_heights = c(1, 1.5)))
-
-```
+          labels = c("", "D"),
+          rel_heights = c(1, 2.5)))
+
+```
+
+One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
+These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
+We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
+The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
+There are more variation in correlation coefficient in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
+Even though SYG1_2 is from Phase 2 but it has the higher median than other assemblages.
 
@@
 Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
+The grey points represent correlation between two attributes.
 #| fig-width:   4.5
-#| fig-height:  4
+#| fig-height:  3.5
 
@@
 
-One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
-These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase
-
 ```{r}
 #| label: fig-corr-raw-ass
-#| fig-cap: A. sth sth B. sth sth.
+#| fig-cap: A.
+Correlation for each site that has more than five stemmed points.
+B.
+Correlation for raw materials that were used for making more than one stemmed poitns.
+The grey points represent correlation between two attributes.
 
@@
        x = "Site") +
-theme(axis.title.y = element_text(size = 7.5)) +
+theme(axis.title.y = element_text(size = 4)) +
   # draw a line to show our threshold value
@@
        x = "Raw material") +
-theme(axis.title.y = element_text(size = 7.5)) +
+theme(axis.title.y = element_text(size = 4)) +
   # draw a line to show our threshold value
@@
 
-## Coefficient of variation
+## Coefficient of variation and Variation over time
 
@@
 
-## Variation Over Time
-
 @fig-plot-cv-over-time shows that the directional trend in CV values of artifact attributes is complicated.
@@
 
-One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
-These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-four-assemblage).
-These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
-The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
-SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure).
 Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width.

### 2023-02-08 analysis/paper/paper.qmd
editing correlation

@@
 ```
+## Correlation between assemblages
 
@@
 These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
-We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
-The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
+We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
 There are more variation in correlation coefficient in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
-Even though SYG1_2 is from Phase 2 but it has the higher median than other assemblages.
+There is no clear pattern between assemblages.
+SYG1_2 is from Phase 2 but it has the higher median than the assemblage from the same site in the later phase.
 
@@
 
-## Correlation between sites and raw materials
-
-@fig-corr-raw-ass
+## Correlation between raw materials
+
+To examine the impact of raw materials on stemmed points, we computed the correlation for different raw materials.
+We excluded raw materials that were used for less than three points.
+@fig-corr-raw-ass shows that stemmed points made out of acidic volcanic rocks are highly correlated between most of the attributes but some of the attributes are negatively related.
+Correlation values for porphyry stemmed points have a narrow distribution.
+Rhyolite stemmed points have widely distributed correlations.
+Some of the attributes are negatively related.
+Like acidic volcanic rocks, the correlation of shale stemmed points is also skewed into two groups.
+Therefore, we assume that the correlation varies depending on raw materials.
 
@@
 #| label: fig-corr-raw-ass
-#| fig-cap: A.
-Correlation for each site that has more than five stemmed points.
-B.
-Correlation for raw materials that were used for making more than one stemmed poitns.
-The grey points represent correlation between two attributes.
+#| fig-cap: Correlation for raw materials that were used for making more than one stemmed poitns. The grey points represent correlation between two attributes.
 
@@
 # correlation by site
-df_sitename_corr_site <- 
-  df_sitename %>%
-  group_by(sitename) %>% 
-  filter(n() > 5) %>% 
-  dplyr::select(-sitename, 
-                -SPstage.Stage) %>% 
-  nest(-sitename) %>% 
-  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
-                                   shave() %>% 
-                                  pivot_longer(-term))) %>% 
-  unnest(corrr) %>% 
-  filter()
-
-correlations_by_site <- 
-  ggplot(df_sitename_corr_site) +
-  aes(sitename,
-      value) +
-  geom_boxplot() +
-  ggbeeswarm::geom_quasirandom(size = 2,
-                               alpha = 0.3) +
-  theme_minimal() +
-  labs(y = "Correlation coefficient",
-       x = "Site") +
-theme(axis.title.y = element_text(size = 4)) +
-  # draw a line to show our threshold value
-  geom_hline(yintercept = threshold_value_cor, colour = "red") +
-  theme_bw()
+# df_sitename_corr_site <- 
+#   df_sitename %>%
+#   group_by(sitename) %>% 
+#   filter(n() > 5) %>% 
+#   dplyr::select(-sitename, 
+#                 -SPstage.Stage) %>% 
+#   nest(-sitename) %>% 
+#   mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+#                                    shave() %>% 
+#                                   pivot_longer(-term))) %>% 
+#   unnest(corrr) %>% 
+#   filter()
+# 
+# correlations_by_site <- 
+#   ggplot(df_sitename_corr_site) +
+#   aes(sitename,
+#       value) +
+#   geom_boxplot() +
+#   ggbeeswarm::geom_quasirandom(size = 2,
+#                                alpha = 0.3) +
+#   theme_minimal() +
+#   labs(y = "Correlation coefficient",
+#        x = "Site") +
+# theme(axis.title.y = element_text(size = 4)) +
+#   # draw a line to show our threshold value
+#   geom_hline(yintercept = threshold_value_cor, colour = "red") +
+#   theme_bw()
 
@@
   group_by(SPstage.Raw_material) %>% 
-  filter(n() > 2) %>% 
+  filter(n() > 3) %>% 
   dplyr::select(-sitename, 
@@
 
-correlations_by_raw_material <- 
+# correlations_by_raw_material <- 
   ggplot(df_sitename_corr_raw_material) +
@@
   geom_hline(yintercept = threshold_value_cor, colour = "red") +
-  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  # scale_x_discrete(guide = guide_axis(n.dodge = 2))+
   theme_bw()
-
-library(cowplot)
-plot_grid(correlations_by_site,
-          correlations_by_raw_material,
-                    ncol = 1,
-                    labels = c("A", "B"))
 
@@
 
-<!-- BM: overall I think the CV values are really not that interesting or informative.
-They are difficult to interpret because they all cluster around 15-30, which seems to me to be the midpoint between 'highly standardized' (3-5) and 'highly different' (40-50).
-So I think we will struggle to convince any readers that we have any strong signals of cultural transmission biases in the CV data.
-Probably we can satisfy most readers that guided variation is the main bias in these data.
-That's my perspective after reading this paper several times, and re-reading several related papers by Bettiner and Eerkens and Lipo, and others  -->
+Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
 
@@
 SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was more influenced by indirect bias than the others.
-
-Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
-Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
 
@@
 For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these attributes.
-
-\newpage
 

### 2023-02-08 analysis/paper/paper.qmd
Done with results!

@@
 To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV).
-We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias over time.
+We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias.
 Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation.
@@
 
-## Correlation coefficients and Variation Over Time
+## Correlation coefficients
 
@@
 
-The correlations between attributes became stronger over time as we can clearly see in the the Panel D of @fig-corr-analysis.
+The correlations between attributes became stronger over time as we can clearly see in the Panel D of @fig-corr-analysis.
 The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold.
@@
 ```
-## Correlation between assemblages
 
@@
 
-## Coefficient of variation and Variation over time
+## Coefficient of variation
 
@@
 
-Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes).
+Panel B of @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher.
@@
 Tang and stem related attributes (e.g.
-SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was more influenced by indirect bias than the others.
+SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes.
+Overall, our results suggests that transmission of stemmed points could be more influenced by guided variation.
 
@@
 
+Similar to Panel D of @fig-corr-analysis, we divided the stemmed points into three chronological phases and computed CV to examine temporal patterns in the modes of cultural transmission.
+Again, we excluded stemmed points for Phase 1 due to the limited number of artifacts.
+
 @fig-plot-cv-over-time shows that the directional trend in CV values of artifact attributes is complicated.
@@
 
-Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width.
+Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width for the assemblages that have more than five stemmed points.
 @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`.
-Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here.
+Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals.
+It is interesting because SYG1_2 has the strong correlation between attributes (@fig-plot-cor-four-assemblage).
+Perhaps it could be related to its small sample size, relative to the other assemblages compared here.
 Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
@@
 
-## Variation Between sites
-
-We examined the CV values per site to address our research question about the cultural transmission over space.
-We excluded assemblages that contain less than five points.
-@fig-plot-cv-by-site shows higher CV values for stem length and stem width, indicating high variability at the stem of the points from these sites.
-Overall size variability is similar between these three assemblages, suggesting a consistent approach to constraining shape variation over space, rather than site-specific variances.
-
-```{r}
-#| label: fig-plot-cv-by-site
-#| fig-cap: CV values per site for each site has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
-#| fig-width:   8.5
-#| fig-height:  4
-
-cv_by_site_df  <-
-  df_full_sitename %>%
-  dplyr::select(-SPstage.Stage, 
-                -lat_dd, 
-                -long_dd, 
-                -sitename, 
-                -SPstage.Raw_material) %>%
-  group_by(full_sitename) %>%
-  add_tally() %>%
-  dplyr::filter( n > 5) %>%
-  dplyr::select(-n) %>%
-  pivot_longer(-full_sitename,
-               names_to = "variable",
-               values_to = "value") %>%
-  group_by(full_sitename, variable) %>%
-  dplyr::summarise(cv_by_site =   sharma_cv(value),
-            cv_low_sharma =  sharma_int_low(value),
-            cv_high_sharma = sharma_int_high(value))
-
-## CV for per each site (n=X) with the full site name : label
-
-cv_by_full_site_df_label <-
-  df_full_sitename %>%
-  dplyr::select(-SPstage.Stage) %>%
-  group_by(full_sitename) %>%
-  add_tally() %>%
-  dplyr::filter( n > 1) %>%
-  mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
-  dplyr::select(-n) %>%
-  nest(-full_sitename, -label) %>%
-  dplyr::mutate(cv_by_site = purrr::map(data, ~map_df(.x, mccv))) %>%
-  dplyr::select(-cv_by_site)
-
-## add label to main dataframe
-
-cv_plot_site_label <- cv_by_site_df %>%
-  left_join(cv_by_full_site_df_label) %>%
-  dplyr::select(- data,-full_sitename)
-
-## facet plot
-
-cv_plot_site_label %>%
-  ggplot(aes(x = variable,
-             y = cv_by_site)) +
-  geom_point() +
-  geom_errorbar(aes(ymin = cv_low_sharma,
-                    ymax = cv_high_sharma),
-                width = 0.2) +
-  xlab("Attribute") +
-  ylab("Coefficient of Variation on Attributes") +
-    # draw a line to show our threshold value
-  geom_hline(yintercept = 25, colour = "red") +
-  facet_wrap( ~ label, scales = "free_y") +
-  theme(legend.position = "none") +
-  coord_cartesian(ylim = c(0,70)) +
-  theme_bw() +
-  theme(strip.text.x = element_text(size = 6.5))
-```
-
 ## Variation Between Raw Materials
@@
 
-# Discussion {.unnumbered}
+# Discussion
 

### 2023-02-09 analysis/paper/paper.qmd
brainstorming for discussion

@@
 This is based on the assumption that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
-In their work, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.
+In their work, @bettinger1999point interpreted correlation coefficient values of around `r threshold_value_cor` and higher as evidence of indirect bias, and lower values as evidence of guided variation.
 In this research, we examine correlation between attributes and explore other variables such as different phases, raw materials and assemblages.
@@
 ```
-
 # Discussion
@@
 We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
-We set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
-If the CV value is lower than 25, we interpreted this as indicating indirect bias while values higher than 25 were interpreted as guided variation.
-
-The CV values for all stemmed point attributes are mostly over 25, except for maximum length, which is just below.
+Following @bettinger1999point, we used `r threshold_value_cor` as a threshold of correlation coefficients to evaluate the two transmission biases.
+Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias.
+Since there is no clear criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
+If the CV value is lower than `r threshold_value`, we interpreted this as indicating indirect bias while values higher than `r threshold_value` were interpreted as guided variation.
+
+Our correlation analyses show that the correlation coefficients of stemmed points are all positive but mostly under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias.
+Some correlations such as body length and maximum length have strong relationship while other correlations such as tang length and body length are much weaker.
+The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but there is no clear pattern and the values vary depending on assemblages (@fig-plot-cor-four-assemblage).
+The CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below.
 Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time).
 Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases.
-Applying the Modified Signed-Likelihood Ratio Test (SLRT) <!-- GP to add a citation for this test here --> to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
+Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
 In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).
@@
 ```
-
-In exploring geographical variation we found no clear pattern variation in CV between sites with more than five stemmed points (@fig-plot-cv-by-site).
-The most striking pattern in this site comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
+In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
+The four assemblages show different distribution patterns.
+
+-SYG1_2 and SYG1_3 are from the same site but different time periods and the correlations decreases over time. SYG1_2 is from earlier period but having the highest median, 
+-two groups of correlations in SYG1_2
+-
+
+The most striking pattern in this assemblage comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
 The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations.
@@
 
-<!-- BM: GP to write up the discussion of the correlation analysis in detail.
-Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology".
-Aim to improve your writing, make it better than your usual style.
--->
-
 # Conclusion {.unnumbered}

### 2023-02-10 analysis/paper/paper.qmd
finish editing the discussion

@@
 
-To examine the impact of raw materials on stemmed points, we computed the correlation for different raw materials.
+To examine the impact of raw materials on the shape of stemmed points, we computed the correlation for different raw materials.
 We excluded raw materials that were used for less than three points.
-@fig-corr-raw-ass shows that stemmed points made out of acidic volcanic rocks are highly correlated between most of the attributes but some of the attributes are negatively related.
+@fig-corr-raw-ass shows that the distribution of correlation coefficients is various depending on raw materials.
+Overall, the materials that used for more and equal to three points have high correlation coefficients.
+The stemmed points made from quartzite have highest correlation coefficients with narrow distribution.
+Acidic volcanic rocks are one of the highly correlated materials and individual correlations are skewed into two groups.
+Some of the correlations are even negatively related.
 Correlation values for porphyry stemmed points have a narrow distribution.
-Rhyolite stemmed points have widely distributed correlations.
-Some of the attributes are negatively related.
+Rhyolite stemmed points have wide distribution and some of the attributes are negatively related.
 Like acidic volcanic rocks, the correlation of shale stemmed points is also skewed into two groups.
-Therefore, we assume that the correlation varies depending on raw materials.
+Tuff stemmed points have wide distribution with higher median value.
 
@@
   group_by(SPstage.Raw_material) %>% 
-  filter(n() > 3) %>% 
+  filter(n() >= 3) %>% 
   dplyr::select(-sitename, 
@@
   geom_hline(yintercept = threshold_value_cor, colour = "red") +
-  # scale_x_discrete(guide = guide_axis(n.dodge = 2))+
+  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
   theme_bw()
@@
 In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
-The four assemblages show different distribution patterns.
-
--SYG1_2 and SYG1_3 are from the same site but different time periods and the correlations decreases over time. SYG1_2 is from earlier period but having the highest median, 
--two groups of correlations in SYG1_2
--
-
-The most striking pattern in this assemblage comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
+We found that each assemblage has different distribution patterns of correlation coefficients and CV values.
+SYG1_2 and SYG1_3 are from the same site but different chronological periods and correlations decreases over time.
+It is likely that correlation or overall shape can be more depending on individual knappers.
+We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation.
+We assume that certain attributes were more carefully transmitted than others.
+We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
 The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations.
@@
 
-The most striking differences in CV values is evident in comparisons of points grouped by raw materials.
-Artifacts made from quartzite have much higher CV values, and wider confidence intervals, than all other raw materials (@fig-plot-cv-by-raw-material).
-This likely reflects the relatively unpredictable knapping quality of quartzite due to the presence of visible grains.
+The most striking differences in both correlation coefficients and CV values is 
+in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
+All distributions are different and there is no clear pattern, indicating that raw materials strongly influence stemmed points.
+We expected that stemmed points with higher correlation coefficients would have lower CV values, but the actual results do not match our expectations.
+For example, artifacts made from acidic quartzite have highest correlation but their CV values are also much higher with wider confidence intervals, than all other raw materials.
+This conflicting results might be related to the relatively unpredictable knapping quality of quartzite.
 These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals.
@@
 
-# Conclusion {.unnumbered}
+# Conclusion
 
@@
 We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error).
-This is indicated by relatively high CV values of around and above 25 and mostly low correlation coefficients of less than 0.5.
+This is indicated by relatively high CV values of around and above `r threshold_value` and mostly low correlation coefficients of less than `r threshold_value_cor`.
+
 A social context that favoured trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
@@
 Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Wolpyeng and Yonsangdong that date to this period.
-One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favoured an expanded range for searching for raw materials, leading to a higher diversity.
+One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
 Similarly, higher mobility may have resulted in more frequent contacts and connections with members of social networks.
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.
-
-<!-- BM: GP to check where we can cite our QSR paper in the conclusion.
-Does what we say here agree with what we said in the QSR paper?
-GP to edit the conclusion here to show good harmony with our previous paper -->
-
-.
 

### 2023-02-10 analysis/paper/paper.qmd
finish editing conclusion

@@
 We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error).
-This is indicated by relatively high CV values of around and above `r threshold_value` and mostly low correlation coefficients of less than `r threshold_value_cor`.
-
-A social context that favoured trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
+This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`.
+A social context that favored trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
 This model proposes that people developed stemmed points by experimenting with existing blade technologies.
@@
 
-The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values at 0.5 and higher in chronological phase 3, after 25 ka.
+The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological phase 3, after 25 ka.
 This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
-This may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model.
-We might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure <!-- GP to find something relevant by Fitzhugh to cite here -->.
+We also confirm a slight decrease in CV values in four attributes.
+This results may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model.
+We might speculate based on the evolutionary ecological assumption that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [@fitzhugh2001risk].
 With more frequent contacts and connections with members of the social network, there may have been more opportunities to learn point-making directly from socially successful individuals.
 
-In comparing sites with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
-We did not find strong support for distinctive forms of cultural transmission of tool-making at different sites.
-Correlation coefficient values across the sites affirm the overall chronological trend indicating indirect bias in phase three.
-Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Wolpyeng and Yonsangdong that date to this period.
+In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
+We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages.
+Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in phase three.
+Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Yonsangdong that date to this period.
 One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
-Similarly, higher mobility may have resulted in more frequent contacts and connections with members of social networks.
+Similarly, change in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.

### 2023-02-13 analysis/paper/paper.qmd
edited up to results

@@
 Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation.
-Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation and then indirect bias was adapted to carry out the technological change.
-We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial part of a successful model to ensure the quality to optimize tool usage.
+Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation <!-- and then indirect bias was adapted to carry out the technological change BM: this is unclear -->.
+We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage.
 As a result, the shape of stemmed points became more standardized among their social groups.
@@
 
-Previous studies of the technological transitions during the Late Paleolithic in Korea have mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia.
+Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points, as part of the discussion of the timing and routes of modern human dispersal in eastern Asia.
 The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
@@
 
-In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
+In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins outside of the Korean peninsula [@bae2010origin; @bae2012nature; @bae2013early].
 While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
-Specifically, the blade technology in claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
+Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
 The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome].
@@
 
-Cultural transmission theory holds that information about how to behave is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
+Cultural transmission theory holds that information about social and technological behaviours is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
 Individuals learn by themselves (e.g.
@@
 Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
-Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
+Garvey measured CV values for weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
 She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e.
@@
 We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness.
-On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.
+On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.
 
@@
 Seong's *in situ* model).
-This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points.
+If this social context was prevalent, we predict this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points.
 If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.
@@
 Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
-In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package (cf.
-Bae's migration model).
+In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package.
+This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf.
+Bae's migration model), where stemmed points first appeared and then spread from.
 In this social context, indirect bias is the dominant influence on transmission of lithic technology.
@@
 To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
-Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
-Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to various research limitations.
-Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
+These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition were used to assign assemblages to a phase.
+Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blade in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
+Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations.
+Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size.
 We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.
@@
 #| label: tbl-korean-chronology
-#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]
+#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]. SP = stemmed point
 
@@
 Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases.
-We exclude phase 1 from our analyses below because of the small sample size to compute correlation and CV.
+We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs.
 Shale is the dominant raw material in both phase 2 and 3.
@@
 
-Similar to the metric attributes from previous studies of cultural transmission and projectile points, i.e.
+Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e.
 maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
-Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes than traditional caliper measurements [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
 We did not include weight and thickness as these measurements were not available to us.
@@
 The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for current and further analyses.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for further analyses.
 
@@
 We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points.
-The CV is the ratio (%) of the standard deviation to the mean.
+The CV is the ratio (usually expressed as a percentage) of the standard deviation to the mean.
 It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc.
-in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence].
+in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization].
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-Based on the Weber Fraction for visual measurement of line length, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
+As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template.
+Among the various artifacts for which \@eerkens2001techniques summarised CV values, the most comparable to stemmed points are Great Basin projectiles points, whose CV values range from 6-55%, with an average of 22%.
+Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].
 
@@
 A variety of methods are available for improving the reliability of CV measurements when using small samples.
-For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e.
+For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller samples (i.e.
 n\<25).
@@
 
-@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
-Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22.
-Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current].
-Unlike correlation, @bettinger1999point had vaguely mentioned about the threshold for CV values.
-Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points.
-We calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point).
-With the exception of weight, the CV range for metric attributes on the points is 17.01 to 24.59.
-We chose higher side on the CV range of Rosegate points as the threshold of our research based on following reasons.
-1) Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
-The duration of transmission process for stemmed points is much longer (approx.30,000yrs) than the case of Rosegate (approx.500yrs in Fremont region) [@bischoff2021rosegate].
-2) The errors are also highly associated with raw materials.
-Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques].
-We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to raw materials for stemmed points such as shale, rhyolite, porphyry, etc.
+One limitation of previous work is the absence of a clear threshold value for interpreting CV values in terms of different transmission biases.
+@bettinger1999point did not identify a threshold for CV values.
+Here we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases.
+We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias.
+With the exception of weight, the CVs for metric attributes on the Rosegate points is 17% to 24%.
+We believe higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on following reasons.
+First, variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
+The duration of transmission process for stemmed points is much longer (in the order of thousands of years) than the case of Rosegate (in the order of hundreds of years in Fremont region) [@bischoff2021rosegate].
+Second, the errors are likely to vary by raw materials.
+Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques].
+We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
 (@fig-raw-materials-by-phase).
-3) Compared to normal CV, corrected CV has slightly higher values [@vanpool2011quantitative].
-Based on the threshold, we premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
+Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values than compared to standard CV values [@vanpool2011quantitative].
+Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likley reflect indirect bias.
 
@@
 ```
+
 # Discussion
@@
 ```
+
 In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
@@
 
-The most striking differences in both correlation coefficients and CV values is 
-in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
+The most striking differences in both correlation coefficients and CV values is in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
 All distributions are different and there is no clear pattern, indicating that raw materials strongly influence stemmed points.

### 2023-02-13 analysis/paper/paper.qmd
BiocManager

@@
 
-Our results show positive correlation between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis).
+Our results show positive correlations between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis).
 Some relationships such as BL and ML have stronger correlation (i.e.
@@
 
-The correlations between attributes became stronger over time as we can clearly see in the Panel D of @fig-corr-analysis.
-The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold.
-Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
+Panel D of @fig-corr-analysis shows that the correlations between attributes became stronger over time.
+The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold.
+value.
+<!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
 On the other hand, Phase 3 could be more associated with indirect bias.
@@
 These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
-We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
+We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
+The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
 There are more variation in correlation coefficient in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.

### 2023-02-14 analysis/paper/paper.qmd
renv

@@
 The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
-There are more variation in correlation coefficient in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
+There is more variation in correlation coefficients in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
 There is no clear pattern between assemblages.

### 2023-02-14 analysis/paper/paper.qmd
renv::init(bioconductor="3.14")

@@
 The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
-There is more variation in correlation coefficients in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
+There is more variation in correlation coefficients in Phase 2 <!-- I don't think this is true, look at the size of the box in the box and whisker plot --> while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
 There is no clear pattern between assemblages.

### 2023-02-14 analysis/paper/paper.qmd
set R version in dockerfile to match R version in renv lockfile

--- 2023-02-14 analysis/paper/paper.qmd
@@
 The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
-There is more variation in correlation coefficients in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
-There is no clear pattern between assemblages.
-SYG1_2 is from Phase 2 but it has the higher median than the assemblage from the same site in the later phase.
+At this finer scale of resolution the picture is more complex.
+Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
+The two assemblages from Phase 3 show similarly correlated attributes, with the median right around our threshold value of `r threshold_value_cor` .
 

### 2023-02-14 analysis/paper/paper.qmd
update dockerfile

@@
 Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
-The two assemblages from Phase 3 show similarly correlated attributes, with the median right around our threshold value of `r threshold_value_cor` .
+The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`, however the spread of correlations is high, suggesting a mixture of transmission biases.
 
@@
 
-To examine the impact of raw materials on the shape of stemmed points, we computed the correlation for different raw materials.
+To examine the impact of raw materials on the shape of stemmed points, we computed the correlations of attributes among raw material groups.
 We excluded raw materials that were used for less than three points.
-@fig-corr-raw-ass shows that the distribution of correlation coefficients is various depending on raw materials.
-Overall, the materials that used for more and equal to three points have high correlation coefficients.
-The stemmed points made from quartzite have highest correlation coefficients with narrow distribution.
-Acidic volcanic rocks are one of the highly correlated materials and individual correlations are skewed into two groups.
-Some of the correlations are even negatively related.
-Correlation values for porphyry stemmed points have a narrow distribution.
-Rhyolite stemmed points have wide distribution and some of the attributes are negatively related.
-Like acidic volcanic rocks, the correlation of shale stemmed points is also skewed into two groups.
-Tuff stemmed points have wide distribution with higher median value.
+@fig-corr-raw-ass shows that the most abundant raw materials have relatively high correlation coefficients.
+The stemmed points made from quartzite and porphyry have the highest correlation coefficients with narrow distributions.
+Acidic volcanic rocks are one of the highly correlated materials, with individual correlations divided into two groups.
+Similarly, the correlation of shale stemmed points is also divided into two groups.
+Rhyolite stemmed points have a wide distribution of correlations, and some of the attributes are negatively correlated.
+Tuff stemmed points have wide distribution with a high median value.
 

### 2023-02-14 analysis/paper/paper.qmd
more dockerfile

@@
 
-# correlation by site
-# df_sitename_corr_site <- 
-#   df_sitename %>%
-#   group_by(sitename) %>% 
-#   filter(n() > 5) %>% 
-#   dplyr::select(-sitename, 
-#                 -SPstage.Stage) %>% 
-#   nest(-sitename) %>% 
-#   mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
-#                                    shave() %>% 
-#                                   pivot_longer(-term))) %>% 
-#   unnest(corrr) %>% 
-#   filter()
-# 
-# correlations_by_site <- 
-#   ggplot(df_sitename_corr_site) +
-#   aes(sitename,
-#       value) +
-#   geom_boxplot() +
-#   ggbeeswarm::geom_quasirandom(size = 2,
-#                                alpha = 0.3) +
-#   theme_minimal() +
-#   labs(y = "Correlation coefficient",
-#        x = "Site") +
-# theme(axis.title.y = element_text(size = 4)) +
-#   # draw a line to show our threshold value
-#   geom_hline(yintercept = threshold_value_cor, colour = "red") +
-#   theme_bw()
-
 # correlation by raw material
@@
   theme_bw()
-
 ```
@@
 
-Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
-Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
-
-Panel B of @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
+In addition to correlation coefficients of point attributes, we measured coefficients of variation (CVs) to quantify variability in individual attributes.
+@fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher.
-Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values.
-The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals.
+The CVs of body length (BL) and maximum length (ML) have the lowest values.
+The low CV and narrow confidence intervals of maximum length indicates that this dimension is highly standardized, relative to the others.
 Tang and stem related attributes (e.g.
 SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes.
-Overall, our results suggests that transmission of stemmed points could be more influenced by guided variation.
+Overall, with all but one attribute having CV values above our threshold value, our results suggests that the transmission of stemmed points was mostly influenced by guided variation.
 
@@
 #| label: fig-plot-cv-all-attributes
-#| fig-cap: A.
-Boxplot for each attribute measured on the stemmed points.
-The black bold horizontal lines in the middle of boxes represent the median value.
-The dots represent individual artifacts.
-B.
-The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
-95% confidence intervals for the CVs were computed using Sharma and Krishna's method.
-The vertical lines indicate the confidence intervals.
+#| fig-cap: The points and numbers represent the CV values for each attribute measured on the stemmed points.
+The vertical lines indicate the 95% confidence intervals for the CV values, which were computed using Sharma and Krishna's method.
 
@@
 
-plot_cv_all_attributes <- 
 ggplot(cv_data_sharma_named,
@@
 
-library(cowplot)
-plot_grid(plot_raw_measurements_for_attributes,
-          plot_cv_all_attributes,
-          ncol = 1, 
-          labels = "AUTO")
-```
-
-Similar to Panel D of @fig-corr-analysis, we divided the stemmed points into three chronological phases and computed CV to examine temporal patterns in the modes of cultural transmission.
-Again, we excluded stemmed points for Phase 1 due to the limited number of artifacts.
+```
+
+@fig-plot-cv-over-time shows CV values for stemmed points grouped into three chronological phases to examine temporal patterns in the modes of cultural transmission.
+We excluded stemmed points for Phase 1 due to the limited number of artifacts dating to this time.
 
 @fig-plot-cv-over-time shows that the directional trend in CV values of artifact attributes is complicated.
-Half of the attributes are below our threshold value of `r threshold_value`, and half are above.
-Only body length crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
-For those attributes where the CV is \<`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
-For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these attributes.
+Half of the attributes are below our threshold value of `r threshold_value`, and half are above for both phases.
+Only body length (BL) crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
+For those attributes where the CV is \<`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width (MW) and tang width (TW), indicating increasing influence of indirect bias for these attributes.
+For stem length (SL), stem width (SW) and tang length (TL), CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these stem and tang attributes.
 
@@
 
-Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width for the assemblages that have more than five stemmed points.
-@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`.
+We further analysed temporal change by looking at specific assemblages in each chronological phase.
+Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length, maximum length, mid width, and tang width for assemblages that have five or more stemmed points.
+@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no directional trend apparent between Phase 2 and Phase 3.
+Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
 Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals.
@@
 Perhaps it could be related to its small sample size, relative to the other assemblages compared here.
-Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
-This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
+These results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
 

### 2023-02-14 analysis/paper/paper.qmd
finished editing discussion section

@@
 Perhaps it could be related to its small sample size, relative to the other assemblages compared here.
-These results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
+The larger assemblages in @fig-plot-cv-four-assemblage indicate guided variation, while the smaller assemblages that dominate @fig-plot-cv-over-time suggest indirect bias.
+Perhaps smaller assemblages represented a social context of higher fidelity copying because the cost of failure was greater due to low social insurance because of low population network sizes.
+Another interpretation is that these results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
 
@@
 
-We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was highly dependent on raw material.
+We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was dependent on raw material.
 We excluded raw materials that were used for less than three points.
 @fig-plot-cv-by-raw-material shows that quartzite, the raw material for the stemmed points at Sachang, has the highest CV values, and also has wide conference intervals.
-Other raw materials are generally stable in variation throughout the attributes.
+Other raw materials are generally low, right around our threshold value, and stable in variation across the attributes.
 
@@
 #| label: fig-plot-cv-by-raw-material
-#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the confidence intervals.
+#| fig-cap: CV values for raw materials that were used for making more than three stemmed points.
+The points represent CV values for each attribute.
+The vertical lines indicate the confidence intervals.
 #| fig-width:   8.5
@@
 
-To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission.
-We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we drew on concepts of cultural transmission.
+We asked three questions to examine cultural transmission process over time: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
 Do the modes of cultural transmission vary over time?
-and Do the modes of cultural transmission vary over space?
-We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
-Following @bettinger1999point, we used `r threshold_value_cor` as a threshold of correlation coefficients to evaluate the two transmission biases.
+and, Do the modes of cultural transmission vary over space?
+We focused on two transmission biases: guided variation (socially isolated groups making stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
+Following @bettinger1999point, we used `r threshold_value_cor` as a threshold value for interpreting correlation coefficients to evaluate which of the two transmission biases was dominant.
 Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias.
-Since there is no clear criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
+Since there is no previously established criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
 If the CV value is lower than `r threshold_value`, we interpreted this as indicating indirect bias while values higher than `r threshold_value` were interpreted as guided variation.
 
-Our correlation analyses show that the correlation coefficients of stemmed points are all positive but mostly under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias.
-Some correlations such as body length and maximum length have strong relationship while other correlations such as tang length and body length are much weaker.
-The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but there is no clear pattern and the values vary depending on assemblages (@fig-plot-cor-four-assemblage).
-The CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below.
+Our correlation analyses show that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias.
+Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker.
+The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
+When all artefacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below.
+However, further analysis by site and raw material show more ambiguous patterns, with CV values close to or below the threshold value.
 Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time).
@@
 Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
-In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).
+In exploring individual assemblages that contain multiple stemmed points we found minimal change in CV values over time (@fig-plot-cv-four-assemblage).
 
@@
 We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation.
-We assume that certain attributes were more carefully transmitted than others.
+Based on these findings, we propose that certain attributes were more carefully transmitted than others.
 We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
-The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations.
-We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape.
+
+Our interpretation is that attributes with lower CV values were be transmitted in a social context dominated by indirect bias, while the higher CV attributes were influenced more by specific manufacturing and maintenance situations.
+We assume that the lower CV values of the four attributes imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape.
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
@@
 The most striking differences in both correlation coefficients and CV values is in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
-All distributions are different and there is no clear pattern, indicating that raw materials strongly influence stemmed points.
-We expected that stemmed points with higher correlation coefficients would have lower CV values, but the actual results do not match our expectations.
+Raw materials strongly influenced stemmed points shape variance.
+We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not match our expectations because of how influential raw materials are in the CV values.
 For example, artifacts made from acidic quartzite have highest correlation but their CV values are also much higher with wider confidence intervals, than all other raw materials.
-This conflicting results might be related to the relatively unpredictable knapping quality of quartzite.
+This complex results might be related to the relatively unpredictable knapping quality of quartzite.
 These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals.

### 2023-02-14 analysis/paper/paper.qmd
finished editing conclusions

@@
 
-We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error).
+We conclude that the dominant mode of cultural transmission for the earliest stemmed points was mostly likely guided variation (trial and error).
 This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`.
-A social context that favored trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
+A social context that favored trial and error for artifact making is most consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
 This model proposes that people developed stemmed points by experimenting with existing blade technologies.
@@
 This results may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model.
-We might speculate based on the evolutionary ecological assumption that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [@fitzhugh2001risk].
-With more frequent contacts and connections with members of the social network, there may have been more opportunities to learn point-making directly from socially successful individuals.
+Drawing on evolutionary ecological theory, we might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
+@fitzhugh2001risk].
+With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.
 
@@
 We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages.
-Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in phase three.
+Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in phase 3.
 Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Yonsangdong that date to this period.

### 2023-02-14 analysis/paper/paper.qmd
commit to merge with Ben's edits

--- 2023-02-14 analysis/paper/paper.qmd
@@
   The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia.
-Previous studies have mainly focussed on possible origin locations of these new technologies.
+Previous studies have mainly focussed on possible origin locations and routes of these new technologies.
 The specific cultural processes of the appearance stemmed points and blades has rarely been considered.
-We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
+In this research, we investigate the potential social contexts related to the technological transition by applying a cultural transmission framework.
 Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic?
-Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points.
-To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV).
-We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias.
-Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation.
-Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation <!-- and then indirect bias was adapted to carry out the technological change BM: this is unclear -->.
-We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage.
-As a result, the shape of stemmed points became more standardized among their social groups.
+Following Bettinger and Eerkens [-@bettinger1999point], we tested models of guided variation and indirect bias using data from the Korean Paleolithic assemblages containing stemmed points.
+We computed correlation coefficients and coefficients of variation (CV) to evaluate the transmission biases.
+Based on our results, it appears that information about the novel technology was transmitted via guided variation with a low impact of indirect bias.
+Some attributes, such as length and width, appear to be transmitted with little variation than others.
+We assume that individuals or groups developed stemmed points by experimenting with blade technologies.
+And then they copied the key elements of a successful model to ensure quality and to optimize tool performance.
+The shape of stemmed points became more standardized among their social groups as a result.
 
@@
 
-The application of evolutionary theory to archaeological research has been productive for the investigation of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
-For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual; @garvey2018current].
-In this paper we used cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
-The primary technological innovation is during this time was the appearance of new tools such as stemmed points and blades.
-The appearance of these new lithic technologies may represent the first arrival of modern humans in this region, and so they may also represent a key event related to modern human dispersal throughout East Asia [@seong2009emergence].
-Previous studies on the origin of stemmed points have mainly focused on their possible origin locations, connecting Korea with global patterns of modern human dispersal.
-However, questions about the specific cultural processes and social contexts of this technological change in Korea remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin].
-In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic.
-Our main question is: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-Additionally we ask: Do the modes of cultural transmission vary over time and space?
+The application of evolutionary theory to archaeological research has been favored for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
+Archaeologists have used evolutionary theories and methods to study human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current].
+In this paper we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
+The primary technological innovation of this period was the introduction of stemmed points and blades.
+These new lithic technologies may represent the first arrival of modern humans in this region, and thus they may also mark an important event in human dispersal through East Asia [@seong2009emergence].
+Most previous studies on stemmed points focused on where they may have originated, connecting Korea with global patterns of modern human dispersal.
+There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea [@seong2008tanged; @bae2010origin; @bae2017origin].
+Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic based on a cultural transmission framework.
+Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+We also ask: Do modes of cultural transmission vary over time and space?
 We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two.
-Our results have implications for determining if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.
+The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.
 
@@
 
-The onset of the Late Paleolithic period in Korea is marked by the appearance of stemmed points around 40-35 ka.
-The stemmed point is a projectile point made out of an elongated flake or blade with slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
-The stemmed point was the first composite tool type appearing on the Korean Peninsula, and represents new hunting strategies there, as well as other adjacent regions where it appeared later [@seong2008tanged; @lee2019were].
-Stemmed points, combined with blade technology, multi-stage production sequences, and evidence of being resharpened and reused, are important aspects of the Late Paleolithic technological innovations in Korea because these features are rarely seen in earlier periods [@chang2013human; @bamforth2009projectile; @seong2015diversity].
-Currently the earliest stemmed points in Northeast Asia are from Yongho-doing site in Korea dated to 38.5ka and made on elongate flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
-After their first appearance in Korea, the stemmed points spread to the Japanese archipelago [@chang2013human].
-
-Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points, as part of the discussion of the timing and routes of modern human dispersal in eastern Asia.
-The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
-The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
-To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea.
-In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one.
-He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention.
-In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies.
-
-In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins outside of the Korean peninsula [@bae2010origin; @bae2012nature; @bae2013early].
+The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka.
+A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
+Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were].
+A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity].
+Currently, the oldest stemmed points in Northeast Asia are from the Yongho-doing site in Korea, dating back to 38.5ka [@seong2009emergence; @seong2015diversity; @bae2017origin].
+Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].
+
+A majority of previous studies of the technological transitions during the Late Paleolithic in Korea have focused on the possible origin locations of stemmed points, as well as the timing and routes of modern human dispersal.
+In order to explain the origin of stemmed points, two competing models can be distinguished: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
+The *in situ* model claims that stemmed points and other Late Paleolithic assemblages, including blade industries, autonomously developed on the Korean peninsula as a result of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
+In order to support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea.
+In his view, the blade industry represents a novel invention, while flake materials indicate a continuously used technology.
+Based on the increased proportion of blades in stone artifact assemblages during the Late Paleolithic, he argues that the new technologies spread throughout the area after its local invention.
+In addition, Seong argues that the emergence of new technologies is a result of gradual, local, evolutionary processes, as evidenced by the increasing number of stemmed points over time.
+
+On the other hand, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers, and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early].
 While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
-Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
+The blade technology was introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake tool tradition came from  southern China [@bae2013early].
 The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome].
 Bae et al.
-[-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
-
-Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development.
-He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size.
+[-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea area more accessible.
+
+According to Lee [-@lee2013current], the transition from the Korean Early Paleolithic to the Korean Late Paleolithic might have been more complicated than either migration models or *in situ* development models.
+Under conditions of low effective population size, he partially agrees with the *in situ* model that simple flake tools had continuously used in Korea due to ancestor-descendant relationships.
 With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration.
@@
 
-To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory.
-This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
-Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
-The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties.
-Social transmission strongly affects these key evolutionary processes [@Whiten_2017], and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
-We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gather populations.
-
-Cultural transmission theory holds that information about social and technological behaviours is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
+We use Darwinian evolutionary theory to measure transmission biases in tool-making.
+The theory has helped archaeologists better understand human behaviors and technological innovations, for example, using human behavioral ecology, phylogenetics, cultural transmission, and niche construction [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+Cultural evolutionary approaches are effective at explaining cultural changes through mechanisms of inheritance, variation-generating processes, and selection.
+Although cultural systems follow similar mechanisms to biological systems, they have unique characteristics.
+Social transmission strongly influences these key evolutionary processes [@Whiten_2017], and can occur through a variety of learning processes, including stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
+In this paper, we examine how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gatherer communities.
+
+######
+Cultural transmission theory holds that information about how to behave is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
 Individuals learn by themselves (e.g.
@@
 Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
-Garvey measured CV values for weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
+Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
 She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e.
@@
 We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness.
-On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.
+On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.
 
@@
 Seong's *in situ* model).
-If this social context was prevalent, we predict this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points.
+This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points.
 If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.
@@
 Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
-In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package.
-This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf.
-Bae's migration model), where stemmed points first appeared and then spread from.
+In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package (cf.
+Bae's migration model).
 In this social context, indirect bias is the dominant influence on transmission of lithic technology.
@@
 To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition were used to assign assemblages to a phase.
-Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blade in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
-Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations.
-Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size.
+These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition.
+Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology).
+Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to various research limitations.
+Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size.
 We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.
@@
 #| label: tbl-korean-chronology
-#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]. SP = stemmed point
+#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]
 
@@
 Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases.
-We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs.
+We exclude phase 1 from our analyses below because of the small sample size to compute correlation and CV.
 Shale is the dominant raw material in both phase 2 and 3.
@@
 
-Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e.
+Similar to the metric attributes from previous studies of cultural transmission and projectile points, i.e.
 maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
-Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes than traditional caliper measurements [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
 We did not include weight and thickness as these measurements were not available to us.
@@
 The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for further analyses.
+We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for current and further analyses.
 
@@
 We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points.
-The CV is the ratio (usually expressed as a percentage) of the standard deviation to the mean.
+The CV is the ratio (%) of the standard deviation to the mean.
 It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc.
-in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization].
+in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence].
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template.
-Among the various artifacts for which \@eerkens2001techniques summarised CV values, the most comparable to stemmed points are Great Basin projectiles points, whose CV values range from 6-55%, with an average of 22%.
-Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].
+Based on the Weber Fraction for visual measurement of line length, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
 
@@
 A variety of methods are available for improving the reliability of CV measurements when using small samples.
-For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller samples (i.e.
+For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e.
 n\<25).
@@
 
-One limitation of previous work is the absence of a clear threshold value for interpreting CV values in terms of different transmission biases.
-@bettinger1999point did not identify a threshold for CV values.
-Here we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases.
-We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias.
-With the exception of weight, the CVs for metric attributes on the Rosegate points is 17% to 24%.
-We believe higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on following reasons.
-First, variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
-The duration of transmission process for stemmed points is much longer (in the order of thousands of years) than the case of Rosegate (in the order of hundreds of years in Fremont region) [@bischoff2021rosegate].
-Second, the errors are likely to vary by raw materials.
-Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques].
-We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
+@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
+Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22.
+Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current].
+Unlike correlation, @bettinger1999point had vaguely mentioned about the threshold for CV values.
+Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points.
+We calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point).
+With the exception of weight, the CV range for metric attributes on the points is 17.01 to 24.59.
+We chose higher side on the CV range of Rosegate points as the threshold of our research based on following reasons.
+1) Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
+The duration of transmission process for stemmed points is much longer (approx.30,000yrs) than the case of Rosegate (approx.500yrs in Fremont region) [@bischoff2021rosegate].
+2) The errors are also highly associated with raw materials.
+Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques].
+We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to raw materials for stemmed points such as shale, rhyolite, porphyry, etc.
 (@fig-raw-materials-by-phase).
-Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values than compared to standard CV values [@vanpool2011quantitative].
-Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likley reflect indirect bias.
+3) Compared to normal CV, corrected CV has slightly higher values [@vanpool2011quantitative].
+Based on the threshold, we premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.
 
@@
 
-Our results show positive correlations between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis).
+Our results show positive correlation between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis).
 Some relationships such as BL and ML have stronger correlation (i.e.
@@
 
-Panel D of @fig-corr-analysis shows that the correlations between attributes became stronger over time.
-The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold.
-value.
-<!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
+The correlations between attributes became stronger over time as we can clearly see in the Panel D of @fig-corr-analysis.
+The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold.
+Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
 On the other hand, Phase 3 could be more associated with indirect bias.
@@
 These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
-We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
-The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
-At this finer scale of resolution the picture is more complex.
-Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
-The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`, however the spread of correlations is high, suggesting a mixture of transmission biases.
+We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS).
+There are more variation in correlation coefficient in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`.
+There is no clear pattern between assemblages.
+SYG1_2 is from Phase 2 but it has the higher median than the assemblage from the same site in the later phase.
 
@@
 
-To examine the impact of raw materials on the shape of stemmed points, we computed the correlations of attributes among raw material groups.
+To examine the impact of raw materials on the shape of stemmed points, we computed the correlation for different raw materials.
 We excluded raw materials that were used for less than three points.
-@fig-corr-raw-ass shows that the most abundant raw materials have relatively high correlation coefficients.
-The stemmed points made from quartzite and porphyry have the highest correlation coefficients with narrow distributions.
-Acidic volcanic rocks are one of the highly correlated materials, with individual correlations divided into two groups.
-Similarly, the correlation of shale stemmed points is also divided into two groups.
-Rhyolite stemmed points have a wide distribution of correlations, and some of the attributes are negatively correlated.
-Tuff stemmed points have wide distribution with a high median value.
+@fig-corr-raw-ass shows that the distribution of correlation coefficients is various depending on raw materials.
+Overall, the materials that used for more and equal to three points have high correlation coefficients.
+The stemmed points made from quartzite have highest correlation coefficients with narrow distribution.
+Acidic volcanic rocks are one of the highly correlated materials and individual correlations are skewed into two groups.
+Some of the correlations are even negatively related.
+Correlation values for porphyry stemmed points have a narrow distribution.
+Rhyolite stemmed points have wide distribution and some of the attributes are negatively related.
+Like acidic volcanic rocks, the correlation of shale stemmed points is also skewed into two groups.
+Tuff stemmed points have wide distribution with higher median value.
 
@@
 #| fig-height:5
+
+# correlation by site
+# df_sitename_corr_site <- 
+#   df_sitename %>%
+#   group_by(sitename) %>% 
+#   filter(n() > 5) %>% 
+#   dplyr::select(-sitename, 
+#                 -SPstage.Stage) %>% 
+#   nest(-sitename) %>% 
+#   mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
+#                                    shave() %>% 
+#                                   pivot_longer(-term))) %>% 
+#   unnest(corrr) %>% 
+#   filter()
+# 
+# correlations_by_site <- 
+#   ggplot(df_sitename_corr_site) +
+#   aes(sitename,
+#       value) +
+#   geom_boxplot() +
+#   ggbeeswarm::geom_quasirandom(size = 2,
+#                                alpha = 0.3) +
+#   theme_minimal() +
+#   labs(y = "Correlation coefficient",
+#        x = "Site") +
+# theme(axis.title.y = element_text(size = 4)) +
+#   # draw a line to show our threshold value
+#   geom_hline(yintercept = threshold_value_cor, colour = "red") +
+#   theme_bw()
 
@@
   theme_bw()
+
 ```
@@
 
-In addition to correlation coefficients of point attributes, we measured coefficients of variation (CVs) to quantify variability in individual attributes.
-@fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
+Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm).
+Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.
+
+Panel B of @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher.
-The CVs of body length (BL) and maximum length (ML) have the lowest values.
-The low CV and narrow confidence intervals of maximum length indicates that this dimension is highly standardized, relative to the others.
+Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values.
+The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals.
 Tang and stem related attributes (e.g.
 SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes.
-Overall, with all but one attribute having CV values above our threshold value, our results suggests that the transmission of stemmed points was mostly influenced by guided variation.
+Overall, our results suggests that transmission of stemmed points could be more influenced by guided variation.
 
@@
 #| label: fig-plot-cv-all-attributes
-#| fig-cap: The points and numbers represent the CV values for each attribute measured on the stemmed points.
-The vertical lines indicate the 95% confidence intervals for the CV values, which were computed using Sharma and Krishna's method.
+#| fig-cap: A.
+Boxplot for each attribute measured on the stemmed points.
+The black bold horizontal lines in the middle of boxes represent the median value.
+The dots represent individual artifacts.
+B.
+The points and blue colored numbers represent CV values for each attribute measured on the stemmed points.
+95% confidence intervals for the CVs were computed using Sharma and Krishna's method.
+The vertical lines indicate the confidence intervals.
 
@@
 
+plot_cv_all_attributes <- 
 ggplot(cv_data_sharma_named,
@@
 
-```
-
-@fig-plot-cv-over-time shows CV values for stemmed points grouped into three chronological phases to examine temporal patterns in the modes of cultural transmission.
-We excluded stemmed points for Phase 1 due to the limited number of artifacts dating to this time.
+library(cowplot)
+plot_grid(plot_raw_measurements_for_attributes,
+          plot_cv_all_attributes,
+          ncol = 1, 
+          labels = "AUTO")
+```
+
+Similar to Panel D of @fig-corr-analysis, we divided the stemmed points into three chronological phases and computed CV to examine temporal patterns in the modes of cultural transmission.
+Again, we excluded stemmed points for Phase 1 due to the limited number of artifacts.
 
 @fig-plot-cv-over-time shows that the directional trend in CV values of artifact attributes is complicated.
-Half of the attributes are below our threshold value of `r threshold_value`, and half are above for both phases.
-Only body length (BL) crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
-For those attributes where the CV is \<`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width (MW) and tang width (TW), indicating increasing influence of indirect bias for these attributes.
-For stem length (SL), stem width (SW) and tang length (TL), CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these stem and tang attributes.
+Half of the attributes are below our threshold value of `r threshold_value`, and half are above.
+Only body length crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
+For those attributes where the CV is \<`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes.
+For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these attributes.
 
@@
 
-We further analysed temporal change by looking at specific assemblages in each chronological phase.
-Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length, maximum length, mid width, and tang width for assemblages that have five or more stemmed points.
-@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no directional trend apparent between Phase 2 and Phase 3.
-Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
+Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width for the assemblages that have more than five stemmed points.
+@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`.
 Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals.
@@
 Perhaps it could be related to its small sample size, relative to the other assemblages compared here.
-These results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
+Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
+This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
 
@@
 ```
-
 In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
@@
 
-The most striking differences in both correlation coefficients and CV values is in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
+The most striking differences in both correlation coefficients and CV values is 
+in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
 All distributions are different and there is no clear pattern, indicating that raw materials strongly influence stemmed points.
@@
 
+The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.??
+
 # Conclusion

### 2023-02-15 analysis/paper/paper.qmd
Chose Ben's edits instead of GP's edits

--- 2023-02-14 analysis/paper/paper.qmd
@@
 
-The onset of the Late Paleolithic period in Korea is marked by the appearance of stemmed points around 40-35 ka.
-The stemmed point is a projectile point made out of an elongated flake or blade with slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
-The stemmed point was the first composite tool type appearing on the Korean Peninsula, and represents new hunting strategies there, as well as other adjacent regions where it appeared later [@seong2008tanged; @lee2019were].
-Stemmed points, combined with blade technology, multi-stage production sequences, and evidence of being resharpened and reused, are important aspects of the Late Paleolithic technological innovations in Korea because these features are rarely seen in earlier periods [@chang2013human; @bamforth2009projectile; @seong2015diversity].
-Currently the earliest stemmed points in Northeast Asia are from Yongho-doing site in Korea dated to 38.5ka and made on elongate flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
-After their first appearance in Korea, the stemmed points spread to the Japanese archipelago [@chang2013human].
+The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka.
+A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
+Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were].
+A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity].
+Currently, the oldest stemmed points in Northeast Asia are from the Yongho-doing site in Korea, dating back to 38.5ka [@seong2009emergence; @seong2015diversity; @bae2017origin].
+Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].
 
@@
 Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two of these biases can be identified in stone artifact assemblages.
-Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
-They equated guided transmission (where individuals learn new behaviors through trial and error) with high metric variation and low correlation between metric attributes.
+Their research focused on the metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600.
+They equated guided transmission (where individuals learn about new behaviors through trial and error) with high metric variation and low correlation between metric attributes.
 Conversely, they inferred indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) from less variation and more correlated variables.
 They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools.
-From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly.
+Based on these results, they inferred that eastern California had a social context of distant and unfamiliar neighbors with little direct contact between groups.
 In this context of limited contact, they argue that the new technologies developed largely by trial and error.
@@
 Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
-Garvey measured CV values for weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site.
+Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site.
 She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e.
 CV = 10%, 5%, and 3%).
-The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying".
+The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying".
 A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
-Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.
+Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.
 
@@
 Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
-They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants.
+They simulated model-based bias in projectile point making by providing the design of a model and information about the model's prior success to their research participants.
 They also simulated guided variation by allowing their participants to explore their own designs.
 They observed that the majority of participants who were able to choose the previous design copied the most-successful model.
-Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
-Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.
+Metric attributes of projectile points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
+Mesoudi and O'Brien's experimental results are critical because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.
 
@@
 Seong's *in situ* model).
-If this social context was prevalent, we predict this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points.
+If this social context was prevalent, we predict that this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points.
 If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.
 
-On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups.
+On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with them.
 Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information.
 Individuals or groups learned the technology of stemmed points by copying a model from another individual or group.
-In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package.
+In this scenario, it is likely that the model, or ideal, stemmed point has been highly successful and frequently chosen, and learners copy all information about the point design as a package.
 This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf.
 Bae's migration model), where stemmed points first appeared and then spread from.
-In this social context, indirect bias is the dominant influence on transmission of lithic technology.
-If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.
+In this social context, indirect bias is the dominant influence on the transmission of lithic technology.
+If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points would be more standardized, attributes on the stemmed points would be more correlated, and assemblages from multiple sites would exhibit less variance between and within them.
 
@@
 
-After the first discovery of stemmed points at the Seokjangri site in the 1960s, around 300 have been found in nearly 30 sites across Korea (@fig-map).
-Most sites contain only a few points but a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
+After the first discovery of stemmed points at the Seokjangri site at the 1960s, around 300 have been found in nearly 30 sites across Korea (@fig-map).
+While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
 Among these stemmed points, we selected those that are unbroken from the tip to the stem.
@@
 This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number).
-The images of the stemmed points were obtained from published excavation reports and by direct photography during our research on local museum collections.
-We defined multiple assemblages in a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
+The images of the stemmed points were obtained through published excavation reports and direct photography during our research in local museums.
+We defined multiple assemblages from a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
 
@@
 To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition were used to assign assemblages to a phase.
-Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blade in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
+These chronologies were constructed based on radiocarbon ages, and assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition.
+Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
 Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations.
 Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size.
-We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.
+We arranged our `r total_number_of_assemblages` assemblages containing stemmed points into these three phases to facilitate observations of change over time.
 
@@
 We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs.
-Shale is the dominant raw material in both phase 2 and 3.
-In phase 3 there is an increase in the proportions of porphyry, driven largely by finds from Yongsangdong, and other raw materials.
-Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at assemblages with more than five stemmed points.
+Shale is the dominant raw material in both Phase 2 and 3.
+In Phase 3 there is an increase in the proportion of porphyry, driven largely by finds from Yongsangdong, and other raw materials.
+Panel B of @fig-raw-materials-by-phase shows the count of stemmed points by raw material type in assemblages with more than five stemmed points.
 Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.
@@
 Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e.
-maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic).
+maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined variations in shape and the relationship between each attribute using morphological attributes on stemmed points (@fig-labelled-schematic).
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
-Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes than traditional caliper measurements [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
-We did not include weight and thickness as these measurements were not available to us.
-For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses.
+Compared to traditional caliper measurements, morphometric data yield more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+We did not include weight and thickness since these measurements were not available to us.
+For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis.
 The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic.
-We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for further analyses.
+Using the point tool in ImageJ [@schneider2012nih] we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analyses.
 
@@
 Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points.
-The correlation of artifact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission.
-This is based on the assumption that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models.
-In their work, @bettinger1999point interpreted correlation coefficient values of around `r threshold_value_cor` and higher as evidence of indirect bias, and lower values as evidence of guided variation.
-In this research, we examine correlation between attributes and explore other variables such as different phases, raw materials and assemblages.
+The correlation between artifact attributes was proposed by @bettinger1999point as a key indicator of indirect bias during cultural transmission.
+It is based on the assumption that artifact-making was transmitted in packages of traits inherited from socially successful individuals who were role models.
+In their work, @bettinger1999point interpreted correlation coefficient values around `r threshold_value_cor` and higher as evidence of indirect bias, and lower values as evidence of guided variation.
+In this research, we examine the correlation between attributes and explore other variables, such as different phases, raw materials and assemblages.
 
 We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points.
-The CV is the ratio (usually expressed as a percentage) of the standard deviation to the mean.
-It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc.
-in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization].
-In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
+The CV is determined by the ratio of the standard deviation to the mean (usually expressed as a percentage).
+The method has been used in various disciplines to calculate standardization, precision, equality, homogeneity, etc.
+[@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization].
+In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
 As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template.
-Among the various artifacts for which \@eerkens2001techniques summarised CV values, the most comparable to stemmed points are Great Basin projectiles points, whose CV values range from 6-55%, with an average of 22%.
+Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable to stemmed points are Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%.
 Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].
@@
 Sample size can impact the accuracy of CV estimates [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative].
-In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
+In previous archaeological studies, sample sizes for CV values have varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new].
 A variety of methods are available for improving the reliability of CV measurements when using small samples.
-For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller samples (i.e.
+@vanpool2011quantitative, for instance, proposed a "corrected CV" for smaller samples (i.e.
 n\<25).
 Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
-Although we have a total of `r total_number_of_sp` artifacts, most of our analyses here involves comparisons of smaller subsets of this sample.
-For example to explore temporal and regional patterns we compared samples of well under 25 pieces.
+Although we have a total of `r total_number_of_sp` artifacts, most of our analyses involve comparisons of smaller subsets.
+As an example, we compared samples of less than 25 pieces in order to explore temporal and regional patterns.
 Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use.
-We also computed confidence intervals using method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].
+We also computed confidence intervals using the method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].
 
@@
 We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias.
-With the exception of weight, the CVs for metric attributes on the Rosegate points is 17% to 24%.
-We believe higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on following reasons.
-First, variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural].
-The duration of transmission process for stemmed points is much longer (in the order of thousands of years) than the case of Rosegate (in the order of hundreds of years in Fremont region) [@bischoff2021rosegate].
+With the exception of weight, the CVs for metric attributes on Rosegate points range from 17% to 24%.
+We believe the higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on the following reasons.
+First, variation is generated by small errors that are transmitted between individuals and the errors get bigger through generations [@eerkens2005cultural].
+The duration of the transmission process for stemmed points is much longer (in the order of thousands of years) than in the case of Rosegate (in the order of hundreds of years in the Fremont region) [@bischoff2021rosegate].
 Second, the errors are likely to vary by raw materials.
@@
 (@fig-raw-materials-by-phase).
-Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values than compared to standard CV values [@vanpool2011quantitative].
-Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likley reflect indirect bias.
+Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values compared to standard CV values [@vanpool2011quantitative].
+Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likely reflect indirect bias.
 
@@
 value.
-<!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
-On the other hand, Phase 3 could be more associated with indirect bias.
 
@@
 Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time).
-Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases.
+Comparing chronological Phases 2 and 3, we found that there are only minor differences between the two phases.
 Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
@@
 
+<!-- GP need to add this thing in the discussion -->
+<!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
+On the other hand, Phase 3 could be more associated with indirect bias.
+The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.??
+
+To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups.
+Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement.
+
 # Conclusion
@@
 
-The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological phase 3, after 25 ka.
+The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka.
 This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
@@
 We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages.
-Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in phase 3.
-Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Yonsangdong that date to this period.
+Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in Phase 3.
+Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
 One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
 Similarly, change in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
-This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.
+This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
 

### 2023-02-15 analysis/paper/paper.qmd
Edited by discussion

@@
 Our results show positive correlations between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis).
-Some relationships such as BL and ML have stronger correlation (i.e.
+Some relationships such as body length (BL) and maximum length (ML) have stronger correlations (i.e.
 darker-blue points) than others (i.e.
@@
 To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarized in @tbl-korean-chronology.
-Among the three chronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.
+Among the three chronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that belong to this phase.
 
 Panel D of @fig-corr-analysis shows that the correlations between attributes became stronger over time.
-The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold.
-value.
+The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold value.
 
@@
 One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
-These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
+These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only on the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
 We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
@@
 Similarly, the correlation of shale stemmed points is also divided into two groups.
-Rhyolite stemmed points have a wide distribution of correlations, and some of the attributes are negatively correlated.
-Tuff stemmed points have wide distribution with a high median value.
+Rhyolite stemmed points have a wide distribution of correlations, and some of the attributes are even negatively correlated.
+Tuff stemmed points have a wide distribution with a high median value.
 
@@
 @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
-Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher.
-The CVs of body length (BL) and maximum length (ML) have the lowest values.
-The low CV and narrow confidence intervals of maximum length indicates that this dimension is highly standardized, relative to the others.
+Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for Korean artifacts is higher.
+The CVs for body length (BL) and maximum length (ML) have the lowest values.
+The low CV and narrow confidence intervals for maximum length indicate that this dimension is highly standardized relative to the others.
 Tang and stem related attributes (e.g.
 SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes.
-Overall, with all but one attribute having CV values above our threshold value, our results suggests that the transmission of stemmed points was mostly influenced by guided variation.
-
+Overall, with all but one attribute having CV values above our threshold value, our results suggest that the transmission of stemmed points was mostly influenced by guided variation.
 ```{r}
@@
 
-@fig-plot-cv-over-time shows CV values for stemmed points grouped into three chronological phases to examine temporal patterns in the modes of cultural transmission.
-We excluded stemmed points for Phase 1 due to the limited number of artifacts dating to this time.
+In @fig-plot-cv-over-time, CV values for stemmed points are grouped into three chronological phases to analyze temporal patterns in cultural transmission.
+We excluded stemmed points from Phase 1 due to the limited number of artifacts dating to this time.
 
@@
 
-We further analysed temporal change by looking at specific assemblages in each chronological phase.
-Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length, maximum length, mid width, and tang width for assemblages that have five or more stemmed points.
+We further analyzed temporal change by looking at specific assemblages in each chronological phase.
+Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) for assemblages that have five or more stemmed points.
 @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no directional trend apparent between Phase 2 and Phase 3.
@@
 Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals.
-It is interesting because SYG1_2 has the strong correlation between attributes (@fig-plot-cor-four-assemblage).
+It is interesting because SYG1_2 has a strong correlation between attributes (@fig-plot-cor-four-assemblage).
 Perhaps it could be related to its small sample size, relative to the other assemblages compared here.
@@
 
-We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was dependent on raw material.
+We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was dependent on raw materials.
 We excluded raw materials that were used for less than three points.
@@
 To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we drew on concepts of cultural transmission.
-We asked three questions to examine cultural transmission process over time: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+We asked three questions to examine the cultural transmission process over time: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
 Do the modes of cultural transmission vary over time?
-and, Do the modes of cultural transmission vary over space?
+And do the modes of cultural transmission vary over space?
 We focused on two transmission biases: guided variation (socially isolated groups making stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
@@
 Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias.
-Since there is no previously established criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
-If the CV value is lower than `r threshold_value`, we interpreted this as indicating indirect bias while values higher than `r threshold_value` were interpreted as guided variation.
+Since there are no previously established criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
+If the CV value is lower than `r threshold_value`, we interpret this as indicating indirect bias while values higher than `r threshold_value` are interpreted as guided variation.
 
@@
 The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
-When all artefacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below.
-However, further analysis by site and raw material show more ambiguous patterns, with CV values close to or below the threshold value.
-Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time).
-Comparing chronological Phases 2 and 3, we found that there are only minor differences between the two phases.
+When all artifacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below.
+However, further analysis by site and raw material shows more ambiguous patterns, with CV values close to or below the threshold value.
+Change in CV values over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time).
+Comparing chronological Phase 2 and 3, we found that there are only minor differences between the two phases.
 Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
-In exploring individual assemblages that contain multiple stemmed points we found minimal change in CV values over time (@fig-plot-cv-four-assemblage).
+In exploring individual assemblages that contain multiple stemmed points we observed minimal changes in CV values over time (@fig-plot-cv-four-assemblage).
 
@@
 
-In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
-We found that each assemblage has different distribution patterns of correlation coefficients and CV values.
-SYG1_2 and SYG1_3 are from the same site but different chronological periods and correlations decreases over time.
-It is likely that correlation or overall shape can be more depending on individual knappers.
-We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation.
+In exploring geographical variation we found no clear pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
+We found that correlation coefficients and CV values are distributed differently in each assemblage.
+While SYG1_2 and SYG1_3 are from the same site, they belong to different chronological periods, and their correlation decreases over time.
+It is likely that the correlation or overall shape can be more depending on individual knappers.
+We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation while the others including stemmed length (SL) and body length (BL) have lower or even negative correlation.
 Based on these findings, we propose that certain attributes were more carefully transmitted than others.
-We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.
+We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW), similar to what is evident in @fig-plot-cv-all-attributes.
 

### 2023-02-16 analysis/paper/paper.qmd
Edited by conclusion (only texts)

@@
 
-Our interpretation is that attributes with lower CV values were be transmitted in a social context dominated by indirect bias, while the higher CV attributes were influenced more by specific manufacturing and maintenance situations.
+In our interpretation, attributes with lower CV values were transmitted in a social context dominated by indirect bias, whereas attributes with high CV values were influenced more by specific manufacturing and maintenance situations.
 We assume that the lower CV values of the four attributes imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape.
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
-This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact.
+This can be contrasted with tang length, stem length, and stem width, which are away from the spot of impact at the end of the artifact.
 These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
 
-The most striking differences in both correlation coefficients and CV values is in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
-Raw materials strongly influenced stemmed points shape variance.
-We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not match our expectations because of how influential raw materials are in the CV values.
-For example, artifacts made from acidic quartzite have highest correlation but their CV values are also much higher with wider confidence intervals, than all other raw materials.
+The most striking differences in both correlation coefficients and CV values are in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
+Raw materials have a significant impact on stemmed points' shape variance.
+We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not match our expectations because of how influential raw materials are on CV values.
+For example, artifacts made from acidic quartzite have the highest correlation, but their CV values are also much higher with wider confidence intervals than all other raw materials.
 This complex results might be related to the relatively unpredictable knapping quality of quartzite.
 These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals.
-This makes is harder for the knapper to repeatedly produce the same size and shape end product.
-This hints at the possibility that physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
+This makes it harder for the knapper to repeatedly produce the same size and shape end product.
+This hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
 However with only three artifacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.
@@
 In this research, we investigated the social context of technological innovation in the Korean Late Paleolithic.
-Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we investigated two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
-We proposed two scenarios for explaining the introduction of new technology: socially isolated groups that made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came in contact with (indirect bias).
+Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
+We proposed two scenarios for explaining the introduction of new technology: socially isolated groups that developed stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came into contact with (indirect bias).
 We asked three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
@@
 This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`.
-A social context that favored trial and error for artifact making is most consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
-This model proposes that people developed stemmed points by experimenting with existing blade technologies.
+A social context that favored trial and error for artifact making is most consistent with Seong's in situ model for the appearance of stemmed points in Korea.
+This model proposes that people developed stemmed points by experimenting with existing enlongated flakes and blades.
 Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.
@@
 The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka.
-This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
-We also confirm a slight decrease in CV values in four attributes.
-This results may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model.
-Drawing on evolutionary ecological theory, we might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
+It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
+We also confirmed a slight decrease in CV values for the four attributes.
+Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission.
+Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
 @fitzhugh2001risk].
@@
 In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
-We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages.
-Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in Phase 3.
+We did not find strong support for distinctive forms of cultural transmission of tool-making in different assemblages.
+Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3.
 Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
 One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
-Similarly, change in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
+Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
-
 # Acknowledgements

### 2023-02-16 analysis/paper/paper.qmd
Dealing with BM's comments

@@
 Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation.
-Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation <!-- and then indirect bias was adapted to carry out the technological change BM: this is unclear -->.
+Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation.
+A slight shift in the social context of transmission towards indirect bias was also observed in the later chronological phase.
 We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage.
@@
 A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity].
-Currently, the oldest stemmed points in Northeast Asia are from the Yongho-doing site in Korea, dating back to 38.5ka [@seong2009emergence; @seong2015diversity; @bae2017origin].
+Currently, the oldest stemmed points in Northeast Asia are from the Yonghodoing site in Korea, dating back to 38.5ka [@seong2008tanged; @seong2015diversity; @bae2012nature].
 Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].
@@
 
-After the first discovery of stemmed points at the Seokjangri site at the 1960s, around 300 have been found in nearly 30 sites across Korea (@fig-map).
+After the first discovery of stemmed points at the Seokjangri site at the 1960s, more than 450 have been found in nearly 30 sites across Korea (@fig-map) [@Sohn_1967; @lee2019were; @Chong_2021].
 While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
@@
 
-<!-- GP need to add this thing in the discussion -->
-<!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation.
-On the other hand, Phase 3 could be more associated with indirect bias.
-The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.??
-
-To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups.
-Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement.
-
 # Conclusion
@@
 A social context that favored trial and error for artifact making is most consistent with Seong's in situ model for the appearance of stemmed points in Korea.
-This model proposes that people developed stemmed points by experimenting with existing enlongated flakes and blades.
+This model proposes that people developed stemmed points by experimenting with existing elongate flakes and blades.
 Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.

### 2023-02-17 analysis/paper/paper.qmd
Edit text more (methods)

@@
 
-After the first discovery of stemmed points at the Seokjangri site at the 1960s, more than 450 have been found in nearly 30 sites across Korea (@fig-map) [@Sohn_1967; @lee2019were; @Chong_2021].
+After the first discovery of stemmed points at the Seokjangri site in the 1960s, more than 450 have been found in nearly 30 sites across Korea (@fig-map) [@Sohn_1967; @lee2019were; @Chong_2021].
 While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].

### 2023-02-17 analysis/paper/paper.qmd
Edited based on proof-reading

@@
 ---
-title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
+title: "Cultural Transmission and Technological Transitions during the Late Paleolithic Period in Korea"
 author:
@@
   The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia.
-Previous studies have mainly focussed on possible origin locations of these new technologies.
-The specific cultural processes of the appearance stemmed points and blades has rarely been considered.
-We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
+Previous studies have mainly focused on possible origin locations of these new technologies.
+The specific cultural processes of the appearance of stemmed points and blades have rarely been considered.
+We investigate these cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
 Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic?
@@
 We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias.
-Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation.
+Some attributes, including length and width, were transmitted with less variation while other attributes appear to have more variation.
 Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation.
@@
 Archaeologists have used evolutionary theories and methods to study human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current].
-In this paper we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
+In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
 The primary technological innovation of this period was the introduction of stemmed points and blades.
-These new lithic technologies may represent the first arrival of modern humans in this region, and thus they may also mark an important event in human dispersal through East Asia [@seong2009emergence].
-Most previous studies on stemmed points focused on where they may have originated, connecting Korea with global patterns of modern human dispersal.
+These new lithic technologies may represent the first arrival of modern humans in this region, and thus, they may also mark an important event in human dispersal through East Asia [@seong2009emergence].
+Most previous studies on stemmed points focused on where these new technologies may have originated, connecting Korea with global patterns of modern human dispersal.
 There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea [@seong2008tanged; @bae2010origin; @bae2017origin].
@@
 The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka.
-A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft.
+A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to the shape of an acute tip, and on the distal end to make a stem, which connects to a shaft.
 Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were].
@@
 
-Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points, as part of the discussion of the timing and routes of modern human dispersal in eastern Asia.
+Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points as part of the discussion of the timing and routes of modern human dispersal in eastern Asia.
 The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin].
-The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
+The *in situ* model claims that stemmed points and other Late Paleolithic assemblages, including blade industries, autonomously emerged in the Korean peninsula as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence].
 To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea.
-In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one.
+In his view, the blade industry represents a new technology, while flakes indicate a continuously-used existing one.
 He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention.
@@
 
-In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins outside of the Korean peninsula [@bae2010origin; @bae2012nature; @bae2013early].
-While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south.
+By contrast, the migration model contends that the new blade industry includes stemmed points, and the earlier flake tool tradition, including large cores, polyhedrals, choppers, and handaxes, originated outside of Korea [@bae2010origin; @bae2012nature; @bae2013early].
+While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that the heterogeneity is the result of the continuous influx of modern human populations from both north and south.
 Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
@@
 Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development.
-He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size.
-With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration.
+He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships under conditions of low effective population size.
+With regard to the blade industry, he claims that the low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration.
 We explore these three options, *in situ*, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.
@@
 To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory.
-This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, such as using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
 Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
 The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties.
-Social transmission strongly affects these key evolutionary processes [@Whiten_2017], and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
-We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gather populations.
+Social transmission strongly affects these key evolutionary processes [@Whiten_2017] and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
+We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gatherer populations.
 
@@
 Individuals learn by themselves (e.g.
-trial and error), or from each other by sharing information.
+trial and error) or from each other by sharing information.
 Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire.
@@
 CV = 10%, 5%, and 3%).
-The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying".
-A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
+The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
 Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.
 
-Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
-Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
-They simulated model-based bias in projectile point making by providing the design of a model and information about the model's prior success to their research participants.
+Garvey, Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
+Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural], who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
+They simulated model-based bias in projectile point-making by providing the design of a model and information about the model's prior success to their research participants.
 They also simulated guided variation by allowing their participants to explore their own designs.
-They observed that the majority of participants who were able to choose the previous design copied the most-successful model.
+They observed that the majority of participants who were able to choose from the previous design ultimately copied the most-successful model.
 Metric attributes of projectile points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
@@
 We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness.
-On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.
-
-To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups.
+On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum, we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.
+
+To contextualize this further, on one end of the spectrum we have socially isolated groups who stayed in physically remote places apart from other groups, or had unfamiliar neighbors with limited contact between groups.
 Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf.
@@
 
-On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with them.
+On the other end of the spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with them.
 Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information.
@@
 This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf.
-Bae's migration model), where stemmed points first appeared and then spread from.
+Bae's migration model), where stemmed points first appeared and then spread from the location(s).
 In this social context, indirect bias is the dominant influence on the transmission of lithic technology.
@@
 While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017].
-Among these stemmed points, we selected those that are unbroken from the tip to the stem.
-We excluded artifacts that were recorded as stemmed points but lack a stem.
+Among these stemmed points, we selected those that were unbroken from the tip to the stem.
+We excluded artifacts that were recorded as stemmed points but lacked a stem.
 This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number).
@@
 To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013].
-These chronologies were constructed based on radiocarbon ages, and assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition.
+These chronologies were constructed based on radiocarbon ages; assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition.
 Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
@@
 Panel B of @fig-raw-materials-by-phase shows the count of stemmed points by raw material type in assemblages with more than five stemmed points.
-Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.
+Much of the raw material diversity comes from isolated finds, with those assemblages containing more than five points, showing high homogeneity in raw materials.
 
@@
 #| label: fig-raw-materials-by-phase
-#| fig-cap: "A: Three chronological phases of the Korean Late Paleolithic, and frequecies of stemmed points and distribution of raw materials in each phase.
+#| fig-cap: "A: Three chronological phases of the Korean Late Paleolithic, and frequencies of stemmed points and distribution of raw materials in each phase.
 B: Frequencies of stemmed points by assemblage and raw material, for assemblages with five or more stemmed points."
@@
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
-Compared to traditional caliper measurements, morphometric data yield more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+Compared to traditional caliper measurements, morphometric data yield interpretable numerical and visual outcomes more easily [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
 We did not include weight and thickness since these measurements were not available to us.
@@
 The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic.
-Using the point tool in ImageJ [@schneider2012nih] we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analyses.
+Using the point tool in ImageJ [@schneider2012nih], we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analysis.
 
@@
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template.
-Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable to stemmed points are Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%.
+As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
+Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable case to stemmed points was Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%.
 Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].
@@
 Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use.
-We also computed confidence intervals using the method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].
+We also computed confidence intervals using the method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and to be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].
 
@@
 @bettinger1999point did not identify a threshold for CV values.
-Here we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases.
-We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias.
-With the exception of weight, the CVs for metric attributes on Rosegate points range from 17% to 24%.
+Here, we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases.
+We used data from @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproduct of indirect bias.
+With the exception of weight, the CVs for metric attributes on Rosegate points ranged from 17% to 24%.
 We believe the higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on the following reasons.
@@
 Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques].
-We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
+We assume that Rosegate projectile points were made from more finer-grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
 (@fig-raw-materials-by-phase).
@@
 Correlation coefficient between attributes for Phase 2.
-The individual point represesnts correlation between two attributes.
+The individual point represents correlation between two attributes.
 Some points are invisible due to their weak relationship.
@@
 Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
-The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`, however the spread of correlations is high, suggesting a mixture of transmission biases.
+The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`; however, the spread of correlations is high, suggesting a mixture of transmission biases.
 
@@
 #| label: fig-corr-raw-ass
-#| fig-cap: Correlation for raw materials that were used for making more than one stemmed poitns. The grey points represent correlation between two attributes.
+#| fig-cap: Correlation for raw materials that were used for making more than one stemmed points. The grey points represent correlation between two attributes.
 
@@
 The points represent CV values for each attribute.
-The vertical lines indicate the confidence interval.
+The vertical lines indicate the confidence intervals.
 #| fig-width:   4.5
@@
 Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker.
-The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
+The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages, the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
 When all artifacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below.
@@
 Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases.
-In exploring individual assemblages that contain multiple stemmed points we observed minimal changes in CV values over time (@fig-plot-cv-four-assemblage).
+In exploring individual assemblages that contain multiple stemmed points, we observed minimal changes in CV values over time (@fig-plot-cv-four-assemblage).
 
@@
 
-In exploring geographical variation we found no clear pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
+In exploring geographical variation, we found no clear pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
 We found that correlation coefficients and CV values are distributed differently in each assemblage.
 While SYG1_2 and SYG1_3 are from the same site, they belong to different chronological periods, and their correlation decreases over time.
-It is likely that the correlation or overall shape can be more depending on individual knappers.
-We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation while the others including stemmed length (SL) and body length (BL) have lower or even negative correlation.
+It is likely that the correlation or overall shape may be more dependent on individual knappers.
+We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation, while the others including stemmed length (SL) and body length (BL) have lower or even negative correlation.
 Based on these findings, we propose that certain attributes were more carefully transmitted than others.
-We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW), similar to what is evident in @fig-plot-cv-all-attributes.
+We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW), which is similar to what is evident in @fig-plot-cv-all-attributes.
 
@@
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
-This can be contrasted with tang length, stem length, and stem width, which are away from the spot of impact at the end of the artifact.
+This can be contrasted with tang length, stem length, and stem width, which are away from the point of impact at the end of the artifact.
 These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
@@
 For example, artifacts made from acidic quartzite have the highest correlation, but their CV values are also much higher with wider confidence intervals than all other raw materials.
-This complex results might be related to the relatively unpredictable knapping quality of quartzite.
+These complex results might be related to the relatively unpredictable knapping quality of quartzite.
 These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals.
@@
 This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`.
-A social context that favored trial and error for artifact making is most consistent with Seong's in situ model for the appearance of stemmed points in Korea.
-This model proposes that people developed stemmed points by experimenting with existing elongate flakes and blades.
+A social context that favored trial and error for artifact making is most consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
+This model proposes that people developed stemmed points by experimenting with existing elongated flakes and blades.
 Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.
@@
 Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
-One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
+One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
 Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
+
 # Acknowledgements

### 2023-02-23 analysis/paper/paper.qmd
editing up through the results section

@@
 ---
-title: "Cultural Transmission and Technological Transitions during the Late Paleolithic Period in Korea"
+title: "Cultural Transmission and Technological Transitions during the Late Paleolithic of Korea"
 author:
@@
 
-The application of evolutionary theory to archaeological research has been favored for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
-Archaeologists have used evolutionary theories and methods to study human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current].
-In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic.
+The application of evolutionary theory to archaeological research has been fruitful for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
+Archaeologists have used evolutionary theories and methods to get insights into human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current].
+In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic (40-35 ka).
 The primary technological innovation of this period was the introduction of stemmed points and blades.
 These new lithic technologies may represent the first arrival of modern humans in this region, and thus, they may also mark an important event in human dispersal through East Asia [@seong2009emergence].
-Most previous studies on stemmed points focused on where these new technologies may have originated, connecting Korea with global patterns of modern human dispersal.
-There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea [@seong2008tanged; @bae2010origin; @bae2017origin].
-Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic based on a cultural transmission framework.
+Previous studies on Korean stemmed points have mostly focused on where these new technologies may have originated, connecting Korea with global patterns of modern human dispersal [@seong2008tanged; @bae2010origin; @bae2017origin].
+There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea.
+Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic, using evolutionary theory to make inferences about cultural transmission from artefact measurements.
 Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-We also ask: Do modes of cultural transmission vary over time and space?
+We also ask: Did modes of cultural transmission vary over time and space?
 We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two.
-The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.
+The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally and independently developed.
 
@@
 
-The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka.
+The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka BP.
 A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to the shape of an acute tip, and on the distal end to make a stem, which connects to a shaft.
@@
 A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity].
-Currently, the oldest stemmed points in Northeast Asia are from the Yonghodoing site in Korea, dating back to 38.5ka [@seong2008tanged; @seong2015diversity; @bae2012nature].
+Currently, the oldest stemmed points in Northeast Asia are from the Yonghodong site in Korea, dating back to 38.5 ka BP [@seong2008tanged; @seong2015diversity; @bae2012nature].
 Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].
@@
 The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome].
-Bae et al.
-[-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
-
-Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development.
+@bae2012nature assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
+
+@lee2013current argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development.
 He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships under conditions of low effective population size.
@@
 To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory.
-This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, such as using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
-Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
-The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties.
+This body of theory has aided archaeologists in understanding a variety of technological innovations and related human behaviors [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+Cultural evolutionary approaches have been effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
+These mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties [@Creanza_Kolodny_Feldman_2017; @Claidiere_Scott_Phillips_Sperber_2014; @Mesoudi_Whiten_Laland_2004; @Acerbi_Mesoudi_2015].
 Social transmission strongly affects these key evolutionary processes [@Whiten_2017] and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying].
@@
 trial and error) or from each other by sharing information.
-Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire.
+Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire [@richerson2008not].
 Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual].
-Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social].
-These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].
+These transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social].
+These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual; @richerson2008not].
 
@@
 In this context of limited contact, they argue that the new technologies developed largely by trial and error.
-On the other hand, in central Nevada they found that metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.
-
-Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
+On the other hand, in central Nevada they found that metric variables were highly correlated, and concluded that the bow and arrow was introduced and spread by indirect bias.
+
+Following Bettinger and Eerkens' framework, @garvey2018current used simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
 Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site.
@@
 CV = 10%, 5%, and 3%).
-The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
+The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is close to the Weber fraction for manually made human artefacts, that is, the threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
 Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.
@@
 Garvey, Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
-Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural], who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
+Direct observations have been reported by @mesoudi2008cultural, who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
 They simulated model-based bias in projectile point-making by providing the design of a model and information about the model's prior success to their research participants.
@@
 Metric attributes of projectile points made when copied from successful models were more highly correlated than attributes of points made by trial and error.
-Mesoudi and O'Brien's experimental results are critical because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.
+Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.
 
@@
 This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf.
-Bae's migration model), where stemmed points first appeared and then spread from the location(s).
+Bae's migration model), where stemmed points first appeared and then spread from.
 In this social context, indirect bias is the dominant influence on the transmission of lithic technology.
@@
 Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
-Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations.
+Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages.
 Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size.
@@
 Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases.
-We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs.
+We exclude Phase 1 from our analyses below because we consider the sample size too small to compute reliable correlations and CVs.
 Shale is the dominant raw material in both Phase 2 and 3.
@@
 We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points.
-Compared to traditional caliper measurements, morphometric data yield interpretable numerical and visual outcomes more easily [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
-We did not include weight and thickness since these measurements were not available to us.
+Compared to traditional caliper measurements, morphometric data yield easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life].
+We did not include weight and thickness because these measurements were not available to us.
 For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis.
@@
 
-Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points.
-The correlation between artifact attributes was proposed by @bettinger1999point as a key indicator of indirect bias during cultural transmission.
+Following @bettinger1999point we computed Pearson's correlation coefficient for all attributes on the stemmed points.
+High correlations between artifact attributes was proposed by @bettinger1999point as a key indicator of indirect bias during cultural transmission.
 It is based on the assumption that artifact-making was transmitted in packages of traits inherited from socially successful individuals who were role models.
@@
 
-We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points.
-The CV is determined by the ratio of the standard deviation to the mean (usually expressed as a percentage).
+To complement the correlation measurements, we chose Coefficient of Variation (CV) to measure the variation among the measurements of the stemmed points.
+The CV is determined by the ratio of the standard deviation to the mean and is usually expressed as a percentage.
 The method has been used in various disciplines to calculate standardization, precision, equality, homogeneity, etc.
@@
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.
-Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable case to stemmed points was Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%.
+As a guide to interpreting CV values, Eerkens and Bettinger @eerkens2001techniques claim that CV values over 57.7% are the result of random production.
+CV values below 1.7% are the byproduct of using a scale or template.
+Among all the various artifacts for which @eerkens2001techniques summarized CV values, the most comparable case to the stemmed points examined here was Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%.
 Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].
@@
 n\<25).
-Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
+Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainties related to sample size [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison].
 Although we have a total of `r total_number_of_sp` artifacts, most of our analyses involve comparisons of smaller subsets.
-As an example, we compared samples of less than 25 pieces in order to explore temporal and regional patterns.
-Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use.
-We also computed confidence intervals using the method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and to be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].
+For example, we compared samples of less than 25 pieces in order to explore temporal and regional patterns.
+Given these relatively small sample sizes in our study, we used the corrected CV formula of @vanpool2011quantitative, which we have implemented here in a function in the R programming language for others to use.
+We also computed confidence intervals using the method proposed by @sharma1994asymptotic to be transparent about the precision of our results [@kelley2007sample; @albatineh2014confidence].
 
 One limitation of previous work is the absence of a clear threshold value for interpreting CV values in terms of different transmission biases.
-@bettinger1999point did not identify a threshold for CV values.
-Here, we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases.
-We used data from @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproduct of indirect bias.
+@bettinger1999point did not identify a threshold for CV values to distinguish guided variation from indirect bias.
+However, one can be inferred from their data.
+We used data from @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which were claimed by Bettinger and Eerkens to result from indirect bias.
 With the exception of weight, the CVs for metric attributes on Rosegate points ranged from 17% to 24%.
-We believe the higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on the following reasons.
+Based on these values, we propose `r threshold_value`% as an approximate threshold to distinguish between guided variation from indirect bias.
+We have three reasons to believe the higher side of the CV range of Rosegate points is appropriate as a threshold value.
 First, variation is generated by small errors that are transmitted between individuals and the errors get bigger through generations [@eerkens2005cultural].
@@
 Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques].
-We assume that Rosegate projectile points were made from more finer-grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
+We assume that Rosegate projectile points were made from more finer-grained raw materials such as flint, compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
 (@fig-raw-materials-by-phase).
 Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values compared to standard CV values [@vanpool2011quantitative].
-Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likely reflect indirect bias.
+Given the tentative identification of this threshold value, we assume that CV values that are lower than `r threshold_value`% likely represent guided variation while CV values over `r threshold_value`% likely reflect indirect bias.
 
 All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam].
-Our R code and data, including the original artifact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].
+Our R code and data, including the original artifact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency, reproducibility, and reuse [@marwick2017computational].
 
@@
 The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold value.
+We interpret Phase 2 as evidence of guided variation and Phase 3 a mixture of guided variation and of indirect bias.
 
@@
 
-all_corr_plot <- rplot(df_sitename_corr, legend = FALSE)
+all_corr_plot <- rplot(df_sitename_corr, 
+                       print_cor = TRUE,
+                       legend = FALSE)
 
@@
 
-p2_corr_plot <- rplot(df_sitename_corr_p2, legend = FALSE)
+p2_corr_plot <- rplot(df_sitename_corr_p2, 
+                      print_cor = TRUE,
+                      legend = FALSE)
 
@@
 p3_corr_plot <-  
-  rplot(df_sitename_corr_p3) +
+  rplot(df_sitename_corr_p3, 
+        print_cor = TRUE) +
    theme(legend.position = c(0.9, 0.9),
@@
 
-One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
+One limitation of this aggregation of stemmed points into chronological phases is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
 These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only on the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
@@
 The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
-At this finer scale of resolution the picture is more complex.
+@fig-plot-cor-four-assemblage shows that at this finer scale of resolution the picture is more complex.
 Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
@@
 
-## Correlation between raw materials
+## Correlations and raw materials
 
@@
 @fig-corr-raw-ass shows that the most abundant raw materials have relatively high correlation coefficients.
-The stemmed points made from quartzite and porphyry have the highest correlation coefficients with narrow distributions.
-Acidic volcanic rocks are one of the highly correlated materials, with individual correlations divided into two groups.
+The stemmed points made from quartzite and porphyry have the highest correlation coefficients with relatively narrow distributions.
+Acidic volcanic rocks are one of the more highly correlated materials, with an outlier group with much lower correlations.
 Similarly, the correlation of shale stemmed points is also divided into two groups.
@@
 In addition to correlation coefficients of point attributes, we measured coefficients of variation (CVs) to quantify variability in individual attributes.
-@fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`.
+@fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv`% to `r max_cv`%.
 Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for Korean artifacts is higher.
@@
 Overall, with all but one attribute having CV values above our threshold value, our results suggest that the transmission of stemmed points was mostly influenced by guided variation.
+
 ```{r}

### 2023-02-24 analysis/paper/paper.qmd
editing discussion

@@
 
-# Introduction {.unnumbered}
+## Introduction {.unnumbered}
 
@@
 
-# The Late Paleolithic of the Korean Peninsula {.unnumbered}
+## The Late Paleolithic of the Korean Peninsula {.unnumbered}
 
@@
 
-# Cultural transmission and transmission biases {.unnumbered}
+## Cultural transmission and transmission biases {.unnumbered}
 
@@
 
-# Modeling the social context of the appearance of stemmed points in the Late Paleolithic of Korea {.unnumbered}
+## Modeling the social context of the appearance of stemmed points in the Late Paleolithic of Korea {.unnumbered}
 
@@
 
-# Materials and methods {.unnumbered}
+## Materials and methods {.unnumbered}
 
@@
 
-## Materials and stemmed points chronology
+### Materials and stemmed points chronology
 
@@
 
-## Methods
+### Methods
 
@@
 Following @bettinger1999point we computed Pearson's correlation coefficient for all attributes on the stemmed points.
-High correlations between artifact attributes was proposed by @bettinger1999point as a key indicator of indirect bias during cultural transmission.
+High correlations between artifact attributes were proposed by @bettinger1999point as a key indicator of indirect bias during cultural transmission.
 It is based on the assumption that artifact-making was transmitted in packages of traits inherited from socially successful individuals who were role models.
@@
 In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current].
-As a guide to interpreting CV values, Eerkens and Bettinger @eerkens2001techniques claim that CV values over 57.7% are the result of random production.
+As a guide to interpreting CV values, @eerkens2001techniques claim that CV values over 57.7% are the result of random production.
 CV values below 1.7% are the byproduct of using a scale or template.
@@
 
-# Results {.unnumbered}
-
-## Correlation coefficients
+## Results {.unnumbered}
+
+### Correlation coefficients
 
@@
 
-## Correlations and raw materials
+### Correlations and raw materials
 
@@
 
-## Coefficient of variation
+### Coefficient of variation
 
@@
                 width = 0.2) +
-  ylab("CV") +
+  ylab("Coefficient of Variation") +
   xlab("Artifact measurement") +
@@
 We excluded stemmed points from Phase 1 due to the limited number of artifacts dating to this time.
-
-@fig-plot-cv-over-time shows that the directional trend in CV values of artifact attributes is complicated.
+@fig-plot-cv-over-time shows that the temporal trend in CV values of artifact attributes is complicated.
 Half of the attributes are below our threshold value of `r threshold_value`, and half are above for both phases.
-Only body length (BL) crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, indicating a shift from indirect bias to guided variation.
+Only body length (BL) crosses the threshold value, changing from \<`r threshold_value` to \>`r threshold_value` over time, suggesting a shift from indirect bias to guided variation.
 For those attributes where the CV is \<`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width (MW) and tang width (TW), indicating increasing influence of indirect bias for these attributes.
@@
 The points represent CV values for each attribute.
-The vertical lines indicate the confidence intervals.
+The vertical lines indicate the 95% confidence intervals for the CV values, which were computed using Sharma and Krishna's method.
 #| fig-width:   4.5
@@
   xlab("Phase") +
-  ylab("Coefficient of Variation for all Attributes") +
+  ylab("Coefficient of Variation") +
     # draw a line to show our threshold value
@@
 We further analyzed temporal change by looking at specific assemblages in each chronological phase.
-Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) for assemblages that have five or more stemmed points.
-@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no directional trend apparent between Phase 2 and Phase 3.
+Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we focused on CV values for body length (BL), maximum length (ML), mid width (MW), and tang width (TW) for assemblages that have five or more stemmed points.
+@fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no temporal trend apparent between Phase 2 and Phase 3.
 Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage.
 Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals.
-It is interesting because SYG1_2 has a strong correlation between attributes (@fig-plot-cor-four-assemblage).
+This is unexpected because SYG1_2 has a strong correlation between attributes (@fig-plot-cor-four-assemblage).
 Perhaps it could be related to its small sample size, relative to the other assemblages compared here.
 The larger assemblages in @fig-plot-cv-four-assemblage indicate guided variation, while the smaller assemblages that dominate @fig-plot-cv-over-time suggest indirect bias.
-Perhaps smaller assemblages represented a social context of higher fidelity copying because the cost of failure was greater due to low social insurance because of low population network sizes.
-Another interpretation is that these results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
+Perhaps smaller assemblages represented a social context of higher fidelity copying because the cost of failure was greater due to low social insurance because of low population network sizes and densities [@fitzhugh2011modeling].
+Another interpretation is that these results suggest that the complex temporal trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
 
@@
   xlab("Attribute") +
-  ylab("Coefficient of Variation on Attributes") +
+  ylab("Coefficient of Variation") +
   facet_wrap( ~ label, scales = "free_y") +
@@
 
-## Variation Between Raw Materials
-
-We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was dependent on raw materials.
+### Variation Between Raw Materials
+
+We examined the relationship between CV values and raw materials to test the hypothesis that the shape variability of stemmed points was dependent on raw materials.
 We excluded raw materials that were used for less than three points.
-@fig-plot-cv-by-raw-material shows that quartzite, the raw material for the stemmed points at Sachang, has the highest CV values, and also has wide conference intervals.
+@fig-plot-cv-by-raw-material shows that quartzite, the raw material for stemmed points at Sachang, has the highest CV values, and also has wide conference intervals.
 Other raw materials are generally low, right around our threshold value, and stable in variation across the attributes.
@@
   geom_hline(yintercept = 25, colour = "red") +
-  xlab("Raw Materials") +
-  ylab("Coefficient of Variation on Attributes") +
+  xlab("Raw Material") +
+  ylab("Coefficient of Variation") +
   facet_wrap( ~ label, scales = "free_y") +
@@
 
-# Discussion
-
-To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we drew on concepts of cultural transmission.
+## Discussion
+
+To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we drew on concepts of cultural transmission from the broader framework of cultural evolutionary theory.
 We asked three questions to examine the cultural transmission process over time: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
@@
 We focused on two transmission biases: guided variation (socially isolated groups making stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others).
-Following @bettinger1999point, we used `r threshold_value_cor` as a threshold value for interpreting correlation coefficients to evaluate which of the two transmission biases was dominant.
+Following @bettinger1999point, we used `r threshold_value_cor` as an approximate threshold value for interpreting correlation coefficients to evaluate which of the two transmission biases was dominant.
 Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias.
-Since there are no previously established criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
+Since there are no previously established criteria for CV, we derived `r threshold_value` from previous work to use as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
 If the CV value is lower than `r threshold_value`, we interpret this as indicating indirect bias while values higher than `r threshold_value` are interpreted as guided variation.
 
-Our correlation analyses show that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias.
+Our correlation analyses found that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias.
 Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker.
+Transmission of the artefact shape as a package seems unlikley.
 The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages, the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
@@
 
-In exploring geographical variation, we found no clear pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
+In exploring variation between sites, as a proxy for spatial variation, we found no readily interpretable pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points.
 We found that correlation coefficients and CV values are distributed differently in each assemblage.
@@
 Based on these findings, we propose that certain attributes were more carefully transmitted than others.
-We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW), which is similar to what is evident in @fig-plot-cv-all-attributes.
-
-In our interpretation, attributes with lower CV values were transmitted in a social context dominated by indirect bias, whereas attributes with high CV values were influenced more by specific manufacturing and maintenance situations.
-We assume that the lower CV values of the four attributes imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape.
+Artefact shape does not appear to have been transmitted as a package, but as a flexible set of attributes that can vary with respect to each other.
+We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW) (cf.
+@fig-plot-cv-all-attributes).
+
+According to our model, attributes with lower CV values were transmitted in a social context dominated by indirect bias, whereas attributes with high CV values were influenced more by specific local manufacturing and maintenance situations.
+We assume that the lower CV values imply that those attributes, BL, ML, MW and TW, were closely associated with the generic projectile function of the tool, Keeping variability low for those attributes would have standardized the shape to ensure predictable performance.
 For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound].
 This can be contrasted with tang length, stem length, and stem width, which are away from the point of impact at the end of the artifact.
-These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
+Our results show these were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
 
@@
 
-# Conclusion
+## Conclusion
 

### 2023-02-24 analysis/paper/paper.qmd
finished editing

@@
 ---
-title: "Cultural Transmission and Technological Transitions during the Late Paleolithic of Korea"
+title: "Cultural Transmission and the Social Contexts of Technological Transitions during the Late Paleolithic of Korea"
 author:
@@
 We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not match our expectations because of how influential raw materials are on CV values.
-For example, artifacts made from acidic quartzite have the highest correlation, but their CV values are also much higher with wider confidence intervals than all other raw materials.
+For example, artifacts made from acidic quartzite have the highest correlations, but their CV values are also much higher with wider confidence intervals than all other raw materials.
 These complex results might be related to the relatively unpredictable knapping quality of quartzite.
-These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals.
+Large grains within quartzite can redirect knapping forces in unexpected directions, resulting in unintended flake removals.
 This makes it harder for the knapper to repeatedly produce the same size and shape end product.
+Variation in raw material nodule size may also contribute to this pattern, for example quartzite nodules may have been found in a much wider range of sizes than nodules of other raw materials.
 This hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
-However with only three artifacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.
+Further analysis of raw material properties is needed to make robust conclusions about their influence here.
 
@@
 It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
-We also confirmed a slight decrease in CV values for the four attributes.
+We also found a slight decrease in CV values for the four attributes.
 Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission.
@@
 In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
-We did not find strong support for distinctive forms of cultural transmission of tool-making in different assemblages.
+We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages.
 Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3.

### 2023-02-24 analysis/paper/paper.qmd
added acknowledgements

@@
 
-<!-- The following line inserts a page break  -->
+We would like to thank Yongjoon Chang, Seonsook Park, Jaeyoung Kim, Soonmi Lee, Taeyoung Kim, and the staff in Korean national museums, and Sangtaek Lim, Chuntaek Seong, Sangmin Cha, Sooa Kim, Gigil Lee, Myeongjin Choi, and Chanho Park in local archaeological institutions and university museums in Korea for permitting access to their collection used in this study and assisting our trip to Korea.
+We thank the Quaternary Research Center, Department of Anthropology and Center for Statistics and the Social Sciences at the University of Washington for supporting our research trip.
+We also appreciate Ben Fitzhugh and Peter Lape for their insightful comments that have greatly improved this manuscript.
 

### 2023-02-24 analysis/paper/paper.qmd
Spelling change for consistency

@@
 There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea.
-Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic, using evolutionary theory to make inferences about cultural transmission from artefact measurements.
+Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic, using evolutionary theory to make inferences about cultural transmission from artifact measurements.
 Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
@@
 CV = 10%, 5%, and 3%).
-The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is close to the Weber fraction for manually made human artefacts, that is, the threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
+The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is close to the Weber fraction for manually made human artifacts, that is, the threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice].
 Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.
@@
 Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker.
-Transmission of the artefact shape as a package seems unlikley.
+Transmission of the artifact shape as a package seems unlikley.
 The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages, the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
@@
 Based on these findings, we propose that certain attributes were more carefully transmitted than others.
-Artefact shape does not appear to have been transmitted as a package, but as a flexible set of attributes that can vary with respect to each other.
+Artifact shape does not appear to have been transmitted as a package, but as a flexible set of attributes that can vary with respect to each other.
 We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW) (cf.

### 2023-03-02 analysis/paper/paper.qmd
created png files for defense slide

@@
         plot.background = element_rect(fill='white', colour="white"))
+
+# ggsave(here::here("analysis/figures/fig-map.png"),
+#        width = 4,
+#        height = 4.5,
+#        units = "in")
+
 ```
@@
           labels = c("A", "B"))
+
+# ggsave(here::here("analysis/figures/fig-raw-materials-by-phase.png"),
+#        width = 4,
+#        height = 3,
+#        units = "in")
+
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-corr-analysis.png"),
+#        width = 4,
+#        height = 3,
+#        units = "in")
+
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-plot-cor-four-assemblage.png"),
+#        width = 4,
+#        height = 3,
+#        units = "in")
+
 ```
@@
   theme_bw()
+
+# ggsave(here::here("analysis/figures/fig-corr-raw-ass.png"),
+#        width = 4,
+#        height = 3,
+#        units = "in")
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-plot-cv-all-attributes.png"),
+#        width = 4.3,
+#        height = 3,
+#        units = "in")
+
 ```
@@
   theme_bw()
+
+# ggsave(here::here("analysis/figures/fig-plot-cv-over-time.png"),
+#        width = 4.3,
+#        height = 3,
+#        units = "in")
+
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-plot-cv-four-assemblage.png"),
+#        width = 4,
+#        height = 3,
+#        units = "in")
+
 ```
@@
   theme_bw()
+
+# ggsave(here::here("analysis/figures/fig-plot-cv-by-raw-material.png"),
+#        width = 4.3,
+#        height = 3,
+#        units = "in")
+
 ```

### 2023-03-09 analysis/paper/paper.qmd
Add Crema's new paper

@@
 To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory.
-This body of theory has aided archaeologists in understanding a variety of technological innovations and related human behaviors [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural].
+This body of theory has aided archaeologists in understanding a variety of technological innovations and related human behaviors [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @crema2023cultural].
 Cultural evolutionary approaches have been effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
@@
 
-Cultural transmission theory holds that information about social and technological behaviours is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
+Cultural transmission theory holds that information about social and technological behaviors is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural].
 Individuals learn by themselves (e.g.

### 2023-04-09 analysis/paper/paper.qmd
Edited based on PL's comments

@@
 The specific cultural processes of the appearance of stemmed points and blades have rarely been considered.
-We investigate these cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
+We explore these cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies.
 Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic?
@@
 Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
-The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome].
+Bae use @shi2005chromosome's genetic studies of the Y chromosome to support the southern route, showing that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP.
 @bae2012nature assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

### 2023-04-09 analysis/paper/paper.qmd
Edited related to PL's comments

@@
 The most striking differences in both correlation coefficients and CV values are in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
-Raw materials have a significant impact on stemmed points' shape variance.
-We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not match our expectations because of how influential raw materials are on CV values.
+Raw materials might have a significant impact on stemmed points' shape variance.
+We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not exactly match our expectations because of how influential raw materials are on CV values.
 For example, artifacts made from acidic quartzite have the highest correlations, but their CV values are also much higher with wider confidence intervals than all other raw materials.
@@
 Variation in raw material nodule size may also contribute to this pattern, for example quartzite nodules may have been found in a much wider range of sizes than nodules of other raw materials.
-This hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
+We could not strongly argue this here due to the small sample size of quartzite to generalize this pattern but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
 Further analysis of raw material properties is needed to make robust conclusions about their influence here.
@@
 Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3.
-Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
-One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
+Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
+One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
 Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].

### 2023-04-11 analysis/paper/paper.qmd
Finished editing

@@
 We excluded raw materials that were used for less than three points.
-@fig-plot-cv-by-raw-material shows that quartzite, the raw material for stemmed points at Sachang, has the highest CV values, and also has wide conference intervals.
+@fig-plot-cv-by-raw-material shows that quartzite, the raw material for stemmed points at Sachang, has the highest CV values, and also has wide confidence intervals.
 Other raw materials are generally low, right around our threshold value, and stable in variation across the attributes.

### 2023-04-11 analysis/paper/paper.qmd
add one para to address Fitzhugh's feedbacks

@@
 
+We concede that our work is rather speculative and provides only a limited explanation of the social contexts surrounding the spread of stemmed points.
+Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span.
+In her simulation study, Garvey was able to derive the population at a particular site based on the number of rooms [@garvey2018current], but our research sites lacked sufficient information to utilize this method.
+Moreover, we recognize that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e.
+the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional].
+Furthermore, we only considered two types of biases, as our primary goal was to understand the general trajectory of technological transition, and we had limited empirical evidence to explore a multiplicity of conjectural scenarios.
+In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points.
+By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.
+
 ## Conclusion

### 2023-04-12 analysis/paper/paper.qmd
Moved the newly added para to conclusion
editing up to discussion

@@
 To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV).
-We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias.
-Some attributes, including length and width, were transmitted with less variation while other attributes appear to have more variation.
+We found that information about the new technology was likely transmitted via guided variation with small contribution from indirect bias.
+Some attributes, including length and width, were transmitted with less variation while other attributes have more variation.
 Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation.
-A slight shift in the social context of transmission towards indirect bias was also observed in the later chronological phase.
+A slight shift in the social context of transmission towards indirect bias was observed in the later chronological phase.
 We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage.
@@
 Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-We also ask: Did modes of cultural transmission vary over time and space?
+We also ask: Did the mode of cultural transmission vary over time and space?
 We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two.
@@
 Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early].
-Bae use @shi2005chromosome's genetic studies of the Y chromosome to support the southern route, showing that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP.
+Bae uses @shi2005chromosome's genetic studies of the Y chromosome to support the southern route, showing that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP.
 @bae2012nature assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.
@@
 To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory.
-This body of theory has aided archaeologists in understanding a variety of technological innovations and related human behaviors [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @crema2023cultural].
+This body of theory has aided archaeologists in understanding a variety of technological innovations and related human behaviors [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @crema2023cultural; @Prentiss_2019].
 Cultural evolutionary approaches have been effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection.
@@
 Following Bettinger and Eerkens' framework, @garvey2018current used simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains.
-Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site.
+Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site in southeastern New Mexico, USA.
 She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e.
@@
 
-Garvey, Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
+Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly.
 Direct observations have been reported by @mesoudi2008cultural, who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study.
@@
 These chronologies were constructed based on radiocarbon ages; assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition.
-Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology).
+Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) assemblages with stemmed points made out of flakes and no blades, 2) assemblages with stemmed points made out of blades or flakes and the existence of blades, and 3) assemblages with stemmed points made out of blades and the existence of micro-blades (@tbl-korean-chronology).
 Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages.
-Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size.
+Inferring approximate ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size.
 We arranged our `r total_number_of_assemblages` assemblages containing stemmed points into these three phases to facilitate observations of change over time.
@@
 Second, the errors are likely to vary by raw materials.
-Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques].
+Some raw materials, such as clay for pots, are easier to control variation in while less controllable materials such as stone for flaked stone artefacts are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques].
 We assume that Rosegate projectile points were made from more finer-grained raw materials such as flint, compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc.
@@
 Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values compared to standard CV values [@vanpool2011quantitative].
-Given the tentative identification of this threshold value, we assume that CV values that are lower than `r threshold_value`% likely represent guided variation while CV values over `r threshold_value`% likely reflect indirect bias.
+Given the tentative identification of this threshold value, we assume that CV values that are higher than `r threshold_value`% likely represent guided variation (trial and error) while CV values less than `r threshold_value`% likely reflect indirect bias (copying from a prestigious or skilled individual).
 
@@
 One limitation of this aggregation of stemmed points into chronological phases is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points.
-These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only on the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
+These isolated finds are ambiguous with respect to local traditions of artifact making, so to further investigate temporal change, we focus only on the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage).
 We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making.
 The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS).
-@fig-plot-cor-four-assemblage shows that at this finer scale of resolution the picture is more complex.
+@fig-plot-cor-four-assemblage shows that at this finer scale of resolution the picture is complex.
 Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
@@
 The vertical lines indicate the 95% confidence intervals for the CV values, which were computed using Sharma and Krishna's method.
+The red line at CV = 25 indicates our approximate threshold to distinguish between guided variation (>25) from indirect bias (<25).
 
@@
 The vertical lines indicate the 95% confidence intervals for the CV values, which were computed using Sharma and Krishna's method.
+The red line at CV = 25 indicates our approximate threshold to distinguish between guided variation (>25) from indirect bias (<25).
 #| fig-width:   4.5
@@
 The vertical lines indicate the confidence intervals.
+The red line at CV = 25 indicates our approximate threshold to distinguish between guided variation (>25) from indirect bias (<25).
 #| fig-width:   4.5
@@
 The vertical lines indicate the confidence intervals.
+The red line at CV = 25 indicates our approximate threshold to distinguish between guided variation (>25) from indirect bias (<25).
 #| fig-width:   8.5
@@
 Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias.
-Since there are no previously established criteria for CV, we derived `r threshold_value` from previous work to use as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
+Since there are no previously established criteria for CV, we derived `r threshold_value`% from previous work to use as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques].
 If the CV value is lower than `r threshold_value`, we interpret this as indicating indirect bias while values higher than `r threshold_value` are interpreted as guided variation.
@@
 Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker.
-Transmission of the artifact shape as a package seems unlikley.
+Given these results, transmission of the artifact shape as a package seems unlikely.
 The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages, the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage).
@@
 
+## Conclusion
+
+In this research, we investigated the social context of technological innovation in the Korean Late Paleolithic.
+Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
+We proposed two scenarios for explaining the introduction of new technology: socially isolated groups that developed stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came into contact with (indirect bias).
+We asked three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
+Do the modes of cultural transmission vary over time?
+And do the modes of cultural transmission vary over space?
+
+We conclude that the dominant mode of cultural transmission for the earliest stemmed points was mostly likely guided variation (trial and error).
+This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`.
+A social context that favored trial and error for artifact making is most consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
+This model proposes that people developed stemmed points by experimenting with existing elongated flakes and blades.
+Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.
+
+The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka.
+It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
+We also found a slight decrease in CV values for the four attributes.
+Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission.
+Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
+@fitzhugh2001risk].
+With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.
+
+In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
+We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages.
+Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3.
+Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
+One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
+Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
+This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
+
 We concede that our work is rather speculative and provides only a limited explanation of the social contexts surrounding the spread of stemmed points.
@@
 
-## Conclusion
-
-In this research, we investigated the social context of technological innovation in the Korean Late Paleolithic.
-Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
-We proposed two scenarios for explaining the introduction of new technology: socially isolated groups that developed stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came into contact with (indirect bias).
-We asked three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic?
-Do the modes of cultural transmission vary over time?
-And do the modes of cultural transmission vary over space?
-
-We conclude that the dominant mode of cultural transmission for the earliest stemmed points was mostly likely guided variation (trial and error).
-This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`.
-A social context that favored trial and error for artifact making is most consistent with Seong's *in situ* model for the appearance of stemmed points in Korea.
-This model proposes that people developed stemmed points by experimenting with existing elongated flakes and blades.
-Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.
-
-The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka.
-It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
-We also found a slight decrease in CV values for the four attributes.
-Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission.
-Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
-@fitzhugh2001risk].
-With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.
-
-In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
-We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages.
-Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3.
-Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
-One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
-Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
-This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
-
 # Acknowledgements

### 2023-04-13 analysis/paper/paper.qmd
editing the conclusion

@@
 It is likely that the correlation or overall shape may be more dependent on individual knappers.
-We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation, while the others including stemmed length (SL) and body length (BL) have lower or even negative correlation.
+We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation, while the others including stem length (SL) and body length (BL) have lower or even negative correlation.
 Based on these findings, we propose that certain attributes were more carefully transmitted than others.
@@
 
-According to our model, attributes with lower CV values were transmitted in a social context dominated by indirect bias, whereas attributes with high CV values were influenced more by specific local manufacturing and maintenance situations.
+According to our model, attributes with lower CV values were transmitted in a social context dominated by indirect bias, whereas attributes with high CV values were likely influenced more by specific local manufacturing and maintenance situations.
 We assume that the lower CV values imply that those attributes, BL, ML, MW and TW, were closely associated with the generic projectile function of the tool, Keeping variability low for those attributes would have standardized the shape to ensure predictable performance.
@@
 This can be contrasted with tang length, stem length, and stem width, which are away from the point of impact at the end of the artifact.
-Our results show these were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
+Our results show these were allowed to vary more freely between specimens, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
 
 The most striking differences in both correlation coefficients and CV values are in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material).
-Raw materials might have a significant impact on stemmed points' shape variance.
+Raw materials likely had a substantial impact on stemmed points' shape variance.
 We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not exactly match our expectations because of how influential raw materials are on CV values.
@@
 Variation in raw material nodule size may also contribute to this pattern, for example quartzite nodules may have been found in a much wider range of sizes than nodules of other raw materials.
-We could not strongly argue this here due to the small sample size of quartzite to generalize this pattern but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
-Further analysis of raw material properties is needed to make robust conclusions about their influence here.
+We could not strongly argue this here due to the small sample size of quartzite to generalize this pattern, but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
+Further analysis of raw material properties and availability is needed to make robust conclusions about their influence here.
 
@@
 
-In this research, we investigated the social context of technological innovation in the Korean Late Paleolithic.
-Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
+In this research we investigated the social contexts of technological innovation in the Korean Late Paleolithic.
+Following previous studies that applied cultural transmission to the introduction of new technologies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined two transmission biases, guided variation (trial and error) and indirect bias (copying a model).
 We proposed two scenarios for explaining the introduction of new technology: socially isolated groups that developed stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came into contact with (indirect bias).

### 2023-04-13 analysis/paper/paper.qmd
Edited in response to Fitzhugh's Fitzhugh's comments

@@
 Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2.
-The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`; however, the spread of correlations is high, suggesting a mixture of transmission biases.
+The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`; however, the spread of correlations is high, suggesting some influence of indirect bias.
 
@@
 @fitzhugh2001risk].
-With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.
+Another possibility is that the cooler climate might result in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources, which would have concentrated them in fewer predictable locations.
+In either case, these conditions might lead to larger networks of individuals interacting with one another and an increased ability to learn from experienced social teachers or successful models.
 
@@
 Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span.
-In her simulation study, Garvey was able to derive the population at a particular site based on the number of rooms [@garvey2018current], but our research sites lacked sufficient information to utilize this method.
-Moreover, we recognize that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e.
+In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current], but our research sites lacked sufficient information to utilize this method.
+In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points.
+We acknowledge that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e.
 the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional].
-Furthermore, we only considered two types of biases, as our primary goal was to understand the general trajectory of technological transition, and we had limited empirical evidence to explore a multiplicity of conjectural scenarios.
+It is our assumption that the majority of stemmed points used in our research have retained their original shape, given that retouch is rare on the tip of stemmed points.
+In our study, we only considered two types of biases as social contexts without considering other factors that impact on morphology of stemmed points such as tool efficiency and manufacturing economization.
+Mainly due to limited empirical evidence to explore a multiplicity of conjectural scenarios, we decided to focus on our primary goal, understanding the general trajectory of technological transition.
 In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points.

### 2023-04-13 analysis/paper/paper.qmd
Added recent papers for citations

@@
 
-The application of evolutionary theory to archaeological research has been fruitful for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural].
-Archaeologists have used evolutionary theories and methods to get insights into human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current].
+The application of evolutionary theory to archaeological research has been fruitful for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural; @walsh2019cultural; @prentiss2020evolutionary; @zwyns2021initial].
+Archaeologists have used evolutionary theories and methods to get insights into human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current; @straffon2019uses; @gjesfjeld2019contributions].
 In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic (40-35 ka).

### 2023-04-22 analysis/paper/paper.qmd
editing the conclusion

@@
 SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes.
-Overall, with all but one attribute having CV values above our threshold value, our results suggest that the transmission of stemmed points was mostly influenced by guided variation.
+Overall, with all but one attribute having CV values above our threshold value at this level of aggregation of assemblages, our results suggest that the transmission of stemmed points was mostly influenced by guided variation.
 
@@
 Variation in raw material nodule size may also contribute to this pattern, for example quartzite nodules may have been found in a much wider range of sizes than nodules of other raw materials.
-We could not strongly argue this here due to the small sample size of quartzite to generalize this pattern, but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
+The small sample size of quartzite artefacts in our study limits how far we can generalize this pattern, but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability.
 Further analysis of raw material properties and availability is needed to make robust conclusions about their influence here.
@@
 We also found a slight decrease in CV values for the four attributes.
-Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission.
-Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
-@fitzhugh2001risk].
-Another possibility is that the cooler climate might result in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources, which would have concentrated them in fewer predictable locations.
-In either case, these conditions might lead to larger networks of individuals interacting with one another and an increased ability to learn from experienced social teachers or successful models.
-
+Those results can be interpreted as a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skill transmission.
 In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
@@
 Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
+
 One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
 Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
+Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf.
+@fitzhugh2001risk].
+Another possibility is that the cooler climate might have resulted in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources <!-- citation needed, or delete this sentence --> , which would have concentrated them in fewer predictable locations.
+In either case, these conditions likely led to larger networks of individuals interacting with one another and an increased opportunities to learn from experienced teachers or successful models.
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
 
-We concede that our work is rather speculative and provides only a limited explanation of the social contexts surrounding the spread of stemmed points.
+A key difference between our work and Bettinger and Eerkens and Garvey is the time span that we are attempting to infer cultural transmission for.
+For Bettinger and Eerkens and Garvey the time span is in the order of 100s of years or dozens of human generations, while our study spans 1000s of years and 100s of human generations.
+In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current].
+
 Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span.
-In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current], but our research sites lacked sufficient information to utilize this method.
+but our research sites lacked sufficient information to utilize this method.
 In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points.

### 2023-04-26 analysis/paper/paper.qmd
Edited conclusion

--- 2023-04-13 analysis/paper/paper.qmd
@@
 One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity.
-Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did].
+In our previous research, we observed that forager groups with stemmed points were able to occupy higher elevation, where the temperature was generally lower [@park2022did].
+The changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of increased social networks.
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.

### 2023-05-06 analysis/paper/paper.qmd
edit the definition of SP

@@
 The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka BP.
-A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to the shape of an acute tip, and on the distal end to make a stem, which connects to a shaft.
+A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the distal end to the shape of an acute tip, and on the proximal end to make a stem, which connects to a shaft.
 Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were].

### 2023-05-11 analysis/paper/paper.qmd
edits to conclusion, references and methods

--- 2023-04-22 analysis/paper/paper.qmd
@@
 Using the point tool in ImageJ [@schneider2012nih], we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analysis.
+We consider that the stone artifacts were generally modified very little during their use life by reuse and resharpening.
+While retouching can alter artefact shapes independent of transmission biases (i.e.
+the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional], retouch is very rare on the margins and tip of stemmed points, so we consider that the majority of stemmed points used in our sample have retained their original shape.
 
@@
 The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka.
-It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
-We also found a slight decrease in CV values for the four attributes.
-Those results can be interpreted as a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skill transmission.
-In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques.
-We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages.
-Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3.
+This is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias.
+We also found a slight decrease in CV values.
+Those results may indicate a shift away from trial and error, towards a greater reliance on copying a model in the social context of tool-making skill transmission.
 Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.
@@
 @fitzhugh2001risk].
-Another possibility is that the cooler climate might have resulted in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources <!-- citation needed, or delete this sentence --> , which would have concentrated them in fewer predictable locations.
-In either case, these conditions likely led to larger networks of individuals interacting with one another and an increased opportunities to learn from experienced teachers or successful models.
+Another possibility is that the cooler climate might have resulted in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources, which would have concentrated them in fewer predictable locations.
+In our previous research, we observed that forager groups with stemmed points were able to occupy higher elevation, where the temperature was generally lower [@park2022did].
+The changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of increased social networks.
 This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
 
-A key difference between our work and Bettinger and Eerkens and Garvey is the time span that we are attempting to infer cultural transmission for.
+A key difference between our work and that of Bettinger and Eerkens and Garvey, which motivated our approach, is the time span that we are attempting to infer cultural transmission for.
 For Bettinger and Eerkens and Garvey the time span is in the order of 100s of years or dozens of human generations, while our study spans 1000s of years and 100s of human generations.
 In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current].
-
-Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span.
-but our research sites lacked sufficient information to utilize this method.
-In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points.
-We acknowledge that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e.
-the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional].
-It is our assumption that the majority of stemmed points used in our research have retained their original shape, given that retouch is rare on the tip of stemmed points.
-In our study, we only considered two types of biases as social contexts without considering other factors that impact on morphology of stemmed points such as tool efficiency and manufacturing economization.
-Mainly due to limited empirical evidence to explore a multiplicity of conjectural scenarios, we decided to focus on our primary goal, understanding the general trajectory of technological transition.
-In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points.
-By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.
+Our findings rely on the untested assumption that similar regularities about cultural transmission are relevant over much longer time periods.
+Are transmission biases scale invariant with respect to time?
+Or is there a time span that when exceeded, patterns in cultural transmission become so overwhelmed with noise such that they are no longer discernible?
+We recognize that the answer to this question is important for determining the validity of our results.
+@Premo_Kuhn_2010 used an agent-based model to investigate Paleolithic cultural change and found that the visibility of cultural change is strongly affected by local extinction rates, in some cases mimicking the results of conformist cultural transmission.
+Longer time periods may increase the probability of extinction events altering the archaeological record in this way.
+Hopefully our claims will stimulate future work on this question to assess the reliability of our assumptions.
+A good entry point to this next phase of work could be adapting Garvey's social learning models to cover a 20,000 year span, such as reported in our study.
+This would require drawing on archaeological and ethnographic information for approximate estimates for the population and generation turnover values required to transmit knowledge of new technology.
+Approximate Bayesian Computation is likely to be a productive computational method for refining these estimates, give how effective it has been in inferring models of cultural transmission from the archaeological record [@Crema_Edinborough_Kerig_Shennan_2014].
 

### 2023-05-11 analysis/paper/paper.qmd
minor edits for text and figure

@@
 
-## Materials and methods {.unnumbered}
+## Methods {.unnumbered}
 
@@
 Shale is the dominant raw material in both Phase 2 and 3.
-In Phase 3 there is an increase in the proportion of porphyry, driven largely by finds from Yongsangdong, and other raw materials.
+In Phase 3 there is an increase in the proportion of porphyry, driven largely by findings from Yongsangdong, and other raw materials.
 Panel B of @fig-raw-materials-by-phase shows the count of stemmed points by raw material type in assemblages with more than five stemmed points.
@@
 # ggsave(here::here("analysis/figures/fig-raw-materials-by-phase.png"),
-#        width = 4,
+#        width = 5,
 #        height = 3,
@@
 
-### Methods
+### Morphometric Analysis
 
@@
 Correlation coefficient for the second and third chronological phases.
-The grey points represent correlation between two attributes."
+The gray points represent correlation between two attributes."
 #| fig-width: 7
@@
 Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
-The grey points represent correlation between two attributes.
+The gray points represent correlation between two attributes.
 #| fig-width:   4.5
@@
 #| label: fig-corr-raw-ass
-#| fig-cap: Correlation for raw materials that were used for making more than one stemmed points. The grey points represent correlation between two attributes.
+#| fig-cap: Correlation for raw materials that were used for making more than one stemmed points. The gray points represent correlation between two attributes.
 
@@
   coord_cartesian(ylim = c(0,250)) +
-  theme_bw()
+  theme_bw()+
+  theme(strip.text.x = element_text(size = 5))
 

### 2023-05-16 analysis/paper/paper.qmd
Changed font size in a plot

@@
   theme_bw()+
-  theme(strip.text.x = element_text(size = 5))
+  theme(strip.text.x = element_text(size = 5),
+        axis.text.x = element_text(size = 5))
 

### 2023-05-17 analysis/paper/paper.qmd
Minor edit for figure 10

--- 2023-05-11 analysis/paper/paper.qmd
@@
   theme_bw()+
-  theme(strip.text.x = element_text(size = 5))
+  theme(strip.text.x = element_text(size = 5),
+        axis.text.x = element_text(size = 4))
 

### 2023-05-18 analysis/paper/paper.qmd
minor edit for submission

--- 2023-05-11 analysis/paper/paper.qmd
@@
  theme(legend.position = "none",
-       axis.text.x = element_text(size = 5)  ) +
+       axis.text.x = element_text(size = 5),
+        axis.text.y = element_text(size = 5)) +
   scale_x_discrete(guide = guide_axis(n.dodge= 2.3)) +
@@
   labs(y = "Correlation coefficient") +
-theme(axis.title.y = element_text(size = 13)) +
+theme(axis.title.y = element_text(size = 11)) +
   # draw a line to show our threshold value
@@
        x = "Raw material") +
-theme(axis.title.y = element_text(size = 4)) +
+theme(axis.title.y = element_text(size = 4),
+      axis.title.x = element_text(size = 2.7)) +
   # draw a line to show our threshold value
@@
   theme_bw()+
-  theme(strip.text.x = element_text(size = 5))
+  theme(strip.text.x = element_text(size = 5),
+        axis.text.x = element_text(size = 4.5))
 
@@
 
-We would like to thank Yongjoon Chang, Seonsook Park, Jaeyoung Kim, Soonmi Lee, Taeyoung Kim, and the staff in Korean national museums, and Sangtaek Lim, Chuntaek Seong, Sangmin Cha, Sooa Kim, Gigil Lee, Myeongjin Choi, and Chanho Park in local archaeological institutions and university museums in Korea for permitting access to their collection used in this study and assisting our trip to Korea.
-We thank the Quaternary Research Center, Department of Anthropology and Center for Statistics and the Social Sciences at the University of Washington for supporting our research trip.
-We also appreciate Ben Fitzhugh and Peter Lape for their insightful comments that have greatly improved this manuscript.
+We thank Yongjoon Chang, Seonsook Park, Jaeyoung Kim, Soonmi Lee, Taeyoung Kim, and the staff in Korean national museums, and Sangtaek Lim, Chuntaek Seong, Sangmin Cha, Sooa Kim, Gigil Lee, Myeongjin Choi, and Chanho Park from local archaeological institutions and university museums in Korea.
+Their assistance and support in accessing the data used in this paper and accommodating our research visits to their collections are greatly appreciated.
+We would also like to acknowledge the financial support provided by the Quaternary Research Center and the travel grants from the Department of Anthropology and the Center for Statistics and the Social Sciences at the University of Washington.
+Furthermore, we express our sincere thanks to Ben Fitzhugh and Peter Lape for their insightful feedback on an early draft, which significantly contributed to the improvement of our work.
 
